### 2020-06-19 manuscript/manuscript_text.Rmd
working draft

---
title: "Palmer Archipelago (Antarctica) penguins: an alternative to Edgar Anderson's iris data"
output: html_document
bibliography: ../bibliography/references.bib
---

```{r setup, include=FALSE}
library(tidyverse)
library(palmerpenguins)
library(kableExtra)
library(patchwork)
library(ggExtra)
library(gt)
library(ggpubr)
library(ggbeeswarm)
knitr::opts_chunk$set(echo = TRUE)
```

## Authors

Allison Horst, Alison Hill, Kristen Gorman

## Keywords

## Abstract

Edgar Anderson's *Iris* data are ubiquitous in statistics and data science education, but their use in Ronald A. Fisher's eugenics research has created momentum to switch to alternative datasets. Size measurements for three penguin species observed on islands in Palmer Archipelago, Antarctica provide a (mostly) suitable, approachable and charismatic alternative to the *Iris* data. Here, we introduce the penguins data, then directly compare the two datasets to demonstrate that educators currently using *Iris* can switch to the Palmer Archipelago penguins for many use cases, particularly in introductory statistics and data science courses.

## Introduction

In 1935, American botanist Edgar Anderson measured petal and sepal dimensions (length and width) for 50 samples from three *Iris* species: *Iris setosa*, *Iris versicolor*, and *Iris virginica* [@anderson]. The manageable but non-trivial size (5 variables and 150 total observations) and characteristics of the data (....SOMETHING HERE) make it amenable for introducing a wide range of statistical methods including data wrangling, visualization, hypothesis testing, regression, machine learning, and more. Anderson's *Iris* dataset is even built into a number of software packages, including in the auto-installed *datasets* package in *R*, Python's *scikit-learn* machine learning library, and the SAS *Sashelp* library, further facilitating its widespread use. As a result, eighty-five years after the data were initially published, Anderson’s Iris measurements are ubiquitous in statistics and data science materials.

*Sounds great. Why consider switching?*

In 1936, one year after its publication in the *Bulletin of the American Iris Society*, Anderson’s *Iris* measurements were further analyzed and published by eugenicist and statistician Ronald A. Fisher in *The Annals of Eugenics* [@fisher]. Regardless of Anderson's initial motivation, the dataset today inextricably linked to Fisher’s eugenics research and even commonly (if unfairly) referred to as "*Fisher's* iris data" (ref). Due to its history in - and association with - Fisher's eugenics research, there is momentum to replace Iris with alternative datasets that offer similar teaching and learning opportunities.

Considering *Iris* usage, we established the following criteria for a suitable alternative dataset:

- Available by an appropriate license, ideally CC0 (no rights reserved)
- Should feature intuitive subjects and variables that are understandable to learners across fields
- Dimensions should be manageable (but not trivial)
- Contain real-world observations for improved understanding and engagement [@cafe]
- Provide similar opportunities for teaching data science skills and statistical methods

Here, we offer an alternative to Anderson's irises that largely satisfies those criteria: an approachable and charismatic dataset containing real-world size measurements for three penguin species observed on islands in Palmer Archipelago, Antarctica. Comparable size, characteristics and learning opportunities make the penguins data uniquely transferable for educators currently using *Iris*. By highlighting similarities in data structure and analyses, as well as several examples for which the penguins are *not* a good alternative, we demonstrate that the penguins can replace irises for many use cases.

## Data source

Size measurements, clutch observations, and blood isotope levels for male and female nesting adult Adélie (*Pygoscelis adeliae*), chinstrap (*Pygoscelis antarcticus*), and gentoo (*Pygoscelis papua*) penguins on three islands in the Palmer Archipelago were collected from 2007 - 2009 by Dr. Kristen Gorman with the [Palmer Station Long Term Ecological Research Program](https://pal.lternet.edu/), part of the [US Long Term Ecological Research Network](https://lternet.edu/). For complete data collection methods see [@gorman_plos]. The data were imported directly from the [Environmental Data Initiative](https://environmentaldatainitiative.org/) (EDI) Data Portal [@adelie_edi; @gentoo_edi; @chinstrap_edi] and are available for use by CC0 license ("No Rights Reserved") in accordance with the [Palmer Station Data Policy](https://pal.lternet.edu/data/policies).

Data for each penguin species (Adélie, chinstrap and gentoo) were imported separately, then combined into a single table stored in the [*palmerpenguins*](https://allisonhorst.github.io/palmerpenguins/) package as object `penguins_raw`. Data in `penguins_raw` are in tidy format [@tidydata] and contain observations for 17 variables (Appendix Table 1) recorded completely or in part for 344 individual penguins.

### Curated `penguins` data

As a direct alternative to Anderson's *Iris* data, we recommend using the curated `penguins` object in the *palmerpenguins* data package. The `penguins` data is a subset of `penguins_raw` retaining all 344 rows (individual penguins) but reduced to the following seven variables:

- **species:** a factor denoting the penguin species (Adelie, Chinstrap, or Gentoo)
- **island:** a factor denoting the island (in Palmer Archipelago, Antarctica) where observed (Biscoe, Dream or Torgersen)
- **bill_length_mm:** a number denoting length of the dorsal ridge of penguin bill (millimeters)
- **bill_depth_mm:** a number denoting the depth of the penguin bill (millimeters)
- **flipper_length_mm:** an integer denoting penguin flipper length (millimeters)
- **body_mass_g:** an integer denoting penguin body mass (grams)
- **sex:** a factor denoting penguin sex (male, female)

The `penguins` subset contains `r sum(complete.cases(penguins))` complete cases, with missing values (n~missing~ = `r sum(is.na(penguins))`) indicated by `NA` throughout. In addition to reduced variables, the curated `penguins` subset is minimally updated from `penguins_raw` as follows:

1. All variable names are converted to lower snake case
2. Factor levels for species are truncated to only include the common name (e.g. "Gentoo", instead of "Gentoo penguin (*Pygoscelis papua*)")
3. Recorded sex for penguin N36A1, originally recorded as ".", is updated to `NA`
4. "culmen_length_mm" and "culmen_depth_mm" variable names are updated to "bill_length_mm" and "bill_depth_mm" to be more intuitive

## Selected comparisons: `penguins` and `iris`

We feel that the `penguins` data is generally useful and approachable for statistics education, and that it is uniquely well-suited to replace `iris`. Comparisons presented here are representative examples, and are not exhaustive.

### Structure and sample size

Both `iris` and `penguins` datasets are tidy [@tidydata], with each column denoting a single variable and each row containing a single observation (Table 1, 2).

```{r, echo = FALSE}
set.seed(1900)
sample_n(iris, 6) %>% 
  kable(caption = "Table 1. Iris dataset, 6 random observations (of 150 total)") %>% 
  kable_styling(full_width = FALSE)
```

```{r, echo = FALSE}
set.seed(1900)
sample_n(penguins, 6) %>% 
  kable(caption = "Table 2. Penguins dataset, 6 random observations (of 344 total)") %>% 
  kable_styling(full_width = FALSE)
```

Overall sizes are comparable: dimensions (# columns $\times$ # rows) are 5 $\times$ 150 and 7 $\times$ 344 for `iris` and `penguins`, respectively, and sample sizes within groups (e.g. by species) are similar (Table 3).

```{r compare_counts, echo = FALSE}

iris_counts <- iris %>% 
  count(Species) %>% 
  rename(`Iris species` = Species)

penguin_counts <- penguins %>% 
  mutate(sex = str_to_title(sex)) %>% 
  rename(`Penguin species` = species) %>% 
  count(`Penguin species`, sex) %>% 
  pivot_wider(names_from = sex, values_from = n)

iris_penguin_n <- cbind(iris_counts, penguin_counts) %>% 
  rename(`Sample size` = n)

iris_penguin_n %>% 
  gt() %>% 
  tab_spanner(
    label = "Iris data",
    columns = vars(`Iris species`, `Sample size`)
  ) %>% 
  tab_spanner(
    label = "Penguin data",
    columns = vars(`Penguin species`, `Female`, `Male`, `NA`)
  ) %>% 
  tab_footnote(
    footnote = "Note: Here, penguin sample sizes are split by species and sex, and could be further divided by island.",
    location = cells_column_labels(
      columns = vars(`Penguin species`)
    )
  )
```

### Correlation

#### Comparison 1: Linear trends across or within species

An overall approximately linear relationship can be explored using petal dimensions from `iris`, and penguin size (flipper length versus body mass) in `penguins` (Figure X). Educators may explore how linear regression across all species of iris or penguins compares to regression performed separately for each species.

```{r, echo = FALSE, warning = FALSE}

penguin_flip_mass_scatter <- ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species), size = 2, alpha = 0.7) +
  theme_minimal() +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  theme(legend.position = c(0.3, 0.8)) +
  labs(x = "Flipper length (mm)", 
       y = "Body mass (g)",
       title = "Penguin size") +
  theme(plot.title.position = "plot")

iris_petal_scatter <- ggplot(iris, aes(x = Petal.Length, y = Petal.Width)) +
  geom_point(aes(color = Species), size = 2, alpha = 0.7) +
  theme_minimal() +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  theme(legend.position = c(0.3, 0.8)) +
  labs(x = "Petal length (cm)",
       y = "Petal width (cm)",
       title = "Iris petal dimensions") +
  theme(plot.title.position = "plot")

penguin_flip_mass_scatter + iris_petal_scatter
```

#### Comparison 2: linear relationships within species

```{r, echo = FALSE, warning = FALSE, message = FALSE}
penguin_bill_scatter <- ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(aes(color = species), size = 2, alpha = 0.7) +
  theme_minimal() +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  theme(legend.position = "bottom") +
  labs(x = "Bill length (mm)", 
       y = "Bill depth (g)",
       title = "Penguin size", 
       color = "Species") +
  theme(plot.title.position = "plot")

iris_sepal_scatter <- ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(aes(color = Species), size = 2, alpha = 0.7) +
  theme_minimal() +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  theme(legend.position = "bottom") +
  labs(x = "Sepal length (cm)",
       y = "Sepal width (cm)",
       title = "Iris sepal dimensions") +
  theme(plot.title.position = "plot")

penguin_bill_scatter + iris_sepal_scatter
```

### Differences between groups

As in `iris`, multiple opportunities for comparisons between groups also exist in `penguins`.

```{r, echo = FALSE, warning = FALSE, message = FALSE}

iris_bee <- ggplot(iris, aes(x = Species, y = Petal.Width)) +
  geom_boxplot(color = "gray80",
               fill = "gray92", 
               outlier.color = NA
               ) +
  geom_beeswarm(aes(color = Species), 
                alpha = 0.6,
                size = 2) + 
  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  theme_minimal() +
  theme(legend.position = c(0.2, 0.8),
        plot.title.position = "plot") +
  labs(x = "Iris species",
       y = "Petal width (cm)",
       title = "Iris petal width")

fpenguins_bee <- penguins %>% 
  ggplot(aes(x = species, y = flipper_length_mm)) +
  geom_boxplot(color = "gray80",
               fill = "gray92", 
               outlier.color = NA
               ) +
  geom_beeswarm(aes(color = species), 
                alpha = 0.6,
                size = 2) + 
  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  theme_minimal() +
  theme(legend.position = c(0.2, 0.8),
        plot.title.position = "plot") +
  labs(x = "Penguin species",
       y = "Flipper length (mm)",
       title = "Penguin flipper length")

fpenguins_bee + iris_bee
```

Compared to the single factor in `iris` (Species), `penguins`' three factors (species, island, sex) increase opportunities for comparisons between groups (Figure X).

```{r, echo = FALSE, warning = FALSE, message = FALSE}

cd <- penguins %>% 
  filter(sex %in% c("male","female")) %>% 
ggplot(aes(x = bill_depth_mm)) +
  geom_histogram(aes(fill = species), alpha = 0.6, position = "identity") +
  scale_fill_manual(values = c("darkorange","darkorchid","cyan4")) +
  facet_wrap(~ sex) +
  theme_minimal()

cl <- penguins %>% 
  filter(sex %in% c("male","female")) %>% 
ggplot(aes(x = bill_length_mm)) +
  geom_histogram(aes(fill = species), alpha = 0.6, position = "identity") +
  scale_fill_manual(values = c("darkorange","darkorchid","cyan4")) +
  facet_wrap(~ sex) +
  theme_minimal()

fl <- penguins %>% 
  filter(sex %in% c("male","female")) %>% 
ggplot(aes(x = flipper_length_mm)) +
  geom_histogram(aes(fill = species), alpha = 0.6, position = "identity") +
  scale_fill_manual(values = c("darkorange","darkorchid","cyan4")) +
  facet_wrap(~ sex) +
  theme_minimal()

bm <- penguins %>% 
  filter(sex %in% c("male","female")) %>% 
ggplot(aes(x = body_mass_g)) +
  geom_histogram(aes(fill = species), alpha = 0.6, position = "identity") +
  scale_fill_manual(values = c("darkorange","darkorchid","cyan4")) +
  facet_wrap(~ sex) +
  theme_minimal()

cd / cl / fl / bm + plot_layout(guides = "collect")
```

#### Comparison 3: Principal components analysis

## Appendix

### Appendix A: Variables in the raw data

- **studyName:** Sampling expedition indicator
- **Sample Number:** Integer denoting the continuous numbering sequence for each penguin
- **Species:** a factor denoting the penguin species (Adelie, Chinstrap, or Gentoo)
- **Region**: A character denoting the region of the Palmer LTER sampling grid
- **Island:** a factor denoting the island where observed (Torgersen, Dream or Biscoe) in Palmer Archipelago
- **Stage:** A character string denoting reproductive stage at sampling
- **Individual ID:** A character string denoting a unique ID for each penguin
- **Clutch Completion:** A character string denoting if the study next observed with a full clutch, i.e. 2 eggs
- **Date Egg:** A date denoting the date study nest observed with 1 egg
- **Culmen Length:** A number denoting the length of the dorsal ridge of a bird's bill (millimeters)
- **Culmen Depth:** A number denoting the depth of the dorsal ridge of a bird's bill (millimeters)
- **Flipper Length:** An integer denoting the penguin flipper length (millimeters)
- **Body Mass:** An integer denoting the penguin body mass (grams)
- **Sex:** A character string denoting the sex of an animal
- **Delta 15 N:** A number denoting the measure of the ratio of stable isotopes 15N:14N
- **Delta 13 C:** A number denoting the measure of the ratio of stable isotopes 13C:12C
- **Comments:** A character string with text providing additional relevant information for data

## Citations

### 2020-06-19 manuscript/manuscript_text.Rmd
describe sample size (#1)

@@
 
-The `penguins` subset contains `r sum(complete.cases(penguins))` complete cases, with missing values (n~missing~ = `r sum(is.na(penguins))`) indicated by `NA` throughout.
+The `penguins` subset contains `r sum(complete.cases(penguins))` complete cases, with missing values (n~missing~ = `r sum(is.na(penguins))`, percent missing is `r round(100*sum(is.na(penguins))/(7*344),2)` %) indicated by `NA` throughout.
 In addition to reduced variables, the curated `penguins` subset is minimally updated from `penguins_raw` as follows:
@@
 
-Both `iris` and `penguins` datasets are tidy [@tidydata], with each column denoting a single variable and each row containing a single observation (Table 1, 2).
+Both `iris` and `penguins` datasets are in tidy format [@tidydata], with each column denoting a single variable and each row containing a single observation (Table 1, 2).
 
@@
 by species) are similar (Table 3).
+Note that the additional factors in `penguins` (species, island and sex) create additional opportunities for grouping, compared to the single factor in `iris` (Species).
 

### 2020-06-19 manuscript/manuscript_text.Rmd

@@
 library(tidyverse)
+library(here)
 library(palmerpenguins)
@@
 
-Allison Horst, Alison Hill, Kristen Gorman
+Allison Horst, Alison Hill, Kristen Gorman (order TBD)
 
@@
 
-In 1936, one year after its publication in the *Bulletin of the American Iris Society*, Anderson’s *Iris* measurements were further analyzed and published by eugenicist and statistician Ronald A.
+In 1936, one year after publication in the *Bulletin of the American Iris Society*, Anderson’s *Iris* measurements were further analyzed and published by eugenicist and statistician Ronald A.
 Fisher in *The Annals of Eugenics* [@fisher].
@@
 - Available by an appropriate license, ideally CC0 (no rights reserved)
-- Should feature intuitive subjects and variables that are understandable to learners across fields
-- Dimensions should be manageable (but not trivial)
-- Contain real-world observations for improved understanding and engagement [@cafe]
+- Feature intuitive subjects and variables understandable to learners across fields
+- Manageable, but not trivial, size
+- Real-world observations for improved understanding and engagement [@cafe]
 - Provide similar opportunities for teaching data science skills and statistical methods
@@
 
-As in `iris`, multiple opportunities for comparisons between groups also exist in `penguins`.
+As in `iris`, multiple opportunities for comparisons between groups also exist in `penguins` (Figure X).
+Notably, more obvious separation between species exists in `iris` (Figure X E & F), compared to any between-species comparisons in `penguins` (Figure X A - D).
 
@@
 
-iris_bee <- ggplot(iris, aes(x = Species, y = Petal.Width)) +
-  geom_boxplot(color = "gray80",
-               fill = "gray92", 
-               outlier.color = NA
-               ) +
-  geom_beeswarm(aes(color = Species), 
-                alpha = 0.6,
-                size = 2) + 
+# Update this with an actual theme FFS and use '&' in patchwork
+
+iris_pw <- ggplot(iris, aes(x = Species, y = Petal.Width)) +
+  geom_boxplot(aes(color = Species),
+               show.legend = FALSE) +
   scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme_minimal() +
-  theme(legend.position = c(0.2, 0.8),
-        plot.title.position = "plot") +
-  labs(x = "Iris species",
-       y = "Petal width (cm)",
-       title = "Iris petal width")
+  theme_bw() +
+  theme(plot.title.position = "plot") +
+  labs(x = "",
+       y = "Petal width (cm)") +
+  coord_flip()
 
-fpenguins_bee <- penguins %>% 
+iris_pl <- ggplot(iris, aes(x = Species, y = Petal.Length)) +
+  geom_boxplot(aes(color = Species),
+               show.legend = FALSE) +
+  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
+  theme_bw() +
+  theme(plot.title.position = "plot") +
+  labs(x = "",
+       y = "Petal length (cm)") +
+  coord_flip()
+
+iris_sl <- ggplot(iris, aes(x = Species, y = Sepal.Length)) +
+  geom_boxplot(aes(color = Species),
+               show.legend = FALSE) +
+  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
+  theme_bw() +
+  theme(plot.title.position = "plot") +
+  labs(x = "",
+       y = "Sepal length (cm)") +
+  coord_flip()
+
+iris_sw <- ggplot(iris, aes(x = Species, y = Sepal.Width)) +
+  geom_boxplot(aes(color = Species),
+               show.legend = FALSE) +
+  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
+  theme_bw() +
+  theme(plot.title.position = "plot") +
+  labs(x = "",
+       y = "Sepal width (cm)") +
+  coord_flip()
+
+penguins_flip <- penguins %>%
   ggplot(aes(x = species, y = flipper_length_mm)) +
-  geom_boxplot(color = "gray80",
-               fill = "gray92", 
-               outlier.color = NA
-               ) +
-  geom_beeswarm(aes(color = species), 
-                alpha = 0.6,
-                size = 2) + 
+  geom_boxplot(aes(color = species),
+               show.legend = FALSE) +
   scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme_minimal() +
-  theme(legend.position = c(0.2, 0.8),
-        plot.title.position = "plot") +
-  labs(x = "Penguin species",
-       y = "Flipper length (mm)",
-       title = "Penguin flipper length")
+  theme_bw() +
+  theme(plot.title.position = "plot") +
+  labs(x = "",
+       y = "Flipper length (mm)") +
+  coord_flip()
 
-fpenguins_bee + iris_bee
+penguins_mass <- penguins %>%
+  ggplot(aes(x = species, y = body_mass_g)) +
+  geom_boxplot(aes(color = species),
+               show.legend = FALSE) +
+  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
+  theme_bw() +
+  theme(plot.title.position = "plot") +
+  labs(x = "",
+       y = "Body mass (g)") +
+  coord_flip()
+
+penguins_bill_length <- penguins %>%
+  ggplot(aes(x = species, y = bill_length_mm)) +
+  geom_boxplot(aes(color = species),
+               show.legend = FALSE) +
+  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
+  theme_bw() +
+  theme(plot.title.position = "plot") +
+  labs(x = "",
+       y = "Bill length (mm)") +
+  coord_flip()
+
+penguins_bill_depth <- penguins %>%
+  ggplot(aes(x = species, y = bill_depth_mm)) +
+  geom_boxplot(aes(color = species),
+               show.legend = FALSE) +
+  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
+  theme_bw() +
+  theme(plot.title.position = "plot") +
+  labs(x = "",
+       y = "Bill depth (mm)") +
+  coord_flip()
+
+penguins_box <- penguins_flip / penguins_mass / penguins_bill_length / penguins_bill_depth + plot_layout(guides = "collect")
+
+iris_box <- iris_pw / iris_pl / iris_sw / iris_sl + plot_layout(guides = "collect")
+
+penguins_box | iris_box + plot_annotation(
+  title = 'Penguin and iris sizes',
+  subtitle = 'Penguin size measurements (left) and iris petal and sepal dimensions (right)'
+)
+
+ggsave(here("figures","penguins_iris_box.png"), height = 7, width = 8)
+
 ```
+
+### Classification?
+
+### Other comparative analyses?
+
+## Selected activities
+
+### Hypothesis testing
+
+- t-tests
+- ANOVAs
+- MANOVAs
+
+### Linear regression
+
+- Application, diagnostics, interpretation & predictions
+
+## Appendix
+
+### Appendix A: Variables in the raw data
+
+- **studyName:** Sampling expedition indicator
+- **Sample Number:** Integer denoting the continuous numbering sequence for each penguin
+- **Species:** a factor denoting the penguin species (Adelie, Chinstrap, or Gentoo)
+- **Region**: A character denoting the region of the Palmer LTER sampling grid
+- **Island:** a factor denoting the island where observed (Torgersen, Dream or Biscoe) in Palmer Archipelago
+- **Stage:** A character string denoting reproductive stage at sampling
+- **Individual ID:** A character string denoting a unique ID for each penguin
+- **Clutch Completion:** A character string denoting if the study next observed with a full clutch, i.e. 2 eggs
+- **Date Egg:** A date denoting the date study nest observed with 1 egg
+- **Culmen Length:** A number denoting the length of the dorsal ridge of a bird's bill (millimeters)
+- **Culmen Depth:** A number denoting the depth of the dorsal ridge of a bird's bill (millimeters)
+- **Flipper Length:** An integer denoting the penguin flipper length (millimeters)
+- **Body Mass:** An integer denoting the penguin body mass (grams)
+- **Sex:** A character string denoting the sex of an animal
+- **Delta 15 N:** A number denoting the measure of the ratio of stable isotopes 15N:14N
+- **Delta 13 C:** A number denoting the measure of the ratio of stable isotopes 13C:12C
+- **Comments:** A character string with text providing additional relevant information for data
+
+## Citations
+
+## Random other stuff
 
@@
 
-#### Comparison 3: Principal components analysis
+## Differences by islands?
 
-## Appendix
+```{r}
+# No difference in Adelie across islands: Chinstrap only on Dream, Gentoo only on Torgerson
+penguins %>% 
+  ggplot(aes(x = island, y = body_mass_g)) +
+  geom_point() +
+  facet_wrap(~species)
 
-### Appendix A: Variables in the raw data
+penguins %>% 
+  ggplot(aes(x = island, y = flipper_length_mm)) +
+  geom_point() +
+  facet_wrap(~species)
 
-- **studyName:** Sampling expedition indicator
-- **Sample Number:** Integer denoting the continuous numbering sequence for each penguin
-- **Species:** a factor denoting the penguin species (Adelie, Chinstrap, or Gentoo)
-- **Region**: A character denoting the region of the Palmer LTER sampling grid
-- **Island:** a factor denoting the island where observed (Torgersen, Dream or Biscoe) in Palmer Archipelago
-- **Stage:** A character string denoting reproductive stage at sampling
-- **Individual ID:** A character string denoting a unique ID for each penguin
-- **Clutch Completion:** A character string denoting if the study next observed with a full clutch, i.e. 2 eggs
-- **Date Egg:** A date denoting the date study nest observed with 1 egg
-- **Culmen Length:** A number denoting the length of the dorsal ridge of a bird's bill (millimeters)
-- **Culmen Depth:** A number denoting the depth of the dorsal ridge of a bird's bill (millimeters)
-- **Flipper Length:** An integer denoting the penguin flipper length (millimeters)
-- **Body Mass:** An integer denoting the penguin body mass (grams)
-- **Sex:** A character string denoting the sex of an animal
-- **Delta 15 N:** A number denoting the measure of the ratio of stable isotopes 15N:14N
-- **Delta 13 C:** A number denoting the measure of the ratio of stable isotopes 13C:12C
-- **Comments:** A character string with text providing additional relevant information for data
+penguins %>% 
+  ggplot(aes(x = island, y = bill_length_mm)) +
+  geom_point() +
+  facet_wrap(~species)
 
-## Citations
+penguins %>% 
+  ggplot(aes(x = island, y = bill_depth_mm)) +
+  geom_point() +
+  facet_wrap(~species)
+```

### 2020-06-20 manuscript/manuscript_text.Rmd
text update 6/20 (#3)

@@
 library(ggbeeswarm)
+library(car)
 knitr::opts_chunk$set(echo = TRUE)
@@
 An overall approximately linear relationship can be explored using petal dimensions from `iris`, and penguin size (flipper length versus body mass) in `penguins` (Figure X).
-Educators may explore how linear regression across all species of iris or penguins compares to regression performed separately for each species.
+Educators may explore how linear regression across all species of iris or penguins (Figure X A,B) compares to regression performed separately for each species (Figure X C,D).
 
-```{r, echo = FALSE, warning = FALSE}
+```{r, echo = FALSE, warning = FALSE, message = FALSE}
 
 penguin_flip_mass_scatter <- ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
-  geom_point(aes(color = species), size = 2, alpha = 0.7) +
+  geom_point(aes(color = species, shape = species), size = 1.5, alpha = 0.7) +
+  geom_smooth(method = "lm", aes(group = species, color = species), se = FALSE) +
   theme_minimal() +
@@
   theme(legend.position = c(0.3, 0.8)) +
+  labs(x = "Flipper length (mm)", 
+       y = "Body mass (g)") +
+  theme(plot.title.position = "plot")
+
+penguin_flip_mass_scatter_all <- ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
+  geom_point(color = "gray50", size = 1.5, alpha = 0.5) +
+  geom_smooth(method = "lm", se = FALSE, color = "black") +
+  theme_minimal() +
   labs(x = "Flipper length (mm)", 
@@
 iris_petal_scatter <- ggplot(iris, aes(x = Petal.Length, y = Petal.Width)) +
-  geom_point(aes(color = Species), size = 2, alpha = 0.7) +
+  geom_point(aes(color = Species, size = Species, shape = Species), size = 1.5, alpha = 0.7) +
+  geom_smooth(aes(group = Species, color = Species), method = "lm", se = FALSE) +
   theme_minimal() +
@@
   theme(legend.position = c(0.3, 0.8)) +
+  labs(x = "Petal length (cm)",
+       y = "Petal width (cm)") +
+  theme(plot.title.position = "plot")
+
+iris_petal_scatter_all <- ggplot(iris, aes(x = Petal.Length, y = Petal.Width)) +
+  geom_point(color = "gray50", size = 1.5, alpha = 0.5) +
+  geom_smooth(method = "lm", se = FALSE, color = "black") +
+  theme_minimal() +
   labs(x = "Petal length (cm)",
@@
 
-penguin_flip_mass_scatter + iris_petal_scatter
+(penguin_flip_mass_scatter_all + iris_petal_scatter_all) / (penguin_flip_mass_scatter + iris_petal_scatter)
+
+ggsave(here("figures","iris_penguins_scatter.png"), height = 6, width = 6)
 ```
@@
 Notably, more obvious separation between species exists in `iris` (Figure X E & F), compared to any between-species comparisons in `penguins` (Figure X A - D).
+
+```{r, include = FALSE}
+
+# CHECK THIS ALL!!! 
+# ANOVA to see if flipper lengths are significantly different...(this says yes, but want other eyes)
+flipper_lm <- lm(flipper_length_mm ~ species, data = penguins)
+
+flipper_anova <- Anova(flipper_lm, type = "III")
+flipper_anova
+
+flip_aov <- aov(flipper_lm)
+summary(flip_aov)
+
+# TukeyHSD(flip_aov) all significantly different
+
+```
 
@@
   theme_bw() +
-  theme(plot.title.position = "plot") +
   labs(x = "",
@@
   scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme_bw() +
-  theme(plot.title.position = "plot") +
+  theme_bw()+
   labs(x = "",
@@
 
-penguins_box <- penguins_flip / penguins_mass / penguins_bill_length / penguins_bill_depth + plot_layout(guides = "collect")
+penguins_box <- penguins_flip / penguins_mass / penguins_bill_length / penguins_bill_depth + plot_layout(guides = "collect") + plot_annotation(title = "Penguins")
 
@@
   title = 'Penguin and iris sizes',
-  subtitle = 'Penguin size measurements (left) and iris petal and sepal dimensions (right)'
+  subtitle = 'Penguin size measurements (left) and iris petal and sepal dimensions (right)',
+  tag_levels = 'A'
 )
@@
 
-## Selected activities
-
 ### Hypothesis testing
 
-- t-tests
-- ANOVAs
-- MANOVAs
+- t-tests?
+- ANOVAs?
+- MANOVAs?
 
-### Linear regression
+## Suggested activities
 
-- Application, diagnostics, interpretation & predictions
+### Activity 1: Predict penguin body mass
 

### 2020-06-25 manuscript/manuscript_text.Rmd
add where else to get penguins

@@
 In 1935, American botanist Edgar Anderson measured petal and sepal dimensions (length and width) for 50 samples from three *Iris* species: *Iris setosa*, *Iris versicolor*, and *Iris virginica* [@anderson].
-The manageable but non-trivial size (5 variables and 150 total observations) and characteristics of the data (....SOMETHING HERE) make it amenable for introducing a wide range of statistical methods including data wrangling, visualization, hypothesis testing, regression, machine learning, and more.
-Anderson's *Iris* dataset is even built into a number of software packages, including in the auto-installed *datasets* package in *R*, Python's *scikit-learn* machine learning library, and the SAS *Sashelp* library, further facilitating its widespread use.
-As a result, eighty-five years after the data were initially published, Anderson’s Iris measurements are ubiquitous in statistics and data science materials.
+The manageable but non-trivial size (5 variables and 150 total observations) and characteristics of the data make it amenable for introducing a wide range of statistical methods including data wrangling, visualization, regression, classification and more.
+Anderson's *Iris* dataset is even built into a number of software packages, including in the auto-installed *datasets* package in *R* (R Core Team), Python's *scikit-learn* machine learning library [@scikit-learn], and the SAS *Sashelp* library, further facilitating its widespread use.
+As a result, eighty-five years after the data were initially published, Anderson’s Iris measurements are ubiquitous in statistics and data science education.
 
@@
 Fisher in *The Annals of Eugenics* [@fisher].
-Regardless of Anderson's initial motivation, the dataset today inextricably linked to Fisher’s eugenics research and even commonly (if unfairly) referred to as "*Fisher's* iris data" (ref).
-Due to its history in - and association with - Fisher's eugenics research, there is momentum to replace Iris with alternative datasets that offer similar teaching and learning opportunities.
+Regardless of Anderson's initial motivation, the dataset today remains inextricably linked to Fisher’s eugenics research and even commonly (if unfairly) referred to as "*Fisher's* iris data" (ref).
+Due to its history in - and strong association with - Fisher's eugenics research, there is momentum to replace the *Iris* data with alternative datasets that offer similar teaching and learning opportunities.
 
-Considering *Iris* usage, we established the following criteria for a suitable alternative dataset:
+Considering *Iris* data usage, we established the following criteria for a suitable alternative dataset:
 
@@
 - Feature intuitive subjects and variables understandable to learners across fields
-- Manageable, but not trivial, size
-- Real-world observations for improved understanding and engagement [@cafe]
+- Manageable - but not trivial - in size
+- Real-world (not manufactured) data for improved understanding and engagement [@cafe]
 - Provide similar opportunities for teaching data science skills and statistical methods
 
-Here, we offer an alternative to Anderson's irises that largely satisfies those criteria: an approachable and charismatic dataset containing real-world size measurements for three penguin species observed on islands in Palmer Archipelago, Antarctica.
+Here, we offer an alternative to Anderson's *Iris* data that largely satisfies those criteria: an approachable and charismatic dataset containing real-world size measurements for three penguin species observed on islands in Palmer Archipelago, Antarctica.
 Comparable size, characteristics and learning opportunities make the penguins data uniquely transferable for educators currently using *Iris*.
@@
 
-As a direct alternative to Anderson's *Iris* data, we recommend using the curated `penguins` object in the *palmerpenguins* data package.
+As a direct alternative to Anderson's *Iris* data, we recommend using the curated `penguins` object in the **palmerpenguins** *R* data package.
 The `penguins` data is a subset of `penguins_raw` retaining all 344 rows (individual penguins) but reduced to the following seven variables:
@@
 
-The `penguins` subset contains `r sum(complete.cases(penguins))` complete cases, with missing values (n~missing~ = `r sum(is.na(penguins))`, percent missing is `r round(100*sum(is.na(penguins))/(7*344),2)` %) indicated by `NA` throughout.
+The `penguins` subset contains `r sum(complete.cases(penguins))` complete cases, with missing values (n~missing~ = `r sum(is.na(penguins))`, or `r round(100*sum(is.na(penguins))/(7*344),2)` % of all values) indicated by `NA` throughout.
 In addition to reduced variables, the curated `penguins` subset is minimally updated from `penguins_raw` as follows:
@@
 Comparisons presented here are representative examples, and are not exhaustive.
+Additional examples and tutorials are shared in the [user-contributed examples](https://allisonhorst.github.io/palmerpenguins/articles/articles/user_contributions.html) article on the package website.
 
@@
 
-### Correlation
+### Linear correlation & regression opportunities
 
@@
 #### Comparison 2: linear relationships within species
+
+More obvious clustering (by penguin species) exists when comparing bill length and depth (Figure X), and the relationship also provides an opportunity to introduce & explore Simpson's Paradox.
 
@@
 
-### Classification?
-
-### Other comparative analyses?
-
 ### Hypothesis testing
 
-- t-tests?
-- ANOVAs?
-- MANOVAs?
+### Principle components analysis
 
-## Suggested activities
+### Clustering (e.g. k-means?)
 
-### Activity 1: Predict penguin body mass
+### Classification (Alison Hill blog post?)
+
+## Penguin data access for non-R users
+
+The Palmer Archipelago penguins data are available to non-R users by direct import from the Environmental Data Initiative, from GitHub, and as a stored dataset in existing packages.
+Below we share several current ways to access the data (as of July 2020 - we expect availability to expand in the future).
+
+### Download or import directly from Environmental Data Initiative
+
+The [Environmental Data Initiative](https://environmentaldatainitiative.org/) is "an NSF-funded project, actively promoting and enabling curation and re-use of environmental data." The raw data for each penguin species (Adélie, Gentoo and Chinstrap) are stored in a separate data file at the following links (see full citations in Citations):
+
+- Adélie data: [KNB-LTER Data Package 219.5](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.219.5)
+- Gentoo data: [KNB-LTER Data Package 220.5](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.220.5)
+- Chinstrap data: [KNB-LTER Data Package 221.5](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.221.6)
+
+The links above contain all metadata information, and provides options to directly download the data as a comma-separated value (CSV) file *or* import for use directly using code provided (e.g.
+for R, Python, SAS, MatLab, or SPSS).
+
+### Download CSV from the palmerpenguins GitHub repo
+
+A CSV containing curated data for all three species (accessed directly from EDI) exists in the 'data-raw' folder in the palmerpenguins repo on GitHub.
+GitHub users can fork / clone the repo, and will then have the cleaned CSV containing all 3 penguin species.
+
+### The penguins in other packages & languages
+
+The Palmer Station penguin data has been added to existing libraries, including (non-exhaustive):
+
+#### Python modules
+
+- Seaborn
+- blah blah
+
+See examples for working with penguins in Python here:
+
+#### Kaggle
+
+- here
+
+#### Elsewhere?
+
+## Conclusion
 
@@
 
-## Citations
-
-## Random other stuff
+# Random other stuff
 
@@
 ```
+
+A Web of Science search reveals that Fisher's 1936 paper has been cited in 7,258 publications (as of June 2020), most often in journals publishing in fields of Computer Science (57% of 7,258 citing references), Mathematics (49%), and Engineering (29%) - most of which we expect are crediting Anderson's *Iris* data, and not Fisher's eugenics work.
+
+# End random other stuff
+
+## Citations

### 2020-06-27 manuscript/manuscript_text.Rmd
add pca & kmeans drafts

@@
 library(car)
+library(GGally)
+library(recipes)
+library(tidytext)
+library(ggrepel)
+library(shadowtext)
+library(glue)
+library(ggdendro)
 knitr::opts_chunk$set(echo = TRUE)
+
+theme_set(theme_minimal())
 ```
@@
 
-Data for each penguin species (Adélie, chinstrap and gentoo) were imported separately, then combined into a single table stored in the [*palmerpenguins*](https://allisonhorst.github.io/palmerpenguins/) package as object `penguins_raw`.
+For easier data access and use we created the [**palmerpenguins**](https://allisonhorst.github.io/palmerpenguins/) data package in *R*, which contains two data objects:
+
+- `penguins_raw`: a table containing all raw data 
+- `penguins`: a simplified table containing only size measurements, sex, island and species
+
+The penguins data can also be imported directly from EDI, or accessed elsewhere (see *Penguin access for non-R users* section for more options).
+
+### Uncurated `penguins_raw` data
+
+Data for each penguin species (Adélie, chinstrap and gentoo) were imported from EDI, then combined into a single table stored in the [*palmerpenguins*](https://allisonhorst.github.io/palmerpenguins/) package as object `penguins_raw`.
 Data in `penguins_raw` are in tidy format [@tidydata] and contain observations for 17 variables (Appendix Table 1) recorded completely or in part for 344 individual penguins.
+For full variable descriptions and metadata, visit the data packages through EDI.
 
@@
 Comparisons presented here are representative examples, and are not exhaustive.
-Additional examples and tutorials are shared in the [user-contributed examples](https://allisonhorst.github.io/palmerpenguins/articles/articles/user_contributions.html) article on the package website.
+Additional examples and tutorials are shared in the [user-contributed examples](https://allisonhorst.github.io/palmerpenguins/articles/articles/user_contributions.html) article on the **palmerpenguins** package website.
 
@@
 
-### Linear correlation & regression opportunities
+### Overview comparison: numeric variables
 
-#### Comparison 1: Linear trends across or within species
+Relationships, distributions and correlations can be explored for the four numeric variables in `iris` (sepal width and length, petal width and length) and `penguins` (flipper length, body mass, bill length and depth) in pair plots (Figure X A,B).
 
-An overall approximately linear relationship can be explored using petal dimensions from `iris`, and penguin size (flipper length versus body mass) in `penguins` (Figure X).
-Educators may explore how linear regression across all species of iris or penguins (Figure X A,B) compares to regression performed separately for each species (Figure X C,D).
+```{r, warning = FALSE, message = FALSE, echo = FALSE, fig.align = "center"}
+penguins %>%
+  select(species, where(is.numeric)) %>% 
+  ggpairs(aes(color = species),
+          columns = c("flipper_length_mm", "body_mass_g", 
+                      "bill_length_mm", "bill_depth_mm")) +
+  scale_colour_manual(values = c("darkorange","purple","cyan4")) +
+  scale_fill_manual(values = c("darkorange","purple","cyan4")) +
+  theme_minimal() +
+  theme(text = element_text(size = 10))
 
-```{r, echo = FALSE, warning = FALSE, message = FALSE}
+ggsave(here("figures","penguins_ggpairs.png"), height = 6, width = 5)
+
+iris %>%
+  ggpairs(aes(color = Species),
+          columns = c("Petal.Length", "Petal.Width", 
+                      "Sepal.Length", "Sepal.Width")) +
+  scale_colour_manual(values = c("darkorange","purple","cyan4")) +
+  scale_fill_manual(values = c("darkorange","purple","cyan4")) +
+  theme_minimal() +
+  theme(text = element_text(size = 10),
+        panel.background = element_rect(fill = "gray90", color = "gray90"))
+
+ggsave(here("figures","iris_ggpairs.png"), height = 6, width = 5)
+
+```
+
+### Linear correlation & regression
+
+As seen in Figures (X,X) there are a number of opportunities to explore linear relationships and correlations in both `penguins` and `iris`.
+Here, we point out one comparison that is uniquely similar:
+petal dimensions (petal length versus petal width) from `iris`, and penguin size (flipper length versus body mass) in `penguins` (Figure X).
+We see that the overall trend across species is relatively linear for both `iris` and `penguins` (Figure X A,B).
+Teachers may encourage students explore how results of regression performed across all three species of iris or penguins (Figure X A,B) compare to results for regression performed within species (Figure X C,D).
+
+Notably, distinctions between species are clearer for `iris` petals - particularly, the much smaller petals for *Iris setosa* - compared to `penguins`, in which Adélie and chinstrap penguins are largely overlapping in body size (body mass and flipper length), and are both generally smaller than gentoos.
+This difference is revisited in the k-means clustering example below.
+
+```{r, echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center"}
 
@@
        y = "Petal width (cm)") +
-  theme(plot.title.position = "plot")
+  theme(plot.title.position = "plot",
+        panel.background = element_rect(fill = "gray90", color = "gray90"))
 
@@
        title = "Iris petal dimensions") +
-  theme(plot.title.position = "plot")
+  theme(plot.title.position = "plot",
+        panel.background = element_rect(fill = "gray90", color = "gray90"))
 
@@
 
-#### Comparison 2: linear relationships within species
+### Hypothesis testing
 
-More obvious clustering (by penguin species) exists when comparing bill length and depth (Figure X), and the relationship also provides an opportunity to introduce & explore Simpson's Paradox.
-
-```{r, echo = FALSE, warning = FALSE, message = FALSE}
-penguin_bill_scatter <- ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
-  geom_point(aes(color = species), size = 2, alpha = 0.7) +
-  theme_minimal() +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme(legend.position = "bottom") +
-  labs(x = "Bill length (mm)", 
-       y = "Bill depth (g)",
-       title = "Penguin size", 
-       color = "Species") +
-  theme(plot.title.position = "plot")
-
-iris_sepal_scatter <- ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) +
-  geom_point(aes(color = Species), size = 2, alpha = 0.7) +
-  theme_minimal() +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme(legend.position = "bottom") +
-  labs(x = "Sepal length (cm)",
-       y = "Sepal width (cm)",
-       title = "Iris sepal dimensions") +
-  theme(plot.title.position = "plot")
-
-penguin_bill_scatter + iris_sepal_scatter
-```
-
-### Differences between groups
-
-As in `iris`, multiple opportunities for comparisons between groups also exist in `penguins` (Figure X).
-Notably, more obvious separation between species exists in `iris` (Figure X E & F), compared to any between-species comparisons in `penguins` (Figure X A - D).
-
-```{r, include = FALSE}
-
-# CHECK THIS ALL!!! 
-# ANOVA to see if flipper lengths are significantly different...(this says yes, but want other eyes)
-flipper_lm <- lm(flipper_length_mm ~ species, data = penguins)
-
-flipper_anova <- Anova(flipper_lm, type = "III")
-flipper_anova
-
-flip_aov <- aov(flipper_lm)
-summary(flip_aov)
-
-# TukeyHSD(flip_aov) all significantly different
-
-```
-
-```{r, echo = FALSE, warning = FALSE, message = FALSE}
-
-# Update this with an actual theme FFS and use '&' in patchwork
-
-iris_pw <- ggplot(iris, aes(x = Species, y = Petal.Width)) +
-  geom_boxplot(aes(color = Species),
-               show.legend = FALSE) +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme_bw() +
-  theme(plot.title.position = "plot") +
-  labs(x = "",
-       y = "Petal width (cm)") +
-  coord_flip()
-
-iris_pl <- ggplot(iris, aes(x = Species, y = Petal.Length)) +
-  geom_boxplot(aes(color = Species),
-               show.legend = FALSE) +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme_bw() +
-  theme(plot.title.position = "plot") +
-  labs(x = "",
-       y = "Petal length (cm)") +
-  coord_flip()
-
-iris_sl <- ggplot(iris, aes(x = Species, y = Sepal.Length)) +
-  geom_boxplot(aes(color = Species),
-               show.legend = FALSE) +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme_bw() +
-  theme(plot.title.position = "plot") +
-  labs(x = "",
-       y = "Sepal length (cm)") +
-  coord_flip()
-
-iris_sw <- ggplot(iris, aes(x = Species, y = Sepal.Width)) +
-  geom_boxplot(aes(color = Species),
-               show.legend = FALSE) +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme_bw() +
-  theme(plot.title.position = "plot") +
-  labs(x = "",
-       y = "Sepal width (cm)") +
-  coord_flip()
-
-penguins_flip <- penguins %>%
-  ggplot(aes(x = species, y = flipper_length_mm)) +
-  geom_boxplot(aes(color = species),
-               show.legend = FALSE) +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme_bw() +
-  labs(x = "",
-       y = "Flipper length (mm)") +
-  coord_flip()
-
-penguins_mass <- penguins %>%
-  ggplot(aes(x = species, y = body_mass_g)) +
-  geom_boxplot(aes(color = species),
-               show.legend = FALSE) +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme_bw() +
-  labs(x = "",
-       y = "Body mass (g)") +
-  coord_flip()
-
-penguins_bill_length <- penguins %>%
-  ggplot(aes(x = species, y = bill_length_mm)) +
-  geom_boxplot(aes(color = species),
-               show.legend = FALSE) +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme_bw()+
-  labs(x = "",
-       y = "Bill length (mm)") +
-  coord_flip()
-
-penguins_bill_depth <- penguins %>%
-  ggplot(aes(x = species, y = bill_depth_mm)) +
-  geom_boxplot(aes(color = species),
-               show.legend = FALSE) +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme_bw() +
-  labs(x = "",
-       y = "Bill depth (mm)") +
-  coord_flip()
-
-penguins_box <- penguins_flip / penguins_mass / penguins_bill_length / penguins_bill_depth + plot_layout(guides = "collect") + plot_annotation(title = "Penguins")
-
-iris_box <- iris_pw / iris_pl / iris_sw / iris_sl + plot_layout(guides = "collect")
-
-penguins_box | iris_box + plot_annotation(
-  title = 'Penguin and iris sizes',
-  subtitle = 'Penguin size measurements (left) and iris petal and sepal dimensions (right)',
-  tag_levels = 'A'
-)
-
-ggsave(here("figures","penguins_iris_box.png"), height = 7, width = 8)
-
-```
-
-### Hypothesis testing
+### Linear discriminant analysis?
 
@@
 
-### Clustering (e.g. k-means?)
+The `iris` dataset appears frequently in examples for multivariate analyses, including principal component analysis (PCA).
+PCA is a dimensional reduction method that can be useful in capturing as much information about patterns in multivariate data as possible in a reduced number of dimensions (e.g., 2-dimensions for visualization).
+PCA with `iris` yields intuitive PC outcomes and clear clustering (Figure X) helpful for learners, as described in detail by others (ref, ref).
 
-### Classification (Alison Hill blog post?)
+PCA with `penguins` reveals a similar outcome (Figure X): two somewhat overlapping species (Adélie and Chinstrap) with one clearly separated (Gentoo), and intuitive loadings that can be related to scatterplots in Figure X.
+Percent of total variance captured by the first and second principal components are very similar between `iris` and `penguins` (Figure X).
+
+```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.align = "center"}
+# Copied from Alison Hill's post!
+
+### PENGUINS PCA: 
+penguin_recipe <-
+  recipe(~., data = penguins) %>% 
+  update_role(species, island, sex, new_role = "id") %>% 
+  step_naomit(all_predictors()) %>% 
+  step_normalize(all_predictors()) %>%
+  step_pca(all_predictors(), id = "pca") %>% 
+  prep()
+
+penguin_pca <- 
+  penguin_recipe %>% 
+  tidy(id = "pca")
+
+penguin_percvar <- penguin_recipe %>% 
+  tidy(id = "pca", type = "variance") %>% 
+  dplyr::filter(terms == "percent variance")
+
+# penguin_pca
+
+# Biplot:
+
+# get pca loadings into wider format
+pca_wider <- penguin_pca %>% 
+  tidyr::pivot_wider(names_from = component, id_cols = terms)
+
+# define arrow style:
+arrow_style <- arrow(length = unit(.05, "inches"),
+                     type = "closed")
+
+# Make the biplot:
+pca_plot <-
+  juice(penguin_recipe) %>%
+  ggplot(aes(PC1, PC2)) +
+  geom_point(aes(color = species, shape = species), 
+             alpha = 0.8, 
+             size = 2) +
+  scale_colour_manual(values = c("darkorange","purple","cyan4"))
+
+penguins_biplot <- pca_plot +
+  geom_segment(data = pca_wider,
+               aes(xend = PC1, yend = PC2), 
+               x = 0, 
+               y = 0, 
+               arrow = arrow_style) + 
+  geom_shadowtext(data = pca_wider,
+            aes(x = PC1, y = PC2, label = terms), 
+            nudge_x = c(0.3,0.3,1.0,0.8), 
+            nudge_y = c(-0.1,-0.2,0.1,-0.1),
+            size = 3, 
+            color = '#0A537D',
+            bg.color = "white") +
+  labs(title = "Penguins PCA biplot") +
+  theme(legend.position = "bottom")
+
+# For positioning (above): 
+# 1: bill_length
+# 2: bill_depth
+# 3: flipper length
+# 4: body mass
+
+penguin_screeplot <- penguin_percvar %>% 
+  ggplot(aes(x = component, y = value)) + 
+  geom_col(fill = "#b6dfe2") + 
+  xlim(c(0, 5)) + 
+  ylim(c(0,80)) +
+  ylab("% of total variance") +
+  labs(title = "Penguins PCA screeplot") +
+  geom_text(aes(label = round(value,2)), vjust=-0.25)
+
+### IRIS PCA: 
+iris_recipe <-
+  recipe(~., data = iris) %>% 
+  update_role(Species, new_role = "id") %>% 
+  step_normalize(all_predictors()) %>%
+  step_pca(all_predictors(), id = "pca") %>% 
+  prep()
+
+iris_pca <- 
+  iris_recipe %>% 
+  tidy(id = "pca")
+
+# Biplot:
+
+# get pca loadings into wider format
+iris_wider <- iris_pca %>% 
+  tidyr::pivot_wider(names_from = component, id_cols = terms)
+
+# define arrow style:
+arrow_style <- arrow(length = unit(.05, "inches"),
+                     type = "closed")
+
+# Make the biplot:
+iris_pca_plot <-
+  juice(iris_recipe) %>%
+  ggplot(aes(PC1, PC2)) +
+  geom_point(aes(color = Species, shape = Species), 
+             alpha = 0.8, 
+             size = 2) +
+  scale_colour_manual(values = c("darkorange","purple","cyan4"))
+
+iris_biplot <- iris_pca_plot +
+  geom_segment(data = iris_wider,
+               aes(xend = PC1, yend = PC2), 
+               x = 0, 
+               y = 0, 
+               arrow = arrow_style) + 
+  geom_shadowtext(data = iris_wider,
+            aes(x = PC1, y = PC2, label = terms), 
+            nudge_x = c(0.3,0.3,0.7,0.6), 
+            nudge_y = c(-0.1,-0.2,0.1,-0.1),
+            size = 3, 
+            color = '#0A537D',
+            bg.color = "white") +
+  labs(title = "Iris PCA biplot") +
+  theme(panel.background = element_rect(fill = "gray90", color = "gray90"),
+        legend.position = "bottom")
+
+iris_percvar <- iris_recipe %>% 
+  tidy(id = "pca", type = "variance") %>% 
+  dplyr::filter(terms == "percent variance")
+
+iris_screeplot <- iris_percvar %>% 
+  ggplot(aes(x = component, y = value)) + 
+  geom_col(fill = "#b6dfe2") + 
+  xlim(c(0, 5)) + 
+  ylim(c(0,80)) +
+  ylab("% of total variance") +
+  labs(title = "Iris PCA screeplot") + 
+  geom_text(aes(label = round(value,2)), vjust=-0.25) +
+  theme(panel.background = element_rect(fill = "gray90", color = "gray90"))
+
+(penguins_biplot | iris_biplot)/(penguin_screeplot | iris_screeplot)
+
+ggsave(here("figures","pca_plots.png"), width = 8, height = 8)
+```
+
+### K-means clustering
+
+Unsupervised clustering by k-means is a common and popular entryway to machine learning and classification, and frequently employs the `iris` data for introductory examples (refs).
+The `penguins` data provides similar opportunities for demonstrating, and illustrating limitations of, k-means clustering.
+
+#### k-means clustering: two variables
+
+First, we compare k-means clustering using two variables for each dataset: for `iris`, petal width and petal length, and for a subset only including male `penguins`, bill length and bill depth.
+For each, we designate three centroids (*k* = 3) since there are three species of both *Iris* and penguins.
+
+Clustering by k-means with *Iris* petal dimension results in perfect separation for *Iris setosa* observations: all 50 *setosa* exist in a single cluster (cluster 1) with no *Iris virginica* or *versicolor* assigned to the cluster (Figure X, Table X).
+There is minimal species misclassification between the *versicolor* (cluster 2) and *versicolor* (cluster 3) species (Table X).
+
+Similarly, Adélie penguins are highly separated from Chinstrap and Gentoo penguins by k-means clustering using only bill length and bill depth: of 73 Adélie penguins in the data, 70 are assigned to a single cluster (Cluster 2), and only 3 are assigned to the cluster containing most Gentoos (Cluster 3).
+No Chinstraps or Gentoos are assigned to Cluster 2 (Figure X).
+The distinction between Chinstraps and Gentoos is less clear, in agreement with overlap in bill dimensions between the species observed visually (Figure X).
+
+```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.align = "center"}
+# TWO VARIABLE k-means comparison
+# Male penguins: bill length vs. bill depth
+# Iris: petal length vs. petal width
+
+penguins_k <- penguins %>% 
+  filter(sex == "male") %>% 
+  select(species, bill_length_mm:body_mass_g) %>% 
+  drop_na()
+
+# Using bill_length & flipper_length
+# THIS ONE SHOWS *pretty* good clustering
+
+bills <- penguins_k %>% select(species, bill_length_mm, bill_depth_mm)
+
+set.seed(1967)
+kclust_bills <- kmeans(bills[,2:3], centers = 3, nstart = 20)
+# kclust_bills
+
+# Table of assigned clusters for penguins
+bills_class <- table(kclust_bills$cluster,bills$species)%>% 
+  as.data.frame() %>% 
+  rename(cluster = Var1, species = Var2, n = Freq) %>% 
+  pivot_wider(names_from = species, values_from = n)
+
+# Same for iris:
+
+petals <- iris %>% select(Species, Petal.Length, Petal.Width)
+
+set.seed(1967)
+kclust_petals <- kmeans(petals[,2:3], centers = 3, nstart = 25)
+# kclust_petals
+
+# Table of assigned clusters for iris petals: 
+petal_class <- table(kclust_petals$cluster, petals$Species) %>% 
+  as.data.frame() %>% 
+  rename(cluster = Var1, Species = Var2, n = Freq) %>% 
+  pivot_wider(names_from = Species, values_from = n)
+
+bills_cluster <- cbind(bills, cluster_no = kclust_bills$cluster)
+
+penguins_kmeans <- ggplot(data = bills_cluster, aes(x = bill_length_mm, 
+                                     bill_depth_mm)) +
+  geom_text(aes(label=cluster_no, color = species), 
+                  size = 4) +
+  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
+  labs(title = "Penguins k-means clustering") +
+  theme(legend.position = "bottom")
+
+iris_cluster <-cbind(petals, cluster_no = kclust_petals$cluster)
+
+iris_kmeans <- ggplot(data = iris_cluster, aes(x = Petal.Length, y = Petal.Width)) +
+  geom_text(aes(label = cluster_no, color = Species), size = 4) +
+  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
+  theme(panel.background = element_rect(fill = "gray90", color = "gray90"),
+        legend.position = "bottom") +
+  labs(title = "Iris k-means clustering")
+
+(penguins_kmeans | iris_kmeans)
+
+# Full table of cluster assignments: 
+combo_cluster <- cbind(bills_class, petal_class) %>% 
+  kable(caption = "Table X. Assigned k-means clusters for results shown in Figure X") %>% 
+  kable_styling(full_width = FALSE) %>% 
+  add_header_above(c("Penguin bill cluster assignments" = 4, "Iris petal cluster assignments" = 4)) %>% 
+  column_spec(c(5:8), background = "gainsboro")
+
+combo_cluster
+```
+#### k-means clustering: all numeric variables
+
+When all four numeric variables for each dataset are included in k-means clustering, *setosa* irises remain distinctly clustered apart from the other two iris species, but there is overlap across all penguin species (Table X).
+
+```{r, echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center"}
+# FOUR VARIABLE k-means comparison
+
+penguins_k_full <- penguins %>% 
+  select(species, bill_length_mm:body_mass_g) %>% 
+  drop_na()
+
+set.seed(1967)
+kclust_penguins <- kmeans(penguins_k_full[,2:5], centers = 3, nstart = 20)
+
+# Table of assigned clusters for penguins
+penguins_class <- table(kclust_penguins$cluster,penguins_k_full$species)%>% 
+  as.data.frame() %>% 
+  rename(cluster = Var1, species = Var2, n = Freq) %>% 
+  pivot_wider(names_from = species, values_from = n)
+
+# penguins_class
+# Same for iris:
+
+set.seed(1967)
+iris_k_full <- kmeans(iris[,1:4], centers = 3, nstart = 25)
+
+# Table of assigned clusters for iris petals: 
+iris_class <- table(iris_k_full$cluster, iris$Species) %>% 
+  as.data.frame() %>% 
+  rename(cluster = Var1, Species = Var2, n = Freq) %>% 
+  pivot_wider(names_from = Species, values_from = n)
+
+# iris_class
+
+# Full table of cluster assignments with all 4 numeric variables: 
+combo_cluster_full <- cbind(penguins_class, iris_class) %>% 
+  kable(caption = "Table X. Assigned k-means clusters for iris and penguins when including all four numeric variables for each") %>% 
+  kable_styling(full_width = FALSE) %>% 
+  add_header_above(c("Penguin cluster assignments" = 4, "Iris cluster assignments" = 4)) %>% 
+  column_spec(c(5:8), background = "gainsboro")
+
+combo_cluster_full
+```
 
@@
 
+Random, maybe include somewhere?
 A Web of Science search reveals that Fisher's 1936 paper has been cited in 7,258 publications (as of June 2020), most often in journals publishing in fields of Computer Science (57% of 7,258 citing references), Mathematics (49%), and Engineering (29%) - most of which we expect are crediting Anderson's *Iris* data, and not Fisher's eugenics work.
+
+### Differences between groups
+
+As in `iris`, multiple opportunities for comparisons between groups also exist in `penguins` (Figure X).
+Notably, more obvious separation between species exists in `iris` (Figure X E & F), compared to any between-species comparisons in `penguins` (Figure X A - D).
+
+```{r, include = FALSE}
+
+# CHECK THIS ALL!!! 
+# ANOVA to see if flipper lengths are significantly different...(this says yes, but want other eyes)
+flipper_lm <- lm(flipper_length_mm ~ species, data = penguins)
+
+flipper_anova <- Anova(flipper_lm, type = "III")
+flipper_anova
+
+flip_aov <- aov(flipper_lm)
+summary(flip_aov)
+
+# TukeyHSD(flip_aov) all significantly different
+
+```
+
+```{r, echo = FALSE, warning = FALSE, message = FALSE}
+
+# Update this with an actual theme FFS and use '&' in patchwork
+
+iris_pw <- ggplot(iris, aes(x = Species, y = Petal.Width)) +
+  geom_boxplot(aes(color = Species),
+               show.legend = FALSE) +
+  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
+  theme_bw() +
+  theme(plot.title.position = "plot") +
+  labs(x = "",
+       y = "Petal width (cm)") +
+  coord_flip()
+
+iris_pl <- ggplot(iris, aes(x = Species, y = Petal.Length)) +
+  geom_boxplot(aes(color = Species),
+               show.legend = FALSE) +
+  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
+  theme_bw() +
+  theme(plot.title.position = "plot") +
+  labs(x = "",
+       y = "Petal length (cm)") +
+  coord_flip()
+
+iris_sl <- ggplot(iris, aes(x = Species, y = Sepal.Length)) +
+  geom_boxplot(aes(color = Species),
+               show.legend = FALSE) +
+  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
+  theme_bw() +
+  theme(plot.title.position = "plot") +
+  labs(x = "",
+       y = "Sepal length (cm)") +
+  coord_flip()
+
+iris_sw <- ggplot(iris, aes(x = Species, y = Sepal.Width)) +
+  geom_boxplot(aes(color = Species),
+               show.legend = FALSE) +
+  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
+  theme_bw() +
+  theme(plot.title.position = "plot") +
+  labs(x = "",
+       y = "Sepal width (cm)") +
+  coord_flip()
+
+penguins_flip <- penguins %>%
+  ggplot(aes(x = species, y = flipper_length_mm)) +
+  geom_boxplot(aes(color = species),
+               show.legend = FALSE) +
+  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
+  theme_bw() +
+  labs(x = "",
+       y = "Flipper length (mm)") +
+  coord_flip()
+
+penguins_mass <- penguins %>%
+  ggplot(aes(x = species, y = body_mass_g)) +
+  geom_boxplot(aes(color = species),
+               show.legend = FALSE) +
+  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
+  theme_bw() +
+  labs(x = "",
+       y = "Body mass (g)") +
+  coord_flip()
+
+penguins_bill_length <- penguins %>%
+  ggplot(aes(x = species, y = bill_length_mm)) +
+  geom_boxplot(aes(color = species),
+               show.legend = FALSE) +
+  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
+  theme_bw()+
+  labs(x = "",
+       y = "Bill length (mm)") +
+  coord_flip()
+
+penguins_bill_depth <- penguins %>%
+  ggplot(aes(x = species, y = bill_depth_mm)) +
+  geom_boxplot(aes(color = species),
+               show.legend = FALSE) +
+  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
+  theme_bw() +
+  labs(x = "",
+       y = "Bill depth (mm)") +
+  coord_flip()
+
+penguins_box <- penguins_flip / penguins_mass / penguins_bill_length / penguins_bill_depth + plot_layout(guides = "collect") + plot_annotation(title = "Penguins")
+
+iris_box <- iris_pw / iris_pl / iris_sw / iris_sl + plot_layout(guides = "collect")
+
+penguins_box | iris_box + plot_annotation(
+  title = 'Penguin and iris sizes',
+  subtitle = 'Penguin size measurements (left) and iris petal and sepal dimensions (right)',
+  tag_levels = 'A'
+)
+
+ggsave(here("figures","penguins_iris_box.png"), height = 7, width = 8)
+
+```
+
+### Hierarchical clustering?
+
+```{r}
+hc_iris <- hclust(dist(iris[, 1:4]))
+ggdendrogram(hc_iris, rotate = TRUE, size = 2)
+
+p_cc <- penguins %>% 
+  drop_na()
+
+hc_penguins <- hclust(dist(p_cc[, 3:6]))
+ggdendrogram(hc_penguins, rotate = TRUE, size = 2)
+
+```
 

### 2020-06-27 code_tests/kmeans_testing.Rmd

--- 2020-06-20 manuscript/manuscript_text.Rmd
@@
 ---
-title: "Palmer Archipelago (Antarctica) penguins: an alternative to Edgar Anderson's iris data"
+title: "k-means clustering with penguins"
+author: "Allison Horst"
+date: "6/26/2020"
 output: html_document
-bibliography: ../bibliography/references.bib
 ---
@@
 library(tidyverse)
-library(here)
+library(tidymodels)
 library(palmerpenguins)
+library(GGally)
+library(shadowtext)
+library(NbClust)
+library(patchwork)
 library(kableExtra)
-library(patchwork)
-library(ggExtra)
-library(gt)
-library(ggpubr)
-library(ggbeeswarm)
-library(car)
 knitr::opts_chunk$set(echo = TRUE)
@@
 
-## Authors
+Following exactly along with: https://www.tidymodels.org/learn/statistics/k-means/
 
-Allison Horst, Alison Hill, Kristen Gorman (order TBD)
+The "old" way, using `kmeans()`:
+```{r}
+penguins_k <- penguins %>% 
+  filter(sex == "male") %>% 
+  select(species, bill_length_mm:body_mass_g) %>% 
+  drop_na()
 
-## Keywords
+# Using bill_length & flipper_length
+# THIS ONE SHOWS *pretty* good clustering
 
-## Abstract
+bills <- penguins_k %>% select(species, bill_length_mm, bill_depth_mm)
 
-Edgar Anderson's *Iris* data are ubiquitous in statistics and data science education, but their use in Ronald A.
-Fisher's eugenics research has created momentum to switch to alternative datasets.
-Size measurements for three penguin species observed on islands in Palmer Archipelago, Antarctica provide a (mostly) suitable, approachable and charismatic alternative to the *Iris* data.
-Here, we introduce the penguins data, then directly compare the two datasets to demonstrate that educators currently using *Iris* can switch to the Palmer Archipelago penguins for many use cases, particularly in introductory statistics and data science courses.
+set.seed(1967)
+kclust_bills <- kmeans(bills[,2:3], centers = 3, nstart = 20)
+kclust_bills
 
-## Introduction
+# Table of assigned clusters for penguins
+bills_class <- table(kclust_bills$cluster,bills$species)%>% 
+  as.data.frame() %>% 
+  rename(cluster = Var1, species = Var2, n = Freq) %>% 
+  pivot_wider(names_from = species, values_from = n)
 
-In 1935, American botanist Edgar Anderson measured petal and sepal dimensions (length and width) for 50 samples from three *Iris* species: *Iris setosa*, *Iris versicolor*, and *Iris virginica* [@anderson].
-The manageable but non-trivial size (5 variables and 150 total observations) and characteristics of the data (....SOMETHING HERE) make it amenable for introducing a wide range of statistical methods including data wrangling, visualization, hypothesis testing, regression, machine learning, and more.
-Anderson's *Iris* dataset is even built into a number of software packages, including in the auto-installed *datasets* package in *R*, Python's *scikit-learn* machine learning library, and the SAS *Sashelp* library, further facilitating its widespread use.
-As a result, eighty-five years after the data were initially published, Anderson’s Iris measurements are ubiquitous in statistics and data science materials.
+# Same for iris:
 
-*Sounds great. Why consider switching?*
+petals <- iris %>% select(Species, Petal.Length, Petal.Width)
 
-In 1936, one year after publication in the *Bulletin of the American Iris Society*, Anderson’s *Iris* measurements were further analyzed and published by eugenicist and statistician Ronald A.
-Fisher in *The Annals of Eugenics* [@fisher].
-Regardless of Anderson's initial motivation, the dataset today inextricably linked to Fisher’s eugenics research and even commonly (if unfairly) referred to as "*Fisher's* iris data" (ref).
-Due to its history in - and association with - Fisher's eugenics research, there is momentum to replace Iris with alternative datasets that offer similar teaching and learning opportunities.
+set.seed(1967)
+kclust_petals <- kmeans(petals[,2:3], centers = 3, nstart = 25)
+kclust_petals
 
-Considering *Iris* usage, we established the following criteria for a suitable alternative dataset:
-
-- Available by an appropriate license, ideally CC0 (no rights reserved)
-- Feature intuitive subjects and variables understandable to learners across fields
-- Manageable, but not trivial, size
-- Real-world observations for improved understanding and engagement [@cafe]
-- Provide similar opportunities for teaching data science skills and statistical methods
-
-Here, we offer an alternative to Anderson's irises that largely satisfies those criteria: an approachable and charismatic dataset containing real-world size measurements for three penguin species observed on islands in Palmer Archipelago, Antarctica.
-Comparable size, characteristics and learning opportunities make the penguins data uniquely transferable for educators currently using *Iris*.
-By highlighting similarities in data structure and analyses, as well as several examples for which the penguins are *not* a good alternative, we demonstrate that the penguins can replace irises for many use cases.
-
-## Data source
-
-Size measurements, clutch observations, and blood isotope levels for male and female nesting adult Adélie (*Pygoscelis adeliae*), chinstrap (*Pygoscelis antarcticus*), and gentoo (*Pygoscelis papua*) penguins on three islands in the Palmer Archipelago were collected from 2007 - 2009 by Dr.
-Kristen Gorman with the [Palmer Station Long Term Ecological Research Program](https://pal.lternet.edu/), part of the [US Long Term Ecological Research Network](https://lternet.edu/).
-For complete data collection methods see [@gorman_plos].
-The data were imported directly from the [Environmental Data Initiative](https://environmentaldatainitiative.org/) (EDI) Data Portal [@adelie_edi; @gentoo_edi; @chinstrap_edi] and are available for use by CC0 license ("No Rights Reserved") in accordance with the [Palmer Station Data Policy](https://pal.lternet.edu/data/policies).
-
-Data for each penguin species (Adélie, chinstrap and gentoo) were imported separately, then combined into a single table stored in the [*palmerpenguins*](https://allisonhorst.github.io/palmerpenguins/) package as object `penguins_raw`.
-Data in `penguins_raw` are in tidy format [@tidydata] and contain observations for 17 variables (Appendix Table 1) recorded completely or in part for 344 individual penguins.
-
-### Curated `penguins` data
-
-As a direct alternative to Anderson's *Iris* data, we recommend using the curated `penguins` object in the *palmerpenguins* data package.
-The `penguins` data is a subset of `penguins_raw` retaining all 344 rows (individual penguins) but reduced to the following seven variables:
-
-- **species:** a factor denoting the penguin species (Adelie, Chinstrap, or Gentoo)
-- **island:** a factor denoting the island (in Palmer Archipelago, Antarctica) where observed (Biscoe, Dream or Torgersen)
-- **bill_length_mm:** a number denoting length of the dorsal ridge of penguin bill (millimeters)
-- **bill_depth_mm:** a number denoting the depth of the penguin bill (millimeters)
-- **flipper_length_mm:** an integer denoting penguin flipper length (millimeters)
-- **body_mass_g:** an integer denoting penguin body mass (grams)
-- **sex:** a factor denoting penguin sex (male, female)
-
-The `penguins` subset contains `r sum(complete.cases(penguins))` complete cases, with missing values (n~missing~ = `r sum(is.na(penguins))`, percent missing is `r round(100*sum(is.na(penguins))/(7*344),2)` %) indicated by `NA` throughout.
-In addition to reduced variables, the curated `penguins` subset is minimally updated from `penguins_raw` as follows:
-
-1. All variable names are converted to lower snake case
-2. Factor levels for species are truncated to only include the common name (e.g. "Gentoo", instead of "Gentoo penguin (*Pygoscelis papua*)")
-3. Recorded sex for penguin N36A1, originally recorded as ".", is updated to `NA`
-4. "culmen_length_mm" and "culmen_depth_mm" variable names are updated to "bill_length_mm" and "bill_depth_mm" to be more intuitive
-
-## Selected comparisons: `penguins` and `iris`
-
-We feel that the `penguins` data is generally useful and approachable for statistics education, and that it is uniquely well-suited to replace `iris`.
-Comparisons presented here are representative examples, and are not exhaustive.
-
-### Structure and sample size
-
-Both `iris` and `penguins` datasets are in tidy format [@tidydata], with each column denoting a single variable and each row containing a single observation (Table 1, 2).
-
-```{r, echo = FALSE}
-set.seed(1900)
-sample_n(iris, 6) %>% 
-  kable(caption = "Table 1. Iris dataset, 6 random observations (of 150 total)") %>% 
-  kable_styling(full_width = FALSE)
-```
-
-```{r, echo = FALSE}
-set.seed(1900)
-sample_n(penguins, 6) %>% 
-  kable(caption = "Table 2. Penguins dataset, 6 random observations (of 344 total)") %>% 
-  kable_styling(full_width = FALSE)
-```
-
-Overall sizes are comparable: dimensions (# columns $\times$ # rows) are 5 $\times$ 150 and 7 $\times$ 344 for `iris` and `penguins`, respectively, and sample sizes within groups (e.g.
-by species) are similar (Table 3).
-Note that the additional factors in `penguins` (species, island and sex) create additional opportunities for grouping, compared to the single factor in `iris` (Species).
-
-```{r compare_counts, echo = FALSE}
-
-iris_counts <- iris %>% 
-  count(Species) %>% 
-  rename(`Iris species` = Species)
-
-penguin_counts <- penguins %>% 
-  mutate(sex = str_to_title(sex)) %>% 
-  rename(`Penguin species` = species) %>% 
-  count(`Penguin species`, sex) %>% 
-  pivot_wider(names_from = sex, values_from = n)
-
-iris_penguin_n <- cbind(iris_counts, penguin_counts) %>% 
-  rename(`Sample size` = n)
-
-iris_penguin_n %>% 
-  gt() %>% 
-  tab_spanner(
-    label = "Iris data",
-    columns = vars(`Iris species`, `Sample size`)
-  ) %>% 
-  tab_spanner(
-    label = "Penguin data",
-    columns = vars(`Penguin species`, `Female`, `Male`, `NA`)
-  ) %>% 
-  tab_footnote(
-    footnote = "Note: Here, penguin sample sizes are split by species and sex, and could be further divided by island.",
-    location = cells_column_labels(
-      columns = vars(`Penguin species`)
-    )
-  )
-```
-
-### Correlation
-
-#### Comparison 1: Linear trends across or within species
-
-An overall approximately linear relationship can be explored using petal dimensions from `iris`, and penguin size (flipper length versus body mass) in `penguins` (Figure X).
-Educators may explore how linear regression across all species of iris or penguins (Figure X A,B) compares to regression performed separately for each species (Figure X C,D).
-
-```{r, echo = FALSE, warning = FALSE, message = FALSE}
-
-penguin_flip_mass_scatter <- ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
-  geom_point(aes(color = species, shape = species), size = 1.5, alpha = 0.7) +
-  geom_smooth(method = "lm", aes(group = species, color = species), se = FALSE) +
-  theme_minimal() +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme(legend.position = c(0.3, 0.8)) +
-  labs(x = "Flipper length (mm)", 
-       y = "Body mass (g)") +
-  theme(plot.title.position = "plot")
-
-penguin_flip_mass_scatter_all <- ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
-  geom_point(color = "gray50", size = 1.5, alpha = 0.5) +
-  geom_smooth(method = "lm", se = FALSE, color = "black") +
-  theme_minimal() +
-  labs(x = "Flipper length (mm)", 
-       y = "Body mass (g)",
-       title = "Penguin size") +
-  theme(plot.title.position = "plot")
-
-iris_petal_scatter <- ggplot(iris, aes(x = Petal.Length, y = Petal.Width)) +
-  geom_point(aes(color = Species, size = Species, shape = Species), size = 1.5, alpha = 0.7) +
-  geom_smooth(aes(group = Species, color = Species), method = "lm", se = FALSE) +
-  theme_minimal() +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme(legend.position = c(0.3, 0.8)) +
-  labs(x = "Petal length (cm)",
-       y = "Petal width (cm)") +
-  theme(plot.title.position = "plot")
-
-iris_petal_scatter_all <- ggplot(iris, aes(x = Petal.Length, y = Petal.Width)) +
-  geom_point(color = "gray50", size = 1.5, alpha = 0.5) +
-  geom_smooth(method = "lm", se = FALSE, color = "black") +
-  theme_minimal() +
-  labs(x = "Petal length (cm)",
-       y = "Petal width (cm)",
-       title = "Iris petal dimensions") +
-  theme(plot.title.position = "plot")
-
-(penguin_flip_mass_scatter_all + iris_petal_scatter_all) / (penguin_flip_mass_scatter + iris_petal_scatter)
-
-ggsave(here("figures","iris_penguins_scatter.png"), height = 6, width = 6)
-```
-
-#### Comparison 2: linear relationships within species
-
-```{r, echo = FALSE, warning = FALSE, message = FALSE}
-penguin_bill_scatter <- ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
-  geom_point(aes(color = species), size = 2, alpha = 0.7) +
-  theme_minimal() +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme(legend.position = "bottom") +
-  labs(x = "Bill length (mm)", 
-       y = "Bill depth (g)",
-       title = "Penguin size", 
-       color = "Species") +
-  theme(plot.title.position = "plot")
-
-iris_sepal_scatter <- ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) +
-  geom_point(aes(color = Species), size = 2, alpha = 0.7) +
-  theme_minimal() +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme(legend.position = "bottom") +
-  labs(x = "Sepal length (cm)",
-       y = "Sepal width (cm)",
-       title = "Iris sepal dimensions") +
-  theme(plot.title.position = "plot")
-
-penguin_bill_scatter + iris_sepal_scatter
-```
-
-### Differences between groups
-
-As in `iris`, multiple opportunities for comparisons between groups also exist in `penguins` (Figure X).
-Notably, more obvious separation between species exists in `iris` (Figure X E & F), compared to any between-species comparisons in `penguins` (Figure X A - D).
-
-```{r, include = FALSE}
-
-# CHECK THIS ALL!!! 
-# ANOVA to see if flipper lengths are significantly different...(this says yes, but want other eyes)
-flipper_lm <- lm(flipper_length_mm ~ species, data = penguins)
-
-flipper_anova <- Anova(flipper_lm, type = "III")
-flipper_anova
-
-flip_aov <- aov(flipper_lm)
-summary(flip_aov)
-
-# TukeyHSD(flip_aov) all significantly different
+# Table of assigned clusters for iris petals: 
+petal_class <- table(kclust_petals$cluster, petals$Species) %>% 
+  as.data.frame() %>% 
+  rename(cluster = Var1, Species = Var2, n = Freq) %>% 
+  pivot_wider(names_from = Species, values_from = n)
 
@@
 
-```{r, echo = FALSE, warning = FALSE, message = FALSE}
+Visualizing it: 
+```{r}
+bills_cluster <- cbind(bills, cluster_no = kclust_bills$cluster)
 
-# Update this with an actual theme FFS and use '&' in patchwork
+penguins_kmeans <- ggplot(data = bills_cluster, aes(x = bill_length_mm, 
+                                     bill_depth_mm)) +
+  geom_text(aes(label=cluster_no, color = species), 
+                  size = 4) +
+  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
+  labs(title = "Penguins k-means clustering") +
+  theme(legend.position = "bottom")
 
-iris_pw <- ggplot(iris, aes(x = Species, y = Petal.Width)) +
-  geom_boxplot(aes(color = Species),
-               show.legend = FALSE) +
+iris_cluster <-cbind(petals, cluster_no = kclust_petals$cluster)
+
+iris_kmeans <- ggplot(data = iris_cluster, aes(x = Petal.Length, y = Petal.Width)) +
+  geom_text(aes(label = cluster_no, color = Species), size = 4) +
   scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme_bw() +
-  theme(plot.title.position = "plot") +
-  labs(x = "",
-       y = "Petal width (cm)") +
-  coord_flip()
+  theme(panel.background = element_rect(fill = "gray90", color = "gray90"),
+        legend.position = "bottom") +
+  labs(title = "Iris k-means clustering")
 
-iris_pl <- ggplot(iris, aes(x = Species, y = Petal.Length)) +
-  geom_boxplot(aes(color = Species),
-               show.legend = FALSE) +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme_bw() +
-  theme(plot.title.position = "plot") +
-  labs(x = "",
-       y = "Petal length (cm)") +
-  coord_flip()
+(penguins_kmeans | iris_kmeans)
 
-iris_sl <- ggplot(iris, aes(x = Species, y = Sepal.Length)) +
-  geom_boxplot(aes(color = Species),
-               show.legend = FALSE) +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme_bw() +
-  theme(plot.title.position = "plot") +
-  labs(x = "",
-       y = "Sepal length (cm)") +
-  coord_flip()
+# Full table of cluster assignments: 
+combo_cluster <- cbind(bills_class, petal_class) %>% 
+  kable() %>% 
+  kable_styling(full_width = FALSE)
 
-iris_sw <- ggplot(iris, aes(x = Species, y = Sepal.Width)) +
-  geom_boxplot(aes(color = Species),
-               show.legend = FALSE) +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme_bw() +
-  theme(plot.title.position = "plot") +
-  labs(x = "",
-       y = "Sepal width (cm)") +
-  coord_flip()
-
-penguins_flip <- penguins %>%
-  ggplot(aes(x = species, y = flipper_length_mm)) +
-  geom_boxplot(aes(color = species),
-               show.legend = FALSE) +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme_bw() +
-  labs(x = "",
-       y = "Flipper length (mm)") +
-  coord_flip()
-
-penguins_mass <- penguins %>%
-  ggplot(aes(x = species, y = body_mass_g)) +
-  geom_boxplot(aes(color = species),
-               show.legend = FALSE) +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme_bw() +
-  labs(x = "",
-       y = "Body mass (g)") +
-  coord_flip()
-
-penguins_bill_length <- penguins %>%
-  ggplot(aes(x = species, y = bill_length_mm)) +
-  geom_boxplot(aes(color = species),
-               show.legend = FALSE) +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme_bw()+
-  labs(x = "",
-       y = "Bill length (mm)") +
-  coord_flip()
-
-penguins_bill_depth <- penguins %>%
-  ggplot(aes(x = species, y = bill_depth_mm)) +
-  geom_boxplot(aes(color = species),
-               show.legend = FALSE) +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme_bw() +
-  labs(x = "",
-       y = "Bill depth (mm)") +
-  coord_flip()
-
-penguins_box <- penguins_flip / penguins_mass / penguins_bill_length / penguins_bill_depth + plot_layout(guides = "collect") + plot_annotation(title = "Penguins")
-
-iris_box <- iris_pw / iris_pl / iris_sw / iris_sl + plot_layout(guides = "collect")
-
-penguins_box | iris_box + plot_annotation(
-  title = 'Penguin and iris sizes',
-  subtitle = 'Penguin size measurements (left) and iris petal and sepal dimensions (right)',
-  tag_levels = 'A'
-)
-
-ggsave(here("figures","penguins_iris_box.png"), height = 7, width = 8)
+combo_cluster
 
 ```
-
-### Classification?
-
-### Other comparative analyses?
-
-### Hypothesis testing
-
-- t-tests?
-- ANOVAs?
-- MANOVAs?
-
-## Suggested activities
-
-### Activity 1: Predict penguin body mass
-
-## Appendix
-
-### Appendix A: Variables in the raw data
-
-- **studyName:** Sampling expedition indicator
-- **Sample Number:** Integer denoting the continuous numbering sequence for each penguin
-- **Species:** a factor denoting the penguin species (Adelie, Chinstrap, or Gentoo)
-- **Region**: A character denoting the region of the Palmer LTER sampling grid
-- **Island:** a factor denoting the island where observed (Torgersen, Dream or Biscoe) in Palmer Archipelago
-- **Stage:** A character string denoting reproductive stage at sampling
-- **Individual ID:** A character string denoting a unique ID for each penguin
-- **Clutch Completion:** A character string denoting if the study next observed with a full clutch, i.e. 2 eggs
-- **Date Egg:** A date denoting the date study nest observed with 1 egg
-- **Culmen Length:** A number denoting the length of the dorsal ridge of a bird's bill (millimeters)
-- **Culmen Depth:** A number denoting the depth of the dorsal ridge of a bird's bill (millimeters)
-- **Flipper Length:** An integer denoting the penguin flipper length (millimeters)
-- **Body Mass:** An integer denoting the penguin body mass (grams)
-- **Sex:** A character string denoting the sex of an animal
-- **Delta 15 N:** A number denoting the measure of the ratio of stable isotopes 15N:14N
-- **Delta 13 C:** A number denoting the measure of the ratio of stable isotopes 13C:12C
-- **Comments:** A character string with text providing additional relevant information for data
-
-## Citations
-
-## Random other stuff
-
-Compared to the single factor in `iris` (Species), `penguins`' three factors (species, island, sex) increase opportunities for comparisons between groups (Figure X).
-
-```{r, echo = FALSE, warning = FALSE, message = FALSE}
-
-cd <- penguins %>% 
-  filter(sex %in% c("male","female")) %>% 
-ggplot(aes(x = bill_depth_mm)) +
-  geom_histogram(aes(fill = species), alpha = 0.6, position = "identity") +
-  scale_fill_manual(values = c("darkorange","darkorchid","cyan4")) +
-  facet_wrap(~ sex) +
-  theme_minimal()
-
-cl <- penguins %>% 
-  filter(sex %in% c("male","female")) %>% 
-ggplot(aes(x = bill_length_mm)) +
-  geom_histogram(aes(fill = species), alpha = 0.6, position = "identity") +
-  scale_fill_manual(values = c("darkorange","darkorchid","cyan4")) +
-  facet_wrap(~ sex) +
-  theme_minimal()
-
-fl <- penguins %>% 
-  filter(sex %in% c("male","female")) %>% 
-ggplot(aes(x = flipper_length_mm)) +
-  geom_histogram(aes(fill = species), alpha = 0.6, position = "identity") +
-  scale_fill_manual(values = c("darkorange","darkorchid","cyan4")) +
-  facet_wrap(~ sex) +
-  theme_minimal()
-
-bm <- penguins %>% 
-  filter(sex %in% c("male","female")) %>% 
-ggplot(aes(x = body_mass_g)) +
-  geom_histogram(aes(fill = species), alpha = 0.6, position = "identity") +
-  scale_fill_manual(values = c("darkorange","darkorchid","cyan4")) +
-  facet_wrap(~ sex) +
-  theme_minimal()
-
-cd / cl / fl / bm + plot_layout(guides = "collect")
-```
-
-## Differences by islands?
-
-```{r}
-# No difference in Adelie across islands: Chinstrap only on Dream, Gentoo only on Torgerson
-penguins %>% 
-  ggplot(aes(x = island, y = body_mass_g)) +
-  geom_point() +
-  facet_wrap(~species)
-
-penguins %>% 
-  ggplot(aes(x = island, y = flipper_length_mm)) +
-  geom_point() +
-  facet_wrap(~species)
-
-penguins %>% 
-  ggplot(aes(x = island, y = bill_length_mm)) +
-  geom_point() +
-  facet_wrap(~species)
-
-penguins %>% 
-  ggplot(aes(x = island, y = bill_depth_mm)) +
-  geom_point() +
-  facet_wrap(~species)
-```
+## Using tidymodels? eeeeehhh we'll see

### 2020-06-28 manuscript/manuscript_text.Rmd
actually deal with conflicts (#6)

@@
   geom_text(aes(label=cluster_no, color = species), 
-                  size = 4) +
+                  size = 4)
+```
+
+```{r, echo = FALSE, warning = FALSE, message = FALSE}
+penguin_bill_scatter <- ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
+  geom_point(aes(color = species), size = 2, alpha = 0.7) +
+  theme_minimal() +
   scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
@@
 ```
+
 #### k-means clustering: all numeric variables
@@
 
-### Hierarchical clustering?
+### Hypothesis testing
+
+### Principle components analysis
+
+### Clustering (e.g. k-means?)
+
+### Classification (Alison Hill blog post?)
+
+## Penguin data access for non-R users
+
+The Palmer Archipelago penguins data are available to non-R users by direct import from the Environmental Data Initiative, from GitHub, and as a stored dataset in existing packages.
+Below we share several current ways to access the data (as of July 2020 - we expect availability to expand in the future).
+
+### Download or import directly from Environmental Data Initiative
+
+The [Environmental Data Initiative](https://environmentaldatainitiative.org/) is "an NSF-funded project, actively promoting and enabling curation and re-use of environmental data." The raw data for each penguin species (Adélie, Gentoo and Chinstrap) are stored in a separate data file at the following links (see full citations in Citations):
+
+- Adélie data: [KNB-LTER Data Package 219.5](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.219.5)
+- Gentoo data: [KNB-LTER Data Package 220.5](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.220.5)
+- Chinstrap data: [KNB-LTER Data Package 221.5](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.221.6)
+
+The links above contain all metadata information, and provides options to directly download the data as a comma-separated value (CSV) file *or* import for use directly using code provided (e.g.
+for R, Python, SAS, MatLab, or SPSS).
+
+### Download CSV from the palmerpenguins GitHub repo
+
+A CSV containing curated data for all three species (accessed directly from EDI) exists in the 'data-raw' folder in the palmerpenguins repo on GitHub.
+GitHub users can fork / clone the repo, and will then have the cleaned CSV containing all 3 penguin species.
+
+### The penguins in other packages & languages
+
+The Palmer Station penguin data has been added to existing libraries, including (non-exhaustive):
+
+#### Python modules
+
+- Seaborn
+- blah blah
+
+See examples for working with penguins in Python here:
+
+#### Kaggle
+
+- here
+
+#### Elsewhere?
+
+## Conclusion
+
+## Appendix
+
+### Appendix A: Variables in the raw data
+
+- **studyName:** Sampling expedition indicator
+- **Sample Number:** Integer denoting the continuous numbering sequence for each penguin
+- **Species:** a factor denoting the penguin species (Adelie, Chinstrap, or Gentoo)
+- **Region**: A character denoting the region of the Palmer LTER sampling grid
+- **Island:** a factor denoting the island where observed (Torgersen, Dream or Biscoe) in Palmer Archipelago
+- **Stage:** A character string denoting reproductive stage at sampling
+- **Individual ID:** A character string denoting a unique ID for each penguin
+- **Clutch Completion:** A character string denoting if the study next observed with a full clutch, i.e. 2 eggs
+- **Date Egg:** A date denoting the date study nest observed with 1 egg
+- **Culmen Length:** A number denoting the length of the dorsal ridge of a bird's bill (millimeters)
+- **Culmen Depth:** A number denoting the depth of the dorsal ridge of a bird's bill (millimeters)
+- **Flipper Length:** An integer denoting the penguin flipper length (millimeters)
+- **Body Mass:** An integer denoting the penguin body mass (grams)
+- **Sex:** A character string denoting the sex of an animal
+- **Delta 15 N:** A number denoting the measure of the ratio of stable isotopes 15N:14N
+- **Delta 13 C:** A number denoting the measure of the ratio of stable isotopes 13C:12C
+- **Comments:** A character string with text providing additional relevant information for data
+
+# Random other stuff
+
+Compared to the single factor in `iris` (Species), `penguins`' three factors (species, island, sex) increase opportunities for comparisons between groups (Figure X).
+
+```{r, echo = FALSE, warning = FALSE, message = FALSE}
+
+cd <- penguins %>% 
+  filter(sex %in% c("male","female")) %>% 
+ggplot(aes(x = bill_depth_mm)) +
+  geom_histogram(aes(fill = species), alpha = 0.6, position = "identity") +
+  scale_fill_manual(values = c("darkorange","darkorchid","cyan4")) +
+  facet_wrap(~ sex) +
+  theme_minimal()
+
+cl <- penguins %>% 
+  filter(sex %in% c("male","female")) %>% 
+ggplot(aes(x = bill_length_mm)) +
+  geom_histogram(aes(fill = species), alpha = 0.6, position = "identity") +
+  scale_fill_manual(values = c("darkorange","darkorchid","cyan4")) +
+  facet_wrap(~ sex) +
+  theme_minimal()
+
+fl <- penguins %>% 
+  filter(sex %in% c("male","female")) %>% 
+ggplot(aes(x = flipper_length_mm)) +
+  geom_histogram(aes(fill = species), alpha = 0.6, position = "identity") +
+  scale_fill_manual(values = c("darkorange","darkorchid","cyan4")) +
+  facet_wrap(~ sex) +
+  theme_minimal()
+
+bm <- penguins %>% 
+  filter(sex %in% c("male","female")) %>% 
+ggplot(aes(x = body_mass_g)) +
+  geom_histogram(aes(fill = species), alpha = 0.6, position = "identity") +
+  scale_fill_manual(values = c("darkorange","darkorchid","cyan4")) +
+  facet_wrap(~ sex) +
+  theme_minimal()
+
+cd / cl / fl / bm + plot_layout(guides = "collect")
+```
+
+## Differences by islands? 
+>>>>>>> 0181779cd02ccdc2413d2698c7029f18572f3757
 
@@
 
+A Web of Science search reveals that Fisher's 1936 paper has been cited in 7,258 publications (as of June 2020), most often in journals publishing in fields of Computer Science (57% of 7,258 citing references), Mathematics (49%), and Engineering (29%) - most of which we expect are crediting Anderson's *Iris* data, and not Fisher's eugenics work.
+
 # End random other stuff

### 2020-07-09 R/penguins_code.Rmd
add k-means clustering & text edits (#7)
* actually deal with conflicts

* update draft w/kmeans

* push recent code ?

---
title: "Palmer penguins paper code"
author: "Allison Horst"
date: "7/2/2020"
output: html_document
---

This document contains accompanying code for all figures, tables and analyses presented in "Palmer Archipelago (Antarctica) penguins data: an alternative to Edgar Anderson's *Iris* data."

## Set-up and required packages:

```{r setup, include = TRUE, message = FALSE, warning = FALSE}

# ---- Code options ---- #
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# ---- Packages required ---- # 
library(tidyverse)
library(here)
library(palmerpenguins)
library(kableExtra)
library(patchwork)
library(gt)
library(GGally)
library(recipes)
library(shadowtext)

# ---- Set default ggplot theme to theme_minimal ---- #
theme_set(theme_minimal())

```

## Table 1: *iris* sample

```{r iris_sample}
set.seed(1900)
sample_n(iris, 6) %>% 
  kable() %>% 
  kable_styling(full_width = FALSE)
```

## Table 2: *penguins* sample

```{r penguin_sample}
set.seed(1900)
sample_n(penguins, 6) %>% 
  kable() %>% 
  kable_styling(full_width = FALSE)
```

## Table 3: Sample size, *iris* (by species) and *penguins* (by species and sex)

### NOTE: I am not happy about this method to convert NA values (for missing factor levels) to 0.

```{r compare_counts}

# ---- Iris counts by species ---- #
iris_counts <- iris %>% 
  count(Species) %>% 
  rename(`Iris species` = Species)

# ---- Penguin counts by species and sex ---- #
penguin_counts <- penguins %>% 
  mutate(sex = str_to_title(sex)) %>% 
  count(species, sex) %>% 
  pivot_wider(names_from = sex, values_from = n) %>% 
  replace_na(list(`NA` = 0)) %>% 
  as.data.frame() %>% 
  rename(`Penguin species` = species)

# ---- Combine iris and penguin counts ---- #

iris_penguin_n <- cbind(iris_counts, penguin_counts) %>% 
  rename(`Sample size` = n)

# ---- Create a finalized table using `gt` package ---- #
iris_penguin_n %>%
  gt() %>%
  tab_spanner(
    label = "Iris sample size (by species)",
    columns = vars(`Iris species`, `Sample size`)
  ) %>%
  tab_spanner(
    label = "Penguin sample size (by species and sex)",
    columns = vars(`Penguin species`, `Female`, `Male`, `NA`)
  ) %>%
  tab_footnote(
    footnote = "Note: Here, penguin sample sizes are split by species and sex, and could be further divided by island. NA column for penguins indicate observations where sex was not recorded.",
    location = cells_column_labels(
      columns = vars(`Penguin species`)
    )
  )

```

## Figure 1: *penguins* pairs plot

```{r penguin_pairs}
penguins %>%
  select(species, where(is.numeric)) %>% 
  ggpairs(aes(color = species, shape = species),
          columns = c("flipper_length_mm", "body_mass_g", 
                      "bill_length_mm", "bill_depth_mm"),
          columnLabels = c("Flipper length (mm)","Body mass (g)", "Bill length (mm)", "Bill depth (mm)"),
          upper = list(continuous = wrap("cor", size = 3))) +
  scale_fill_manual(values = c("darkorange","purple","cyan4")) +
  scale_color_manual(values = c("darkorange","purple","cyan4")) +
  scale_shape_manual(values = c(15,16,17)) +
  theme_minimal() +
  theme(
        text = element_text(size = 10, family = "Arial"),
        panel.grid.major = element_line(colour = NA),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(color = "gray80", fill = NA)
        )
```

## Figure 2: *iris* pairs plot

```{r iris_pairs}
iris %>%
  ggpairs(aes(color = Species),
          columns = c("Petal.Length", "Petal.Width", 
                      "Sepal.Length", "Sepal.Width"),
          columnLabels = c("Petal length (cm)","Petal width (cm)", "Sepal length (cm)", "Sepal width (cm)"),
          upper = list(continuous = wrap("cor", size = 3))) +
  scale_colour_manual(values = c("darkorange","purple","cyan4")) +
  scale_fill_manual(values = c("darkorange","purple","cyan4")) +
  theme_minimal() +
  theme(
        text = element_text(size = 10, family = "Arial"),
        panel.grid.major = element_line(colour = NA),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "gray90", color = "gray90")
        )
```

## Figure 3: Visualize selected linear relationships

```{r linear_viz, warning = FALSE, message = FALSE, fig.height = 6, fig.width = 6}

# ---- Penguin scatterplot, species included ---- #

penguin_flip_mass_scatter <- ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = species), size = 1.5, alpha = 0.5) +
  geom_smooth(method = "lm", aes(group = species, color = species), se = FALSE, show.legend = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  theme(legend.position = "bottom") +
  labs(x = "Flipper length (mm)", 
       y = "Body mass (g)") +
  theme(plot.title.position = "plot",
        panel.border = element_rect(fill = NA, color = "gray70"))

# ----- Penguin scatterplot, species omitted ---- #

penguin_flip_mass_scatter_all <- ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(color = "gray50", size = 1.5, alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  theme_minimal() +
  labs(x = "Flipper length (mm)", 
       y = "Body mass (g)") +
  theme(plot.title.position = "plot",
        panel.border = element_rect(fill = NA, color = "gray70"))

# ---- Iris scatterplot, species included ---- #

iris_petal_scatter <- ggplot(iris, aes(x = Petal.Length, y = Petal.Width)) +
  geom_point(aes(color = Species, size = Species, shape = Species), size = 1.5, alpha = 0.5) +
  geom_smooth(aes(group = Species, color = Species), method = "lm", se = FALSE, show.legend = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  theme(legend.position = "bottom") +
  labs(x = "Petal length (cm)",
       y = "Petal width (cm)") +
  theme(plot.title.position = "plot",
        panel.background = element_rect(fill = "gray90", color = "gray90"))

# ---- Iris scatterplot, species omitted ---- #

iris_petal_scatter_all <- ggplot(iris, aes(x = Petal.Length, y = Petal.Width)) +
  geom_point(color = "gray50", size = 1.5, alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  theme_minimal() +
  labs(x = "Petal length (cm)",
       y = "Petal width (cm)") +
  theme(plot.title.position = "plot",
        panel.background = element_rect(fill = "gray90", color = "gray90"))

(penguin_flip_mass_scatter_all + iris_petal_scatter_all) / (penguin_flip_mass_scatter + iris_petal_scatter) + plot_annotation(tag_levels = 'A')
```

## Figure 4: Simpson's Paradox example

```{r simpsons_paradox, fig.width = 8, fig.height = 4}
# Simpson's Paradox example (bill dimensions)
simpson_nospecies <- ggplot(data = penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(color = "gray40", alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  theme_minimal() +
  theme(panel.border = element_rect(fill = NA, color = "gray70")) +
  labs(x = "Bill length (mm)", y = "Bill depth (mm)")

simpson_wspecies <- ggplot(data = penguins, aes(x = bill_length_mm, y = bill_depth_mm, group = species)) +
  geom_point(aes(color = species, shape = species), size = 2, alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, aes(color = species)) +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  theme_minimal() +
  theme(panel.border = element_rect(fill = NA, color = "gray70"),
        legend.position = c(0.16,0.14)) +
  labs(x = "Bill length (mm)", y = "Bill depth (mm)")

(simpson_nospecies | simpson_wspecies) + plot_annotation(tag_levels = 'A')

```

## Figure 5: PCA

```{r pca, fig.height = 8, fig.width = 8}

# ---- Penguins data prep ---- #
penguin_recipe <-
  recipe(~., data = penguins) %>% 
  update_role(species, island, sex, new_role = "id") %>% 
  step_naomit(all_predictors()) %>% 
  step_normalize(all_predictors()) %>%
  step_pca(all_predictors(), id = "pca") %>% 
  prep()

# ---- Penguin PCA ---- #
penguin_pca <- 
  penguin_recipe %>% 
  tidy(id = "pca")

# ---- Get penguins % variance ---- #

penguin_percvar <- penguin_recipe %>% 
  tidy(id = "pca", type = "variance") %>% 
  dplyr::filter(terms == "percent variance")

# ---- Clean penguins PCA outcome ---- #

pca_wider <- penguin_pca %>% 
  tidyr::pivot_wider(names_from = component, id_cols = terms)

# ---- Define arrow style for biplot ---- #
arrow_style <- arrow(length = unit(.05, "inches"),
                     type = "closed")

# ---- Make penguins biplot ---- #
pca_plot <-
  juice(penguin_recipe) %>%
  ggplot(aes(PC1, PC2)) +
  geom_point(aes(color = species, shape = species), 
             alpha = 0.5, 
             size = 2) +
  scale_colour_manual(values = c("darkorange","purple","cyan4"))

penguins_biplot <- pca_plot +
  geom_segment(data = pca_wider,
               aes(xend = PC1, yend = PC2), 
               x = 0, 
               y = 0, 
               arrow = arrow_style) + 
  geom_shadowtext(data = pca_wider,
            aes(x = PC1, y = PC2, label = terms), 
            nudge_x = c(0.7,0.7,1.7,1.2), 
            nudge_y = c(-0.1,-0.2,0.1,-0.1),
            size = 4, 
            color = "black",
            bg.color = "white") +
  theme(legend.position = "bottom",
        panel.border = element_rect(color = "gray70", fill = NA))

# For positioning (above): 
# 1: bill_length
# 2: bill_depth
# 3: flipper length
# 4: body mass

# ---- Penguins scree plot ---- #
penguin_screeplot <- penguin_percvar %>% 
  ggplot(aes(x = component, y = value)) + 
  geom_col(fill = "gray50") + 
  scale_x_continuous(limits = c(0, 5), expand = c(0,0)) + 
  scale_y_continuous(limits = c(0,100), expand = c(0,0)) +
  ylab("% of total variance") +
  geom_text(aes(label = round(value,2)), vjust=-0.25) +
  theme(panel.border = element_rect(color = "gray70", fill = NA))

#############################

# Iris PCA

# ---- Prep iris data for PCA ---- #

iris_recipe <-
  recipe(~., data = iris) %>% 
  update_role(Species, new_role = "id") %>% 
  step_normalize(all_predictors()) %>%
  step_pca(all_predictors(), id = "pca") %>% 
  prep()

# ---- Iris PCA ---- #

iris_pca <- 
  iris_recipe %>% 
  tidy(id = "pca")

# ---- Iris PCA loadings into wider format ---- #
iris_wider <- iris_pca %>% 
  tidyr::pivot_wider(names_from = component, id_cols = terms)

# ---- Make iris PCA biplot ---- #

iris_pca_plot <-
  juice(iris_recipe) %>%
  ggplot(aes(PC1, PC2)) +
  geom_point(aes(color = Species, shape = Species), 
             alpha = 0.5, 
             size = 2) +
  scale_colour_manual(values = c("darkorange","purple","cyan4"))

iris_biplot <- iris_pca_plot +
  geom_segment(data = iris_wider,
               aes(xend = PC1, yend = PC2), 
               x = 0, 
               y = 0, 
               arrow = arrow_style) + 
  geom_shadowtext(data = iris_wider,
            aes(x = PC1, y = PC2, label = terms), 
            nudge_x = c(0.5,0.3,1,1.2), 
            nudge_y = c(-0.1,-0.2,0.1,-0.1),
            size = 4, 
            color = "black",
            bg.color = "white") +
  theme(panel.background = element_rect(fill = "gray90", color = "gray90"),
        legend.position = "bottom")

# ---- Get iris % variance ---- #
iris_percvar <- iris_recipe %>% 
  tidy(id = "pca", type = "variance") %>% 
  dplyr::filter(terms == "percent variance")

# ---- Make iris scree plot ---- #
iris_screeplot <- iris_percvar %>% 
  ggplot(aes(x = component, y = value)) + 
  geom_col(fill = "gray50") + 
  scale_x_continuous(limits = c(0, 5), expand = c(0,0)) + 
  scale_y_continuous(limits = c(0,100), expand = c(0,0)) +
  ylab("% of total variance") +
  geom_text(aes(label = round(value,2)), vjust=-0.25) +
  theme(panel.background = element_rect(fill = "gray90", color = "gray90"))

# ---- Create compound figure (uses patchwork) ---- #
(penguins_biplot | iris_biplot) / (penguin_screeplot | iris_screeplot) + plot_annotation(tag_levels = 'A')
```

## Figure 6: k-means

Simplified to two variables for each dataset (could be performed for all 4 numeric variables instead).

### NOTE: I am not happy about this method to convert NA values (for missing factor levels) to 0.

```{r penguin_kmeans}

# ---- PENGUIN BILL DIMENSIONS K-MEANS ---- #

# ---- Simplify & drop missings ---- #

p_bills <- penguins %>% 
  select(species, bill_length_mm, bill_depth_mm) %>% 
  drop_na()

# ---- Remove species factor & scale ---- # 
p_bills_nospecies <- p_bills %>% 
  select(-species) %>% 
  scale() %>% 
  as.data.frame()

# ---- K-means on penguin bill dimensions (k = 3, w/20 centroid starts) ---- #

p_bills_kmeans <- kmeans(p_bills_nospecies, centers = 3, nstart = 20)

# ---- Bind cluster assignments to the data ---- #

p_kmeans_clusters <- cbind(p_bills_nospecies, cluster = as.factor(p_bills_kmeans$cluster), species = p_bills$species)

# ---- Plot clusters ---- #

p_kmeans_gg <- ggplot(p_kmeans_clusters, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_text(aes(label = cluster, color = species)) +
  theme(legend.position = "bottom") +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4"))

# ---- Get counts in each cluster by species ---- #
p_cluster_counts <- p_kmeans_clusters %>% 
  count(species, cluster) %>% 
  pivot_wider(names_from = species, values_from = n) %>% 
  arrange(cluster)

```

```{r iris_kmeans}

# ---- IRIS PETALS K-MEANS ---- #

# ---- Subset iris species & petal dimensions ---- #
iris_petals <- iris %>% 
  select(Species, Petal.Length, Petal.Width)

# ---- remove species factor & scale petal dimensions ---- #

ip_nospecies <- iris_petals %>% 
  select(-Species) %>% 
  scale() %>% 
  as.data.frame()

# ---- k-means on iris petal dimensions (k = 3, w/20 centroid starts) ---- #

set.seed(1967)
ip_kmeans <- kmeans(ip_nospecies, centers = 3, nstart = 20)

# ---- Bind clusters to the data ---- #

iris_kmeans_clusters <- cbind(ip_nospecies, cluster = as.factor(ip_kmeans$cluster), species = iris$Species)

# ---- Plot clusters ---- #  
iris_kmeans_gg <- ggplot(iris_kmeans_clusters, aes(x = Petal.Length, y = Petal.Width)) +
  geom_text(aes(label = cluster, color = species)) +
   theme(legend.position = "bottom",
         panel.background = element_rect(fill = "gray92", color = "gray90")) +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4"))

# ---- Get counts in each cluster by species ---- #
iris_cluster_counts <- iris_kmeans_clusters %>% 
  count(species, cluster) %>% 
  pivot_wider(names_from = species, values_from = n) %>% 
  arrange(cluster)

```

```{r kmeans_figs}

# ---- Combine k-means plots with patchwork ---- #

(p_kmeans_gg | iris_kmeans_gg) + plot_annotation(tag_levels = "A")

# ---- Make table of cluster assignments ---- #

kmeans_2var_table <- cbind(p_cluster_counts, iris_cluster_counts) %>% 
  replace_na(list(`Adelie` = 0, `setosa` = 0, `versicolor` = 0, `virginica` = 0)) %>% 
  kable() %>% 
  kable_styling(full_width = FALSE) %>% 
  column_spec(c(5:8), background = "gainsboro") %>% 
  add_header_above(c("Penguins cluster assignments" = 4, "Iris cluster assignments" = 4))

# ---- Return table ---- #

kmeans_2var_table
```

### 2020-07-09 code_tests/kmeans_testing.Rmd
add k-means clustering & text edits (#7)
* actually deal with conflicts

* update draft w/kmeans

* push recent code ?

@@
 library(kableExtra)
+library(mclust)
 knitr::opts_chunk$set(echo = TRUE)
@@
 
-The "old" way, using `kmeans()`:
+Using `kmeans()`:
+
+Example 1: Male penguins bill dimensions
 ```{r}
-penguins_k <- penguins %>% 
+mp_bills <- penguins %>% 
   filter(sex == "male") %>% 
-  select(species, bill_length_mm:body_mass_g) %>% 
+  select(species, bill_length_mm, bill_depth_mm)
+
+mp_bills_nospecies <- mp_bills %>% 
+  select(-species) %>% 
+  scale() %>% 
+  as.data.frame()
+
+# Perform k-means on male penguin bill dimensions (k = 3, w/20 centroid starts)
+mp_bills_kmeans <- kmeans(mp_bills_nospecies, centers = 3, nstart = 20)
+mp_bills_kmeans
+
+# Bind clusters to the data
+mp_kmeans_clusters <- cbind(mp_bills_nospecies, cluster = mp_bills_kmeans$cluster, species = mp_bills$species)
+
+ggplot(mp_kmeans_clusters, aes(x = bill_length_mm, y = bill_depth_mm)) +
+  geom_text(aes(label = cluster, color = species))
+
+male_cluster_counts <- mp_kmeans_clusters %>% 
+  complete() %>% 
+  count(species, cluster) %>% 
+  pivot_wider(names_from = species, values_from = n)
+
+male_cluster_counts
+# SERIOUSLY WTF...those 3s are NOT closer to the 3 centroid...
+# 
+# plot(mp_bills_nospecies, col = mp_bills_kmeans$cluster)
+# points(mp_bills_kmeans$centers, col = 1:2, pch = 8, cex = 2)
+# 
+# # Testing some tidymodels methods instead...
+# test_cluster <- augment(mp_bills_kmeans, mp_bills_nospecies) %>% 
+#   rename(cluster_number = .cluster)
+# 
+# ggplot(data = test_cluster, aes(x = bill_length_mm, y = bill_depth_mm, color = cluster_number)) +
+#   geom_point()
+```
+
+# With all penguins (bill dimensions)
+
+Example 2: All penguins bill dimensions
+```{r}
+p_bills <- penguins %>% 
+  select(species, bill_length_mm, bill_depth_mm) %>% 
   drop_na()
 
-# Using bill_length & flipper_length
-# THIS ONE SHOWS *pretty* good clustering
+p_bills_nospecies <- p_bills %>% 
+  select(-species) %>% 
+  scale() %>% 
+  as.data.frame()
 
-bills <- penguins_k %>% select(species, bill_length_mm, bill_depth_mm)
+# Perform k-means on male penguin bill dimensions (k = 3, w/20 centroid starts)
+p_bills_kmeans <- kmeans(p_bills_nospecies, centers = 3, nstart = 20)
+p_bills_kmeans
 
-set.seed(1967)
-kclust_bills <- kmeans(bills[,2:3], centers = 3, nstart = 20)
-kclust_bills
+# Bind clusters to the data
+p_kmeans_clusters <- cbind(p_bills_nospecies, cluster = p_bills_kmeans$cluster, species = p_bills$species)
 
-# Table of assigned clusters for penguins
-bills_class <- table(kclust_bills$cluster,bills$species)%>% 
-  as.data.frame() %>% 
-  rename(cluster = Var1, species = Var2, n = Freq) %>% 
+ggplot(p_kmeans_clusters, aes(x = bill_length_mm, y = bill_depth_mm)) +
+  geom_text(aes(label = cluster, color = species))
+
+p_cluster_counts <- p_kmeans_clusters %>% 
+  complete() %>% 
+  count(species, cluster) %>% 
   pivot_wider(names_from = species, values_from = n)
 
-# Same for iris:
-
-petals <- iris %>% select(Species, Petal.Length, Petal.Width)
-
-set.seed(1967)
-kclust_petals <- kmeans(petals[,2:3], centers = 3, nstart = 25)
-kclust_petals
-
-# Table of assigned clusters for iris petals: 
-petal_class <- table(kclust_petals$cluster, petals$Species) %>% 
-  as.data.frame() %>% 
-  rename(cluster = Var1, Species = Var2, n = Freq) %>% 
-  pivot_wider(names_from = Species, values_from = n)
-
+p_cluster_counts
 ```
 
-Visualizing it: 
+# Trying EM using `mclust` package for fun
+
+See: https://en.wikibooks.org/wiki/Data_Mining_Algorithms_In_R/Clustering/Expectation_Maximization_%28EM%29#Executing_the_Algorithm
+
 ```{r}
-bills_cluster <- cbind(bills, cluster_no = kclust_bills$cluster)
+BIC <- mclustBIC(mp_bills_nospecies)
+plot(BIC)
+summary(BIC)
 
-penguins_kmeans <- ggplot(data = bills_cluster, aes(x = bill_length_mm, 
-                                     bill_depth_mm)) +
-  geom_text(aes(label=cluster_no, color = species), 
-                  size = 4) +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  labs(title = "Penguins k-means clustering") +
-  theme(legend.position = "bottom")
+mod1 <- Mclust(mp_bills_nospecies, G = 3, x = BIC)
+summary(mod1, parameters = TRUE)
 
-iris_cluster <-cbind(petals, cluster_no = kclust_petals$cluster)
-
-iris_kmeans <- ggplot(data = iris_cluster, aes(x = Petal.Length, y = Petal.Width)) +
-  geom_text(aes(label = cluster_no, color = Species), size = 4) +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme(panel.background = element_rect(fill = "gray90", color = "gray90"),
-        legend.position = "bottom") +
-  labs(title = "Iris k-means clustering")
-
-(penguins_kmeans | iris_kmeans)
-
-# Full table of cluster assignments: 
-combo_cluster <- cbind(bills_class, petal_class) %>% 
-  kable() %>% 
-  kable_styling(full_width = FALSE)
-
-combo_cluster
-
+plot(mod1, what = "classification")
 ```
-## Using tidymodels? eeeeehhh we'll see

### 2020-07-09 manuscript/manuscript_text.Rmd
add k-means clustering & text edits (#7)
* actually deal with conflicts

* update draft w/kmeans

* push recent code ?

--- 2020-06-25 manuscript/manuscript_text.Rmd
@@
 ---
-title: "Palmer Archipelago (Antarctica) penguins: an alternative to Edgar Anderson's iris data"
-output: html_document
+title: 'Palmer Archipelago (Antarctica) penguins: an alternative to Edgar Anderson''s
+  iris data'
+output:
+  html_document:
+    number_sections: yes
+  pdf_document: default
 bibliography: ../bibliography/references.bib
@@
 library(patchwork)
-library(ggExtra)
 library(gt)
-library(ggpubr)
-library(ggbeeswarm)
-library(car)
+library(GGally)
+library(recipes)
+library(shadowtext)
 knitr::opts_chunk$set(echo = TRUE)
+
+theme_set(theme_minimal())
 ```
 
-## Authors
+# Authors (placeholder order TBD)
 
-Allison Horst, Alison Hill, Kristen Gorman (order TBD)
+Allison Horst^1^, Alison Hill^2^, Kristen Gorman^3^
 
-## Keywords
+^1^ Bren School of Environmental Science and Management, University of California, Santa Barbara
 
-## Abstract
+^2^ RStudio
+
+^3^ College of Fisheries and Ocean Sciences, University of Alaska Fairbanks
+
+# Keywords
+
+Multivariate analysis, cluster analysis, correlation, classification, zoology, penguin morphology
+
+# Abstract
 
@@
 Fisher's eugenics research has created momentum to switch to alternative datasets.
-Size measurements for three penguin species observed on islands in Palmer Archipelago, Antarctica provide a (mostly) suitable, approachable and charismatic alternative to the *Iris* data.
-Here, we introduce the penguins data, then directly compare the two datasets to demonstrate that educators currently using *Iris* can switch to the Palmer Archipelago penguins for many use cases, particularly in introductory statistics and data science courses.
+Size measurements for three penguin species observed on islands in Palmer Archipelago, Antarctica provide a suitable, approachable and charismatic alternative to the *Iris* data.
+Here, we introduce the penguin data, then directly compare the two datasets to demonstrate that educators currently using *Iris* can switch to the Palmer Archipelago penguins for many use cases including data wrangling and visualization, correlation, regression, multivariate analysis (e.g.
+PCA), cluster analysis and classification (e.g.
+by k-means).
 
-## Introduction
+# Introduction
 
 In 1935, American botanist Edgar Anderson measured petal and sepal dimensions (length and width) for 50 samples from three *Iris* species: *Iris setosa*, *Iris versicolor*, and *Iris virginica* [@anderson].
-The manageable but non-trivial size (5 variables and 150 total observations) and characteristics of the data make it amenable for introducing a wide range of statistical methods including data wrangling, visualization, regression, classification and more.
+The manageable but non-trivial size (5 variables and 150 total observations) and characteristics of the data make it amenable for introducing a wide range of statistical methods including data wrangling, visualization, regression, cluster analysis, machine learning, and more.
 Anderson's *Iris* dataset is even built into a number of software packages, including in the auto-installed *datasets* package in *R* (R Core Team), Python's *scikit-learn* machine learning library [@scikit-learn], and the SAS *Sashelp* library, further facilitating its widespread use.
-As a result, eighty-five years after the data were initially published, Anderson’s Iris measurements are ubiquitous in statistics and data science education.
+As a result, eighty-five years after the data were initially published, Anderson’s *Iris* measurements are ubiquitous in statistics and data science education.
 
@@
 In 1936, one year after publication in the *Bulletin of the American Iris Society*, Anderson’s *Iris* measurements were further analyzed and published by eugenicist and statistician Ronald A.
-Fisher in *The Annals of Eugenics* [@fisher].
-Regardless of Anderson's initial motivation, the dataset today remains inextricably linked to Fisher’s eugenics research and even commonly (if unfairly) referred to as "*Fisher's* iris data" (ref).
-Due to its history in - and strong association with - Fisher's eugenics research, there is momentum to replace the *Iris* data with alternative datasets that offer similar teaching and learning opportunities.
+Fisher in *Annals of Eugenics* [@fisher].
+Regardless of Anderson's initial motivation, the data today remain inextricably linked to Fisher’s eugenics research and are even commonly (if unfairly) referred to as "*Fisher's* iris data".
+Due to its history in - and strong association with - Fisher's work, there is momentum for teachers to (1) address the history of the *Iris* data in eugenics research when teaching with it, or (2) replace the *Iris* data with an alternative dataset that offers similar teaching and learning opportunities.
+This paper focuses on (2): a replacement for the *Iris* data.
 
-Considering *Iris* data usage, we established the following criteria for a suitable alternative dataset:
+Considering *Iris* data usage in statistics education, we established the following criteria for a suitable replacement dataset:
 
@@
 Comparable size, characteristics and learning opportunities make the penguins data uniquely transferable for educators currently using *Iris*.
-By highlighting similarities in data structure and analyses, as well as several examples for which the penguins are *not* a good alternative, we demonstrate that the penguins can replace irises for many use cases.
+By comparing data structure, size, and a range of analyses side-by-side, we demonstrate that the Palmer Archipelago penguin measurements can replace Anderson's iris data for many use cases.
 
-## Data source
+# Data source
 
@@
 Kristen Gorman with the [Palmer Station Long Term Ecological Research Program](https://pal.lternet.edu/), part of the [US Long Term Ecological Research Network](https://lternet.edu/).
-For complete data collection methods see [@gorman_plos].
-The data were imported directly from the [Environmental Data Initiative](https://environmentaldatainitiative.org/) (EDI) Data Portal [@adelie_edi; @gentoo_edi; @chinstrap_edi] and are available for use by CC0 license ("No Rights Reserved") in accordance with the [Palmer Station Data Policy](https://pal.lternet.edu/data/policies).
+For complete data collection methods and published analyses see [@gorman_plos].
 
-Data for each penguin species (Adélie, chinstrap and gentoo) were imported separately, then combined into a single table stored in the [*palmerpenguins*](https://allisonhorst.github.io/palmerpenguins/) package as object `penguins_raw`.
+Data were imported from the [Environmental Data Initiative](https://environmentaldatainitiative.org/) (EDI) Data Portal [@adelie_edi; @gentoo_edi; @chinstrap_edi](see description below), and are generously available for use by CC0 license ("No Rights Reserved") in accordance with the [Palmer Station Data Policy](https://pal.lternet.edu/data/policies) and the [LTER Data Access Policy](https://lternet.edu/data-access-policy/) for Type I data.
+
+# How to get the Palmer Archipelago penguin data
+
+The combined raw data for all three penguin species (penguins_raw.csv) and the curated penguin data (penguins.csv), which are identical to the `penguins_raw` and `penguins` data objects in the **palmerpenguins** *R* package described below, are available for download from the Journal of Statistics Education site.
+The data can be also accessed through EDI directly, or via existing packages for different programming languages.
+Options included here are based on availability as of July 2020, and may be updated or deleted by package maintainers.
+
+## Download or import from EDI
+
+The raw data for each penguin species (Adélie, Gentoo and Chinstrap) are stored in a separate data package, accessed through the EDI repository:
+
+- Adélie data: [KNB-LTER Data Package 219.5](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.219.5)
+- Gentoo data: [KNB-LTER Data Package 220.5](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.220.5)
+- Chinstrap data: [KNB-LTER Data Package 221.5](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.221.6)
+
+The links above contain all metadata information, and provides options to directly download the data as a comma-separated value (CSV) file *or* import for use directly using code provided (e.g.
+for R, Python, SAS, MatLab, or SPSS).
+
+## R: **palmerpenguins** package and GitHub repo
+
+For easier data access and use in *R*, we created the [**palmerpenguins**](https://allisonhorst.github.io/palmerpenguins/) data package.
+*R* users can install the development version of the package using:
+
+```
+library(remotes)
+remotes::install_github("allisonhorst/palmerpenguins")
+```
+
+The **palmerpenguins** package contains two data objects, described below:
+
+- `penguins_raw`: a table containing all raw data 
+- `penguins`: a simplified table containing only size measurements, sex, island and species
+
+### Raw data: `penguins_raw`
+
+Data for each penguin species (Adélie, chinstrap and gentoo) were imported from EDI, then the raw data were combined into a single table stored in the [*palmerpenguins*](https://allisonhorst.github.io/palmerpenguins/) package as object `penguins_raw`.
 Data in `penguins_raw` are in tidy format [@tidydata] and contain observations for 17 variables (Appendix Table 1) recorded completely or in part for 344 individual penguins.
+For full variable descriptions and metadata, visit the data packages through EDI.
 
@@
 
-## Selected comparisons: `penguins` and `iris`
+The package repository on GitHub also contains a comma separated value (CSV) file containing the curated `penguins` data.
+Users can access and import the penguins.csv file directly from the data-raw/ folder in the repository:
 
-We feel that the `penguins` data is generally useful and approachable for statistics education, and that it is uniquely well-suited to replace `iris`.
-Comparisons presented here are representative examples, and are not exhaustive.
-Additional examples and tutorials are shared in the [user-contributed examples](https://allisonhorst.github.io/palmerpenguins/articles/articles/user_contributions.html) article on the package website.
+## Penguins in other computing languages:
 
-### Structure and sample size
+### Python: `seaborn` data vizualization library
+
+### Julia: `PalmerPenguins.jl` package
+
+# Selected comparisons: `penguins` and `iris`
+
+We feel that the `penguins` dataset in the **palmerpenguins** *R* package is generally useful and approachable for statistics education, and that it is uniquely well-suited to replace `iris`.
+Comparisons presented here between `penguins` and `iris` are select examples for common `iris` uses in statistics education, and are not exhaustive.
+
+## Structure and sample size
 
@@
 by species) are similar (Table 3).
-Note that the additional factors in `penguins` (species, island and sex) create additional opportunities for grouping, compared to the single factor in `iris` (Species).
+Notably, sample sizes differ for the three penguin species, while sample sizes are equal in `iris`, and multiple factors in `penguins` (species, island and sex) create additional opportunities for grouping, compared to the single species factor in `iris`.
+We feel that the presence of missing values and unequal sample sizes in `penguins` creates opportunities for data exploration and analyses that are closer to what students might expect for data "in the wild."
 
@@
   mutate(sex = str_to_title(sex)) %>% 
-  rename(`Penguin species` = species) %>% 
-  count(`Penguin species`, sex) %>% 
-  pivot_wider(names_from = sex, values_from = n)
+  group_by(species, sex) %>% 
+  tally() %>% 
+  pivot_wider(names_from = sex, values_from = n) %>% 
+  replace_na(list(`NA` = 0)) %>% 
+  as.data.frame() %>% 
+  rename(`Penguin species` = species)
 
@@
   tab_spanner(
-    label = "Iris data",
+    label = "Iris sample size (by species)",
     columns = vars(`Iris species`, `Sample size`)
@@
   tab_spanner(
-    label = "Penguin data",
+    label = "Penguin sample size (by species and sex)",
     columns = vars(`Penguin species`, `Female`, `Male`, `NA`)
@@
   tab_footnote(
-    footnote = "Note: Here, penguin sample sizes are split by species and sex, and could be further divided by island.",
+    footnote = "Note: Here, penguin sample sizes are split by species and sex, and could be further divided by island. NA column for penguins indicate observations where sex was not recorded.",
     location = cells_column_labels(
@@
   )
+
 ```
 
-### Linear correlation & regression opportunities
+## Overview comparison: numeric variables
 
-#### Comparison 1: Linear trends across or within species
+Distributions and relationships between variables and groups can be similarly explored for the four numeric variables in `iris` (sepal width and length, petal width and length) and `penguins` (flipper length, body mass, bill length and depth) (Figure X A,B).
 
-An overall approximately linear relationship can be explored using petal dimensions from `iris`, and penguin size (flipper length versus body mass) in `penguins` (Figure X).
-Educators may explore how linear regression across all species of iris or penguins (Figure X A,B) compares to regression performed separately for each species (Figure X C,D).
+```{r, warning = FALSE, message = FALSE, echo = FALSE, fig.align = "center"}
+penguins %>%
+  select(species, where(is.numeric)) %>% 
+  ggpairs(aes(color = species, shape = species),
+          columns = c("flipper_length_mm", "body_mass_g", 
+                      "bill_length_mm", "bill_depth_mm"),
+          columnLabels = c("Flipper length (mm)","Body mass (g)", "Bill length (mm)", "Bill depth (mm)"),
+          upper = list(continuous = wrap("cor", size = 3))) +
+  scale_fill_manual(values = c("darkorange","purple","cyan4")) +
+  scale_color_manual(values = c("darkorange","purple","cyan4")) +
+  scale_shape_manual(values = c(15,16,17)) +
+  theme_minimal() +
+  theme(
+        text = element_text(size = 10, family = "Arial"),
+        panel.grid.major = element_line(colour = NA),
+        panel.grid.minor = element_blank(),
+        panel.border = element_rect(color = "gray80", fill = NA)
+        )
 
-```{r, echo = FALSE, warning = FALSE, message = FALSE}
+ggsave(here("figures","penguins_ggpairs.png"), height = 5, width = 9)
+
+iris %>%
+  ggpairs(aes(color = Species),
+          columns = c("Petal.Length", "Petal.Width", 
+                      "Sepal.Length", "Sepal.Width"),
+          columnLabels = c("Petal length (cm)","Petal width (cm)", "Sepal length (cm)", "Sepal width (cm)"),
+          upper = list(continuous = wrap("cor", size = 3))) +
+  scale_colour_manual(values = c("darkorange","purple","cyan4")) +
+  scale_fill_manual(values = c("darkorange","purple","cyan4")) +
+  theme_minimal() +
+  theme(
+        text = element_text(size = 10, family = "Arial"),
+        panel.grid.major = element_line(colour = NA),
+        panel.grid.minor = element_blank(),
+        panel.background = element_rect(fill = "gray90", color = "gray90")
+        )
+
+ggsave(here("figures","iris_ggpairs.png"), height = 5, width = 9)
+
+```
+
+## Linear correlation & regression
+
+As seen in Figures (X,X) there are a number of opportunities to explore linear relationships and correlations in both `penguins` and `iris`.
+Here, we point out one comparison that is uniquely similar:
+petal dimensions (petal length versus petal width) from `iris`, and penguin size (flipper length versus body mass) in `penguins` (Figure X).
+We see that the overall trend across species is relatively linear for both `iris` and `penguins` (Figure X A,B).
+Teachers may encourage students explore how results of regression performed across all three species of iris or penguins (Figure X A,B) compare to results for regression performed within species (Figure X C,D).
+
+Notably, distinctions between species are clearer for `iris` petals - particularly, the much smaller petals for *Iris setosa* - compared to `penguins`, in which Adélie and chinstrap penguins are largely overlapping in body size (body mass and flipper length), and are both generally smaller than gentoos.
+This difference is revisited in the k-means clustering example below.
+
+```{r, echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center"}
 
 penguin_flip_mass_scatter <- ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
-  geom_point(aes(color = species, shape = species), size = 1.5, alpha = 0.7) +
-  geom_smooth(method = "lm", aes(group = species, color = species), se = FALSE) +
+  geom_point(aes(color = species, shape = species), size = 1.5, alpha = 0.5) +
+  geom_smooth(method = "lm", aes(group = species, color = species), se = FALSE, show.legend = FALSE) +
   theme_minimal() +
   scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme(legend.position = c(0.3, 0.8)) +
+  theme(legend.position = "bottom") +
   labs(x = "Flipper length (mm)", 
        y = "Body mass (g)") +
-  theme(plot.title.position = "plot")
+  theme(plot.title.position = "plot",
+        panel.border = element_rect(fill = NA, color = "gray70"))
 
@@
   labs(x = "Flipper length (mm)", 
-       y = "Body mass (g)",
-       title = "Penguin size") +
-  theme(plot.title.position = "plot")
+       y = "Body mass (g)") +
+  theme(plot.title.position = "plot",
+        panel.border = element_rect(fill = NA, color = "gray70"))
 
 iris_petal_scatter <- ggplot(iris, aes(x = Petal.Length, y = Petal.Width)) +
-  geom_point(aes(color = Species, size = Species, shape = Species), size = 1.5, alpha = 0.7) +
-  geom_smooth(aes(group = Species, color = Species), method = "lm", se = FALSE) +
+  geom_point(aes(color = Species, size = Species, shape = Species), size = 1.5, alpha = 0.5) +
+  geom_smooth(aes(group = Species, color = Species), method = "lm", se = FALSE, show.legend = FALSE) +
   theme_minimal() +
   scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme(legend.position = c(0.3, 0.8)) +
+  theme(legend.position = "bottom") +
   labs(x = "Petal length (cm)",
        y = "Petal width (cm)") +
-  theme(plot.title.position = "plot")
+  theme(plot.title.position = "plot",
+        panel.background = element_rect(fill = "gray90", color = "gray90"))
 
@@
   labs(x = "Petal length (cm)",
-       y = "Petal width (cm)",
-       title = "Iris petal dimensions") +
-  theme(plot.title.position = "plot")
+       y = "Petal width (cm)") +
+  theme(plot.title.position = "plot",
+        panel.background = element_rect(fill = "gray90", color = "gray90"))
 
-(penguin_flip_mass_scatter_all + iris_petal_scatter_all) / (penguin_flip_mass_scatter + iris_petal_scatter)
+(penguin_flip_mass_scatter_all + iris_petal_scatter_all) / (penguin_flip_mass_scatter + iris_petal_scatter) + plot_annotation(tag_levels = 'A')
 
-ggsave(here("figures","iris_penguins_scatter.png"), height = 6, width = 6)
+ggsave(here("figures","iris_penguins_scatter.png"), height = 6, width = 8)
+
+# Simpson's Paradox example (bill dimensions)
+simpson_nospecies <- ggplot(data = penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
+  geom_point(color = "gray40", alpha = 0.6, size = 2) +
+  geom_smooth(method = "lm", se = FALSE, color = "black") +
+  theme_minimal() +
+  theme(panel.border = element_rect(fill = NA, color = "gray70")) +
+  labs(x = "Bill length (mm)", y = "Bill depth (mm)")
+
+simpson_wspecies <- ggplot(data = penguins, aes(x = bill_length_mm, y = bill_depth_mm, group = species)) +
+  geom_point(aes(color = species, shape = species), size = 2, alpha = 0.5) +
+  geom_smooth(method = "lm", se = FALSE, aes(color = species)) +
+  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
+  theme_minimal() +
+  theme(panel.border = element_rect(fill = NA, color = "gray70"),
+        legend.position = c(0.16,0.14)) +
+  labs(x = "Bill length (mm)", y = "Bill depth (mm)")
+
+simpson_gg <- (simpson_nospecies | simpson_wspecies) + plot_annotation(tag_levels = 'A')
+
+simpson_gg
+
+ggsave(here("figures","simpson_gg.png"), width = 8, height = 4)
 ```
 
-#### Comparison 2: linear relationships within species
+## Hypothesis testing
 
-More obvious clustering (by penguin species) exists when comparing bill length and depth (Figure X), and the relationship also provides an opportunity to introduce & explore Simpson's Paradox.
+Opportunities for hypothesis tests are similar using the four continuous quantitative variables in `penguins` and `iris`.
+Here, we describe replacement options for teachers currently using `iris` means comparisons using t-tests and one-way analysis of variance (ANOVA).
 
-```{r, echo = FALSE, warning = FALSE, message = FALSE}
-penguin_bill_scatter <- ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
-  geom_point(aes(color = species), size = 2, alpha = 0.7) +
-  theme_minimal() +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme(legend.position = "bottom") +
-  labs(x = "Bill length (mm)", 
-       y = "Bill depth (g)",
-       title = "Penguin size", 
-       color = "Species") +
-  theme(plot.title.position = "plot")
+### Two-sample t-tests
 
-iris_sepal_scatter <- ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) +
-  geom_point(aes(color = Species), size = 2, alpha = 0.7) +
-  theme_minimal() +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme(legend.position = "bottom") +
-  labs(x = "Sepal length (cm)",
-       y = "Sepal width (cm)",
-       title = "Iris sepal dimensions") +
-  theme(plot.title.position = "plot")
+TODO
 
-penguin_bill_scatter + iris_sepal_scatter
+### One-way analysis of variance (ANOVA)
+
+TODO
+
+## Principal components analysis
+
+The `iris` dataset appears frequently in examples for multivariate analyses, including principal component analysis (PCA).
+PCA is a dimensional reduction method that can be useful in capturing as much information about patterns in multivariate data as possible in a reduced number of dimensions (e.g., 2-dimensions for visualization).
+PCA with `iris` yields intuitive PC outcomes and clear clustering (Figure X) helpful for learners, as described in detail by others (ref, ref).
+
+PCA with `penguins` reveals a similar outcome (Figure X): two somewhat overlapping species (Adélie and Chinstrap) with one clearly separated (Gentoo), and intuitive loadings that can be related to scatterplots in Figure X.
+Percent of total variance captured by the first and second principal components are very similar between `iris` and `penguins` (Figure X).
+
+```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.align = "center"}
+# Copied from Alison Hill's post!
+
+### PENGUINS PCA: 
+penguin_recipe <-
+  recipe(~., data = penguins) %>% 
+  update_role(species, island, sex, new_role = "id") %>% 
+  step_naomit(all_predictors()) %>% 
+  step_normalize(all_predictors()) %>%
+  step_pca(all_predictors(), id = "pca") %>% 
+  prep()
+
+penguin_pca <- 
+  penguin_recipe %>% 
+  tidy(id = "pca")
+
+penguin_percvar <- penguin_recipe %>% 
+  tidy(id = "pca", type = "variance") %>% 
+  dplyr::filter(terms == "percent variance")
+
+# penguin_pca
+
+# Biplot:
+
+# get pca loadings into wider format
+pca_wider <- penguin_pca %>% 
+  tidyr::pivot_wider(names_from = component, id_cols = terms)
+
+# define arrow style:
+arrow_style <- arrow(length = unit(.05, "inches"),
+                     type = "closed")
+
+# Make the biplot:
+pca_plot <-
+  juice(penguin_recipe) %>%
+  ggplot(aes(PC1, PC2)) +
+  geom_point(aes(color = species, shape = species), 
+             alpha = 0.5, 
+             size = 2) +
+  scale_colour_manual(values = c("darkorange","purple","cyan4"))
+
+penguins_biplot <- pca_plot +
+  geom_segment(data = pca_wider,
+               aes(xend = PC1, yend = PC2), 
+               x = 0, 
+               y = 0, 
+               arrow = arrow_style) + 
+  geom_shadowtext(data = pca_wider,
+            aes(x = PC1, y = PC2, label = terms), 
+            nudge_x = c(0.7,0.7,1.7,1.2), 
+            nudge_y = c(-0.1,-0.2,0.1,-0.1),
+            size = 4, 
+            color = "black",
+            bg.color = "white") +
+  theme(legend.position = "bottom",
+        panel.border = element_rect(color = "gray70", fill = NA))
+
+# For positioning (above): 
+# 1: bill_length
+# 2: bill_depth
+# 3: flipper length
+# 4: body mass
+
+penguin_screeplot <- penguin_percvar %>% 
+  ggplot(aes(x = component, y = value)) + 
+  geom_col(fill = "gray50") + 
+  scale_x_continuous(limits = c(0, 5), expand = c(0,0)) + 
+  scale_y_continuous(limits = c(0,100), expand = c(0,0)) +
+  ylab("% of total variance") +
+  geom_text(aes(label = round(value,2)), vjust=-0.25) +
+  theme(panel.border = element_rect(color = "gray70", fill = NA))
+
+### IRIS PCA: 
+iris_recipe <-
+  recipe(~., data = iris) %>% 
+  update_role(Species, new_role = "id") %>% 
+  step_normalize(all_predictors()) %>%
+  step_pca(all_predictors(), id = "pca") %>% 
+  prep()
+
+iris_pca <- 
+  iris_recipe %>% 
+  tidy(id = "pca")
+
+# Biplot:
+
+# get pca loadings into wider format
+iris_wider <- iris_pca %>% 
+  tidyr::pivot_wider(names_from = component, id_cols = terms)
+
+# define arrow style:
+arrow_style <- arrow(length = unit(.05, "inches"),
+                     type = "closed")
+
+# Make the biplot:
+iris_pca_plot <-
+  juice(iris_recipe) %>%
+  ggplot(aes(PC1, PC2)) +
+  geom_point(aes(color = Species, shape = Species), 
+             alpha = 0.5, 
+             size = 2) +
+  scale_colour_manual(values = c("darkorange","purple","cyan4"))
+
+iris_biplot <- iris_pca_plot +
+  geom_segment(data = iris_wider,
+               aes(xend = PC1, yend = PC2), 
+               x = 0, 
+               y = 0, 
+               arrow = arrow_style) + 
+  geom_shadowtext(data = iris_wider,
+            aes(x = PC1, y = PC2, label = terms), 
+            nudge_x = c(0.5,0.3,1,1.2), 
+            nudge_y = c(-0.1,-0.2,0.1,-0.1),
+            size = 4, 
+            color = "black",
+            bg.color = "white") +
+  theme(panel.background = element_rect(fill = "gray90", color = "gray90"),
+        legend.position = "bottom")
+
+iris_percvar <- iris_recipe %>% 
+  tidy(id = "pca", type = "variance") %>% 
+  dplyr::filter(terms == "percent variance")
+
+iris_screeplot <- iris_percvar %>% 
+  ggplot(aes(x = component, y = value)) + 
+  geom_col(fill = "gray50") + 
+  scale_x_continuous(limits = c(0, 5), expand = c(0,0)) + 
+  scale_y_continuous(limits = c(0,100), expand = c(0,0)) +
+  ylab("% of total variance") +
+  geom_text(aes(label = round(value,2)), vjust=-0.25) +
+  theme(panel.background = element_rect(fill = "gray90", color = "gray90"))
+
+(penguins_biplot | iris_biplot) / (penguin_screeplot | iris_screeplot) + plot_annotation(tag_levels = 'A')
+
+ggsave(here("figures","pca_plots.png"), width = 8, height = 8)
 ```
 
-### Differences between groups
+## K-means clustering
 
-As in `iris`, multiple opportunities for comparisons between groups also exist in `penguins` (Figure X).
-Notably, more obvious separation between species exists in `iris` (Figure X E & F), compared to any between-species comparisons in `penguins` (Figure X A - D).
+Unsupervised clustering by k-means is a common and popular entryway to machine learning and classification, and frequently employs the `iris` data for introductory examples (refs).
+The `penguins` data provides similar opportunities for demonstrating, and illustrating limitations of, k-means clustering.
 
-```{r, include = FALSE}
+Here, we compare k-means clustering using only two variables for each dataset: for `iris`, petal width and petal length, and for `penguins`, bill length and bill depth.
+All included variables are scaled prior to k-means by Z-score standardization.
+Three clusters (*k* = 3) are specified for each since there are three species of both *Iris* (*setosa*, *versicolor*, and *virginica*) and penguins (Adélie, chinstrap and gentoo).
 
-# CHECK THIS ALL!!! 
-# ANOVA to see if flipper lengths are significantly different...(this says yes, but want other eyes)
-flipper_lm <- lm(flipper_length_mm ~ species, data = penguins)
+K-means with `penguins` and `iris` bill and petal dimensions, respectively, yields largely distinct clusters each dominated by one species (Figure X, Table X).
 
-flipper_anova <- Anova(flipper_lm, type = "III")
-flipper_anova
+```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.align = "center"}
+# TWO VARIABLE k-means comparison
+# Penguins: Bill length vs. bill depth
+# Iris: petal length vs. petal width
 
-flip_aov <- aov(flipper_lm)
-summary(flip_aov)
+p_bills <- penguins %>% 
+  select(species, bill_length_mm, bill_depth_mm) %>% 
+  drop_na()
 
-# TukeyHSD(flip_aov) all significantly different
+p_bills_nospecies <- p_bills %>% 
+  select(-species) %>% 
+  scale() %>% 
+  as.data.frame()
+
+# Perform k-means on penguin bill dimensions (k = 3, w/20 centroid starts)
+p_bills_kmeans <- kmeans(p_bills_nospecies, centers = 3, nstart = 20)
+
+# Bind clusters to the data
+p_kmeans_clusters <- cbind(p_bills_nospecies, cluster = as.factor(p_bills_kmeans$cluster), species = p_bills$species)
+
+# Plot clusters
+p_kmeans_gg <- ggplot(p_kmeans_clusters, aes(x = bill_length_mm, y = bill_depth_mm)) +
+  geom_text(aes(label = cluster, color = species)) +
+  theme(legend.position = "bottom") +
+  scale_color_manual(values = c("darkorange","darkorchid","cyan4"))
+
+# Get counts in each cluster by species
+p_cluster_counts <- p_kmeans_clusters %>% 
+  count(species, cluster) %>% 
+  pivot_wider(names_from = species, values_from = n) %>% 
+  arrange(cluster)
 
@@
 
-```{r, echo = FALSE, warning = FALSE, message = FALSE}
+```{r, echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center"}
 
-# Update this with an actual theme FFS and use '&' in patchwork
+iris_petals <- iris %>% 
+  select(Species, Petal.Length, Petal.Width)
 
-iris_pw <- ggplot(iris, aes(x = Species, y = Petal.Width)) +
-  geom_boxplot(aes(color = Species),
-               show.legend = FALSE) +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme_bw() +
-  theme(plot.title.position = "plot") +
-  labs(x = "",
-       y = "Petal width (cm)") +
-  coord_flip()
+# remove species factor & scale petal dimensions: 
+ip_nospecies <- iris_petals %>% 
+  select(-Species) %>% 
+  scale() %>% 
+  as.data.frame()
 
-iris_pl <- ggplot(iris, aes(x = Species, y = Petal.Length)) +
-  geom_boxplot(aes(color = Species),
-               show.legend = FALSE) +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme_bw() +
-  theme(plot.title.position = "plot") +
-  labs(x = "",
-       y = "Petal length (cm)") +
-  coord_flip()
+# Perform k-means on iris petal dimensions (k = 3, w/20 centroid starts)
+set.seed(1967)
+ip_kmeans <- kmeans(ip_nospecies, centers = 3, nstart = 20)
 
-iris_sl <- ggplot(iris, aes(x = Species, y = Sepal.Length)) +
-  geom_boxplot(aes(color = Species),
-               show.legend = FALSE) +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme_bw() +
-  theme(plot.title.position = "plot") +
-  labs(x = "",
-       y = "Sepal length (cm)") +
-  coord_flip()
+# Bind clusters to the data
+iris_kmeans_clusters <- cbind(ip_nospecies, cluster = as.factor(ip_kmeans$cluster), species = iris$Species)
 
-iris_sw <- ggplot(iris, aes(x = Species, y = Sepal.Width)) +
-  geom_boxplot(aes(color = Species),
-               show.legend = FALSE) +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme_bw() +
-  theme(plot.title.position = "plot") +
-  labs(x = "",
-       y = "Sepal width (cm)") +
-  coord_flip()
+# Plot clusters: 
+iris_kmeans_gg <- ggplot(iris_kmeans_clusters, aes(x = Petal.Length, y = Petal.Width)) +
+  geom_text(aes(label = cluster, color = species)) +
+   theme(legend.position = "bottom",
+         panel.background = element_rect(fill = "gray92", color = "gray90")) +
+  scale_color_manual(values = c("darkorange","darkorchid","cyan4"))
 
-penguins_flip <- penguins %>%
-  ggplot(aes(x = species, y = flipper_length_mm)) +
-  geom_boxplot(aes(color = species),
-               show.legend = FALSE) +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme_bw() +
-  labs(x = "",
-       y = "Flipper length (mm)") +
-  coord_flip()
-
-penguins_mass <- penguins %>%
-  ggplot(aes(x = species, y = body_mass_g)) +
-  geom_boxplot(aes(color = species),
-               show.legend = FALSE) +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme_bw() +
-  labs(x = "",
-       y = "Body mass (g)") +
-  coord_flip()
-
-penguins_bill_length <- penguins %>%
-  ggplot(aes(x = species, y = bill_length_mm)) +
-  geom_boxplot(aes(color = species),
-               show.legend = FALSE) +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme_bw()+
-  labs(x = "",
-       y = "Bill length (mm)") +
-  coord_flip()
-
-penguins_bill_depth <- penguins %>%
-  ggplot(aes(x = species, y = bill_depth_mm)) +
-  geom_boxplot(aes(color = species),
-               show.legend = FALSE) +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme_bw() +
-  labs(x = "",
-       y = "Bill depth (mm)") +
-  coord_flip()
-
-penguins_box <- penguins_flip / penguins_mass / penguins_bill_length / penguins_bill_depth + plot_layout(guides = "collect") + plot_annotation(title = "Penguins")
-
-iris_box <- iris_pw / iris_pl / iris_sw / iris_sl + plot_layout(guides = "collect")
-
-penguins_box | iris_box + plot_annotation(
-  title = 'Penguin and iris sizes',
-  subtitle = 'Penguin size measurements (left) and iris petal and sepal dimensions (right)',
-  tag_levels = 'A'
-)
-
-ggsave(here("figures","penguins_iris_box.png"), height = 7, width = 8)
-
+# Get counts in each cluster by species
+iris_cluster_counts <- iris_kmeans_clusters %>% 
+  count(species, cluster) %>% 
+  pivot_wider(names_from = species, values_from = n) %>% 
+  arrange(cluster)
 ```
 
-### Hypothesis testing
+```{r, echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center"}
+(p_kmeans_gg | iris_kmeans_gg) + plot_annotation(tag_levels = "A")
 
-### Principle components analysis
+kmeans_2var_table <- cbind(p_cluster_counts, iris_cluster_counts) %>% 
+  kable() %>% 
+  kable_styling(full_width = FALSE) %>% 
+  column_spec(c(5:8), background = "gainsboro") %>% 
+  add_header_above(c("Penguins cluster assignments" = 4, "Iris cluster assignments" = 4))
 
-### Clustering (e.g. k-means?)
+kmeans_2var_table
+```
 
-### Classification (Alison Hill blog post?)
+## Machine learning (species classification)
 
-## Penguin data access for non-R users
+# Conclusion
 
-The Palmer Archipelago penguins data are available to non-R users by direct import from the Environmental Data Initiative, from GitHub, and as a stored dataset in existing packages.
-Below we share several current ways to access the data (as of July 2020 - we expect availability to expand in the future).
+Through side-by-side comparisons of data structure, size, and analysis outcomes, we have shown that the Palmer Archipelago penguins dataset can serve as a direct replacement for `iris` for many use cases in statistics education.
 
-### Download or import directly from Environmental Data Initiative
+# Acknowledgements
 
-The [Environmental Data Initiative](https://environmentaldatainitiative.org/) is "an NSF-funded project, actively promoting and enabling curation and re-use of environmental data." The raw data for each penguin species (Adélie, Gentoo and Chinstrap) are stored in a separate data file at the following links (see full citations in Citations):
+We would like to thank the following people for their contributions: Marty Downs (NCEAS) for help with the Palmer Archipelago penguins data access and license information, Dr.
+Julien Brun (NCEAS) for code to directly import the data from EDI into *R*, and Dr.
+Hadley Wickham for contributions to the **palmerpenguins** package.
+[Just a placeholder - any updates / additions / changes go ahead & add...]
 
-- Adélie data: [KNB-LTER Data Package 219.5](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.219.5)
-- Gentoo data: [KNB-LTER Data Package 220.5](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.220.5)
-- Chinstrap data: [KNB-LTER Data Package 221.5](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.221.6)
+# Appendix
 
-The links above contain all metadata information, and provides options to directly download the data as a comma-separated value (CSV) file *or* import for use directly using code provided (e.g.
-for R, Python, SAS, MatLab, or SPSS).
-
-### Download CSV from the palmerpenguins GitHub repo
-
-A CSV containing curated data for all three species (accessed directly from EDI) exists in the 'data-raw' folder in the palmerpenguins repo on GitHub.
-GitHub users can fork / clone the repo, and will then have the cleaned CSV containing all 3 penguin species.
-
-### The penguins in other packages & languages
-
-The Palmer Station penguin data has been added to existing libraries, including (non-exhaustive):
-
-#### Python modules
-
-- Seaborn
-- blah blah
-
-See examples for working with penguins in Python here:
-
-#### Kaggle
-
-- here
-
-#### Elsewhere?
-
-## Conclusion
-
-## Appendix
-
-### Appendix A: Variables in the raw data
+## Appendix A: Variables in the raw data
 
@@
 
-# Random other stuff
-
-Compared to the single factor in `iris` (Species), `penguins`' three factors (species, island, sex) increase opportunities for comparisons between groups (Figure X).
-
-```{r, echo = FALSE, warning = FALSE, message = FALSE}
-
-cd <- penguins %>% 
-  filter(sex %in% c("male","female")) %>% 
-ggplot(aes(x = bill_depth_mm)) +
-  geom_histogram(aes(fill = species), alpha = 0.6, position = "identity") +
-  scale_fill_manual(values = c("darkorange","darkorchid","cyan4")) +
-  facet_wrap(~ sex) +
-  theme_minimal()
-
-cl <- penguins %>% 
-  filter(sex %in% c("male","female")) %>% 
-ggplot(aes(x = bill_length_mm)) +
-  geom_histogram(aes(fill = species), alpha = 0.6, position = "identity") +
-  scale_fill_manual(values = c("darkorange","darkorchid","cyan4")) +
-  facet_wrap(~ sex) +
-  theme_minimal()
-
-fl <- penguins %>% 
-  filter(sex %in% c("male","female")) %>% 
-ggplot(aes(x = flipper_length_mm)) +
-  geom_histogram(aes(fill = species), alpha = 0.6, position = "identity") +
-  scale_fill_manual(values = c("darkorange","darkorchid","cyan4")) +
-  facet_wrap(~ sex) +
-  theme_minimal()
-
-bm <- penguins %>% 
-  filter(sex %in% c("male","female")) %>% 
-ggplot(aes(x = body_mass_g)) +
-  geom_histogram(aes(fill = species), alpha = 0.6, position = "identity") +
-  scale_fill_manual(values = c("darkorange","darkorchid","cyan4")) +
-  facet_wrap(~ sex) +
-  theme_minimal()
-
-cd / cl / fl / bm + plot_layout(guides = "collect")
-```
-
-## Differences by islands?
-
-```{r}
-# No difference in Adelie across islands: Chinstrap only on Dream, Gentoo only on Torgerson
-penguins %>% 
-  ggplot(aes(x = island, y = body_mass_g)) +
-  geom_point() +
-  facet_wrap(~species)
-
-penguins %>% 
-  ggplot(aes(x = island, y = flipper_length_mm)) +
-  geom_point() +
-  facet_wrap(~species)
-
-penguins %>% 
-  ggplot(aes(x = island, y = bill_length_mm)) +
-  geom_point() +
-  facet_wrap(~species)
-
-penguins %>% 
-  ggplot(aes(x = island, y = bill_depth_mm)) +
-  geom_point() +
-  facet_wrap(~species)
-```
-
-A Web of Science search reveals that Fisher's 1936 paper has been cited in 7,258 publications (as of June 2020), most often in journals publishing in fields of Computer Science (57% of 7,258 citing references), Mathematics (49%), and Engineering (29%) - most of which we expect are crediting Anderson's *Iris* data, and not Fisher's eugenics work.
-
-# End random other stuff
-
-## Citations
+# Citations

### 2020-07-11 manuscript/manuscript_text.Rmd
add tidy kmeans

@@
 library(shadowtext)
+library(broom)
 knitr::opts_chunk$set(echo = TRUE)
@@
 
-p_bills <- penguins %>% 
-  select(species, bill_length_mm, bill_depth_mm) %>% 
+pb_species <- penguins %>% 
+  select(species, starts_with("bill")) %>% 
   drop_na()
 
-p_bills_nospecies <- p_bills %>% 
+pb_nospecies <- pb_species %>% 
   select(-species) %>% 
-  scale() %>% 
-  as.data.frame()
+  recipe() %>% 
+  step_normalize(all_numeric()) %>% 
+  prep() %>% 
+  juice()
 
 # Perform k-means on penguin bill dimensions (k = 3, w/20 centroid starts)
-p_bills_kmeans <- kmeans(p_bills_nospecies, centers = 3, nstart = 20)
-
-# Bind clusters to the data
-p_kmeans_clusters <- cbind(p_bills_nospecies, cluster = as.factor(p_bills_kmeans$cluster), species = p_bills$species)
+# save augmented data
+pb_clust <- 
+  pb_nospecies %>% 
+  kmeans(centers = 3, nstart = 20) %>% 
+  broom::augment(pb_species)
 
 # Plot clusters
-p_kmeans_gg <- ggplot(p_kmeans_clusters, aes(x = bill_length_mm, y = bill_depth_mm)) +
-  geom_text(aes(label = cluster, color = species)) +
+pb_kmeans_gg <- 
+  pb_clust %>% 
+  ggplot(aes(x = bill_length_mm, y = bill_depth_mm)) +
+  geom_text(aes(label = .cluster, color = species),
+            key_glyph = draw_key_rect) +
   theme(legend.position = "bottom") +
@@
 # Get counts in each cluster by species
-p_cluster_counts <- p_kmeans_clusters %>% 
-  count(species, cluster) %>% 
-  pivot_wider(names_from = species, values_from = n) %>% 
-  arrange(cluster)
-
+pb_clust_n <- pb_clust %>% 
+  count(species, .cluster) %>% 
+  pivot_wider(names_from = species, values_from = n, names_sort = TRUE) %>% 
+  arrange(.cluster)
 ```
@@
 
-iris_petals <- iris %>% 
-  select(Species, Petal.Length, Petal.Width)
+ip_species <- iris %>% 
+  select(Species, starts_with("Petal"))
 
 # remove species factor & scale petal dimensions: 
-ip_nospecies <- iris_petals %>% 
+ip_nospecies <- ip_species %>% 
   select(-Species) %>% 
-  scale() %>% 
-  as.data.frame()
+  recipe() %>% 
+  step_normalize(all_numeric()) %>% 
+  prep() %>% 
+  juice()
 
 # Perform k-means on iris petal dimensions (k = 3, w/20 centroid starts)
-set.seed(1967)
-ip_kmeans <- kmeans(ip_nospecies, centers = 3, nstart = 20)
+ip_clust <- 
+  ip_nospecies %>% 
+  kmeans(centers = 3, nstart = 20) %>% 
+  broom::augment(ip_species)
 
-# Bind clusters to the data
-iris_kmeans_clusters <- cbind(ip_nospecies, cluster = as.factor(ip_kmeans$cluster), species = iris$Species)
-
-# Plot clusters: 
-iris_kmeans_gg <- ggplot(iris_kmeans_clusters, aes(x = Petal.Length, y = Petal.Width)) +
-  geom_text(aes(label = cluster, color = species)) +
+# Plot clusters
+ip_kmeans_gg <- 
+  ip_clust %>% 
+  ggplot(aes(x = Petal.Length, y = Petal.Width)) +
+  geom_text(aes(label = .cluster, color = Species), 
+            key_glyph = draw_key_rect) +
   theme(legend.position = "bottom",
@@
 # Get counts in each cluster by species
-iris_cluster_counts <- iris_kmeans_clusters %>% 
-  count(species, cluster) %>% 
-  pivot_wider(names_from = species, values_from = n) %>% 
-  arrange(cluster)
+ip_clust_n <- ip_clust %>% 
+  count(Species, .cluster) %>% 
+  pivot_wider(names_from = Species, values_from = n, names_sort = TRUE) %>% 
+  arrange(.cluster)
 ```
@@
 ```{r, echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center"}
-(p_kmeans_gg | iris_kmeans_gg) + plot_annotation(tag_levels = "A")
+(pb_kmeans_gg | ip_kmeans_gg) + plot_annotation(tag_levels = "A")
 
-kmeans_2var_table <- cbind(p_cluster_counts, iris_cluster_counts) %>% 
+kmeans_2var_table <- cbind(pb_clust_n, ip_clust_n) %>% 
   kable() %>% 

### 2020-07-12 manuscript/manuscript_text.Rmd
remove text (code only version)

@@
 ---
-title: 'Palmer Archipelago (Antarctica) penguins: an alternative to Edgar Anderson''s
-  iris data'
+title: 'Palmer Penguins manuscript code'
 output:
@@
 
-# Keywords
+# MANUSCRIPT CODE
 
-Multivariate analysis, cluster analysis, correlation, classification, zoology, penguin morphology
-
-# Abstract
-
-Edgar Anderson's *Iris* data are ubiquitous in statistics and data science education, but their use in Ronald A.
-Fisher's eugenics research has created momentum to switch to alternative datasets.
-Size measurements for three penguin species observed on islands in Palmer Archipelago, Antarctica provide a suitable, approachable and charismatic alternative to the *Iris* data.
-Here, we introduce the penguin data, then directly compare the two datasets to demonstrate that educators currently using *Iris* can switch to the Palmer Archipelago penguins for many use cases including data wrangling and visualization, correlation, regression, multivariate analysis (e.g.
-PCA), cluster analysis and classification (e.g.
-by k-means).
-
-# Introduction
-
-In 1935, American botanist Edgar Anderson measured petal and sepal dimensions (length and width) for 50 samples from three *Iris* species: *Iris setosa*, *Iris versicolor*, and *Iris virginica* [@anderson].
-The manageable but non-trivial size (5 variables and 150 total observations) and characteristics of the data make it amenable for introducing a wide range of statistical methods including data wrangling, visualization, regression, cluster analysis, machine learning, and more.
-Anderson's *Iris* dataset is even built into a number of software packages, including in the auto-installed *datasets* package in *R* (R Core Team), Python's *scikit-learn* machine learning library [@scikit-learn], and the SAS *Sashelp* library, further facilitating its widespread use.
-As a result, eighty-five years after the data were initially published, Anderson’s *Iris* measurements are ubiquitous in statistics and data science education.
-
-*Sounds great. Why consider switching?*
-
-In 1936, one year after publication in the *Bulletin of the American Iris Society*, Anderson’s *Iris* measurements were further analyzed and published by eugenicist and statistician Ronald A.
-Fisher in *Annals of Eugenics* [@fisher].
-Regardless of Anderson's initial motivation, the data today remain inextricably linked to Fisher’s eugenics research and are even commonly (if unfairly) referred to as "*Fisher's* iris data".
-Due to its history in - and strong association with - Fisher's work, there is momentum for teachers to (1) address the history of the *Iris* data in eugenics research when teaching with it, or (2) replace the *Iris* data with an alternative dataset that offers similar teaching and learning opportunities.
-This paper focuses on (2): a replacement for the *Iris* data.
-
-Considering *Iris* data usage in statistics education, we established the following criteria for a suitable replacement dataset:
-
-- Available by an appropriate license, ideally CC0 (no rights reserved)
-- Feature intuitive subjects and variables understandable to learners across fields
-- Manageable - but not trivial - in size
-- Real-world (not manufactured) data for improved understanding and engagement [@cafe]
-- Provide similar opportunities for teaching data science skills and statistical methods
-
-Here, we offer an alternative to Anderson's *Iris* data that largely satisfies those criteria: an approachable and charismatic dataset containing real-world size measurements for three penguin species observed on islands in Palmer Archipelago, Antarctica.
-Comparable size, characteristics and learning opportunities make the penguins data uniquely transferable for educators currently using *Iris*.
-By comparing data structure, size, and a range of analyses side-by-side, we demonstrate that the Palmer Archipelago penguin measurements can replace Anderson's iris data for many use cases.
-
-# Data source
-
-Size measurements, clutch observations, and blood isotope levels for male and female nesting adult Adélie (*Pygoscelis adeliae*), chinstrap (*Pygoscelis antarcticus*), and gentoo (*Pygoscelis papua*) penguins on three islands in the Palmer Archipelago were collected from 2007 - 2009 by Dr.
-Kristen Gorman with the [Palmer Station Long Term Ecological Research Program](https://pal.lternet.edu/), part of the [US Long Term Ecological Research Network](https://lternet.edu/).
-For complete data collection methods and published analyses see [@gorman_plos].
-
-Data were imported from the [Environmental Data Initiative](https://environmentaldatainitiative.org/) (EDI) Data Portal [@adelie_edi; @gentoo_edi; @chinstrap_edi](see description below), and are generously available for use by CC0 license ("No Rights Reserved") in accordance with the [Palmer Station Data Policy](https://pal.lternet.edu/data/policies) and the [LTER Data Access Policy](https://lternet.edu/data-access-policy/) for Type I data.
-
-# How to get the Palmer Archipelago penguin data
-
-The combined raw data for all three penguin species (penguins_raw.csv) and the curated penguin data (penguins.csv), which are identical to the `penguins_raw` and `penguins` data objects in the **palmerpenguins** *R* package described below, are available for download from the Journal of Statistics Education site.
-The data can be also accessed through EDI directly, or via existing packages for different programming languages.
-Options included here are based on availability as of July 2020, and may be updated or deleted by package maintainers.
-
-## Download or import from EDI
-
-The raw data for each penguin species (Adélie, Gentoo and Chinstrap) are stored in a separate data package, accessed through the EDI repository:
-
-- Adélie data: [KNB-LTER Data Package 219.5](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.219.5)
-- Gentoo data: [KNB-LTER Data Package 220.5](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.220.5)
-- Chinstrap data: [KNB-LTER Data Package 221.5](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.221.6)
-
-The links above contain all metadata information, and provides options to directly download the data as a comma-separated value (CSV) file *or* import for use directly using code provided (e.g.
-for R, Python, SAS, MatLab, or SPSS).
-
-## R: **palmerpenguins** package and GitHub repo
-
-For easier data access and use in *R*, we created the [**palmerpenguins**](https://allisonhorst.github.io/palmerpenguins/) data package.
-*R* users can install the development version of the package using:
-
-```
-library(remotes)
-remotes::install_github("allisonhorst/palmerpenguins")
-```
-
-The **palmerpenguins** package contains two data objects, described below:
-
-- `penguins_raw`: a table containing all raw data 
-- `penguins`: a simplified table containing only size measurements, sex, island and species
-
-### Raw data: `penguins_raw`
-
-Data for each penguin species (Adélie, chinstrap and gentoo) were imported from EDI, then the raw data were combined into a single table stored in the [*palmerpenguins*](https://allisonhorst.github.io/palmerpenguins/) package as object `penguins_raw`.
-Data in `penguins_raw` are in tidy format [@tidydata] and contain observations for 17 variables (Appendix Table 1) recorded completely or in part for 344 individual penguins.
-For full variable descriptions and metadata, visit the data packages through EDI.
-
-### Curated `penguins` data
-
-As a direct alternative to Anderson's *Iris* data, we recommend using the curated `penguins` object in the **palmerpenguins** *R* data package.
-The `penguins` data is a subset of `penguins_raw` retaining all 344 rows (individual penguins) but reduced to the following seven variables:
-
-- **species:** a factor denoting the penguin species (Adelie, Chinstrap, or Gentoo)
-- **island:** a factor denoting the island (in Palmer Archipelago, Antarctica) where observed (Biscoe, Dream or Torgersen)
-- **bill_length_mm:** a number denoting length of the dorsal ridge of penguin bill (millimeters)
-- **bill_depth_mm:** a number denoting the depth of the penguin bill (millimeters)
-- **flipper_length_mm:** an integer denoting penguin flipper length (millimeters)
-- **body_mass_g:** an integer denoting penguin body mass (grams)
-- **sex:** a factor denoting penguin sex (male, female)
-
-The `penguins` subset contains `r sum(complete.cases(penguins))` complete cases, with missing values (n~missing~ = `r sum(is.na(penguins))`, or `r round(100*sum(is.na(penguins))/(7*344),2)` % of all values) indicated by `NA` throughout.
-In addition to reduced variables, the curated `penguins` subset is minimally updated from `penguins_raw` as follows:
-
-1. All variable names are converted to lower snake case
-2. Factor levels for species are truncated to only include the common name (e.g. "Gentoo", instead of "Gentoo penguin (*Pygoscelis papua*)")
-3. Recorded sex for penguin N36A1, originally recorded as ".", is updated to `NA`
-4. "culmen_length_mm" and "culmen_depth_mm" variable names are updated to "bill_length_mm" and "bill_depth_mm" to be more intuitive
-
-The package repository on GitHub also contains a comma separated value (CSV) file containing the curated `penguins` data.
-Users can access and import the penguins.csv file directly from the data-raw/ folder in the repository:
-
-## Penguins in other computing languages:
-
-### Python: `seaborn` data vizualization library
-
-### Julia: `PalmerPenguins.jl` package
-
-# Selected comparisons: `penguins` and `iris`
-
-We feel that the `penguins` dataset in the **palmerpenguins** *R* package is generally useful and approachable for statistics education, and that it is uniquely well-suited to replace `iris`.
-Comparisons presented here between `penguins` and `iris` are select examples for common `iris` uses in statistics education, and are not exhaustive.
+### See shared googledoc for MS text
 
 ## Structure and sample size
-
-Both `iris` and `penguins` datasets are in tidy format [@tidydata], with each column denoting a single variable and each row containing a single observation (Table 1, 2).
 
@@
 sample_n(iris, 6) %>% 
-  kable(caption = "Table 1. Iris dataset, 6 random observations (of 150 total)") %>% 
+  kable() %>% 
   kable_styling(full_width = FALSE)
@@
 sample_n(penguins, 6) %>% 
-  kable(caption = "Table 2. Penguins dataset, 6 random observations (of 344 total)") %>% 
+  kable() %>% 
   kable_styling(full_width = FALSE)
 ```
-
-Overall sizes are comparable: dimensions (# columns $\times$ # rows) are 5 $\times$ 150 and 7 $\times$ 344 for `iris` and `penguins`, respectively, and sample sizes within groups (e.g.
-by species) are similar (Table 3).
-Notably, sample sizes differ for the three penguin species, while sample sizes are equal in `iris`, and multiple factors in `penguins` (species, island and sex) create additional opportunities for grouping, compared to the single species factor in `iris`.
-We feel that the presence of missing values and unequal sample sizes in `penguins` creates opportunities for data exploration and analyses that are closer to what students might expect for data "in the wild."
 
@@
 ## Overview comparison: numeric variables
-
-Distributions and relationships between variables and groups can be similarly explored for the four numeric variables in `iris` (sepal width and length, petal width and length) and `penguins` (flipper length, body mass, bill length and depth) (Figure X A,B).
 
@@
 ## Linear correlation & regression
-
-As seen in Figures (X,X) there are a number of opportunities to explore linear relationships and correlations in both `penguins` and `iris`.
-Here, we point out one comparison that is uniquely similar:
-petal dimensions (petal length versus petal width) from `iris`, and penguin size (flipper length versus body mass) in `penguins` (Figure X).
-We see that the overall trend across species is relatively linear for both `iris` and `penguins` (Figure X A,B).
-Teachers may encourage students explore how results of regression performed across all three species of iris or penguins (Figure X A,B) compare to results for regression performed within species (Figure X C,D).
-
-Notably, distinctions between species are clearer for `iris` petals - particularly, the much smaller petals for *Iris setosa* - compared to `penguins`, in which Adélie and chinstrap penguins are largely overlapping in body size (body mass and flipper length), and are both generally smaller than gentoos.
-This difference is revisited in the k-means clustering example below.
 
@@
 
-## Hypothesis testing
-
-Opportunities for hypothesis tests are similar using the four continuous quantitative variables in `penguins` and `iris`.
-Here, we describe replacement options for teachers currently using `iris` means comparisons using t-tests and one-way analysis of variance (ANOVA).
-
-### Two-sample t-tests
-
-TODO
-
-### One-way analysis of variance (ANOVA)
-
-TODO
-
-## Principal components analysis
-
-The `iris` dataset appears frequently in examples for multivariate analyses, including principal component analysis (PCA).
-PCA is a dimensional reduction method that can be useful in capturing as much information about patterns in multivariate data as possible in a reduced number of dimensions (e.g., 2-dimensions for visualization).
-PCA with `iris` yields intuitive PC outcomes and clear clustering (Figure X) helpful for learners, as described in detail by others (ref, ref).
-
-PCA with `penguins` reveals a similar outcome (Figure X): two somewhat overlapping species (Adélie and Chinstrap) with one clearly separated (Gentoo), and intuitive loadings that can be related to scatterplots in Figure X.
-Percent of total variance captured by the first and second principal components are very similar between `iris` and `penguins` (Figure X).
+## PCA
 
@@
 
-Unsupervised clustering by k-means is a common and popular entryway to machine learning and classification, and frequently employs the `iris` data for introductory examples (refs).
-The `penguins` data provides similar opportunities for demonstrating, and illustrating limitations of, k-means clustering.
-
-Here, we compare k-means clustering using only two variables for each dataset: for `iris`, petal width and petal length, and for `penguins`, bill length and bill depth.
-All included variables are scaled prior to k-means by Z-score standardization.
-Three clusters (*k* = 3) are specified for each since there are three species of both *Iris* (*setosa*, *versicolor*, and *virginica*) and penguins (Adélie, chinstrap and gentoo).
-
-K-means with `penguins` and `iris` bill and petal dimensions, respectively, yields largely distinct clusters each dominated by one species (Figure X, Table X).
-
 ```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.align = "center"}
@@
 ```
-
-## Machine learning (species classification)
-
-# Conclusion
-
-Through side-by-side comparisons of data structure, size, and analysis outcomes, we have shown that the Palmer Archipelago penguins dataset can serve as a direct replacement for `iris` for many use cases in statistics education.
-
-# Acknowledgements
-
-We would like to thank the following people for their contributions: Marty Downs (NCEAS) for help with the Palmer Archipelago penguins data access and license information, Dr.
-Julien Brun (NCEAS) for code to directly import the data from EDI into *R*, and Dr.
-Hadley Wickham for contributions to the **palmerpenguins** package.
-[Just a placeholder - any updates / additions / changes go ahead & add...]
-
-# Appendix
-
-## Appendix A: Variables in the raw data
-
-- **studyName:** Sampling expedition indicator
-- **Sample Number:** Integer denoting the continuous numbering sequence for each penguin
-- **Species:** a factor denoting the penguin species (Adelie, Chinstrap, or Gentoo)
-- **Region**: A character denoting the region of the Palmer LTER sampling grid
-- **Island:** a factor denoting the island where observed (Torgersen, Dream or Biscoe) in Palmer Archipelago
-- **Stage:** A character string denoting reproductive stage at sampling
-- **Individual ID:** A character string denoting a unique ID for each penguin
-- **Clutch Completion:** A character string denoting if the study next observed with a full clutch, i.e. 2 eggs
-- **Date Egg:** A date denoting the date study nest observed with 1 egg
-- **Culmen Length:** A number denoting the length of the dorsal ridge of a bird's bill (millimeters)
-- **Culmen Depth:** A number denoting the depth of the dorsal ridge of a bird's bill (millimeters)
-- **Flipper Length:** An integer denoting the penguin flipper length (millimeters)
-- **Body Mass:** An integer denoting the penguin body mass (grams)
-- **Sex:** A character string denoting the sex of an animal
-- **Delta 15 N:** A number denoting the measure of the ratio of stable isotopes 15N:14N
-- **Delta 13 C:** A number denoting the measure of the ratio of stable isotopes 13C:12C
-- **Comments:** A character string with text providing additional relevant information for data
-
-# Citations

### 2020-07-20 penguins.Rmd
add R Journal rticle template

---
title: Capitalized Title Here
author:
  - name: Allison Horst
    affiliation: Bren School of Environmental Science and Management, University of California, Santa Barbara
    address:
    - line 1
    - line 2
    email:  ahorst@ucsb.edu
  - name: Alison Hill
    affiliation: RStudio, PBC
    address:
    - line 1
    - line 2
    email:  alison@rstudio.com
  - name: Kristen Gorman
    affiliation: University of Alaska Fairbanks College of Fisheries and Ocean Sciences
    address:
    - line 1
    - line 2
    email:  kbgorman@alaska.edu
abstract: >
  An abstract of less than 150 words.
preamble: |
  % Any extra LaTeX you need in the preamble
output: rticles::rjournal_article
---

## Introduction

Introductory section which may include references in parentheses
[@R], or cite a reference such as @R in the text.

## Section title in sentence case

This section may contain a figure such as Figure \ref{fig:Rlogo}.

```{r, Rlogo, echo=FALSE, fig.cap='The logo of R.', out.width='2in', fig.align='center', fig.pos='htbp'}
knitr::include_graphics('palmerpenguins-logo.png')
```

## Another section

There will likely be several sections, perhaps including code snippets, such as:

```{r}
x <- 1:10
x
```

## Summary

This file is only a basic article template. For full details of _The R Journal_ style and information on how to prepare your article for submission, see the [Instructions for Authors](https://journal.r-project.org/share/author-guide.pdf).

\bibliography{RJreferences}

### 2020-07-20 penguins.tex
add R Journal rticle template (#10)

% !TeX root = RJwrapper.tex
\title{Capitalized Title Here}
\author{by Allison Horst, Alison Hill, Kristen Gorman}

\maketitle

\abstract{%
An abstract of less than 150 words.
}

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

Introductory section which may include references in parentheses
\citep{R}, or cite a reference such as \citet{R} in the text.

\hypertarget{section-title-in-sentence-case}{%
\subsection{Section title in sentence
case}\label{section-title-in-sentence-case}}

This section may contain a figure such as Figure \ref{fig:Rlogo}.

\begin{Schunk}
\begin{figure}[htbp]

{\centering \includegraphics[width=2in]{palmerpenguins-logo}

}

\caption[The logo of R]{The logo of R.}\label{fig:Rlogo}
\end{figure}
\end{Schunk}

\hypertarget{another-section}{%
\subsection{Another section}\label{another-section}}

There will likely be several sections, perhaps including code snippets,
such as:

\begin{Schunk}
\begin{Sinput}
x <- 1:10
x
\end{Sinput}
\begin{Soutput}
#>  [1]  1  2  3  4  5  6  7  8  9 10
\end{Soutput}
\end{Schunk}

\hypertarget{summary}{%
\subsection{Summary}\label{summary}}

This file is only a basic article template. For full details of
\emph{The R Journal} style and information on how to prepare your
article for submission, see the
\href{https://journal.r-project.org/share/author-guide.pdf}{Instructions
for Authors}.

\bibliography{RJreferences}

\address{%
Allison Horst\\
Bren School of Environmental Science and Management, University of
California, Santa Barbara\\
line 1\\ line 2\\
}
\href{mailto:ahorst@ucsb.edu}{\nolinkurl{ahorst@ucsb.edu}}

\address{%
Alison Hill\\
RStudio, PBC\\
line 1\\ line 2\\
}
\href{mailto:alison@rstudio.com}{\nolinkurl{alison@rstudio.com}}

\address{%
Kristen Gorman\\
University of Alaska Fairbanks College of Fisheries and Ocean Sciences\\
line 1\\ line 2\\
}
\href{mailto:kbgorman@alaska.edu}{\nolinkurl{kbgorman@alaska.edu}}

### 2020-08-03 R/manuscript_code.Rmd
clean up repo, start populating .Rmd & bib

--- 2020-07-12 manuscript/manuscript_text.Rmd
@@
 
-# Authors (placeholder order TBD)
-
-Allison Horst^1^, Alison Hill^2^, Kristen Gorman^3^
-
-^1^ Bren School of Environmental Science and Management, University of California, Santa Barbara
-
-^2^ RStudio
-
-^3^ College of Fisheries and Ocean Sciences, University of Alaska Fairbanks
-
 # MANUSCRIPT CODE
-
-### See shared googledoc for MS text
 
@@
 
-```{r, echo = FALSE}
+```{r, iris-table, echo = FALSE}
 set.seed(1900)
 sample_n(iris, 6) %>% 
-  kable() %>% 
-  kable_styling(full_width = FALSE)
+  kable(caption = "Six randomly selected rows from the iris dataset (of 150 total rows). Values for petal and sepal dimensions are in centimeters.")
 ```
 
-```{r, echo = FALSE}
+```{r, penguins-table}
 set.seed(1900)
 sample_n(penguins, 6) %>% 
-  kable() %>% 
-  kable_styling(full_width = FALSE)
+  kable(caption = "Six randomly selected rows from the penguins dataset (of 344 total rows).
+Units of the four numeric penguin measurements are included as suffixes in variable names, with \"_mm\" indicating millimeters, and \"_g\" indicating grams.")
 ```
@@
 penguin_counts <- penguins %>% 
+  mutate(species = as.character(species)) %>% 
+  mutate(species = case_when(
+    species == "Adelie" ~ "Adélie", 
+    TRUE ~ species
+  )) %>% 
   mutate(sex = str_to_title(sex)) %>% 
@@
    tab_footnote(
-    footnote = "Note: Here, penguin sample sizes are split by species and sex, and could be further divided by island. NA column for penguins indicate observations where sex was not recorded.",
+     footnote = "Note: Here, penguin sample sizes are split by species and sex, and could be further divided by island and year.
+NA column for penguins indicate observations where sex was not recorded.",
      location = cells_column_labels(
@@
    )
-
 ```

### 2020-08-03 penguins.Rmd
clean up repo, start populating .Rmd & bib

@@
 ---
-title: Capitalized Title Here
+title: Palmer Archipelago Penguins Data in the palmerpenguins R Package - An Alternative to Anderson's *Irises*
 author:
-  - name: Allison Horst
-    affiliation: Bren School of Environmental Science and Management, University of California, Santa Barbara
+  - name: Allison M. Horst
+    affiliation: Bren School of Environmental Science and Management
     address:
-    - line 1
-    - line 2
+    - University of California, Santa Barbara
+    - Santa Barbara, CA 93106-5131
     email:  ahorst@ucsb.edu
-  - name: Alison Hill
+  - name: Alison P. Hill
     affiliation: RStudio, PBC
     address:
-    - line 1
-    - line 2
+    - 250 Northern Ave
+    - Boston, MA 02210
     email:  alison@rstudio.com
-  - name: Kristen Gorman
+  - name: Kristen B. Gorman
     affiliation: University of Alaska Fairbanks College of Fisheries and Ocean Sciences
     address:
-    - line 1
-    - line 2
+    - 2150 Koyukuk Drive
+    - 245 O’Neill Building
+    - Fairbanks, AK 99775-7220
     email:  kbgorman@alaska.edu
 abstract: >
-  An abstract of less than 150 words.
+
+ The famous *Iris* dataset, which exists in the base R **datasets** package as **iris**, is ubiquitous in statistics and data science education.
+While the **iris** measurements were originally collected by botanist Edgar Anderson, their use in Ronald A.
+Fisher's eugenics research has created momentum to address the dataset's racist use and/or replace **iris** with alternate data in teaching materials.
+Penguin morphological measurements for three species of *Pygoscelis* penguins that breed on islands throughout the Palmer Archipelago, Antarctica, provide an approachable, charismatic and near drop-in replacement for the **iris** data.
+Here, we introduce the **palmerpenguins** R package and describe the included **penguins** data.
+We directly compare the two datasets for selected analyses to demonstrate that R users, in particular teachers and learners currently using the **iris** data, can switch to the Palmer Archipelago penguins for many use cases including data wrangling and visualization, correlation, regression, hypothesis testing, multivariate analysis (e.g.
+PCA), cluster analysis and classification (e.g.
+by k-means).
+
 preamble: |
@@
 
+```{r setup, include=FALSE}
+library(tidyverse)
+library(here)
+library(palmerpenguins)
+library(kableExtra)
+library(patchwork)
+library(gt)
+library(GGally)
+library(recipes)
+library(shadowtext)
+library(broom)
+knitr::opts_chunk$set(echo = TRUE)
+
+theme_set(theme_minimal())
+```
+
 ## Introduction
 
-Introductory section which may include references in parentheses
-[@R], or cite a reference such as @R in the text.
+<!---Introductory section which may include references in parentheses
+[@R], or cite a reference such as @R in the text.--->
 
-## Section title in sentence case
+In 1935, American botanist Edgar Anderson measured petal and sepal structural dimensions (length and width) for 50 flowers from three Iris species: *Iris setosa*, *Iris versicolor*, and *Iris virginica* [@anderson_irises_1935].
+The manageable but non-trivial size (5 variables and 150 total observations) and characteristics of the data make it amenable for introducing a wide range of statistical methods including data wrangling, visualization, regression, multivariate exploration, and machine learning.
+Anderson's *Iris* dataset is built into a number of software packages including the auto-installed datasets package in R [@r_core_team_r_2019], Python's scikit-learn machine learning library [@pedregosa_scikit-learn_2011], and the SAS Sashelp library (SAS Institute, Cary NC), which have facilitated widespread use of the data.
+As a result, eighty-five years after the data were initially published, Anderson’s *Iris* measurements are ubiquitous in statistics and data science course materials and tutorials.
+**Sounds great.
+Why consider replacing iris?**
 
-This section may contain a figure such as Figure \ref{fig:Rlogo}.
+In 1936, one year after appearing in the Bulletin of the American Iris Society, Anderson’s *Iris* measurements were published by eugenicist and statistician Ronald A.
+Fisher in Annals of Eugenics [@fisher_use_1936].
+Regardless of Anderson's initial motivation, the data today remain inextricably linked to Fisher’s eugenics research and are even commonly, if unfairly, referred to as "Fisher's iris data.” For example, an August 2020 search for “Iris flower data set” in Wikipedia returns an article beginning with “The iris flower data set or Fisher’s iris data…”, then later in the article: “It is sometimes called Anderson's Iris data set” [@wikipedia_iris_2020].
+The *Iris* dataset is similarly credited to Fisher in statistical computing literature [@trendafilov_simple_2009, @wang_matlab_2015, @woods_how_2015, @chen_unsupervised_2018].
+
+There is recent momentum to address Fisher’s racist legacy in statistics.
+In June 2020, the Committee of Presidents of Statistical Societies’ R.A.
+Fisher Award and Lectureship was replaced by the Distinguished Achievement Award and Lectureship, to “advance a more just, equitable, diverse, and inclusive statistical community” [@noauthor_institute_2020].
+Cambridge University’s Gonville and Caius College recently announced plans to remove a stained-glass window celebrating Fisher from a campus dining hall [@noauthor_sir_2020].
+In the R community, there are calls to (1) address the *Iris* dataset's use in Fisher’s eugenics work, and/or (2) consider using a different dataset [@poisot_timothee_its_2020, @garrick_aden-buie_lets_2020].
+This paper describes (2): a suitable alternative dataset for educators wanting to replace the *Iris* data in their teaching materials.
+
+Considering *Iris* data usage in data science and statistics materials, we established the following criteria for a suitable replacement dataset:
+
+- Available by appropriate license
+- Features intuitive subjects and variables understandable to learners across fields
+- Manageable (but not trivial) in size
+- Minimal data cleaning and pre-processing required for most analyses
+- Real-world (not manufactured) data
+- Similar opportunities for teaching and learning R, data science, and statistical skills
+
+Here, we describe an alternative to Anderson's *Iris* data that largely satisfies these criteria: an approachable and charismatic dataset containing real-world morphological data for three *Pygoscelis* penguin species that breed throughout the Western Antarctic Peninsula region, made available through the Long-Term Ecological Research Network (US LTER).
+By comparing data structure, size, and a range of analyses side-by-side for the two datasets, we demonstrate that the Palmer Archipelago penguin measurements can replace Anderson's *Iris* data for many use cases in statistics and data science education.
+
+## Data source
+
+Body size measurements, clutch (i.e., egg laying) observations (e.g., date of first egg laid, and clutch completion), and carbon (^13^C/^12^C, $\delta$^13^C) and nitrogen (^15^N/^14^N, $\delta$^15^N) stable isotope values of red blood cells for male and female adult Adélie (*P.
+adeliae*), chinstrap (*P.
+antarcticus*), and gentoo (*P.
+papua*) penguins on three islands (Biscoe, Dream and Torgersen) in the Palmer Archipelago were collected from 2007 - 2009 by Dr.
+Kristen Gorman in collaboration with the [Palmer Station LTER](https://pal.lternet.edu/), part of the [US LTER Network](https://lternet.edu/).
+For complete data collection methods and published analyses see Gorman et al.
+(2014).
+
+The data in the **palmerpenguins** R package are available for use by CC0 license ("No Rights Reserved") in accordance with the [Palmer Station LTER Data Policy](https://pal.lternet.edu/data/policies) and the [LTER Data Access Policy](https://lternet.edu/data-access-policy/), and were imported from the [Environmental Data Initiative (EDI) Data Portal](https://environmentaldatainitiative.org/) at the links below:
+
+- Adélie penguin data (LTER and Gorman 2020a): [KNB-LTER Data Package 219.5](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.219.5)
+- Chinstrap penguin data (LTER and Gorman 2020b): [KNB-LTER Data Package 221.5](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.221.6)
+- Gentoo penguin data (LTER and Gorman 2020c): [KNB-LTER Data Package 220.5](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.220.5)
+
+## R package: **palmerpenguins**
+
+R users can install the **palmerpenguins** package from CRAN by:
+
+      `install.packages("palmerpenguins")`
+
+Alternatively, the development version of the package can be installed from GitHub:
+
+      `remotes::install_github("allisonhorst/palmerpenguins")`.
+
+Information, examples, and links to community-contributed materials are available on the **palmerpenguins** package website: https://allisonhorst.github.io/palmerpenguins/.
+
+The **palmerpenguins** R package contains two data objects: **penguins_raw** and **penguins**.
+The **penguins_raw** data consists of all raw data for 17 variables (Appendix Table 1), recorded completely or in part for 344 individual penguins, accessed directly from EDI.
+As a direct alternative to Anderson's *Iris* data we recommend using the curated data in **penguins**, which is a subset of **penguins_raw** retaining all 344 observations, minimally updated (Appendix B) and reduced to the following eight variables:
+
+- *species:* a factor denoting the penguin species (Adélie, chinstrap, or gentoo)
+- *island:* a factor denoting the island (in Palmer Archipelago, Antarctica) where observed (Biscoe, Dream or Torgersen)
+- *bill_length_mm:* a number denoting length of the dorsal ridge of a penguin bill (millimeters)
+- *bill_depth_mm:* a number denoting the depth of a penguin bill (millimeters)
+- *flipper_length_mm:* an integer denoting the length of a penguin flipper (millimeters)
+- *body_mass_g:* an integer denoting the weight of a penguin’s body (grams)
+- *sex:* a factor denoting the sex of a penguin sex (male, female) based on molecular data
+- *year:* an integer denoting the year of study (2007, 2008 or 2009)
+
+The same data exist as comma-separated value (CSV) files in the package ("penguins_raw.csv" and "penguins.csv"), and can be read in using the built-in `path_to_file()` function in **palmerpenguins**.
+For example,
+
+```
+    library(tidyverse)
+    library(palmerpenguins)
+    df <- read_csv(path_to_file(“penguins.csv”))
+```
+
+will read in "penguins.csv" as if from an external file, thus automatically parsing *species*, *island* and *sex* variables as characters.
+This option allows users opportunities to practice or demonstrate reading in data from a CSV, then updating variable class (e.g.
+characters to factors).
+
+## Other data access options
+
+Python: Python users can access the penguins data in the **seaborn** data visualization library (Waskom et al. 2017). Example code to load the data in Python:
+
+```
+    import seaborn as sns
+    df = sns.load_dataset('penguins') 
+```
+
+Julia: Julia users can access the penguins data in the **PalmerPenguins.jl** package. Example code to import the penguins data through **PalmerPenguins.jl**:
+
+```
+    julia> using DataFrames
+    julia> df = DataFrame(table)
+```
+
+## Software and code
+
+All analyses were performed in the R language environment using version 3.6.2 (R Core Team 2019).
+Complete code for this paper is shared in the Supplemental Material.
+We acknowledge the following R packages used in analyses, with gratitude to developers and contributors:
+
+- **tidyverse** (Wickham et al. 2019), for data import and cleaning
+- **ggplot2** (Wickham 2016:2), for data visualizations
+- **here** (Müller 2017), for file path control
+- **kableExtra** (Zhu 2019) for finalized tables
+- **gt** (Iannone, Cheng, and Schloerke 2020) for finalized tables
+- **GGally** (Schloerke et al. 2020), for pairs plots
+- **patchwork** (Pedersen 2019), for compound figures
+- **shadowtext** (Yu 2019), to add a background color to text labels
+- **recipes** (Kuhn and Wickham 2020), for data pre-processing
+- **base** and **stats** (R Core Team 2019) for various analyses throughout
+- **pkgdown** (ref) was used to build the package website
+
+## Selected comparisons between **iris** and **penguins**
+
+The **penguins** data in **palmerpenguins** is generally useful and approachable for data science and statistics education, and is uniquely well-suited to replace the **iris** dataset.
+Comparisons presented are selected examples for common **iris** uses, and are not exhaustive.
+
+### Data structure and sample size
+
+Both **iris** and **penguins** are in tidy format [ref] with each column denoting a single variable and each row containing measurements for a single *Iris* flower or penguin.
+The two datasets are comparable in size: dimensions (columns × rows) are 5 × 150 and 8 × 344 for **iris** and **penguins**, respectively, and sample sizes within species are similar (Table 1).
+Notably, sample sizes differ for the three penguin species, while sample sizes in **iris** are equal (n = 50 for each Iris species).
+Multiple factor variables in **penguins** (*species*, *island* and *sex*) along with *year* create additional opportunities for grouping, compared to the single factor (*species*) in **iris**.
+
+Unlike **iris**, which contains only complete cases, the **penguins** dataset contains a small number of missing values (n~missing~ = 19, out of 2,752 total values).
+Missing values and unequal sample sizes are common in real-world data, and we believe add learning value to **penguins**.
+
+### Continuous quantitative variables: overview
+
+Distributions, relationships between variables and clustering can be visually explored between species for the four structural size measurements in **iris** (sepal width and length, petal width and length) and **penguins** (flipper length, body mass, bill length and depth).
+It is worth noting that the correlation coefficients presented are Pearson’s *r*, developed by and named for another famous eugenicist and statistician, Karl Pearson [ref].
+
+### Linear correlation, regression, and Simpson’s Paradox
+
+There exist numerous opportunities to explore linear relationships and correlations in penguins and iris both within and across species (Figures 1 & 2).
+Here, we highlight one comparison that is uniquely similar: petal dimensions (petal length versus petal width) from iris, and penguin size (flipper length versus body mass) in penguins (Figure 3).
+The overall trend across all three species is approximately linear for both iris and penguins (Figure 3A,B).
+Teachers may encourage students to explore how simple linear regression results and predictions differ when the species variable is omitted (Figure 3A,B), compared to multiple linear regression with species included (Figure 3C,D).
+Notably, distinctions between species are clearer for iris petals - particularly, the much smaller petals for Iris setosa - compared to penguins, in which Adélie and chinstrap penguins are largely overlapping in body size (body mass and flipper length), and are both generally smaller than gentoos.
+
+### Penguins bonus: Simpson’s Paradox
+
+Simpson’s Paradox is a data phenomenon in which a trend observed between variables is reversed when data are pooled, omitting a meaningful variable.
+While often taught and discussed in statistics courses, finding a real-world and approachable example of Simpson’s paradox can be a challenge.
+Here, we show one (of several possible; see Figure XX) Simpson’s Paradox examples in **penguins**: exploring bill dimensions with and without *species* included (Figure XX).
+When penguin *species* is omitted (Figure XX), the bill length and depth appear negatively correlated overall.
+The trend is reversed when *species* is included, revealing a positive correlation between bill length and bill depth (Figure XX)
+
+### Principal component analysis
+
+Principal component analysis (PCA) is a dimensional reduction method commonly used to explore patterns in multivariate data.
+PCA tutorials frequently employ the **iris** dataset, which is useful for teaching the method due to multivariate normality, and clear, approachable outcomes for variable loadings and clustering.
+
+A comparison of PCA with the four variables of structural size measurements in **penguins** and **iris** (both normalized prior to PCA) reveals highly similar results (Figure XX).
+For both datasets, one species is distinct (gentoo penguins, and setosa irises) while the other two species (chinstrap/Adélie and versicolor/virginica) appear somewhat overlapping in the first two principal components (Figure XX).
+Variance explained by each principal component (PC) is similar, particularly for PC1 and PC2: for **penguins**, 88.15% of total variance is captured by the first two PCs, compared to 95.81% for **iris**, with a similarly large percentage of variance captured by PC1 and PC2 in each (Figure XX).
+
+### K-means clustering
+
+Unsupervised clustering by k-means is a common and popular entryway to machine learning and classification, and frequently employs the **iris** data for introductory examples.
+The **penguins** data provides similar opportunities for introducing k-means clustering.
+For simplicity, we compare k-means clustering using only two variables for each dataset: for **iris**, petal width and petal length, and for **penguins**, bill length and bill depth.
+All variables are scaled prior to k-means.
+Three clusters (*k* = 3) are specified for each since there are three species of both *Iris* (*setosa*, *versicolor*, and *virginica*) and penguins (Adélie, chinstrap and gentoo).
+
+K-means clustering with penguin bill dimensions and iris petal dimensions yields largely distinct clusters each dominated by one species (Figure XX, Table XX).
+For iris petal dimensions, k-means yields a perfectly separated cluster (Cluster 1) containing all 50 setosa iris observations and zero misclassified virginica or versicolor irises (Table 4).
+While clustering is not perfectly distinct for any penguin species, each species is largely contained within a single cluster, with little overlap from the other two species.
+For example, considering Adélie penguins (orange observations in Figure 6A): 147 (out of 151) Adélie penguins are assigned to Cluster 1, zero are assigned to Cluster 2, and 4 are assigned to the chinstrap-dominated Cluster 3 (Table 4).
+Only 5 (of 68) chinstrap penguins and 1 (of 123) gentoo penguins are assigned to the Adélie-dominated Cluster 1 (Table 4).
+
+## Other alternatives to iris
+
+## Discussion
+
+Here, we have shown that the **penguins** data, available in the **palmerpenguins** R package, can replace **iris** for a number of common use cases in data science and statistics education including exploratory data visualization, linear correlation and regression, PCA and clustering by k-means.
+In addition, teaching and learning opportunities in **penguins** are increased due to a greater number of variables, missing values, unequal sample sizes, and Simpson's paradox examples.
+
+## Conclusion
+
+## Appendix
+
+### Appendix A
+### Appendix B
+
+Data in the **penguins** object have been minimally updated from **penguins_raw** as follows:
+
+- All variable names are converted to lower snake case
+- Entries in *species* are truncated to only include the common name (e.g. "gentoo", instead of "gentoo penguin (Pygoscelis papua)")
+- Recorded sex for penguin N36A1, originally recorded as ".", is updated to `NA`
+- *culmen_length_mm* and *culmen_depth_mm* variable names are updated to *bill_length_mm* and *bill_depth_mm*, respectively
+- Class for categorical variables (*species*, *island*, *sex*) is updated to factor
+- Variable *year* was pulled from clutch observations
+
+<!---This section may contain a figure such as Figure \ref{fig:Rlogo}.
 
@@
 ``` 
+--->
 
+<!---
 ## Another section
@@
 ```
+--->
 
+<!----
 ## Summary
@@
 
+--->
+
 \bibliography{RJreferences}

### 2020-08-03 penguins.tex
clean up repo, start populating .Rmd & bib

@@
 % !TeX root = RJwrapper.tex
-\title{Capitalized Title Here}
-\author{by Allison Horst, Alison Hill, Kristen Gorman}
+\title{Palmer Archipelago Penguins Data in the palmerpenguins R Package - An
+Alternative to Anderson's \emph{Irises}}
+\author{by Allison M. Horst, Alison P. Hill, Kristen B. Gorman}
 
@@
 \abstract{%
-An abstract of less than 150 words.
+The famous \emph{Iris} dataset, which exists in the base R
+\textbf{datasets} package as \textbf{iris}, is ubiquitous in statistics
+and data science education. While the \textbf{iris} measurements were
+originally collected by botanist Edgar Anderson, their use in Ronald A.
+Fisher's eugenics research has created momentum to address the dataset's
+racist use and/or replace \textbf{iris} with alternate data in teaching
+materials. Penguin morphological measurements for three species of
+\emph{Pygoscelis} penguins that breed on islands throughout the Palmer
+Archipelago, Antarctica, provide an approachable, charismatic and near
+drop-in replacement for the \textbf{iris} data. Here, we introduce the
+\textbf{palmerpenguins} R package and describe the included
+\textbf{penguins} data. We directly compare the two datasets for
+selected analyses to demonstrate that R users, in particular teachers
+and learners currently using the \textbf{iris} data, can switch to the
+Palmer Archipelago penguins for many use cases including data wrangling
+and visualization, correlation, regression, hypothesis testing,
+multivariate analysis (e.g.~PCA), cluster analysis and classification
+(e.g.~by k-means).
 }
@@
 
-Introductory section which may include references in parentheses
-\citep{R}, or cite a reference such as \citet{R} in the text.
+In 1935, American botanist Edgar Anderson measured petal and sepal
+structural dimensions (length and width) for 50 flowers from three Iris
+species: \emph{Iris setosa}, \emph{Iris versicolor}, and \emph{Iris
+virginica} \citep{anderson_irises_1935}. The manageable but non-trivial
+size (5 variables and 150 total observations) and characteristics of the
+data make it amenable for introducing a wide range of statistical
+methods including data wrangling, visualization, regression,
+multivariate exploration, and machine learning. Anderson's \emph{Iris}
+dataset is built into a number of software packages including the
+auto-installed datasets package in R \citep{r_core_team_r_2019},
+Python's scikit-learn machine learning library
+\citep{pedregosa_scikit-learn_2011}, and the SAS Sashelp library (SAS
+Institute, Cary NC), which have facilitated widespread use of the data.
+As a result, eighty-five years after the data were initially published,
+Anderson's \emph{Iris} measurements are ubiquitous in statistics and
+data science course materials and tutorials. \textbf{Sounds great. Why
+consider replacing iris?}
 
-\hypertarget{section-title-in-sentence-case}{%
-\subsection{Section title in sentence
-case}\label{section-title-in-sentence-case}}
+In 1936, one year after appearing in the Bulletin of the American Iris
+Society, Anderson's \emph{Iris} measurements were published by
+eugenicist and statistician Ronald A. Fisher in Annals of Eugenics
+\citep{fisher_use_1936}. Regardless of Anderson's initial motivation,
+the data today remain inextricably linked to Fisher's eugenics research
+and are even commonly, if unfairly, referred to as ``Fisher's iris
+data.'' For example, an August 2020 search for ``Iris flower data set''
+in Wikipedia returns an article beginning with ``The iris flower data
+set or Fisher's iris data\ldots{}'', then later in the article: ``It is
+sometimes called Anderson's Iris data set'' \citep{wikipedia_iris_2020}.
+The \emph{Iris} dataset is similarly credited to Fisher in statistical
+computing literature \citep[\citet{wang_matlab_2015},
+\citet{woods_how_2015},
+\citet{chen_unsupervised_2018}]{trendafilov_simple_2009}.
 
-This section may contain a figure such as Figure \ref{fig:Rlogo}.
+There is recent momentum to address Fisher's racist legacy in
+statistics. In June 2020, the Committee of Presidents of Statistical
+Societies' R.A. Fisher Award and Lectureship was replaced by the
+Distinguished Achievement Award and Lectureship, to ``advance a more
+just, equitable, diverse, and inclusive statistical community''
+\citep{noauthor_institute_2020}. Cambridge University's Gonville and
+Caius College recently announced plans to remove a stained-glass window
+celebrating Fisher from a campus dining hall \citep{noauthor_sir_2020}.
+In the R community, there are calls to (1) address the \emph{Iris}
+dataset's use in Fisher's eugenics work, and/or (2) consider using a
+different dataset
+\citep[\citet{garrick_aden-buie_lets_2020}]{poisot_timothee_its_2020}.
+This paper describes (2): a suitable alternative dataset for educators
+wanting to replace the \emph{Iris} data in their teaching materials.
 
-\begin{Schunk}
-\begin{figure}[htbp]
+Considering \emph{Iris} data usage in data science and statistics
+materials, we established the following criteria for a suitable
+replacement dataset:
 
-{\centering \includegraphics[width=2in]{palmerpenguins-logo}
+\begin{itemize}
+\tightlist
+\item
+  Available by appropriate license
+\item
+  Features intuitive subjects and variables understandable to learners
+  across fields
+\item
+  Manageable (but not trivial) in size
+\item
+  Minimal data cleaning and pre-processing required for most analyses
+\item
+  Real-world (not manufactured) data
+\item
+  Similar opportunities for teaching and learning R, data science, and
+  statistical skills
+\end{itemize}
 
-}
+Here, we describe an alternative to Anderson's \emph{Iris} data that
+largely satisfies these criteria: an approachable and charismatic
+dataset containing real-world morphological data for three
+\emph{Pygoscelis} penguin species that breed throughout the Western
+Antarctic Peninsula region, made available through the Long-Term
+Ecological Research Network (US LTER). By comparing data structure,
+size, and a range of analyses side-by-side for the two datasets, we
+demonstrate that the Palmer Archipelago penguin measurements can replace
+Anderson's \emph{Iris} data for many use cases in statistics and data
+science education.
 
-\caption[The logo of R]{The logo of R.}\label{fig:Rlogo}
-\end{figure}
-\end{Schunk}
+\hypertarget{data-source}{%
+\subsection{Data source}\label{data-source}}
 
-\hypertarget{another-section}{%
-\subsection{Another section}\label{another-section}}
+Body size measurements, clutch (i.e., egg laying) observations (e.g.,
+date of first egg laid, and clutch completion), and carbon
+(\textsuperscript{13}C/\textsuperscript{12}C,
+\(\delta\)\textsuperscript{13}C) and nitrogen
+(\textsuperscript{15}N/\textsuperscript{14}N,
+\(\delta\)\textsuperscript{15}N) stable isotope values of red blood
+cells for male and female adult Adélie (\emph{P. adeliae}), chinstrap
+(\emph{P. antarcticus}), and gentoo (\emph{P. papua}) penguins on three
+islands (Biscoe, Dream and Torgersen) in the Palmer Archipelago were
+collected from 2007 - 2009 by Dr.~Kristen Gorman in collaboration with
+the \href{https://pal.lternet.edu/}{Palmer Station LTER}, part of the
+\href{https://lternet.edu/}{US LTER Network}. For complete data
+collection methods and published analyses see Gorman et al.~(2014).
 
-There will likely be several sections, perhaps including code snippets,
-such as:
+The data in the \textbf{palmerpenguins} R package are available for use
+by CC0 license (``No Rights Reserved'') in accordance with the
+\href{https://pal.lternet.edu/data/policies}{Palmer Station LTER Data
+Policy} and the \href{https://lternet.edu/data-access-policy/}{LTER Data
+Access Policy}, and were imported from the
+\href{https://environmentaldatainitiative.org/}{Environmental Data
+Initiative (EDI) Data Portal} at the links below:
 
-\begin{Schunk}
-\begin{Sinput}
-x <- 1:10
-x
-\end{Sinput}
-\begin{Soutput}
-#>  [1]  1  2  3  4  5  6  7  8  9 10
-\end{Soutput}
-\end{Schunk}
+\begin{itemize}
+\tightlist
+\item
+  Adélie penguin data (LTER and Gorman 2020a):
+  \href{https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.219.5}{KNB-LTER
+  Data Package 219.5}
+\item
+  Chinstrap penguin data (LTER and Gorman 2020b):
+  \href{https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.221.6}{KNB-LTER
+  Data Package 221.5}
+\item
+  Gentoo penguin data (LTER and Gorman 2020c):
+  \href{https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.220.5}{KNB-LTER
+  Data Package 220.5}
+\end{itemize}
 
-\hypertarget{summary}{%
-\subsection{Summary}\label{summary}}
+\hypertarget{r-package-palmerpenguins}{%
+\subsection{\texorpdfstring{R package:
+\textbf{palmerpenguins}}{R package: palmerpenguins}}\label{r-package-palmerpenguins}}
 
-This file is only a basic article template. For full details of
-\emph{The R Journal} style and information on how to prepare your
-article for submission, see the
-\href{https://journal.r-project.org/share/author-guide.pdf}{Instructions
-for Authors}.
+R users can install the \textbf{palmerpenguins} package from CRAN by:
+
+\begin{verbatim}
+  `install.packages("palmerpenguins")`
+\end{verbatim}
+
+Alternatively, the development version of the package can be installed
+from GitHub:
+
+\begin{verbatim}
+  `remotes::install_github("allisonhorst/palmerpenguins")`. 
+\end{verbatim}
+
+Information, examples, and links to community-contributed materials are
+available on the \textbf{palmerpenguins} package website:
+\url{https://allisonhorst.github.io/palmerpenguins/}.
+
+The \textbf{palmerpenguins} R package contains two data objects:
+\textbf{penguins\_raw} and \textbf{penguins}. The \textbf{penguins\_raw}
+data consists of all raw data for 17 variables (Appendix Table 1),
+recorded completely or in part for 344 individual penguins, accessed
+directly from EDI. As a direct alternative to Anderson's \emph{Iris}
+data we recommend using the curated data in \textbf{penguins}, which is
+a subset of \textbf{penguins\_raw} retaining all 344 observations,
+minimally updated (Appendix B) and reduced to the following eight
+variables:
+
+\begin{itemize}
+\tightlist
+\item
+  \emph{species:} a factor denoting the penguin species (Adélie,
+  chinstrap, or gentoo)
+\item
+  \emph{island:} a factor denoting the island (in Palmer Archipelago,
+  Antarctica) where observed (Biscoe, Dream or Torgersen)
+\item
+  \emph{bill\_length\_mm:} a number denoting length of the dorsal ridge
+  of a penguin bill (millimeters)
+\item
+  \emph{bill\_depth\_mm:} a number denoting the depth of a penguin bill
+  (millimeters)
+\item
+  \emph{flipper\_length\_mm:} an integer denoting the length of a
+  penguin flipper (millimeters)
+\item
+  \emph{body\_mass\_g:} an integer denoting the weight of a penguin's
+  body (grams)
+\item
+  \emph{sex:} a factor denoting the sex of a penguin sex (male, female)
+  based on molecular data
+\item
+  \emph{year:} an integer denoting the year of study (2007, 2008 or
+  2009)
+\end{itemize}
+
+The same data exist as comma-separated value (CSV) files in the package
+(``penguins\_raw.csv'' and ``penguins.csv''), and can be read in using
+the built-in \texttt{path\_to\_file()} function in
+\textbf{palmerpenguins}. For example,
+
+\begin{verbatim}
+    library(tidyverse)
+    library(palmerpenguins)
+    df <- read_csv(path_to_file(“penguins.csv”))
+\end{verbatim}
+
+will read in ``penguins.csv'' as if from an external file, thus
+automatically parsing \emph{species}, \emph{island} and \emph{sex}
+variables as characters. This option allows users opportunities to
+practice or demonstrate reading in data from a CSV, then updating
+variable class (e.g.~characters to factors).
+
+\hypertarget{other-data-access-options}{%
+\subsection{Other data access options}\label{other-data-access-options}}
+
+Python: Python users can access the penguins data in the
+\textbf{seaborn} data visualization library (Waskom et al.~2017).
+Example code to load the data in Python:
+
+\begin{verbatim}
+    import seaborn as sns
+    df = sns.load_dataset('penguins') 
+\end{verbatim}
+
+Julia: Julia users can access the penguins data in the
+\textbf{PalmerPenguins.jl} package. Example code to import the penguins
+data through \textbf{PalmerPenguins.jl}:
+
+\begin{verbatim}
+    julia> using DataFrames
+    julia> df = DataFrame(table)
+\end{verbatim}
+
+\hypertarget{software-and-code}{%
+\subsection{Software and code}\label{software-and-code}}
+
+All analyses were performed in the R language environment using version
+3.6.2 (R Core Team 2019). Complete code for this paper is shared in the
+Supplemental Material. We acknowledge the following R packages used in
+analyses, with gratitude to developers and contributors:
+
+\begin{itemize}
+\tightlist
+\item
+  \textbf{tidyverse} (Wickham et al.~2019), for data import and cleaning
+\item
+  \textbf{ggplot2} (Wickham 2016:2), for data visualizations
+\item
+  \textbf{here} (Müller 2017), for file path control
+\item
+  \textbf{kableExtra} (Zhu 2019) for finalized tables
+\item
+  \textbf{gt} (Iannone, Cheng, and Schloerke 2020) for finalized tables
+\item
+  \textbf{GGally} (Schloerke et al.~2020), for pairs plots
+\item
+  \textbf{patchwork} (Pedersen 2019), for compound figures
+\item
+  \textbf{shadowtext} (Yu 2019), to add a background color to text
+  labels
+\item
+  \textbf{recipes} (Kuhn and Wickham 2020), for data pre-processing
+\item
+  \textbf{base} and \textbf{stats} (R Core Team 2019) for various
+  analyses throughout
+\item
+  \textbf{pkgdown} (ref) was used to build the package website
+\end{itemize}
+
+\hypertarget{selected-comparisons-between-iris-and-penguins}{%
+\subsection{\texorpdfstring{Selected comparisons between \textbf{iris}
+and
+\textbf{penguins}}{Selected comparisons between iris and penguins}}\label{selected-comparisons-between-iris-and-penguins}}
+
+The \textbf{penguins} data in \textbf{palmerpenguins} is generally
+useful and approachable for data science and statistics education, and
+is uniquely well-suited to replace the \textbf{iris} dataset.
+Comparisons presented are selected examples for common \textbf{iris}
+uses, and are not exhaustive.
+
+\hypertarget{data-structure-and-sample-size}{%
+\subsubsection{Data structure and sample
+size}\label{data-structure-and-sample-size}}
+
+Both \textbf{iris} and \textbf{penguins} are in tidy format {[}ref{]}
+with each column denoting a single variable and each row containing
+measurements for a single \emph{Iris} flower or penguin. The two
+datasets are comparable in size: dimensions (columns × rows) are 5 × 150
+and 8 × 344 for \textbf{iris} and \textbf{penguins}, respectively, and
+sample sizes within species are similar (Table 1). Notably, sample sizes
+differ for the three penguin species, while sample sizes in
+\textbf{iris} are equal (n = 50 for each Iris species). Multiple factor
+variables in \textbf{penguins} (\emph{species}, \emph{island} and
+\emph{sex}) along with \emph{year} create additional opportunities for
+grouping, compared to the single factor (\emph{species}) in
+\textbf{iris}.
+
+Unlike \textbf{iris}, which contains only complete cases, the
+\textbf{penguins} dataset contains a small number of missing values
+(n\textsubscript{missing} = 19, out of 2,752 total values). Missing
+values and unequal sample sizes are common in real-world data, and we
+believe add learning value to \textbf{penguins}.
+
+\hypertarget{continuous-quantitative-variables-overview}{%
+\subsubsection{Continuous quantitative variables:
+overview}\label{continuous-quantitative-variables-overview}}
+
+Distributions, relationships between variables and clustering can be
+visually explored between species for the four structural size
+measurements in \textbf{iris} (sepal width and length, petal width and
+length) and \textbf{penguins} (flipper length, body mass, bill length
+and depth). It is worth noting that the correlation coefficients
+presented are Pearson's \emph{r}, developed by and named for another
+famous eugenicist and statistician, Karl Pearson {[}ref{]}.
+
+\hypertarget{linear-correlation-regression-and-simpsons-paradox}{%
+\subsubsection{Linear correlation, regression, and Simpson's
+Paradox}\label{linear-correlation-regression-and-simpsons-paradox}}
+
+There exist numerous opportunities to explore linear relationships and
+correlations in penguins and iris both within and across species
+(Figures 1 \& 2). Here, we highlight one comparison that is uniquely
+similar: petal dimensions (petal length versus petal width) from iris,
+and penguin size (flipper length versus body mass) in penguins (Figure
+3). The overall trend across all three species is approximately linear
+for both iris and penguins (Figure 3A,B). Teachers may encourage
+students to explore how simple linear regression results and predictions
+differ when the species variable is omitted (Figure 3A,B), compared to
+multiple linear regression with species included (Figure 3C,D). Notably,
+distinctions between species are clearer for iris petals - particularly,
+the much smaller petals for Iris setosa - compared to penguins, in which
+Adélie and chinstrap penguins are largely overlapping in body size (body
+mass and flipper length), and are both generally smaller than gentoos.
+
+\hypertarget{penguins-bonus-simpsons-paradox}{%
+\subsubsection{Penguins bonus: Simpson's
+Paradox}\label{penguins-bonus-simpsons-paradox}}
+
+Simpson's Paradox is a data phenomenon in which a trend observed between
+variables is reversed when data are pooled, omitting a meaningful
+variable. While often taught and discussed in statistics courses,
+finding a real-world and approachable example of Simpson's paradox can
+be a challenge. Here, we show one (of several possible; see Figure XX)
+Simpson's Paradox examples in \textbf{penguins}: exploring bill
+dimensions with and without \emph{species} included (Figure XX). When
+penguin \emph{species} is omitted (Figure XX), the bill length and depth
+appear negatively correlated overall. The trend is reversed when
+\emph{species} is included, revealing a positive correlation between
+bill length and bill depth (Figure XX)
+
+\hypertarget{principal-component-analysis}{%
+\subsubsection{Principal component
+analysis}\label{principal-component-analysis}}
+
+Principal component analysis (PCA) is a dimensional reduction method
+commonly used to explore patterns in multivariate data. PCA tutorials
+frequently employ the \textbf{iris} dataset, which is useful for
+teaching the method due to multivariate normality, and clear,
+approachable outcomes for variable loadings and clustering.
+
+A comparison of PCA with the four variables of structural size
+measurements in \textbf{penguins} and \textbf{iris} (both normalized
+prior to PCA) reveals highly similar results (Figure XX). For both
+datasets, one species is distinct (gentoo penguins, and setosa irises)
+while the other two species (chinstrap/Adélie and versicolor/virginica)
+appear somewhat overlapping in the first two principal components
+(Figure XX). Variance explained by each principal component (PC) is
+similar, particularly for PC1 and PC2: for \textbf{penguins}, 88.15\% of
+total variance is captured by the first two PCs, compared to 95.81\% for
+\textbf{iris}, with a similarly large percentage of variance captured by
+PC1 and PC2 in each (Figure XX).
+
+\hypertarget{k-means-clustering}{%
+\subsubsection{K-means clustering}\label{k-means-clustering}}
+
+Unsupervised clustering by k-means is a common and popular entryway to
+machine learning and classification, and frequently employs the
+\textbf{iris} data for introductory examples. The \textbf{penguins} data
+provides similar opportunities for introducing k-means clustering. For
+simplicity, we compare k-means clustering using only two variables for
+each dataset: for \textbf{iris}, petal width and petal length, and for
+\textbf{penguins}, bill length and bill depth. All variables are scaled
+prior to k-means. Three clusters (\emph{k} = 3) are specified for each
+since there are three species of both \emph{Iris} (\emph{setosa},
+\emph{versicolor}, and \emph{virginica}) and penguins (Adélie, chinstrap
+and gentoo).
+
+K-means clustering with penguin bill dimensions and iris petal
+dimensions yields largely distinct clusters each dominated by one
+species (Figure XX, Table XX). For iris petal dimensions, k-means yields
+a perfectly separated cluster (Cluster 1) containing all 50 setosa iris
+observations and zero misclassified virginica or versicolor irises
+(Table 4). While clustering is not perfectly distinct for any penguin
+species, each species is largely contained within a single cluster, with
+little overlap from the other two species. For example, considering
+Adélie penguins (orange observations in Figure 6A): 147 (out of 151)
+Adélie penguins are assigned to Cluster 1, zero are assigned to Cluster
+2, and 4 are assigned to the chinstrap-dominated Cluster 3 (Table 4).
+Only 5 (of 68) chinstrap penguins and 1 (of 123) gentoo penguins are
+assigned to the Adélie-dominated Cluster 1 (Table 4).
+
+\hypertarget{other-alternatives-to-iris}{%
+\subsection{Other alternatives to
+iris}\label{other-alternatives-to-iris}}
+
+\hypertarget{discussion}{%
+\subsection{Discussion}\label{discussion}}
+
+Here, we have shown that the \textbf{penguins} data, available in the
+\textbf{palmerpenguins} R package, can replace \textbf{iris} for a
+number of common use cases in data science and statistics education
+including exploratory data visualization, linear correlation and
+regression, PCA and clustering by k-means. In addition, we feel that
+teaching and learning opportunities in \textbf{penguins} are increased
+due to a greater number of variables, missing values, unequal sample
+sizes, and Simpson's paradox examples.
+
+\hypertarget{conclusion}{%
+\subsection{Conclusion}\label{conclusion}}
+
+\hypertarget{appendix}{%
+\subsection{Appendix}\label{appendix}}
+
+\hypertarget{appendix-a}{%
+\subsubsection{Appendix A}\label{appendix-a}}
+
+\hypertarget{appendix-b}{%
+\subsubsection{Appendix B}\label{appendix-b}}
+
+Data in the \textbf{penguins} object have been minimally updated from
+\textbf{penguins\_raw} as follows:
+
+\begin{itemize}
+\tightlist
+\item
+  All variable names are converted to lower snake case
+\item
+  Entries in \emph{species} are truncated to only include the common
+  name (e.g.~``gentoo'', instead of ``gentoo penguin (Pygoscelis
+  papua)'')
+\item
+  Recorded sex for penguin N36A1, originally recorded as ``.'', is
+  updated to \texttt{NA}
+\item
+  \emph{culmen\_length\_mm} and \emph{culmen\_depth\_mm} variable names
+  are updated to \emph{bill\_length\_mm} and \emph{bill\_depth\_mm},
+  respectively
+\item
+  Class for categorical variables (\emph{species}, \emph{island},
+  \emph{sex}) is updated to factor
+\item
+  Variable \emph{year} was pulled from clutch observations
+\end{itemize}
 
@@
 \address{%
-Allison Horst\\
-Bren School of Environmental Science and Management, University of
-California, Santa Barbara\\
-line 1\\ line 2\\
+Allison M. Horst\\
+Bren School of Environmental Science and Management\\
+University of California, Santa Barbara\\ Santa Barbara, CA 93106-5131\\
 }
@@
 \address{%
-Alison Hill\\
+Alison P. Hill\\
 RStudio, PBC\\
-line 1\\ line 2\\
+250 Northern Ave\\ Boston, MA 02210\\
 }
@@
 \address{%
-Kristen Gorman\\
+Kristen B. Gorman\\
 University of Alaska Fairbanks College of Fisheries and Ocean Sciences\\
-line 1\\ line 2\\
+2150 Koyukuk Drive\\ 245 O'Neill Building\\ Fairbanks, AK 99775-7220\\
 }

### 2020-08-04 penguins.Rmd
update penguins paper

@@
 
-In 1936, one year after appearing in the Bulletin of the American Iris Society, Anderson’s *Iris* measurements were published by eugenicist and statistician Ronald A.
-Fisher in Annals of Eugenics [@fisher_use_1936].
+In 1936, one year after being described in the Bulletin of the American Iris Society, Anderson’s *Iris* measurements were published in full by eugenicist and statistician Ronald A.
+Fisher in *Annals of Eugenics* [@fisher_use_1936].
 Regardless of Anderson's initial motivation, the data today remain inextricably linked to Fisher’s eugenics research and are even commonly, if unfairly, referred to as "Fisher's iris data.” For example, an August 2020 search for “Iris flower data set” in Wikipedia returns an article beginning with “The iris flower data set or Fisher’s iris data…”, then later in the article: “It is sometimes called Anderson's Iris data set” [@wikipedia_iris_2020].
@@
 
-      `remotes::install_github("allisonhorst/palmerpenguins")`.
+      `remotes::install_github("allisonhorst/palmerpenguins")`
 
@@
 
-### Continuous quantitative variables: overview
+```{r, echo = FALSE}
+iris_counts <- iris %>% 
+  count(Species) %>% 
+  rename(`Iris species` = Species)
 
-Distributions, relationships between variables and clustering can be visually explored between species for the four structural size measurements in **iris** (sepal width and length, petal width and length) and **penguins** (flipper length, body mass, bill length and depth).
-It is worth noting that the correlation coefficients presented are Pearson’s *r*, developed by and named for another famous eugenicist and statistician, Karl Pearson [ref].
+penguin_counts <- penguins %>% 
+  mutate(species = as.character(species)) %>% 
+  mutate(species = case_when(
+    species == "Adelie" ~ "Adélie", 
+    species == "Chinstrap" ~ "chinstrap",
+    species == "Gentoo" ~ "gentoo",
+    TRUE ~ species
+  )) %>% 
+  mutate(sex = str_to_title(sex)) %>% 
+  group_by(species, sex) %>% 
+  tally() %>% 
+  pivot_wider(names_from = sex, values_from = n) %>% 
+  replace_na(list(`NA` = 0)) %>% 
+  as.data.frame() %>% 
+  rename(`Penguin species` = species)
 
-### Linear correlation, regression, and Simpson’s Paradox
+iris_penguin_n <- cbind(iris_counts, penguin_counts) %>%
+   rename(`Sample size` = n)
 
-There exist numerous opportunities to explore linear relationships and correlations in penguins and iris both within and across species (Figures 1 & 2).
+iris_penguin_n %>% 
+   kable(caption = "Grouped sample size for iris (by species; n = 150 total) and penguins (by species and sex; n = 344 total).
+Penguins can be further grouped by variables for island and study year.", align = "lclccc") %>%
+  add_header_above(c("Iris sample size (by species)" = 2, "Penguin sample size (by species and sex)" = 4 ))
+
+```
+
+### Continuous quantitative variables
+
+Distributions, relationships between variables and clustering can be visually explored between species for the four structural size measurements in **penguins** (flipper length, body mass, bill length and depth; Figure \ref{fig:penguins-pairs}) and **iris** (sepal width and length, petal width and length; Figure \ref{fig:iris-pairs}).
+
+```{r, penguins-pairs, fig.cap = "Distribution and correlations for numeric variables in the penguins data (flipper length (mm), body mass, (g) bill length (mm) and bill depth (mm)) for the three observed species: gentoo (green, triangles); chinstrap (blue, circles); and Adélie (orange, squares).
+Correlations are Pearson's r (*p < 0.05; **p < 0.01; ***p < 0.001).", warning = FALSE, message = FALSE, echo = FALSE}
+penguin_pairs <- penguins %>%
+  mutate(species = as.character(species)) %>% 
+  mutate(species = case_when(
+    species == "Adelie" ~ "Adélie", 
+    TRUE ~ species
+  )) %>% 
+  select(species, where(is.numeric)) %>% 
+  ggpairs(aes(color = species, shape = species),
+          columns = c("flipper_length_mm", "body_mass_g", 
+                      "bill_length_mm", "bill_depth_mm"),
+          columnLabels = c("Flipper length (mm)","Body mass (g)", "Bill length (mm)", "Bill depth (mm)"),
+          upper = list(continuous = wrap("cor", size = 2.7, color = "black")),
+          lower = list(continuous = wrap(ggally_points, size = 1.3))) +
+  scale_fill_manual(values = c("#e59f00","#0072b2","#009e73")) +
+  scale_color_manual(values = c("#e59f00","#0072b2","#009e73")) +
+  scale_shape_manual(values = c(15,16,17)) +
+  theme_minimal() +
+  theme(
+        text = element_text(size = 9),
+        panel.grid.major = element_line(colour = NA),
+        panel.grid.minor = element_blank(),
+        panel.border = element_rect(color = "gray80", fill = NA)
+        )
+
+penguin_pairs
+```
+
+```{r, iris-pairs, fig.cap = "Distribution and correlations for numeric variables in the iris data (petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm)) for the three included Iris species: setosa (light gray, circles); versicolor (dark gray, triangles); and virginica (black, squares).
+Correlations are Pearson's r (*p < 0.05; **p < 0.01; ***p < 0.001).", warning = FALSE, message = FALSE, echo = FALSE}
+iris_pairs <- iris %>%
+  ggpairs(aes(color = Species, shape = Species),
+          columns = c("Petal.Length", "Petal.Width", 
+                      "Sepal.Length", "Sepal.Width"),
+          columnLabels = c("Petal length (cm)","Petal width (cm)", "Sepal length (cm)", "Sepal width (cm)"),
+          upper = list(continuous = wrap("cor", size = 2.7, color = "black"))) +
+  scale_colour_manual(values = c("gray70","gray40","black")) +
+  scale_fill_manual(values = c("gray70","gray40","black")) +
+  theme_minimal() +
+  theme(
+        text = element_text(size = 9),
+        panel.grid.major = element_line(colour = NA),
+        panel.grid.minor = element_blank(),
+        panel.background = element_rect(fill = "white", color = "gray80")
+        )
+
+iris_pairs
+
+```
+
+There exist numerous opportunities to explore linear relationships and correlations in **penguins** and **iris** both within and across species (Figures \ref{fig:penguins-pairs} & \ref{fig:iris-pairs}).
 Here, we highlight one comparison that is uniquely similar: petal dimensions (petal length versus petal width) from iris, and penguin size (flipper length versus body mass) in penguins (Figure 3).
@@
 Teachers may encourage students to explore how simple linear regression results and predictions differ when the species variable is omitted (Figure 3A,B), compared to multiple linear regression with species included (Figure 3C,D).
-Notably, distinctions between species are clearer for iris petals - particularly, the much smaller petals for Iris setosa - compared to penguins, in which Adélie and chinstrap penguins are largely overlapping in body size (body mass and flipper length), and are both generally smaller than gentoos.
+
+Notably, distinctions between species are clearer for iris petals - particularly, the much smaller petals for *Iris setosa* - compared to penguins, in which Adélie and chinstrap penguins are largely overlapping in body size (body mass and flipper length), and are both generally smaller than gentoos.
 
@@
 
-Here, we have shown that the **penguins** data, available in the **palmerpenguins** R package, can replace **iris** for a number of common use cases in data science and statistics education including exploratory data visualization, linear correlation and regression, PCA and clustering by k-means.
+Educators have three options regarding **iris** use in course materials:
+
+1. Use **iris** without addressing its use in Fisher's eugenics research
+2. Use **iris** *and* address its use in Fisher's eugenics research
+3. Replace **iris** with an alternative dataset offering similar teaching and learning opportunities
+
+Here, we have shown that structural size measurements for Palmer Archipelago *Pygoscelis* penguins, available as **penguins** in the **palmerpenguins** R package, can replace **iris** for a number of common use cases in data science and statistics education including exploratory data visualization, linear correlation and regression, PCA, and clustering by k-means.
 In addition, teaching and learning opportunities in **penguins** are increased due to a greater number of variables, missing values, unequal sample sizes, and Simpson's paradox examples.
+
+The **iris** dataset is widespread today because it was employed by Fisher to advance statistical methods for eugenics research.
+While the *Irises* were described in Anderson's 1935 article in *The Bulletin of the American Iris Society* [@anderson_irises_1935], the measurements were published in full in Fisher's 1936 *Annals of Eugenics* publication [@fisher_use_1936] - a journal for which Fisher served as editor before becoming chair of the Department of Eugenics at University College London.
 

### 2020-08-04 penguins.tex
update penguins paper

@@
 
-In 1936, one year after appearing in the Bulletin of the American Iris
-Society, Anderson's \emph{Iris} measurements were published by
-eugenicist and statistician Ronald A. Fisher in Annals of Eugenics
-\citep{fisher_use_1936}. Regardless of Anderson's initial motivation,
-the data today remain inextricably linked to Fisher's eugenics research
-and are even commonly, if unfairly, referred to as ``Fisher's iris
-data.'' For example, an August 2020 search for ``Iris flower data set''
-in Wikipedia returns an article beginning with ``The iris flower data
-set or Fisher's iris data\ldots{}'', then later in the article: ``It is
-sometimes called Anderson's Iris data set'' \citep{wikipedia_iris_2020}.
-The \emph{Iris} dataset is similarly credited to Fisher in statistical
-computing literature \citep[\citet{wang_matlab_2015},
-\citet{woods_how_2015},
+In 1936, one year after being described in the Bulletin of the American
+Iris Society, Anderson's \emph{Iris} measurements were published in full
+by eugenicist and statistician Ronald A. Fisher in \emph{Annals of
+Eugenics} \citep{fisher_use_1936}. Regardless of Anderson's initial
+motivation, the data today remain inextricably linked to Fisher's
+eugenics research and are even commonly, if unfairly, referred to as
+``Fisher's iris data.'' For example, an August 2020 search for ``Iris
+flower data set'' in Wikipedia returns an article beginning with ``The
+iris flower data set or Fisher's iris data\ldots{}'', then later in the
+article: ``It is sometimes called Anderson's Iris data set''
+\citep{wikipedia_iris_2020}. The \emph{Iris} dataset is similarly
+credited to Fisher in statistical computing literature
+\citep[\citet{wang_matlab_2015}, \citet{woods_how_2015},
 \citet{chen_unsupervised_2018}]{trendafilov_simple_2009}.
@@
 \begin{verbatim}
-  `remotes::install_github("allisonhorst/palmerpenguins")`. 
+  `remotes::install_github("allisonhorst/palmerpenguins")`
 \end{verbatim}
@@
 
-\hypertarget{continuous-quantitative-variables-overview}{%
-\subsubsection{Continuous quantitative variables:
-overview}\label{continuous-quantitative-variables-overview}}
+\begin{Schunk}
+\begin{table}
+
+\caption{\label{tab:unnamed-chunk-1}Grouped sample size for iris (by species; n = 150 total) and penguins (by species and sex; n = 344 total).
+Penguins can be further grouped by variables for island and study year.}
+\centering
+\begin{tabular}[t]{l|c|l|c|c|c}
+\hline
+\multicolumn{2}{c|}{Iris sample size (by species)} & \multicolumn{4}{c}{Penguin sample size (by species and sex)} \\
+\cline{1-2} \cline{3-6}
+Iris species & Sample size & Penguin species & Female & Male & NA\\
+\hline
+setosa & 50 & Adélie & 73 & 73 & 6\\
+\hline
+versicolor & 50 & chinstrap & 34 & 34 & 0\\
+\hline
+virginica & 50 & gentoo & 58 & 61 & 5\\
+\hline
+\end{tabular}
+\end{table}
+
+\end{Schunk}
+
+\hypertarget{continuous-quantitative-variables}{%
+\subsubsection{Continuous quantitative
+variables}\label{continuous-quantitative-variables}}
 
@@
 visually explored between species for the four structural size
-measurements in \textbf{iris} (sepal width and length, petal width and
-length) and \textbf{penguins} (flipper length, body mass, bill length
-and depth). It is worth noting that the correlation coefficients
-presented are Pearson's \emph{r}, developed by and named for another
-famous eugenicist and statistician, Karl Pearson {[}ref{]}.
+measurements in \textbf{penguins} (flipper length, body mass, bill
+length and depth; Figure \ref{fig:penguins-pairs}) and \textbf{iris}
+(sepal width and length, petal width and length; Figure
+\ref{fig:iris-pairs}).
 
-\hypertarget{linear-correlation-regression-and-simpsons-paradox}{%
-\subsubsection{Linear correlation, regression, and Simpson's
-Paradox}\label{linear-correlation-regression-and-simpsons-paradox}}
+\begin{Schunk}
+\begin{figure}
+\includegraphics{penguins_files/figure-latex/penguins-pairs-1} \caption[Distribution and correlations for numeric variables in the penguins data (flipper length (mm), body mass, (g) bill length (mm) and bill depth (mm)) for the three observed species]{Distribution and correlations for numeric variables in the penguins data (flipper length (mm), body mass, (g) bill length (mm) and bill depth (mm)) for the three observed species: gentoo (green, triangles); chinstrap (blue, circles); and Adélie (orange, squares).
+Correlations are Pearson's r (*p < 0.05; **p < 0.01; ***p < 0.001).}\label{fig:penguins-pairs}
+\end{figure}
+\end{Schunk}
+
+\begin{Schunk}
+\begin{figure}
+\includegraphics{penguins_files/figure-latex/iris-pairs-1} \caption[Distribution and correlations for numeric variables in the iris data (petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm)) for the three included Iris species]{Distribution and correlations for numeric variables in the iris data (petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm)) for the three included Iris species: setosa (light gray, circles); versicolor (dark gray, triangles); and virginica (black, squares).
+Correlations are Pearson's r (*p < 0.05; **p < 0.01; ***p < 0.001).}\label{fig:iris-pairs}
+\end{figure}
+\end{Schunk}
 
 There exist numerous opportunities to explore linear relationships and
-correlations in penguins and iris both within and across species
-(Figures 1 \& 2). Here, we highlight one comparison that is uniquely
-similar: petal dimensions (petal length versus petal width) from iris,
-and penguin size (flipper length versus body mass) in penguins (Figure
-3). The overall trend across all three species is approximately linear
-for both iris and penguins (Figure 3A,B). Teachers may encourage
-students to explore how simple linear regression results and predictions
-differ when the species variable is omitted (Figure 3A,B), compared to
-multiple linear regression with species included (Figure 3C,D). Notably,
-distinctions between species are clearer for iris petals - particularly,
-the much smaller petals for Iris setosa - compared to penguins, in which
-Adélie and chinstrap penguins are largely overlapping in body size (body
-mass and flipper length), and are both generally smaller than gentoos.
+correlations in \textbf{penguins} and \textbf{iris} both within and
+across species (Figures \ref{fig:penguins-pairs} \&
+\ref{fig:iris-pairs}). Here, we highlight one comparison that is
+uniquely similar: petal dimensions (petal length versus petal width)
+from iris, and penguin size (flipper length versus body mass) in
+penguins (Figure 3). The overall trend across all three species is
+approximately linear for both iris and penguins (Figure 3A,B). Teachers
+may encourage students to explore how simple linear regression results
+and predictions differ when the species variable is omitted (Figure
+3A,B), compared to multiple linear regression with species included
+(Figure 3C,D).
+
+Notably, distinctions between species are clearer for iris petals -
+particularly, the much smaller petals for \emph{Iris setosa} - compared
+to penguins, in which Adélie and chinstrap penguins are largely
+overlapping in body size (body mass and flipper length), and are both
+generally smaller than gentoos.
 
@@
 
-Here, we have shown that the \textbf{penguins} data, available in the
-\textbf{palmerpenguins} R package, can replace \textbf{iris} for a
-number of common use cases in data science and statistics education
+Educators have three options regarding \textbf{iris} use in course
+materials:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Use \textbf{iris} without addressing its use in Fisher's eugenics
+  research
+\item
+  Use \textbf{iris} \emph{and} address its use in Fisher's eugenics
+  research
+\item
+  Replace \textbf{iris} with an alternative dataset offering similar
+  teaching and learning opportunities
+\end{enumerate}
+
+Here, we have shown that structural size measurements for Palmer
+Archipelago \emph{Pygoscelis} penguins, available as \textbf{penguins}
+in the \textbf{palmerpenguins} R package, can replace \textbf{iris} for
+a number of common use cases in data science and statistics education
 including exploratory data visualization, linear correlation and
-regression, PCA and clustering by k-means. In addition, we feel that
-teaching and learning opportunities in \textbf{penguins} are increased
-due to a greater number of variables, missing values, unequal sample
-sizes, and Simpson's paradox examples.
+regression, PCA, and clustering by k-means. In addition, teaching and
+learning opportunities in \textbf{penguins} are increased due to a
+greater number of variables, missing values, unequal sample sizes, and
+Simpson's paradox examples.
+
+The \textbf{iris} dataset is widespread today because it was employed by
+Fisher to advance statistical methods for eugenics research. While the
+\emph{Irises} were described in Anderson's 1935 article in \emph{The
+Bulletin of the American Iris Society} \citep{anderson_irises_1935}, the
+measurements were published in full in Fisher's 1936 \emph{Annals of
+Eugenics} publication \citep{fisher_use_1936} - a journal for which
+Fisher served as editor before becoming chair of the Department of
+Eugenics at University College London.
 

### 2020-08-05 R/manuscript_code.Rmd
Move to R Journal formatting, update figures

@@
    )
+
+ iris_penguin_n <- cbind(iris_counts, penguin_counts) %>%
+   rename(`Sample size` = n)
+
+ iris_penguin_n %>% 
+   kable() %>% 
+   add_header_above(c("Iris sample size (by species)" = 2, "Penguin sample size (by species and sex)" = 4 )) %>% 
+   kable_styling()
+
 ```
@@
 ```{r, warning = FALSE, message = FALSE, echo = FALSE, fig.align = "center"}
-penguins %>%
+penguin_pairs <- penguins %>%
+  mutate(species = as.character(species)) %>% 
+  mutate(species = case_when(
+    species == "Adelie" ~ "Adélie", 
+    TRUE ~ species
+  )) %>% 
   select(species, where(is.numeric)) %>% 
@@
           columnLabels = c("Flipper length (mm)","Body mass (g)", "Bill length (mm)", "Bill depth (mm)"),
-          upper = list(continuous = wrap("cor", size = 3))) +
-  scale_fill_manual(values = c("darkorange","purple","cyan4")) +
-  scale_color_manual(values = c("darkorange","purple","cyan4")) +
+          upper = list(continuous = wrap("cor", size = 2.7)),
+          lower = list(continuous = wrap(ggally_points, size = 1.3))) +
+  scale_fill_manual(values = c("darkorange","#0072b2","#009e73")) +
+  scale_color_manual(values = c("darkorange","#0072b2","#009e73")) +
   scale_shape_manual(values = c(15,16,17)) +
@@
   theme(
-        text = element_text(size = 10, family = "Arial"),
+        text = element_text(size = 9),
         panel.grid.major = element_line(colour = NA),
@@
 
-ggsave(here("figures","penguins_ggpairs.png"), height = 5, width = 9)
+penguin_pairs
 
-iris %>%
-  ggpairs(aes(color = Species),
+iris_pairs <- iris %>%
+  ggpairs(aes(color = Species, shape = Species),
           columns = c("Petal.Length", "Petal.Width", 
@@
           columnLabels = c("Petal length (cm)","Petal width (cm)", "Sepal length (cm)", "Sepal width (cm)"),
-          upper = list(continuous = wrap("cor", size = 3))) +
-  scale_colour_manual(values = c("darkorange","purple","cyan4")) +
-  scale_fill_manual(values = c("darkorange","purple","cyan4")) +
+          upper = list(continuous = wrap("cor", size = 2.7, color = "black"))) +
+  scale_colour_manual(values = c("gray70","gray40","black")) +
+  scale_fill_manual(values = c("gray70","gray40","black")) +
   theme_minimal() +
   theme(
-        text = element_text(size = 10, family = "Arial"),
+        text = element_text(size = 9),
         panel.grid.major = element_line(colour = NA),
         panel.grid.minor = element_blank(),
-        panel.background = element_rect(fill = "gray90", color = "gray90")
+        panel.background = element_rect(fill = "white", color = "gray80")
         )
 
-ggsave(here("figures","iris_ggpairs.png"), height = 5, width = 9)
+iris_pairs
 
@@
 
-penguin_flip_mass_scatter <- ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
-  geom_point(aes(color = species, shape = species), size = 1.5, alpha = 0.5) +
+penguin_flip_mass_scatter <- 
+  penguins %>% 
+  mutate(species = as.character(species)) %>% 
+  mutate(species = case_when(
+    species == "Adelie" ~ "Adélie", 
+    species == "Chinstrap" ~ "chinstrap",
+    species == "Gentoo" ~ "gentoo",
+    TRUE ~ species
+  )) %>%
+  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
+  geom_point(aes(color = species, shape = species), size = 1.5) +
   geom_smooth(method = "lm", aes(group = species, color = species), se = FALSE, show.legend = FALSE) +
   theme_minimal() +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme(legend.position = "bottom") +
+  scale_color_manual(values = c("darkorange","#0072b2","#009e73")) +
+  theme(legend.position = c(0.2,0.85)) +
   labs(x = "Flipper length (mm)", 
@@
   theme(plot.title.position = "plot",
-        panel.border = element_rect(fill = NA, color = "gray70"))
+        panel.grid = element_blank(),
+        panel.border = element_rect(fill = NA, color = "gray70"),
+        legend.title = element_blank())
 
 penguin_flip_mass_scatter_all <- ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
-  geom_point(color = "gray50", size = 1.5, alpha = 0.5) +
+  geom_point(color = "gray50", size = 1.5) +
   geom_smooth(method = "lm", se = FALSE, color = "black") +
@@
   theme(plot.title.position = "plot",
+        panel.grid = element_blank(),
         panel.border = element_rect(fill = NA, color = "gray70"))
@@
 iris_petal_scatter <- ggplot(iris, aes(x = Petal.Length, y = Petal.Width)) +
-  geom_point(aes(color = Species, size = Species, shape = Species), size = 1.5, alpha = 0.5) +
+  geom_point(aes(color = Species, size = Species, shape = Species), size = 1.5) +
   geom_smooth(aes(group = Species, color = Species), method = "lm", se = FALSE, show.legend = FALSE) +
   theme_minimal() +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme(legend.position = "bottom") +
+  scale_color_manual(values = c("gray70","gray40","black")) +
+  theme(legend.position = c(0.2, 0.85)) +
   labs(x = "Petal length (cm)",
@@
   theme(plot.title.position = "plot",
-        panel.background = element_rect(fill = "gray90", color = "gray90"))
+        panel.grid = element_blank(),
+        panel.background = element_rect(fill = NA, color = "gray70"),
+        legend.title = element_blank())
 
 iris_petal_scatter_all <- ggplot(iris, aes(x = Petal.Length, y = Petal.Width)) +
-  geom_point(color = "gray50", size = 1.5, alpha = 0.5) +
+  geom_point(color = "gray50", size = 1.5) +
   geom_smooth(method = "lm", se = FALSE, color = "black") +
@@
   theme(plot.title.position = "plot",
-        panel.background = element_rect(fill = "gray90", color = "gray90"))
+        panel.grid = element_blank(),
+        panel.background = element_rect(fill = NA, color = "gray70"))
 
-(penguin_flip_mass_scatter_all + iris_petal_scatter_all) / (penguin_flip_mass_scatter + iris_petal_scatter) + plot_annotation(tag_levels = 'A')
+linear_example <- (penguin_flip_mass_scatter + iris_petal_scatter) + plot_annotation(tag_levels = 'A')
 
-ggsave(here("figures","iris_penguins_scatter.png"), height = 6, width = 8)
+ggsave(here("fig","linear_example.png"), width = 6, height = 3)
+```
 
+## Simpson's Paradox
+
+```{r}
 # Simpson's Paradox example (bill dimensions)
@@
 
-simpson_wspecies <- ggplot(data = penguins, aes(x = bill_length_mm, y = bill_depth_mm, group = species)) +
-  geom_point(aes(color = species, shape = species), size = 2, alpha = 0.5) +
+simpson_wspecies <- 
+  penguins %>% 
+  mutate(species = as.character(species)) %>% 
+  mutate(species = case_when(
+    species == "Adelie" ~ "Adélie",
+    TRUE ~ species)
+    ) %>% 
+    ggplot(aes(x = bill_length_mm, y = bill_depth_mm, group = species)) +
+  geom_point(aes(color = species, shape = species), size = 2, alpha = 0.8) +
   geom_smooth(method = "lm", se = FALSE, aes(color = species)) +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
+  scale_color_manual(values = c("darkorange","#0072b2","#009e73")) +
   theme_minimal() +
-  theme(panel.border = element_rect(fill = NA, color = "gray70"),
-        legend.position = c(0.16,0.14)) +
+  theme(panel.border = element_rect(fill = NA, color = "gray70")) +
   labs(x = "Bill length (mm)", y = "Bill depth (mm)")
@@
 
-simpson_gg
-
-ggsave(here("figures","simpson_gg.png"), width = 8, height = 4)
+ggsave(here("fig","simpson_gg.png"), width = 8, height = 3.5)
 ```
@@
 # Copied from Alison Hill's post! 
+### PENGUINS PCA:
 
-### PENGUINS PCA: 
+# Omit year
+penguins_noyr <- penguins %>% 
+  select(-year)
+
 penguin_recipe <-
-  recipe(~., data = penguins) %>% 
+  recipe(~., data = penguins_noyr) %>% 
   update_role(species, island, sex, new_role = "id") %>% 
@@
 # penguin_pca
-
 # Biplot: 
-
 # get pca loadings into wider format
@@
   tidyr::pivot_wider(names_from = component, id_cols = terms)
-
 # define arrow style:
@@
                      type = "closed")
-
 # Make the biplot:
@@
   geom_point(aes(color = species, shape = species), 
-             alpha = 0.5, 
+             alpha = 0.7, 
              size = 2) +
-  scale_colour_manual(values = c("darkorange","purple","cyan4"))
+  scale_colour_manual(values = c("darkorange","#0072b2","#009e73"))
 
@@
   prep()
-
 iris_pca <- 
@@
   tidy(id = "pca") 
-
 # Biplot: 
-
 # get pca loadings into wider format
@@
   geom_point(aes(color = Species, shape = Species), 
-             alpha = 0.5, 
+             alpha = 0.8, 
              size = 2) +
-  scale_colour_manual(values = c("darkorange","purple","cyan4"))
-
+  scale_colour_manual(values = c("gray70","gray40","black")) 
 iris_biplot <- iris_pca_plot +
@@
             bg.color = "white") +
-  theme(panel.background = element_rect(fill = "gray90", color = "gray90"),
+  theme(panel.background = element_rect(fill = NA, color = "gray70"),
         legend.position = "bottom")
@@
   geom_text(aes(label = round(value,2)), vjust=-0.25) +
-  theme(panel.background = element_rect(fill = "gray90", color = "gray90"))
+  theme(panel.border = element_rect(color = "gray70", fill = NA))
 
@@
 
-ggsave(here("figures","pca_plots.png"), width = 8, height = 8)
+ggsave(here("fig","pca_plots.png"), width = 8, height = 8)
 ```
@@
 ```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.align = "center"}
+
 # TWO VARIABLE k-means comparison
@@
             key_glyph = draw_key_rect) +
-  theme(legend.position = "bottom") +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4"))
+  scale_color_manual(values = c("darkorange","#0072b2","#009e73"))+
+  theme(legend.position = "bottom",
+         panel.border = element_rect(fill = NA, color = "gray70"))
 
@@
   arrange(.cluster)
-```
 
-```{r, echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center"}
+### Iris k-means
 
@@
   theme(legend.position = "bottom",
-         panel.background = element_rect(fill = "gray92", color = "gray90")) +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4"))
+         panel.border = element_rect(fill = NA, color = "gray70")) +
+  scale_color_manual(values = c("gray70","gray50","black"))
 
@@
   arrange(.cluster)
-```
 
-```{r, echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center"}
+# Combine kmeans plots for penguins & iris
 (pb_kmeans_gg | ip_kmeans_gg) + plot_annotation(tag_levels = "A")
+
+ggsave(here("fig","kmeans.png"), width = 8, height = 4.5)
 

### 2020-08-05 penguins.Rmd
move over to R Journal formattng (#11)
* remove text (code only version)

* clean up repo, start populating .Rmd & bib

* update penguins paper

* Move to R Journal formatting, update figures

@@
           columnLabels = c("Flipper length (mm)","Body mass (g)", "Bill length (mm)", "Bill depth (mm)"),
-          upper = list(continuous = wrap("cor", size = 2.7, color = "black")),
+          upper = list(continuous = wrap("cor", size = 2.7)),
           lower = list(continuous = wrap(ggally_points, size = 1.3))) +
-  scale_fill_manual(values = c("#e59f00","#0072b2","#009e73")) +
-  scale_color_manual(values = c("#e59f00","#0072b2","#009e73")) +
+  scale_fill_manual(values = c("darkorange","#0072b2","#009e73")) +
+  scale_color_manual(values = c("darkorange","#0072b2","#009e73")) +
   scale_shape_manual(values = c(15,16,17)) +
@@
 There exist numerous opportunities to explore linear relationships and correlations in **penguins** and **iris** both within and across species (Figures \ref{fig:penguins-pairs} & \ref{fig:iris-pairs}).
-Here, we highlight one comparison that is uniquely similar: petal dimensions (petal length versus petal width) from iris, and penguin size (flipper length versus body mass) in penguins (Figure 3).
-The overall trend across all three species is approximately linear for both iris and penguins (Figure 3A,B).
-Teachers may encourage students to explore how simple linear regression results and predictions differ when the species variable is omitted (Figure 3A,B), compared to multiple linear regression with species included (Figure 3C,D).
+Here, we highlight one comparison that is uniquely similar: petal dimensions (petal length versus petal width) from **iris**, and penguin size (flipper length versus body mass) in **penguins** (\ref{fig:linear-example}).
+The overall trend across all three species is approximately linear for both **iris** and **penguins**.
+Teachers may encourage students to explore how simple linear regression results and predictions differ when the species variable is omitted, compared to multiple linear regression with species included (\ref{fig:linear-example}).
+
+```{r, linear-example, fig.cap = "Representative linear relationships for penguin flipper length (mm) and body mass (g) for Adélie (orange circles), chinstrap (blue triangles), and gentoo (green squares) penguins (A), and iris petal length (cm) and width (cm) for setosa (light gray circles), versicolor (dark gray triangles) and virginica (black squares) irises (B).
+Within-species linear model is visualized for each penguin or iris species.", echo = FALSE, out.width = '6in', fig.align='center', fig.pos='htbp'}
+
+knitr::include_graphics('fig/linear_example.png')
+
+```
 
@@
 
-### Penguins bonus: Simpson’s Paradox
-
 Simpson’s Paradox is a data phenomenon in which a trend observed between variables is reversed when data are pooled, omitting a meaningful variable.
 While often taught and discussed in statistics courses, finding a real-world and approachable example of Simpson’s paradox can be a challenge.
-Here, we show one (of several possible; see Figure XX) Simpson’s Paradox examples in **penguins**: exploring bill dimensions with and without *species* included (Figure XX).
-When penguin *species* is omitted (Figure XX), the bill length and depth appear negatively correlated overall.
-The trend is reversed when *species* is included, revealing a positive correlation between bill length and bill depth (Figure XX)
+Here, we show one (of several possible; see Figure \ref{fig:penguins-pairs}) Simpson’s Paradox examples in **penguins**: exploring bill dimensions with and without *species* included (Figure \ref{fig:simpsons}).
+When penguin *species* is omitted (Figure \ref{fig:simpsons} A), bill length and depth appear negatively correlated overall.
+The trend is reversed when *species* is included, revealing an obviously positive correlation between bill length and bill depth within species (Figure \ref{fig:simpsons} B).
+
+```{r, simpsons, fig.cap = "Trends for penguin bill dimensions (bill length and bill depth, millimeters) if the ‘species’ variable is excluded (A) or included (B), illustrating Simpson’s Paradox.
+Note: linear regression for bill dimensions without including species in (A) is ill-advised; the linear trendline is only included to visualize trend reversal for Simpson’s Paradox when compared to (B).", echo = FALSE, out.width = '6in', fig.align='center', fig.pos='htbp'}
+
+knitr::include_graphics('fig/simpson_gg.png')
+
+```
 
@@
 
-A comparison of PCA with the four variables of structural size measurements in **penguins** and **iris** (both normalized prior to PCA) reveals highly similar results (Figure XX).
-For both datasets, one species is distinct (gentoo penguins, and setosa irises) while the other two species (chinstrap/Adélie and versicolor/virginica) appear somewhat overlapping in the first two principal components (Figure XX).
-Variance explained by each principal component (PC) is similar, particularly for PC1 and PC2: for **penguins**, 88.15% of total variance is captured by the first two PCs, compared to 95.81% for **iris**, with a similarly large percentage of variance captured by PC1 and PC2 in each (Figure XX).
+A comparison of PCA with the four variables of structural size measurements in **penguins** and **iris** (both normalized prior to PCA) reveals highly similar results (Figure \ref{fig:pca}).
+For both datasets, one species is distinct (gentoo penguins, and setosa irises) while the other two species (chinstrap/Adélie and versicolor/virginica) appear somewhat overlapping in the first two principal components (Figure \ref{fig:pca} A,B).
+Variance explained by each principal component (PC) is similar, particularly for PC1 and PC2: for **penguins**, 88.15% of total variance is captured by the first two PCs, compared to 95.81% for **iris**, with a similarly large percentage of variance captured by PC1 and PC2 in each (Figure \ref{fig:pca} C,D).
+
+```{r, pca, fig.cap = "Principal component analysis biplots and scree plots for structural size measurements in penguins (A,C) and iris (B,D), revealing similarities in multivariate patterns, variable loadings, and variance explained by each component.
+For penguins, variables are flipper length (mm), body mass (g), bill length (mm) and bill depth (mm); groups are visualized by species (Adélie = orange circles, chinstrap = blue triangles, gentoo = green squares).
+For iris, variables are petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm); groups are visualized by species (setosa = light gray circles, versicolor = dark gray triangles, virginica = black squares).
+Values above scree plot columns (C,D) indicate percent of total variance explained by each of the four principal components.", echo = FALSE, out.width = '6in', fig.align='center', fig.pos='htbp'}
+
+knitr::include_graphics('fig/pca_plots.png')
+
+```
 
@@
 
-K-means clustering with penguin bill dimensions and iris petal dimensions yields largely distinct clusters each dominated by one species (Figure XX, Table XX).
-For iris petal dimensions, k-means yields a perfectly separated cluster (Cluster 1) containing all 50 setosa iris observations and zero misclassified virginica or versicolor irises (Table 4).
+K-means clustering with penguin bill dimensions and iris petal dimensions yields largely distinct clusters each dominated by one species (\ref{fig:kmeans}).
+For iris petal dimensions, k-means yields a perfectly separated cluster (Cluster 1) containing all 50 setosa iris observations and zero misclassified virginica or versicolor irises (Table 2).
 While clustering is not perfectly distinct for any penguin species, each species is largely contained within a single cluster, with little overlap from the other two species.
-For example, considering Adélie penguins (orange observations in Figure 6A): 147 (out of 151) Adélie penguins are assigned to Cluster 1, zero are assigned to Cluster 2, and 4 are assigned to the chinstrap-dominated Cluster 3 (Table 4).
-Only 5 (of 68) chinstrap penguins and 1 (of 123) gentoo penguins are assigned to the Adélie-dominated Cluster 1 (Table 4).
+For example, considering Adélie penguins (orange observations in Figure \ref{fig:kmeans} A): 147 (out of 151) Adélie penguins are assigned to Cluster 1, zero are assigned to Cluster 2, and 4 are assigned to the chinstrap-dominated Cluster 3 (Table 2).
+Only 5 (of 68) chinstrap penguins and 1 (of 123) gentoo penguins are assigned to the Adélie-dominated Cluster 1 (Table 2).
+
+```{r, include = FALSE, echo = FALSE}
+# Including so kable table can be created here
+
+# TWO VARIABLE k-means comparison
+# Penguins: Bill length vs. bill depth
+# Iris: petal length vs. petal width
+
+pb_species <- penguins %>% 
+  select(species, starts_with("bill")) %>% 
+  drop_na()
+
+pb_nospecies <- pb_species %>% 
+  select(-species) %>% 
+  recipe() %>% 
+  step_normalize(all_numeric()) %>% 
+  prep() %>% 
+  juice()
+
+# Perform k-means on penguin bill dimensions (k = 3, w/20 centroid starts)
+# save augmented data
+pb_clust <- 
+  pb_nospecies %>% 
+  kmeans(centers = 3, nstart = 20) %>% 
+  broom::augment(pb_species)
+
+# Plot clusters
+pb_kmeans_gg <- 
+  pb_clust %>% 
+  ggplot(aes(x = bill_length_mm, y = bill_depth_mm)) +
+  geom_text(aes(label = .cluster, color = species),
+            key_glyph = draw_key_rect) +
+  scale_color_manual(values = c("darkorange","#0072b2","#009e73"))+
+  theme(legend.position = "bottom",
+         panel.border = element_rect(fill = NA, color = "gray70"))
+
+# Get counts in each cluster by species
+pb_clust_n <- pb_clust %>% 
+  count(species, .cluster) %>% 
+  pivot_wider(names_from = species, values_from = n, names_sort = TRUE) %>% 
+  arrange(.cluster)
+
+### Iris k-means
+
+ip_species <- iris %>% 
+  select(Species, starts_with("Petal"))
+
+# remove species factor & scale petal dimensions: 
+ip_nospecies <- ip_species %>% 
+  select(-Species) %>% 
+  recipe() %>% 
+  step_normalize(all_numeric()) %>% 
+  prep() %>% 
+  juice()
+
+# Perform k-means on iris petal dimensions (k = 3, w/20 centroid starts)
+ip_clust <- 
+  ip_nospecies %>% 
+  kmeans(centers = 3, nstart = 20) %>% 
+  broom::augment(ip_species)
+
+# Plot clusters
+ip_kmeans_gg <- 
+  ip_clust %>% 
+  ggplot(aes(x = Petal.Length, y = Petal.Width)) +
+  geom_text(aes(label = .cluster, color = Species), 
+            key_glyph = draw_key_rect) +
+  theme(legend.position = "bottom",
+         panel.border = element_rect(fill = NA, color = "gray70")) +
+  scale_color_manual(values = c("gray70","gray50","black"))
+
+# Get counts in each cluster by species
+ip_clust_n <- ip_clust %>% 
+  count(Species, .cluster) %>% 
+  pivot_wider(names_from = Species, values_from = n, names_sort = TRUE) %>% 
+  arrange(.cluster)
+
+# Combine kmeans plots for penguins & iris
+(pb_kmeans_gg | ip_kmeans_gg) + plot_annotation(tag_levels = "A")
+
+# ggsave(here("fig","kmeans.png"), width = 8, height = 4.5)
+
+kmeans_2var_table <- cbind(pb_clust_n, ip_clust_n) %>% 
+  kable(caption = "Grouped sample size for iris (by species; n = 150 total) and penguins (by species and sex; n = 344 total).
+Penguins can be further grouped by variables for island and study year.", align = "lclccc") %>%
+  kable_styling(full_width = FALSE) %>% 
+  column_spec(c(5:8), background = "gainsboro") %>% 
+  add_header_above(c("Penguins cluster assignments" = 4, "Iris cluster assignments" = 4))
+```
+
+```{r, kmeans, fig.cap = "K-means clustering outcomes for penguin bill dimensions (A) and iris petal dimensions (B).
+Numbers indicate the cluster to which an observation was assigned, revealing a high degree of separation between species for both penguins and iris.", echo = FALSE, out.width = '6in', fig.align='center', fig.pos='htbp'}
+
+knitr::include_graphics('fig/kmeans.png')
+```
+
+```{r, echo = FALSE}
+kmeans_2var_table <- cbind(pb_clust_n, ip_clust_n) %>% 
+  kable(col.names = c("Cluster", "Adélie", "chinstrap", "gentoo", "Cluster", "setosa", "versicolor", "virginica"),
+        caption = "K-means cluster assignments by species based on penguin bill length (mm) and depth (mm), and iris petal length (cm) and width (cm).", align = "cccccc") %>% 
+  kable_styling(full_width = FALSE) %>% 
+  add_header_above(c("Penguins cluster assignments" = 4, "Iris cluster assignments" = 4))
+
+kmeans_2var_table
+```
 

### 2020-08-05 penguins.tex
move over to R Journal formattng (#11)
* remove text (code only version)

* clean up repo, start populating .Rmd & bib

* update penguins paper

* Move to R Journal formatting, update figures

@@
 uniquely similar: petal dimensions (petal length versus petal width)
-from iris, and penguin size (flipper length versus body mass) in
-penguins (Figure 3). The overall trend across all three species is
-approximately linear for both iris and penguins (Figure 3A,B). Teachers
-may encourage students to explore how simple linear regression results
-and predictions differ when the species variable is omitted (Figure
-3A,B), compared to multiple linear regression with species included
-(Figure 3C,D).
+from \textbf{iris}, and penguin size (flipper length versus body mass)
+in \textbf{penguins} (\ref{fig:linear-example}). The overall trend
+across all three species is approximately linear for both \textbf{iris}
+and \textbf{penguins}. Teachers may encourage students to explore how
+simple linear regression results and predictions differ when the species
+variable is omitted, compared to multiple linear regression with species
+included (\ref{fig:linear-example}).
+
+\begin{Schunk}
+\begin{figure}[htbp]
+
+{\centering \includegraphics[width=6in]{fig/linear_example}
+
+}
+
+\caption[Representative linear relationships for penguin flipper length (mm) and body mass (g) for Adélie (orange circles), chinstrap (blue triangles), and gentoo (green squares) penguins (A), and iris petal length (cm) and width (cm) for setosa (light gray circles), versicolor (dark gray triangles) and virginica (black squares) irises (B)]{Representative linear relationships for penguin flipper length (mm) and body mass (g) for Adélie (orange circles), chinstrap (blue triangles), and gentoo (green squares) penguins (A), and iris petal length (cm) and width (cm) for setosa (light gray circles), versicolor (dark gray triangles) and virginica (black squares) irises (B).
+Within-species linear model is visualized for each penguin or iris species.}\label{fig:linear-example}
+\end{figure}
+\end{Schunk}
 
@@
 
-\hypertarget{penguins-bonus-simpsons-paradox}{%
-\subsubsection{Penguins bonus: Simpson's
-Paradox}\label{penguins-bonus-simpsons-paradox}}
-
 Simpson's Paradox is a data phenomenon in which a trend observed between
@@
 finding a real-world and approachable example of Simpson's paradox can
-be a challenge. Here, we show one (of several possible; see Figure XX)
-Simpson's Paradox examples in \textbf{penguins}: exploring bill
-dimensions with and without \emph{species} included (Figure XX). When
-penguin \emph{species} is omitted (Figure XX), the bill length and depth
-appear negatively correlated overall. The trend is reversed when
-\emph{species} is included, revealing a positive correlation between
-bill length and bill depth (Figure XX)
+be a challenge. Here, we show one (of several possible; see Figure
+\ref{fig:penguins-pairs}) Simpson's Paradox examples in
+\textbf{penguins}: exploring bill dimensions with and without
+\emph{species} included (Figure \ref{fig:simpsons}). When penguin
+\emph{species} is omitted (Figure \ref{fig:simpsons} A), bill length and
+depth appear negatively correlated overall. The trend is reversed when
+\emph{species} is included, revealing an obviously positive correlation
+between bill length and bill depth within species (Figure
+\ref{fig:simpsons} B).
+
+\begin{Schunk}
+\begin{figure}[htbp]
+
+{\centering \includegraphics[width=6in]{fig/simpson_gg}
+
+}
+
+\caption[Trends for penguin bill dimensions (bill length and bill depth, millimeters) if the ‘species’ variable is excluded (A) or included (B), illustrating Simpson’s Paradox]{Trends for penguin bill dimensions (bill length and bill depth, millimeters) if the ‘species’ variable is excluded (A) or included (B), illustrating Simpson’s Paradox.
+Note: linear regression for bill dimensions without including species in (A) is ill-advised; the linear trendline is only included to visualize trend reversal for Simpson’s Paradox when compared to (B).}\label{fig:simpsons}
+\end{figure}
+\end{Schunk}
 
@@
 measurements in \textbf{penguins} and \textbf{iris} (both normalized
-prior to PCA) reveals highly similar results (Figure XX). For both
-datasets, one species is distinct (gentoo penguins, and setosa irises)
-while the other two species (chinstrap/Adélie and versicolor/virginica)
-appear somewhat overlapping in the first two principal components
-(Figure XX). Variance explained by each principal component (PC) is
-similar, particularly for PC1 and PC2: for \textbf{penguins}, 88.15\% of
-total variance is captured by the first two PCs, compared to 95.81\% for
-\textbf{iris}, with a similarly large percentage of variance captured by
-PC1 and PC2 in each (Figure XX).
+prior to PCA) reveals highly similar results (Figure \ref{fig:pca}). For
+both datasets, one species is distinct (gentoo penguins, and setosa
+irises) while the other two species (chinstrap/Adélie and
+versicolor/virginica) appear somewhat overlapping in the first two
+principal components (Figure \ref{fig:pca} A,B). Variance explained by
+each principal component (PC) is similar, particularly for PC1 and PC2:
+for \textbf{penguins}, 88.15\% of total variance is captured by the
+first two PCs, compared to 95.81\% for \textbf{iris}, with a similarly
+large percentage of variance captured by PC1 and PC2 in each (Figure
+\ref{fig:pca} C,D).
+
+\begin{Schunk}
+\begin{figure}[htbp]
+
+{\centering \includegraphics[width=6in]{fig/pca_plots}
+
+}
+
+\caption[Principal component analysis biplots and scree plots for structural size measurements in penguins (A,C) and iris (B,D), revealing similarities in multivariate patterns, variable loadings, and variance explained by each component]{Principal component analysis biplots and scree plots for structural size measurements in penguins (A,C) and iris (B,D), revealing similarities in multivariate patterns, variable loadings, and variance explained by each component.
+For penguins, variables are flipper length (mm), body mass (g), bill length (mm) and bill depth (mm); groups are visualized by species (Adélie = orange circles, chinstrap = blue triangles, gentoo = green squares).
+For iris, variables are petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm); groups are visualized by species (setosa = light gray circles, versicolor = dark gray triangles, virginica = black squares).
+Values above scree plot columns (C,D) indicate percent of total variance explained by each of the four principal components.}\label{fig:pca}
+\end{figure}
+\end{Schunk}
 
@@
 dimensions yields largely distinct clusters each dominated by one
-species (Figure XX, Table XX). For iris petal dimensions, k-means yields
-a perfectly separated cluster (Cluster 1) containing all 50 setosa iris
+species (\ref{fig:kmeans}). For iris petal dimensions, k-means yields a
+perfectly separated cluster (Cluster 1) containing all 50 setosa iris
 observations and zero misclassified virginica or versicolor irises
-(Table 4). While clustering is not perfectly distinct for any penguin
+(Table 2). While clustering is not perfectly distinct for any penguin
 species, each species is largely contained within a single cluster, with
 little overlap from the other two species. For example, considering
-Adélie penguins (orange observations in Figure 6A): 147 (out of 151)
-Adélie penguins are assigned to Cluster 1, zero are assigned to Cluster
-2, and 4 are assigned to the chinstrap-dominated Cluster 3 (Table 4).
-Only 5 (of 68) chinstrap penguins and 1 (of 123) gentoo penguins are
-assigned to the Adélie-dominated Cluster 1 (Table 4).
+Adélie penguins (orange observations in Figure \ref{fig:kmeans} A): 147
+(out of 151) Adélie penguins are assigned to Cluster 1, zero are
+assigned to Cluster 2, and 4 are assigned to the chinstrap-dominated
+Cluster 3 (Table 2). Only 5 (of 68) chinstrap penguins and 1 (of 123)
+gentoo penguins are assigned to the Adélie-dominated Cluster 1 (Table
+2).
+
+\begin{Schunk}
+\begin{figure}[htbp]
+
+{\centering \includegraphics[width=6in]{fig/kmeans}
+
+}
+
+\caption[K-means clustering outcomes for penguin bill dimensions (A) and iris petal dimensions (B)]{K-means clustering outcomes for penguin bill dimensions (A) and iris petal dimensions (B).
+Numbers indicate the cluster to which an observation was assigned, revealing a high degree of separation between species for both penguins and iris.}\label{fig:kmeans}
+\end{figure}
+\end{Schunk}
+
+\begin{Schunk}
+\begin{table}
+
+\caption{\label{tab:unnamed-chunk-3}K-means cluster assignments by species based on penguin bill length (mm) and depth (mm), and iris petal length (cm) and width (cm).}
+\centering
+\begin{tabular}[t]{c|c|c|c|c|c|c|c}
+\hline
+\multicolumn{4}{c|}{Penguins cluster assignments} & \multicolumn{4}{c}{Iris cluster assignments} \\
+\cline{1-4} \cline{5-8}
+Cluster & Adélie & chinstrap & gentoo & Cluster & setosa & versicolor & virginica\\
+\hline
+1 & NA & 9 & 116 & 1 & NA & 48 & 4\\
+\hline
+2 & 4 & 54 & 6 & 2 & NA & 2 & 46\\
+\hline
+3 & 147 & 5 & 1 & 3 & 50 & NA & NA\\
+\hline
+\end{tabular}
+\end{table}
+
+\end{Schunk}
 

### 2020-08-06 penguins.Rmd
add bookdown to enable cross references in text

@@
 ---
-title: Palmer Archipelago Penguins Data in the palmerpenguins R Package - An Alternative to Anderson's *Irises*
+title: Palmer Archipelago Penguins Data in the palmerpenguins R Package - An Alternative
+  to Anderson's *Irises*
 author:
@@
   email: kbgorman@alaska.edu
-abstract: >
-
- The famous *Iris* dataset, which exists in the base R **datasets** package as **iris**, is ubiquitous in statistics and data science education.
-While the **iris** measurements were originally collected by botanist Edgar Anderson, their use in Ronald A.
-Fisher's eugenics research has created momentum to address the dataset's racist use and/or replace **iris** with alternate data in teaching materials.
-Penguin morphological measurements for three species of *Pygoscelis* penguins that breed on islands throughout the Palmer Archipelago, Antarctica, provide an approachable, charismatic and near drop-in replacement for the **iris** data.
-Here, we introduce the **palmerpenguins** R package and describe the included **penguins** data.
-We directly compare the two datasets for selected analyses to demonstrate that R users, in particular teachers and learners currently using the **iris** data, can switch to the Palmer Archipelago penguins for many use cases including data wrangling and visualization, correlation, regression, hypothesis testing, multivariate analysis (e.g.
-PCA), cluster analysis and classification (e.g.
-by k-means).
-
+abstract: "\nThe famous *Iris* dataset, which exists in the base R **datasets** package
+  as **iris**, is ubiquitous in statistics and data science education. While the **iris**
+  measurements were originally collected by botanist Edgar Anderson, their use in
+  Ronald A. Fisher's eugenics research has created momentum to address the dataset's
+  racist use and/or replace **iris** with alternate data in teaching materials. Penguin
+  morphological measurements for three species of *Pygoscelis* penguins that breed
+  on islands throughout the Palmer Archipelago, Antarctica, provide an approachable,
+  charismatic and near drop-in replacement for the **iris** data. Here, we introduce
+  the **palmerpenguins** R package and describe the included **penguins** data. We
+  directly compare the two datasets for selected analyses to demonstrate that R users,
+  in particular teachers and learners currently using the **iris** data, can switch
+  to the Palmer Archipelago penguins for many use cases including data wrangling and
+  visualization, correlation, regression, hypothesis testing, multivariate analysis
+  (e.g. PCA), cluster analysis and classification (e.g. by k-means). \n"
+output:
+  bookdown::pdf_book:
+    base_format: rticles::rjournal_article
 preamble: |
   % Any extra LaTeX you need in the preamble
-output: rticles::rjournal_article
 ---

### 2020-08-06 penguins.tex
add bookdown to enable cross references in text

@@
 % !TeX root = RJwrapper.tex
-\title{Palmer Archipelago Penguins Data in the palmerpenguins R Package - An
-Alternative to Anderson's \emph{Irises}}
+\title{Palmer Archipelago Penguins Data in the palmerpenguins R Package - An Alternative to Anderson's \emph{Irises}}
 \author{by Allison M. Horst, Alison P. Hill, Kristen B. Gorman}
@@
 \abstract{%
-The famous \emph{Iris} dataset, which exists in the base R
-\textbf{datasets} package as \textbf{iris}, is ubiquitous in statistics
-and data science education. While the \textbf{iris} measurements were
-originally collected by botanist Edgar Anderson, their use in Ronald A.
-Fisher's eugenics research has created momentum to address the dataset's
-racist use and/or replace \textbf{iris} with alternate data in teaching
-materials. Penguin morphological measurements for three species of
-\emph{Pygoscelis} penguins that breed on islands throughout the Palmer
-Archipelago, Antarctica, provide an approachable, charismatic and near
-drop-in replacement for the \textbf{iris} data. Here, we introduce the
-\textbf{palmerpenguins} R package and describe the included
-\textbf{penguins} data. We directly compare the two datasets for
-selected analyses to demonstrate that R users, in particular teachers
-and learners currently using the \textbf{iris} data, can switch to the
-Palmer Archipelago penguins for many use cases including data wrangling
-and visualization, correlation, regression, hypothesis testing,
-multivariate analysis (e.g.~PCA), cluster analysis and classification
-(e.g.~by k-means).
+The famous \emph{Iris} dataset, which exists in the base R \textbf{datasets} package as \textbf{iris}, is ubiquitous in statistics and data science education.
+While the \textbf{iris} measurements were originally collected by botanist Edgar Anderson, their use in Ronald A.
+Fisher's eugenics research has created momentum to address the dataset's racist use and/or replace \textbf{iris} with alternate data in teaching materials.
+Penguin morphological measurements for three species of \emph{Pygoscelis} penguins that breed on islands throughout the Palmer Archipelago, Antarctica, provide an approachable, charismatic and near drop-in replacement for the \textbf{iris} data.
+Here, we introduce the \textbf{palmerpenguins} R package and describe the included \textbf{penguins} data.
+We directly compare the two datasets for selected analyses to demonstrate that R users, in particular teachers and learners currently using the \textbf{iris} data, can switch to the Palmer Archipelago penguins for many use cases including data wrangling and visualization, correlation, regression, hypothesis testing, multivariate analysis (e.g.~PCA), cluster analysis and classification (e.g.~by k-means).
 }
@@
 
-In 1935, American botanist Edgar Anderson measured petal and sepal
-structural dimensions (length and width) for 50 flowers from three Iris
-species: \emph{Iris setosa}, \emph{Iris versicolor}, and \emph{Iris
-virginica} \citep{anderson_irises_1935}. The manageable but non-trivial
-size (5 variables and 150 total observations) and characteristics of the
-data make it amenable for introducing a wide range of statistical
-methods including data wrangling, visualization, regression,
-multivariate exploration, and machine learning. Anderson's \emph{Iris}
-dataset is built into a number of software packages including the
-auto-installed datasets package in R \citep{r_core_team_r_2019},
-Python's scikit-learn machine learning library
-\citep{pedregosa_scikit-learn_2011}, and the SAS Sashelp library (SAS
-Institute, Cary NC), which have facilitated widespread use of the data.
-As a result, eighty-five years after the data were initially published,
-Anderson's \emph{Iris} measurements are ubiquitous in statistics and
-data science course materials and tutorials. \textbf{Sounds great. Why
-consider replacing iris?}
+In 1935, American botanist Edgar Anderson measured petal and sepal structural dimensions (length and width) for 50 flowers from three Iris species: \emph{Iris setosa}, \emph{Iris versicolor}, and \emph{Iris virginica} \citep{anderson_irises_1935}.
+The manageable but non-trivial size (5 variables and 150 total observations) and characteristics of the data make it amenable for introducing a wide range of statistical methods including data wrangling, visualization, regression, multivariate exploration, and machine learning.
+Anderson's \emph{Iris} dataset is built into a number of software packages including the auto-installed datasets package in R \citep{r_core_team_r_2019}, Python's scikit-learn machine learning library \citep{pedregosa_scikit-learn_2011}, and the SAS Sashelp library (SAS Institute, Cary NC), which have facilitated widespread use of the data.
+As a result, eighty-five years after the data were initially published, Anderson's \emph{Iris} measurements are ubiquitous in statistics and data science course materials and tutorials.
+\textbf{Sounds great.
+Why consider replacing iris?}
 
-In 1936, one year after being described in the Bulletin of the American
-Iris Society, Anderson's \emph{Iris} measurements were published in full
-by eugenicist and statistician Ronald A. Fisher in \emph{Annals of
-Eugenics} \citep{fisher_use_1936}. Regardless of Anderson's initial
-motivation, the data today remain inextricably linked to Fisher's
-eugenics research and are even commonly, if unfairly, referred to as
-``Fisher's iris data.'' For example, an August 2020 search for ``Iris
-flower data set'' in Wikipedia returns an article beginning with ``The
-iris flower data set or Fisher's iris data\ldots{}'', then later in the
-article: ``It is sometimes called Anderson's Iris data set''
-\citep{wikipedia_iris_2020}. The \emph{Iris} dataset is similarly
-credited to Fisher in statistical computing literature
-\citep[\citet{wang_matlab_2015}, \citet{woods_how_2015},
-\citet{chen_unsupervised_2018}]{trendafilov_simple_2009}.
+In 1936, one year after being described in the Bulletin of the American Iris Society, Anderson's \emph{Iris} measurements were published in full by eugenicist and statistician Ronald A.
+Fisher in \emph{Annals of Eugenics} \citep{fisher_use_1936}.
+Regardless of Anderson's initial motivation, the data today remain inextricably linked to Fisher's eugenics research and are even commonly, if unfairly, referred to as ``Fisher's iris data.'' For example, an August 2020 search for ``Iris flower data set'' in Wikipedia returns an article beginning with ``The iris flower data set or Fisher's iris data\ldots{}'', then later in the article: ``It is sometimes called Anderson's Iris data set'' \citep{wikipedia_iris_2020}.
+The \emph{Iris} dataset is similarly credited to Fisher in statistical computing literature \citep[\citet{wang_matlab_2015}, \citet{woods_how_2015}, \citet{chen_unsupervised_2018}]{trendafilov_simple_2009}.
 
-There is recent momentum to address Fisher's racist legacy in
-statistics. In June 2020, the Committee of Presidents of Statistical
-Societies' R.A. Fisher Award and Lectureship was replaced by the
-Distinguished Achievement Award and Lectureship, to ``advance a more
-just, equitable, diverse, and inclusive statistical community''
-\citep{noauthor_institute_2020}. Cambridge University's Gonville and
-Caius College recently announced plans to remove a stained-glass window
-celebrating Fisher from a campus dining hall \citep{noauthor_sir_2020}.
-In the R community, there are calls to (1) address the \emph{Iris}
-dataset's use in Fisher's eugenics work, and/or (2) consider using a
-different dataset
-\citep[\citet{garrick_aden-buie_lets_2020}]{poisot_timothee_its_2020}.
-This paper describes (2): a suitable alternative dataset for educators
-wanting to replace the \emph{Iris} data in their teaching materials.
+There is recent momentum to address Fisher's racist legacy in statistics.
+In June 2020, the Committee of Presidents of Statistical Societies' R.A.
+Fisher Award and Lectureship was replaced by the Distinguished Achievement Award and Lectureship, to ``advance a more just, equitable, diverse, and inclusive statistical community'' \citep{noauthor_institute_2020}.
+Cambridge University's Gonville and Caius College recently announced plans to remove a stained-glass window celebrating Fisher from a campus dining hall \citep{noauthor_sir_2020}.
+In the R community, there are calls to (1) address the \emph{Iris} dataset's use in Fisher's eugenics work, and/or (2) consider using a different dataset \citep[\citet{garrick_aden-buie_lets_2020}]{poisot_timothee_its_2020}.
+This paper describes (2): a suitable alternative dataset for educators wanting to replace the \emph{Iris} data in their teaching materials.
 
-Considering \emph{Iris} data usage in data science and statistics
-materials, we established the following criteria for a suitable
-replacement dataset:
+Considering \emph{Iris} data usage in data science and statistics materials, we established the following criteria for a suitable replacement dataset:
 
@@
 \item
-  Features intuitive subjects and variables understandable to learners
-  across fields
+  Features intuitive subjects and variables understandable to learners across fields
 \item
@@
 \item
-  Similar opportunities for teaching and learning R, data science, and
-  statistical skills
+  Similar opportunities for teaching and learning R, data science, and statistical skills
 \end{itemize}
 
-Here, we describe an alternative to Anderson's \emph{Iris} data that
-largely satisfies these criteria: an approachable and charismatic
-dataset containing real-world morphological data for three
-\emph{Pygoscelis} penguin species that breed throughout the Western
-Antarctic Peninsula region, made available through the Long-Term
-Ecological Research Network (US LTER). By comparing data structure,
-size, and a range of analyses side-by-side for the two datasets, we
-demonstrate that the Palmer Archipelago penguin measurements can replace
-Anderson's \emph{Iris} data for many use cases in statistics and data
-science education.
+Here, we describe an alternative to Anderson's \emph{Iris} data that largely satisfies these criteria: an approachable and charismatic dataset containing real-world morphological data for three \emph{Pygoscelis} penguin species that breed throughout the Western Antarctic Peninsula region, made available through the Long-Term Ecological Research Network (US LTER).
+By comparing data structure, size, and a range of analyses side-by-side for the two datasets, we demonstrate that the Palmer Archipelago penguin measurements can replace Anderson's \emph{Iris} data for many use cases in statistics and data science education.
 
@@
 
-Body size measurements, clutch (i.e., egg laying) observations (e.g.,
-date of first egg laid, and clutch completion), and carbon
-(\textsuperscript{13}C/\textsuperscript{12}C,
-\(\delta\)\textsuperscript{13}C) and nitrogen
-(\textsuperscript{15}N/\textsuperscript{14}N,
-\(\delta\)\textsuperscript{15}N) stable isotope values of red blood
-cells for male and female adult Adélie (\emph{P. adeliae}), chinstrap
-(\emph{P. antarcticus}), and gentoo (\emph{P. papua}) penguins on three
-islands (Biscoe, Dream and Torgersen) in the Palmer Archipelago were
-collected from 2007 - 2009 by Dr.~Kristen Gorman in collaboration with
-the \href{https://pal.lternet.edu/}{Palmer Station LTER}, part of the
-\href{https://lternet.edu/}{US LTER Network}. For complete data
-collection methods and published analyses see Gorman et al.~(2014).
+Body size measurements, clutch (i.e., egg laying) observations (e.g., date of first egg laid, and clutch completion), and carbon (\textsuperscript{13}C/\textsuperscript{12}C, \(\delta\)\textsuperscript{13}C) and nitrogen (\textsuperscript{15}N/\textsuperscript{14}N, \(\delta\)\textsuperscript{15}N) stable isotope values of red blood cells for male and female adult Adélie (\emph{P.
+adeliae}), chinstrap (\emph{P.
+antarcticus}), and gentoo (\emph{P.
+papua}) penguins on three islands (Biscoe, Dream and Torgersen) in the Palmer Archipelago were collected from 2007 - 2009 by Dr.~Kristen Gorman in collaboration with the \href{https://pal.lternet.edu/}{Palmer Station LTER}, part of the \href{https://lternet.edu/}{US LTER Network}.
+For complete data collection methods and published analyses see Gorman et al.~(2014).
 
-The data in the \textbf{palmerpenguins} R package are available for use
-by CC0 license (``No Rights Reserved'') in accordance with the
-\href{https://pal.lternet.edu/data/policies}{Palmer Station LTER Data
-Policy} and the \href{https://lternet.edu/data-access-policy/}{LTER Data
-Access Policy}, and were imported from the
-\href{https://environmentaldatainitiative.org/}{Environmental Data
-Initiative (EDI) Data Portal} at the links below:
+The data in the \textbf{palmerpenguins} R package are available for use by CC0 license (``No Rights Reserved'') in accordance with the \href{https://pal.lternet.edu/data/policies}{Palmer Station LTER Data Policy} and the \href{https://lternet.edu/data-access-policy/}{LTER Data Access Policy}, and were imported from the \href{https://environmentaldatainitiative.org/}{Environmental Data Initiative (EDI) Data Portal} at the links below:
 
@@
 \item
-  Adélie penguin data (LTER and Gorman 2020a):
-  \href{https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.219.5}{KNB-LTER
-  Data Package 219.5}
+  Adélie penguin data (LTER and Gorman 2020a): \href{https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.219.5}{KNB-LTER Data Package 219.5}
 \item
-  Chinstrap penguin data (LTER and Gorman 2020b):
-  \href{https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.221.6}{KNB-LTER
-  Data Package 221.5}
+  Chinstrap penguin data (LTER and Gorman 2020b): \href{https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.221.6}{KNB-LTER Data Package 221.5}
 \item
-  Gentoo penguin data (LTER and Gorman 2020c):
-  \href{https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.220.5}{KNB-LTER
-  Data Package 220.5}
+  Gentoo penguin data (LTER and Gorman 2020c): \href{https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.220.5}{KNB-LTER Data Package 220.5}
 \end{itemize}
@@
 \hypertarget{r-package-palmerpenguins}{%
-\subsection{\texorpdfstring{R package:
-\textbf{palmerpenguins}}{R package: palmerpenguins}}\label{r-package-palmerpenguins}}
+\subsection{\texorpdfstring{R package: \textbf{palmerpenguins}}{R package: palmerpenguins}}\label{r-package-palmerpenguins}}
 
@@
 
-Alternatively, the development version of the package can be installed
-from GitHub:
+Alternatively, the development version of the package can be installed from GitHub:
 
@@
 
-Information, examples, and links to community-contributed materials are
-available on the \textbf{palmerpenguins} package website:
-\url{https://allisonhorst.github.io/palmerpenguins/}.
+Information, examples, and links to community-contributed materials are available on the \textbf{palmerpenguins} package website: \url{https://allisonhorst.github.io/palmerpenguins/}.
 
-The \textbf{palmerpenguins} R package contains two data objects:
-\textbf{penguins\_raw} and \textbf{penguins}. The \textbf{penguins\_raw}
-data consists of all raw data for 17 variables (Appendix Table 1),
-recorded completely or in part for 344 individual penguins, accessed
-directly from EDI. As a direct alternative to Anderson's \emph{Iris}
-data we recommend using the curated data in \textbf{penguins}, which is
-a subset of \textbf{penguins\_raw} retaining all 344 observations,
-minimally updated (Appendix B) and reduced to the following eight
-variables:
+The \textbf{palmerpenguins} R package contains two data objects: \textbf{penguins\_raw} and \textbf{penguins}.
+The \textbf{penguins\_raw} data consists of all raw data for 17 variables (Appendix Table 1), recorded completely or in part for 344 individual penguins, accessed directly from EDI.
+As a direct alternative to Anderson's \emph{Iris} data we recommend using the curated data in \textbf{penguins}, which is a subset of \textbf{penguins\_raw} retaining all 344 observations, minimally updated (Appendix B) and reduced to the following eight variables:
 
@@
 \item
-  \emph{species:} a factor denoting the penguin species (Adélie,
-  chinstrap, or gentoo)
+  \emph{species:} a factor denoting the penguin species (Adélie, chinstrap, or gentoo)
 \item
-  \emph{island:} a factor denoting the island (in Palmer Archipelago,
-  Antarctica) where observed (Biscoe, Dream or Torgersen)
+  \emph{island:} a factor denoting the island (in Palmer Archipelago, Antarctica) where observed (Biscoe, Dream or Torgersen)
 \item
-  \emph{bill\_length\_mm:} a number denoting length of the dorsal ridge
-  of a penguin bill (millimeters)
+  \emph{bill\_length\_mm:} a number denoting length of the dorsal ridge of a penguin bill (millimeters)
 \item
-  \emph{bill\_depth\_mm:} a number denoting the depth of a penguin bill
-  (millimeters)
+  \emph{bill\_depth\_mm:} a number denoting the depth of a penguin bill (millimeters)
 \item
-  \emph{flipper\_length\_mm:} an integer denoting the length of a
-  penguin flipper (millimeters)
+  \emph{flipper\_length\_mm:} an integer denoting the length of a penguin flipper (millimeters)
 \item
-  \emph{body\_mass\_g:} an integer denoting the weight of a penguin's
-  body (grams)
+  \emph{body\_mass\_g:} an integer denoting the weight of a penguin's body (grams)
 \item
-  \emph{sex:} a factor denoting the sex of a penguin sex (male, female)
-  based on molecular data
+  \emph{sex:} a factor denoting the sex of a penguin sex (male, female) based on molecular data
 \item
-  \emph{year:} an integer denoting the year of study (2007, 2008 or
-  2009)
+  \emph{year:} an integer denoting the year of study (2007, 2008 or 2009)
 \end{itemize}
 
-The same data exist as comma-separated value (CSV) files in the package
-(``penguins\_raw.csv'' and ``penguins.csv''), and can be read in using
-the built-in \texttt{path\_to\_file()} function in
-\textbf{palmerpenguins}. For example,
+The same data exist as comma-separated value (CSV) files in the package (``penguins\_raw.csv'' and ``penguins.csv''), and can be read in using the built-in \texttt{path\_to\_file()} function in \textbf{palmerpenguins}.
+For example,
 
@@
 
-will read in ``penguins.csv'' as if from an external file, thus
-automatically parsing \emph{species}, \emph{island} and \emph{sex}
-variables as characters. This option allows users opportunities to
-practice or demonstrate reading in data from a CSV, then updating
-variable class (e.g.~characters to factors).
+will read in ``penguins.csv'' as if from an external file, thus automatically parsing \emph{species}, \emph{island} and \emph{sex} variables as characters.
+This option allows users opportunities to practice or demonstrate reading in data from a CSV, then updating variable class (e.g.~characters to factors).
 
@@
 
-Python: Python users can access the penguins data in the
-\textbf{seaborn} data visualization library (Waskom et al.~2017).
-Example code to load the data in Python:
+Python: Python users can access the penguins data in the \textbf{seaborn} data visualization library (Waskom et al.~2017). Example code to load the data in Python:
 
@@
 
-Julia: Julia users can access the penguins data in the
-\textbf{PalmerPenguins.jl} package. Example code to import the penguins
-data through \textbf{PalmerPenguins.jl}:
+Julia: Julia users can access the penguins data in the \textbf{PalmerPenguins.jl} package. Example code to import the penguins data through \textbf{PalmerPenguins.jl}:
 
@@
 
-All analyses were performed in the R language environment using version
-3.6.2 (R Core Team 2019). Complete code for this paper is shared in the
-Supplemental Material. We acknowledge the following R packages used in
-analyses, with gratitude to developers and contributors:
+All analyses were performed in the R language environment using version 3.6.2 (R Core Team 2019).
+Complete code for this paper is shared in the Supplemental Material.
+We acknowledge the following R packages used in analyses, with gratitude to developers and contributors:
 
@@
 \item
-  \textbf{shadowtext} (Yu 2019), to add a background color to text
-  labels
+  \textbf{shadowtext} (Yu 2019), to add a background color to text labels
 \item
@@
 \item
-  \textbf{base} and \textbf{stats} (R Core Team 2019) for various
-  analyses throughout
+  \textbf{base} and \textbf{stats} (R Core Team 2019) for various analyses throughout
 \item
@@
 \hypertarget{selected-comparisons-between-iris-and-penguins}{%
-\subsection{\texorpdfstring{Selected comparisons between \textbf{iris}
-and
-\textbf{penguins}}{Selected comparisons between iris and penguins}}\label{selected-comparisons-between-iris-and-penguins}}
+\subsection{\texorpdfstring{Selected comparisons between \textbf{iris} and \textbf{penguins}}{Selected comparisons between iris and penguins}}\label{selected-comparisons-between-iris-and-penguins}}
 
-The \textbf{penguins} data in \textbf{palmerpenguins} is generally
-useful and approachable for data science and statistics education, and
-is uniquely well-suited to replace the \textbf{iris} dataset.
-Comparisons presented are selected examples for common \textbf{iris}
-uses, and are not exhaustive.
+The \textbf{penguins} data in \textbf{palmerpenguins} is generally useful and approachable for data science and statistics education, and is uniquely well-suited to replace the \textbf{iris} dataset.
+Comparisons presented are selected examples for common \textbf{iris} uses, and are not exhaustive.
 
 \hypertarget{data-structure-and-sample-size}{%
-\subsubsection{Data structure and sample
-size}\label{data-structure-and-sample-size}}
+\subsubsection{Data structure and sample size}\label{data-structure-and-sample-size}}
 
-Both \textbf{iris} and \textbf{penguins} are in tidy format {[}ref{]}
-with each column denoting a single variable and each row containing
-measurements for a single \emph{Iris} flower or penguin. The two
-datasets are comparable in size: dimensions (columns × rows) are 5 × 150
-and 8 × 344 for \textbf{iris} and \textbf{penguins}, respectively, and
-sample sizes within species are similar (Table 1). Notably, sample sizes
-differ for the three penguin species, while sample sizes in
-\textbf{iris} are equal (n = 50 for each Iris species). Multiple factor
-variables in \textbf{penguins} (\emph{species}, \emph{island} and
-\emph{sex}) along with \emph{year} create additional opportunities for
-grouping, compared to the single factor (\emph{species}) in
-\textbf{iris}.
+Both \textbf{iris} and \textbf{penguins} are in tidy format {[}ref{]} with each column denoting a single variable and each row containing measurements for a single \emph{Iris} flower or penguin.
+The two datasets are comparable in size: dimensions (columns × rows) are 5 × 150 and 8 × 344 for \textbf{iris} and \textbf{penguins}, respectively, and sample sizes within species are similar (Table 1).
+Notably, sample sizes differ for the three penguin species, while sample sizes in \textbf{iris} are equal (n = 50 for each Iris species).
+Multiple factor variables in \textbf{penguins} (\emph{species}, \emph{island} and \emph{sex}) along with \emph{year} create additional opportunities for grouping, compared to the single factor (\emph{species}) in \textbf{iris}.
 
-Unlike \textbf{iris}, which contains only complete cases, the
-\textbf{penguins} dataset contains a small number of missing values
-(n\textsubscript{missing} = 19, out of 2,752 total values). Missing
-values and unequal sample sizes are common in real-world data, and we
-believe add learning value to \textbf{penguins}.
+Unlike \textbf{iris}, which contains only complete cases, the \textbf{penguins} dataset contains a small number of missing values (n\textsubscript{missing} = 19, out of 2,752 total values).
+Missing values and unequal sample sizes are common in real-world data, and we believe add learning value to \textbf{penguins}.
 
@@
 
-\caption{\label{tab:unnamed-chunk-1}Grouped sample size for iris (by species; n = 150 total) and penguins (by species and sex; n = 344 total).
+\caption{(\#tab:unnamed-chunk-1)Grouped sample size for iris (by species; n = 150 total) and penguins (by species and sex; n = 344 total).
 Penguins can be further grouped by variables for island and study year.}
@@
 \hypertarget{continuous-quantitative-variables}{%
-\subsubsection{Continuous quantitative
-variables}\label{continuous-quantitative-variables}}
+\subsubsection{Continuous quantitative variables}\label{continuous-quantitative-variables}}
 
-Distributions, relationships between variables and clustering can be
-visually explored between species for the four structural size
-measurements in \textbf{penguins} (flipper length, body mass, bill
-length and depth; Figure \ref{fig:penguins-pairs}) and \textbf{iris}
-(sepal width and length, petal width and length; Figure
-\ref{fig:iris-pairs}).
+Distributions, relationships between variables and clustering can be visually explored between species for the four structural size measurements in \textbf{penguins} (flipper length, body mass, bill length and depth; Figure \ref{fig:penguins-pairs}) and \textbf{iris} (sepal width and length, petal width and length; Figure \ref{fig:iris-pairs}).
 
@@
 \includegraphics{penguins_files/figure-latex/penguins-pairs-1} \caption[Distribution and correlations for numeric variables in the penguins data (flipper length (mm), body mass, (g) bill length (mm) and bill depth (mm)) for the three observed species]{Distribution and correlations for numeric variables in the penguins data (flipper length (mm), body mass, (g) bill length (mm) and bill depth (mm)) for the three observed species: gentoo (green, triangles); chinstrap (blue, circles); and Adélie (orange, squares).
-Correlations are Pearson's r (*p < 0.05; **p < 0.01; ***p < 0.001).}\label{fig:penguins-pairs}
+Correlations are Pearson's r (*p < 0.05; **p < 0.01; ***p < 0.001).}(\#fig:penguins-pairs)
 \end{figure}
@@
 \includegraphics{penguins_files/figure-latex/iris-pairs-1} \caption[Distribution and correlations for numeric variables in the iris data (petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm)) for the three included Iris species]{Distribution and correlations for numeric variables in the iris data (petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm)) for the three included Iris species: setosa (light gray, circles); versicolor (dark gray, triangles); and virginica (black, squares).
-Correlations are Pearson's r (*p < 0.05; **p < 0.01; ***p < 0.001).}\label{fig:iris-pairs}
+Correlations are Pearson's r (*p < 0.05; **p < 0.01; ***p < 0.001).}(\#fig:iris-pairs)
 \end{figure}
@@
 
-There exist numerous opportunities to explore linear relationships and
-correlations in \textbf{penguins} and \textbf{iris} both within and
-across species (Figures \ref{fig:penguins-pairs} \&
-\ref{fig:iris-pairs}). Here, we highlight one comparison that is
-uniquely similar: petal dimensions (petal length versus petal width)
-from \textbf{iris}, and penguin size (flipper length versus body mass)
-in \textbf{penguins} (\ref{fig:linear-example}). The overall trend
-across all three species is approximately linear for both \textbf{iris}
-and \textbf{penguins}. Teachers may encourage students to explore how
-simple linear regression results and predictions differ when the species
-variable is omitted, compared to multiple linear regression with species
-included (\ref{fig:linear-example}).
+There exist numerous opportunities to explore linear relationships and correlations in \textbf{penguins} and \textbf{iris} both within and across species (Figures \ref{fig:penguins-pairs} \& \ref{fig:iris-pairs}).
+Here, we highlight one comparison that is uniquely similar: petal dimensions (petal length versus petal width) from \textbf{iris}, and penguin size (flipper length versus body mass) in \textbf{penguins} (\ref{fig:linear-example}).
+The overall trend across all three species is approximately linear for both \textbf{iris} and \textbf{penguins}.
+Teachers may encourage students to explore how simple linear regression results and predictions differ when the species variable is omitted, compared to multiple linear regression with species included (\ref{fig:linear-example}).
 
@@
 \caption[Representative linear relationships for penguin flipper length (mm) and body mass (g) for Adélie (orange circles), chinstrap (blue triangles), and gentoo (green squares) penguins (A), and iris petal length (cm) and width (cm) for setosa (light gray circles), versicolor (dark gray triangles) and virginica (black squares) irises (B)]{Representative linear relationships for penguin flipper length (mm) and body mass (g) for Adélie (orange circles), chinstrap (blue triangles), and gentoo (green squares) penguins (A), and iris petal length (cm) and width (cm) for setosa (light gray circles), versicolor (dark gray triangles) and virginica (black squares) irises (B).
-Within-species linear model is visualized for each penguin or iris species.}\label{fig:linear-example}
+Within-species linear model is visualized for each penguin or iris species.}(\#fig:linear-example)
 \end{figure}
@@
 
-Notably, distinctions between species are clearer for iris petals -
-particularly, the much smaller petals for \emph{Iris setosa} - compared
-to penguins, in which Adélie and chinstrap penguins are largely
-overlapping in body size (body mass and flipper length), and are both
-generally smaller than gentoos.
+Notably, distinctions between species are clearer for iris petals - particularly, the much smaller petals for \emph{Iris setosa} - compared to penguins, in which Adélie and chinstrap penguins are largely overlapping in body size (body mass and flipper length), and are both generally smaller than gentoos.
 
-Simpson's Paradox is a data phenomenon in which a trend observed between
-variables is reversed when data are pooled, omitting a meaningful
-variable. While often taught and discussed in statistics courses,
-finding a real-world and approachable example of Simpson's paradox can
-be a challenge. Here, we show one (of several possible; see Figure
-\ref{fig:penguins-pairs}) Simpson's Paradox examples in
-\textbf{penguins}: exploring bill dimensions with and without
-\emph{species} included (Figure \ref{fig:simpsons}). When penguin
-\emph{species} is omitted (Figure \ref{fig:simpsons} A), bill length and
-depth appear negatively correlated overall. The trend is reversed when
-\emph{species} is included, revealing an obviously positive correlation
-between bill length and bill depth within species (Figure
-\ref{fig:simpsons} B).
+Simpson's Paradox is a data phenomenon in which a trend observed between variables is reversed when data are pooled, omitting a meaningful variable.
+While often taught and discussed in statistics courses, finding a real-world and approachable example of Simpson's paradox can be a challenge.
+Here, we show one (of several possible; see Figure \ref{fig:penguins-pairs}) Simpson's Paradox examples in \textbf{penguins}: exploring bill dimensions with and without \emph{species} included (Figure \ref{fig:simpsons}).
+When penguin \emph{species} is omitted (Figure \ref{fig:simpsons} A), bill length and depth appear negatively correlated overall.
+The trend is reversed when \emph{species} is included, revealing an obviously positive correlation between bill length and bill depth within species (Figure \ref{fig:simpsons} B).
 
@@
 \caption[Trends for penguin bill dimensions (bill length and bill depth, millimeters) if the ‘species’ variable is excluded (A) or included (B), illustrating Simpson’s Paradox]{Trends for penguin bill dimensions (bill length and bill depth, millimeters) if the ‘species’ variable is excluded (A) or included (B), illustrating Simpson’s Paradox.
-Note: linear regression for bill dimensions without including species in (A) is ill-advised; the linear trendline is only included to visualize trend reversal for Simpson’s Paradox when compared to (B).}\label{fig:simpsons}
+Note: linear regression for bill dimensions without including species in (A) is ill-advised; the linear trendline is only included to visualize trend reversal for Simpson’s Paradox when compared to (B).}(\#fig:simpsons)
 \end{figure}
@@
 \hypertarget{principal-component-analysis}{%
-\subsubsection{Principal component
-analysis}\label{principal-component-analysis}}
+\subsubsection{Principal component analysis}\label{principal-component-analysis}}
 
-Principal component analysis (PCA) is a dimensional reduction method
-commonly used to explore patterns in multivariate data. PCA tutorials
-frequently employ the \textbf{iris} dataset, which is useful for
-teaching the method due to multivariate normality, and clear,
-approachable outcomes for variable loadings and clustering.
+Principal component analysis (PCA) is a dimensional reduction method commonly used to explore patterns in multivariate data.
+PCA tutorials frequently employ the \textbf{iris} dataset, which is useful for teaching the method due to multivariate normality, and clear, approachable outcomes for variable loadings and clustering.
 
-A comparison of PCA with the four variables of structural size
-measurements in \textbf{penguins} and \textbf{iris} (both normalized
-prior to PCA) reveals highly similar results (Figure \ref{fig:pca}). For
-both datasets, one species is distinct (gentoo penguins, and setosa
-irises) while the other two species (chinstrap/Adélie and
-versicolor/virginica) appear somewhat overlapping in the first two
-principal components (Figure \ref{fig:pca} A,B). Variance explained by
-each principal component (PC) is similar, particularly for PC1 and PC2:
-for \textbf{penguins}, 88.15\% of total variance is captured by the
-first two PCs, compared to 95.81\% for \textbf{iris}, with a similarly
-large percentage of variance captured by PC1 and PC2 in each (Figure
-\ref{fig:pca} C,D).
+A comparison of PCA with the four variables of structural size measurements in \textbf{penguins} and \textbf{iris} (both normalized prior to PCA) reveals highly similar results (Figure \ref{fig:pca}).
+For both datasets, one species is distinct (gentoo penguins, and setosa irises) while the other two species (chinstrap/Adélie and versicolor/virginica) appear somewhat overlapping in the first two principal components (Figure \ref{fig:pca} A,B).
+Variance explained by each principal component (PC) is similar, particularly for PC1 and PC2: for \textbf{penguins}, 88.15\% of total variance is captured by the first two PCs, compared to 95.81\% for \textbf{iris}, with a similarly large percentage of variance captured by PC1 and PC2 in each (Figure \ref{fig:pca} C,D).
 
@@
 For iris, variables are petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm); groups are visualized by species (setosa = light gray circles, versicolor = dark gray triangles, virginica = black squares).
-Values above scree plot columns (C,D) indicate percent of total variance explained by each of the four principal components.}\label{fig:pca}
+Values above scree plot columns (C,D) indicate percent of total variance explained by each of the four principal components.}(\#fig:pca)
 \end{figure}
@@
 
-Unsupervised clustering by k-means is a common and popular entryway to
-machine learning and classification, and frequently employs the
-\textbf{iris} data for introductory examples. The \textbf{penguins} data
-provides similar opportunities for introducing k-means clustering. For
-simplicity, we compare k-means clustering using only two variables for
-each dataset: for \textbf{iris}, petal width and petal length, and for
-\textbf{penguins}, bill length and bill depth. All variables are scaled
-prior to k-means. Three clusters (\emph{k} = 3) are specified for each
-since there are three species of both \emph{Iris} (\emph{setosa},
-\emph{versicolor}, and \emph{virginica}) and penguins (Adélie, chinstrap
-and gentoo).
+Unsupervised clustering by k-means is a common and popular entryway to machine learning and classification, and frequently employs the \textbf{iris} data for introductory examples.
+The \textbf{penguins} data provides similar opportunities for introducing k-means clustering.
+For simplicity, we compare k-means clustering using only two variables for each dataset: for \textbf{iris}, petal width and petal length, and for \textbf{penguins}, bill length and bill depth.
+All variables are scaled prior to k-means.
+Three clusters (\emph{k} = 3) are specified for each since there are three species of both \emph{Iris} (\emph{setosa}, \emph{versicolor}, and \emph{virginica}) and penguins (Adélie, chinstrap and gentoo).
 
-K-means clustering with penguin bill dimensions and iris petal
-dimensions yields largely distinct clusters each dominated by one
-species (\ref{fig:kmeans}). For iris petal dimensions, k-means yields a
-perfectly separated cluster (Cluster 1) containing all 50 setosa iris
-observations and zero misclassified virginica or versicolor irises
-(Table 2). While clustering is not perfectly distinct for any penguin
-species, each species is largely contained within a single cluster, with
-little overlap from the other two species. For example, considering
-Adélie penguins (orange observations in Figure \ref{fig:kmeans} A): 147
-(out of 151) Adélie penguins are assigned to Cluster 1, zero are
-assigned to Cluster 2, and 4 are assigned to the chinstrap-dominated
-Cluster 3 (Table 2). Only 5 (of 68) chinstrap penguins and 1 (of 123)
-gentoo penguins are assigned to the Adélie-dominated Cluster 1 (Table
-2).
+K-means clustering with penguin bill dimensions and iris petal dimensions yields largely distinct clusters each dominated by one species (\ref{fig:kmeans}).
+For iris petal dimensions, k-means yields a perfectly separated cluster (Cluster 1) containing all 50 setosa iris observations and zero misclassified virginica or versicolor irises (Table 2).
+While clustering is not perfectly distinct for any penguin species, each species is largely contained within a single cluster, with little overlap from the other two species.
+For example, considering Adélie penguins (orange observations in Figure \ref{fig:kmeans} A): 147 (out of 151) Adélie penguins are assigned to Cluster 1, zero are assigned to Cluster 2, and 4 are assigned to the chinstrap-dominated Cluster 3 (Table 2).
+Only 5 (of 68) chinstrap penguins and 1 (of 123) gentoo penguins are assigned to the Adélie-dominated Cluster 1 (Table 2).
 
@@
 \caption[K-means clustering outcomes for penguin bill dimensions (A) and iris petal dimensions (B)]{K-means clustering outcomes for penguin bill dimensions (A) and iris petal dimensions (B).
-Numbers indicate the cluster to which an observation was assigned, revealing a high degree of separation between species for both penguins and iris.}\label{fig:kmeans}
+Numbers indicate the cluster to which an observation was assigned, revealing a high degree of separation between species for both penguins and iris.}(\#fig:kmeans)
 \end{figure}
@@
 
-\caption{\label{tab:unnamed-chunk-3}K-means cluster assignments by species based on penguin bill length (mm) and depth (mm), and iris petal length (cm) and width (cm).}
+\caption{(\#tab:unnamed-chunk-3)K-means cluster assignments by species based on penguin bill length (mm) and depth (mm), and iris petal length (cm) and width (cm).}
 \centering
@@
 \hline
-1 & NA & 9 & 116 & 1 & NA & 48 & 4\\
+1 & 4 & 54 & 6 & 1 & 50 & NA & NA\\
 \hline
-2 & 4 & 54 & 6 & 2 & NA & 2 & 46\\
+2 & NA & 9 & 116 & 2 & NA & 2 & 46\\
 \hline
-3 & 147 & 5 & 1 & 3 & 50 & NA & NA\\
+3 & 147 & 5 & 1 & 3 & NA & 48 & 4\\
 \hline
@@
 \hypertarget{other-alternatives-to-iris}{%
-\subsection{Other alternatives to
-iris}\label{other-alternatives-to-iris}}
+\subsection{Other alternatives to iris}\label{other-alternatives-to-iris}}
 
@@
 
-Educators have three options regarding \textbf{iris} use in course
-materials:
+Educators have three options regarding \textbf{iris} use in course materials:
 
@@
 \item
-  Use \textbf{iris} without addressing its use in Fisher's eugenics
-  research
+  Use \textbf{iris} without addressing its use in Fisher's eugenics research
 \item
-  Use \textbf{iris} \emph{and} address its use in Fisher's eugenics
-  research
+  Use \textbf{iris} \emph{and} address its use in Fisher's eugenics research
 \item
-  Replace \textbf{iris} with an alternative dataset offering similar
-  teaching and learning opportunities
+  Replace \textbf{iris} with an alternative dataset offering similar teaching and learning opportunities
 \end{enumerate}
 
-Here, we have shown that structural size measurements for Palmer
-Archipelago \emph{Pygoscelis} penguins, available as \textbf{penguins}
-in the \textbf{palmerpenguins} R package, can replace \textbf{iris} for
-a number of common use cases in data science and statistics education
-including exploratory data visualization, linear correlation and
-regression, PCA, and clustering by k-means. In addition, teaching and
-learning opportunities in \textbf{penguins} are increased due to a
-greater number of variables, missing values, unequal sample sizes, and
-Simpson's paradox examples.
+Here, we have shown that structural size measurements for Palmer Archipelago \emph{Pygoscelis} penguins, available as \textbf{penguins} in the \textbf{palmerpenguins} R package, can replace \textbf{iris} for a number of common use cases in data science and statistics education including exploratory data visualization, linear correlation and regression, PCA, and clustering by k-means.
+In addition, teaching and learning opportunities in \textbf{penguins} are increased due to a greater number of variables, missing values, unequal sample sizes, and Simpson's paradox examples.
 
-The \textbf{iris} dataset is widespread today because it was employed by
-Fisher to advance statistical methods for eugenics research. While the
-\emph{Irises} were described in Anderson's 1935 article in \emph{The
-Bulletin of the American Iris Society} \citep{anderson_irises_1935}, the
-measurements were published in full in Fisher's 1936 \emph{Annals of
-Eugenics} publication \citep{fisher_use_1936} - a journal for which
-Fisher served as editor before becoming chair of the Department of
-Eugenics at University College London.
+The \textbf{iris} dataset is widespread today because it was employed by Fisher to advance statistical methods for eugenics research.
+While the \emph{Irises} were described in Anderson's 1935 article in \emph{The Bulletin of the American Iris Society} \citep{anderson_irises_1935}, the measurements were published in full in Fisher's 1936 \emph{Annals of Eugenics} publication \citep{fisher_use_1936} - a journal for which Fisher served as editor before becoming chair of the Department of Eugenics at University College London.
 
@@
 
-Data in the \textbf{penguins} object have been minimally updated from
-\textbf{penguins\_raw} as follows:
+Data in the \textbf{penguins} object have been minimally updated from \textbf{penguins\_raw} as follows:
 
@@
 \item
-  Entries in \emph{species} are truncated to only include the common
-  name (e.g.~``gentoo'', instead of ``gentoo penguin (Pygoscelis
-  papua)'')
+  Entries in \emph{species} are truncated to only include the common name (e.g.~``gentoo'', instead of ``gentoo penguin (Pygoscelis papua)'')
 \item
-  Recorded sex for penguin N36A1, originally recorded as ``.'', is
-  updated to \texttt{NA}
+  Recorded sex for penguin N36A1, originally recorded as ``.'', is updated to \texttt{NA}
 \item
-  \emph{culmen\_length\_mm} and \emph{culmen\_depth\_mm} variable names
-  are updated to \emph{bill\_length\_mm} and \emph{bill\_depth\_mm},
-  respectively
+  \emph{culmen\_length\_mm} and \emph{culmen\_depth\_mm} variable names are updated to \emph{bill\_length\_mm} and \emph{bill\_depth\_mm}, respectively
 \item
-  Class for categorical variables (\emph{species}, \emph{island},
-  \emph{sex}) is updated to factor
+  Class for categorical variables (\emph{species}, \emph{island}, \emph{sex}) is updated to factor
 \item

### 2020-09-12 penguins.Rmd
update intro

--- 2020-08-05 penguins.Rmd
@@
 
- The famous *Iris* dataset, which exists in the base R **datasets** package as **iris**, is ubiquitous in statistics and data science education.
-While the **iris** measurements were originally collected by botanist Edgar Anderson, their use in Ronald A.
-Fisher's eugenics research has created momentum to address the dataset's racist use and/or replace **iris** with alternate data in teaching materials.
+ Edgar Anderson's famous 1935 *Iris* dataset, which exists in the base R **datasets** package as **iris**, is ubiquitous in statistics and data science education.
+Due to a lack of metadata, non-intuitive variables (e.g.
+Sepal.Width), its publication in (and inextricable connection with) Fisher's 1936 paper in *Annals of Eugenics*, and perfect structure (e.g.
+no missing values) unrepresentative of most real-world environmental data, a modern alternative to **iris** is needed.
 Penguin morphological measurements for three species of *Pygoscelis* penguins that breed on islands throughout the Palmer Archipelago, Antarctica, provide an approachable, charismatic and near drop-in replacement for the **iris** data.
@@
 Anderson's *Iris* dataset is built into a number of software packages including the auto-installed datasets package in R [@r_core_team_r_2019], Python's scikit-learn machine learning library [@pedregosa_scikit-learn_2011], and the SAS Sashelp library (SAS Institute, Cary NC), which have facilitated widespread use of the data.
-As a result, eighty-five years after the data were initially published, Anderson’s *Iris* measurements are ubiquitous in statistics and data science course materials and tutorials.
-**Sounds great.
-Why consider replacing iris?**
+As a result, eighty-five years after the data were initially published, Anderson’s *Iris* measurements are ubiquitous in statistics, computational methods, and data science courses and materials.
+
+There are several reasons that practitioners and educators may want to retire Anderson's *Iris* dataset in modern data science materials, including:
+
+- Anderson's  *Iris* dataset entirely lacks metadata, limiting meaningful interpretation and discussion of research methods, analyses, and outcomes
+- Two of its five variables, `Sepal.Width` and `Sepal.Length`, are not intuitive to most users
+- In 1936, the *Iris* data were published by eugenicist R.A.
+Fisher in *Annals of Eugenics*, and are now inextricably linked to Fisher's eugenics work; in fact, the dataset is commonly referred to as *Fisher's Iris data*
+- The *Iris* dataset contains only complete cases (i.e.
+no missing values) and equal sample sizes (*n* = 50 for each *Iris* species) - more perfect than real-world datasets learners are likely to encounter "in the wild"
+- The single categorical variable (`Species`) limits options for analyses
+
+<!---**Sounds great. Why consider replacing iris?**
 
@@
 In the R community, there are calls to (1) address the *Iris* dataset's use in Fisher’s eugenics work, and/or (2) consider using a different dataset [@poisot_timothee_its_2020, @garrick_aden-buie_lets_2020].
-This paper describes (2): a suitable alternative dataset for educators wanting to replace the *Iris* data in their teaching materials.
+--->
 
-Considering *Iris* data usage in data science and statistics materials, we established the following criteria for a suitable replacement dataset:
+Given these motivations to retire *Iris*, we sought to identify, then make readily available, a suitable alternative dataset.
+Considering *Iris* usage in data science and statistics materials, we established the following criteria for a suitable alternative:
 
-- Available by appropriate license
-- Features intuitive subjects and variables understandable to learners across fields
+- Available by appropriate license (ideally, CC0 "no rights reserved")
+- Features intuitive subjects and variables that are interesting and understandable to learners across fields
+- Complete metadata and documentation
 - Manageable (but not trivial) in size
 - Minimal data cleaning and pre-processing required for most analyses
-- Real-world (not manufactured) data
+- Real-world (not manufactured), modern data
 - Similar opportunities for teaching and learning R, data science, and statistical skills
+- Can easily replace *Iris* for most use cases
 
-Here, we describe an alternative to Anderson's *Iris* data that largely satisfies these criteria: an approachable and charismatic dataset containing real-world morphological data for three *Pygoscelis* penguin species that breed throughout the Western Antarctic Peninsula region, made available through the Long-Term Ecological Research Network (US LTER).
-By comparing data structure, size, and a range of analyses side-by-side for the two datasets, we demonstrate that the Palmer Archipelago penguin measurements can replace Anderson's *Iris* data for many use cases in statistics and data science education.
+Here, we describe an alternative to Anderson's *Iris* data that largely satisfies these criteria: a refreshing, approachable and charismatic dataset containing real-world morphological data for three *Pygoscelis* penguin species that breed throughout the Western Antarctic Peninsula region, made available through the Long-Term Ecological Research Network (US LTER).
+By comparing data structure, size, and a range of analyses side-by-side for the two datasets, we demonstrate that the Palmer Archipelago penguin measurements are an ideal substitute for Anderson's *Iris* data for many use cases in statistics and data science education.
 
@@
 
-## Other alternatives to iris
+## Conclusion
 
-## Discussion
-
-Educators have three options regarding **iris** use in course materials:
-
-1. Use **iris** without addressing its use in Fisher's eugenics research
-2. Use **iris** *and* address its use in Fisher's eugenics research
-3. Replace **iris** with an alternative dataset offering similar teaching and learning opportunities
-
-Here, we have shown that structural size measurements for Palmer Archipelago *Pygoscelis* penguins, available as **penguins** in the **palmerpenguins** R package, can replace **iris** for a number of common use cases in data science and statistics education including exploratory data visualization, linear correlation and regression, PCA, and clustering by k-means.
+Here, we have shown that structural size measurements for Palmer Archipelago *Pygoscelis* penguins, available as **penguins** in the **palmerpenguins** R package, offer a near drop-in replacement for **iris** in a number of common use cases for data science and statistics education including exploratory data visualization, linear correlation and regression, PCA, and clustering by k-means.
 In addition, teaching and learning opportunities in **penguins** are increased due to a greater number of variables, missing values, unequal sample sizes, and Simpson's paradox examples.
-
-The **iris** dataset is widespread today because it was employed by Fisher to advance statistical methods for eugenics research.
-While the *Irises* were described in Anderson's 1935 article in *The Bulletin of the American Iris Society* [@anderson_irises_1935], the measurements were published in full in Fisher's 1936 *Annals of Eugenics* publication [@fisher_use_1936] - a journal for which Fisher served as editor before becoming chair of the Department of Eugenics at University College London.
-
-## Conclusion
 

### 2020-09-12 penguins.tex
update intro

--- 2020-08-05 penguins.tex
@@
 \abstract{%
-The famous \emph{Iris} dataset, which exists in the base R
-\textbf{datasets} package as \textbf{iris}, is ubiquitous in statistics
-and data science education. While the \textbf{iris} measurements were
-originally collected by botanist Edgar Anderson, their use in Ronald A.
-Fisher's eugenics research has created momentum to address the dataset's
-racist use and/or replace \textbf{iris} with alternate data in teaching
-materials. Penguin morphological measurements for three species of
-\emph{Pygoscelis} penguins that breed on islands throughout the Palmer
-Archipelago, Antarctica, provide an approachable, charismatic and near
-drop-in replacement for the \textbf{iris} data. Here, we introduce the
-\textbf{palmerpenguins} R package and describe the included
-\textbf{penguins} data. We directly compare the two datasets for
-selected analyses to demonstrate that R users, in particular teachers
-and learners currently using the \textbf{iris} data, can switch to the
-Palmer Archipelago penguins for many use cases including data wrangling
-and visualization, correlation, regression, hypothesis testing,
-multivariate analysis (e.g.~PCA), cluster analysis and classification
-(e.g.~by k-means).
+Edgar Anderson's famous 1935 \emph{Iris} dataset, which exists in the
+base R \textbf{datasets} package as \textbf{iris}, is ubiquitous in
+statistics and data science education. Due to a lack of metadata,
+non-intuitive variables (e.g.~Sepal.Width), its publication in (and
+inextricable connection with) Fisher's 1936 paper in \emph{Annals of
+Eugenics}, and perfect structure (e.g.~no missing values)
+unrepresentative of most real-world environmental data, a modern
+alternative to \textbf{iris} is needed. Penguin morphological
+measurements for three species of \emph{Pygoscelis} penguins that breed
+on islands throughout the Palmer Archipelago, Antarctica, provide an
+approachable, charismatic and near drop-in replacement for the
+\textbf{iris} data. Here, we introduce the \textbf{palmerpenguins} R
+package and describe the included \textbf{penguins} data. We directly
+compare the two datasets for selected analyses to demonstrate that R
+users, in particular teachers and learners currently using the
+\textbf{iris} data, can switch to the Palmer Archipelago penguins for
+many use cases including data wrangling and visualization, correlation,
+regression, hypothesis testing, multivariate analysis (e.g.~PCA),
+cluster analysis and classification (e.g.~by k-means).
 }
@@
 As a result, eighty-five years after the data were initially published,
-Anderson's \emph{Iris} measurements are ubiquitous in statistics and
-data science course materials and tutorials. \textbf{Sounds great. Why
-consider replacing iris?}
+Anderson's \emph{Iris} measurements are ubiquitous in statistics,
+computational methods, and data science courses and materials.
 
-In 1936, one year after being described in the Bulletin of the American
-Iris Society, Anderson's \emph{Iris} measurements were published in full
-by eugenicist and statistician Ronald A. Fisher in \emph{Annals of
-Eugenics} \citep{fisher_use_1936}. Regardless of Anderson's initial
-motivation, the data today remain inextricably linked to Fisher's
-eugenics research and are even commonly, if unfairly, referred to as
-``Fisher's iris data.'' For example, an August 2020 search for ``Iris
-flower data set'' in Wikipedia returns an article beginning with ``The
-iris flower data set or Fisher's iris data\ldots{}'', then later in the
-article: ``It is sometimes called Anderson's Iris data set''
-\citep{wikipedia_iris_2020}. The \emph{Iris} dataset is similarly
-credited to Fisher in statistical computing literature
-\citep[\citet{wang_matlab_2015}, \citet{woods_how_2015},
-\citet{chen_unsupervised_2018}]{trendafilov_simple_2009}.
-
-There is recent momentum to address Fisher's racist legacy in
-statistics. In June 2020, the Committee of Presidents of Statistical
-Societies' R.A. Fisher Award and Lectureship was replaced by the
-Distinguished Achievement Award and Lectureship, to ``advance a more
-just, equitable, diverse, and inclusive statistical community''
-\citep{noauthor_institute_2020}. Cambridge University's Gonville and
-Caius College recently announced plans to remove a stained-glass window
-celebrating Fisher from a campus dining hall \citep{noauthor_sir_2020}.
-In the R community, there are calls to (1) address the \emph{Iris}
-dataset's use in Fisher's eugenics work, and/or (2) consider using a
-different dataset
-\citep[\citet{garrick_aden-buie_lets_2020}]{poisot_timothee_its_2020}.
-This paper describes (2): a suitable alternative dataset for educators
-wanting to replace the \emph{Iris} data in their teaching materials.
-
-Considering \emph{Iris} data usage in data science and statistics
-materials, we established the following criteria for a suitable
-replacement dataset:
+There are several reasons that practitioners and educators may want to
+retire Anderson's \emph{Iris} dataset in modern data science materials,
+including:
 
@@
 \item
-  Available by appropriate license
+  Anderson's \emph{Iris} dataset entirely lacks metadata, limiting
+  meaningful interpretation and discussion of research methods,
+  analyses, and outcomes
 \item
-  Features intuitive subjects and variables understandable to learners
-  across fields
+  Two of its five variables, \texttt{Sepal.Width} and
+  \texttt{Sepal.Length}, are not intuitive to most users
+\item
+  In 1936, the \emph{Iris} data were published by eugenicist R.A. Fisher
+  in \emph{Annals of Eugenics}, and are now inextricably linked to
+  Fisher's eugenics work; in fact, the dataset is commonly referred to
+  as \emph{Fisher's Iris data}
+\item
+  The \emph{Iris} dataset contains only complete cases (i.e.~no missing
+  values) and equal sample sizes (\emph{n} = 50 for each \emph{Iris}
+  species) - more perfect than real-world datasets learners are likely
+  to encounter ``in the wild''
+\item
+  The single categorical variable (\texttt{Species}) limits options for
+  analyses
+\end{itemize}
+
+Given these motivations to retire \emph{Iris}, we sought to identify,
+then make readily available, a suitable alternative dataset. Considering
+\emph{Iris} usage in data science and statistics materials, we
+established the following criteria for a suitable alternative:
+
+\begin{itemize}
+\tightlist
+\item
+  Available by appropriate license (ideally, CC0 ``no rights reserved'')
+\item
+  Features intuitive subjects and variables that are interesting and
+  understandable to learners across fields
+\item
+  Complete metadata and documentation
 \item
@@
 \item
-  Real-world (not manufactured) data
+  Real-world (not manufactured), modern data
 \item
@@
   statistical skills
+\item
+  Can easily replace \emph{Iris} for most use cases
 \end{itemize}
@@
 Here, we describe an alternative to Anderson's \emph{Iris} data that
-largely satisfies these criteria: an approachable and charismatic
-dataset containing real-world morphological data for three
+largely satisfies these criteria: a refreshing, approachable and
+charismatic dataset containing real-world morphological data for three
 \emph{Pygoscelis} penguin species that breed throughout the Western
@@
 size, and a range of analyses side-by-side for the two datasets, we
-demonstrate that the Palmer Archipelago penguin measurements can replace
-Anderson's \emph{Iris} data for many use cases in statistics and data
-science education.
+demonstrate that the Palmer Archipelago penguin measurements are an
+ideal substitute for Anderson's \emph{Iris} data for many use cases in
+statistics and data science education.
 
@@
 \hline
-1 & NA & 9 & 116 & 1 & NA & 48 & 4\\
+1 & 4 & 54 & 6 & 1 & 50 & NA & NA\\
 \hline
-2 & 4 & 54 & 6 & 2 & NA & 2 & 46\\
+2 & 147 & 5 & 1 & 2 & NA & 48 & 4\\
 \hline
-3 & 147 & 5 & 1 & 3 & 50 & NA & NA\\
+3 & NA & 9 & 116 & 3 & NA & 2 & 46\\
 \hline
@@
 
-\hypertarget{other-alternatives-to-iris}{%
-\subsection{Other alternatives to
-iris}\label{other-alternatives-to-iris}}
-
-\hypertarget{discussion}{%
-\subsection{Discussion}\label{discussion}}
-
-Educators have three options regarding \textbf{iris} use in course
-materials:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Use \textbf{iris} without addressing its use in Fisher's eugenics
-  research
-\item
-  Use \textbf{iris} \emph{and} address its use in Fisher's eugenics
-  research
-\item
-  Replace \textbf{iris} with an alternative dataset offering similar
-  teaching and learning opportunities
-\end{enumerate}
+\hypertarget{conclusion}{%
+\subsection{Conclusion}\label{conclusion}}
 
@@
 Archipelago \emph{Pygoscelis} penguins, available as \textbf{penguins}
-in the \textbf{palmerpenguins} R package, can replace \textbf{iris} for
-a number of common use cases in data science and statistics education
-including exploratory data visualization, linear correlation and
-regression, PCA, and clustering by k-means. In addition, teaching and
-learning opportunities in \textbf{penguins} are increased due to a
-greater number of variables, missing values, unequal sample sizes, and
-Simpson's paradox examples.
-
-The \textbf{iris} dataset is widespread today because it was employed by
-Fisher to advance statistical methods for eugenics research. While the
-\emph{Irises} were described in Anderson's 1935 article in \emph{The
-Bulletin of the American Iris Society} \citep{anderson_irises_1935}, the
-measurements were published in full in Fisher's 1936 \emph{Annals of
-Eugenics} publication \citep{fisher_use_1936} - a journal for which
-Fisher served as editor before becoming chair of the Department of
-Eugenics at University College London.
-
-\hypertarget{conclusion}{%
-\subsection{Conclusion}\label{conclusion}}
+in the \textbf{palmerpenguins} R package, offer a near drop-in
+replacement for \textbf{iris} in a number of common use cases for data
+science and statistics education including exploratory data
+visualization, linear correlation and regression, PCA, and clustering by
+k-means. In addition, teaching and learning opportunities in
+\textbf{penguins} are increased due to a greater number of variables,
+missing values, unequal sample sizes, and Simpson's paradox examples.
 

### 2020-09-13 penguins.Rmd
minor intro updates

@@
 In 1935, American botanist Edgar Anderson measured petal and sepal structural dimensions (length and width) for 50 flowers from three Iris species: *Iris setosa*, *Iris versicolor*, and *Iris virginica* [@anderson_irises_1935].
-The manageable but non-trivial size (5 variables and 150 total observations) and characteristics of the data make it amenable for introducing a wide range of statistical methods including data wrangling, visualization, regression, multivariate exploration, and machine learning.
+The manageable but non-trivial size (5 variables and 150 total observations) and characteristics of the data, including linear relationships and multivariate normality, make it amenable for introducing a wide range of statistical methods including data wrangling, visualization, regression, multivariate exploration, and machine learning.
 Anderson's *Iris* dataset is built into a number of software packages including the auto-installed datasets package in R [@r_core_team_r_2019], Python's scikit-learn machine learning library [@pedregosa_scikit-learn_2011], and the SAS Sashelp library (SAS Institute, Cary NC), which have facilitated widespread use of the data.
@@
 
-There are several reasons that practitioners and educators may want to retire Anderson's *Iris* dataset in modern data science materials, including:
-
-- Anderson's  *Iris* dataset entirely lacks metadata, limiting meaningful interpretation and discussion of research methods, analyses, and outcomes
-- Two of its five variables, `Sepal.Width` and `Sepal.Length`, are not intuitive to most users
-- In 1936, the *Iris* data were published by eugenicist R.A.
-Fisher in *Annals of Eugenics*, and are now inextricably linked to Fisher's eugenics work; in fact, the dataset is commonly referred to as *Fisher's Iris data*
-- The *Iris* dataset contains only complete cases (i.e.
-no missing values) and equal sample sizes (*n* = 50 for each *Iris* species) - more perfect than real-world datasets learners are likely to encounter "in the wild"
-- The single categorical variable (`Species`) limits options for analyses
+There are a number of reasons that practitioners and educators may want to retire Anderson's *Iris* dataset in their modern data science materials.
+Anderson's *Iris* dataset lacks metadata almost entirely [@anderson_irises_1935], limiting meaningful interpretation and discussion of research methods, analyses, and outcomes.
+Of five variables in the dataset, two (`Sepal.Width` and `Sepal.Length`) are unintuitive for most non-botanists, and even with explanation the difference between *Iris* *petal* and *sepal* dimensions is not obvious.
+The *Iris* dataset contains only complete cases (there are no missing values) and equal sample size for all 3 species (*n* = 50 for all), which is cleaner than most real-world datasets learners are likely to encounter "in the wild." The single factor (`Species`) in the *Iris* data limits options for analyses.
+Finally, due to their publication in (and inextricable connection to) eugenicist and statistician R.A.
+Fisher's 1936 paper in *Annals of Eugenics* [@fisher_use_1936], the *Iris* dataset is burdened by a history in racist research.
 
@@
 Given these motivations to retire *Iris*, we sought to identify, then make readily available, a suitable alternative dataset.
-Considering *Iris* usage in data science and statistics materials, we established the following criteria for a suitable alternative:
+Considering the positive and negative characteristics of the *Iris* dataset in data science and statistics materials, we established the following criteria for a suitable alternative:
 

### 2020-09-13 penguins.tex
minor intro updates

@@
 size (5 variables and 150 total observations) and characteristics of the
-data make it amenable for introducing a wide range of statistical
-methods including data wrangling, visualization, regression,
-multivariate exploration, and machine learning. Anderson's \emph{Iris}
-dataset is built into a number of software packages including the
-auto-installed datasets package in R \citep{r_core_team_r_2019},
-Python's scikit-learn machine learning library
-\citep{pedregosa_scikit-learn_2011}, and the SAS Sashelp library (SAS
-Institute, Cary NC), which have facilitated widespread use of the data.
-As a result, eighty-five years after the data were initially published,
-Anderson's \emph{Iris} measurements are ubiquitous in statistics,
-computational methods, and data science courses and materials.
+data, including linear relationships and multivariate normality, make it
+amenable for introducing a wide range of statistical methods including
+data wrangling, visualization, regression, multivariate exploration, and
+machine learning. Anderson's \emph{Iris} dataset is built into a number
+of software packages including the auto-installed datasets package in R
+\citep{r_core_team_r_2019}, Python's scikit-learn machine learning
+library \citep{pedregosa_scikit-learn_2011}, and the SAS Sashelp library
+(SAS Institute, Cary NC), which have facilitated widespread use of the
+data. As a result, eighty-five years after the data were initially
+published, Anderson's \emph{Iris} measurements are ubiquitous in
+statistics, computational methods, and data science courses and
+materials.
 
-There are several reasons that practitioners and educators may want to
-retire Anderson's \emph{Iris} dataset in modern data science materials,
-including:
-
-\begin{itemize}
-\tightlist
-\item
-  Anderson's \emph{Iris} dataset entirely lacks metadata, limiting
-  meaningful interpretation and discussion of research methods,
-  analyses, and outcomes
-\item
-  Two of its five variables, \texttt{Sepal.Width} and
-  \texttt{Sepal.Length}, are not intuitive to most users
-\item
-  In 1936, the \emph{Iris} data were published by eugenicist R.A. Fisher
-  in \emph{Annals of Eugenics}, and are now inextricably linked to
-  Fisher's eugenics work; in fact, the dataset is commonly referred to
-  as \emph{Fisher's Iris data}
-\item
-  The \emph{Iris} dataset contains only complete cases (i.e.~no missing
-  values) and equal sample sizes (\emph{n} = 50 for each \emph{Iris}
-  species) - more perfect than real-world datasets learners are likely
-  to encounter ``in the wild''
-\item
-  The single categorical variable (\texttt{Species}) limits options for
-  analyses
-\end{itemize}
+There are a number of reasons that practitioners and educators may want
+to retire Anderson's \emph{Iris} dataset in their modern data science
+materials. Anderson's \emph{Iris} dataset lacks metadata almost entirely
+\citep{anderson_irises_1935}, limiting meaningful interpretation and
+discussion of research methods, analyses, and outcomes. Of five
+variables in the dataset, two (\texttt{Sepal.Width} and
+\texttt{Sepal.Length}) are unintuitive for most non-botanists, and even
+with explanation the difference between \emph{Iris} \emph{petal} and
+\emph{sepal} dimensions is not obvious. The \emph{Iris} dataset contains
+only complete cases (there are no missing values) and equal sample size
+for all 3 species (\emph{n} = 50 for all), which is cleaner than most
+real-world datasets learners are likely to encounter ``in the wild.''
+The single factor (\texttt{Species}) in the \emph{Iris} data limits
+options for analyses. Finally, due to their publication in (and
+inextricable connection to) eugenicist and statistician R.A. Fisher's
+1936 paper in \emph{Annals of Eugenics} \citep{fisher_use_1936}, the
+\emph{Iris} dataset is burdened by a history in racist research.
 
@@
 then make readily available, a suitable alternative dataset. Considering
-\emph{Iris} usage in data science and statistics materials, we
-established the following criteria for a suitable alternative:
+the positive and negative characteristics of the \emph{Iris} dataset in
+data science and statistics materials, we established the following
+criteria for a suitable alternative:
 
@@
 \hline
-1 & 4 & 54 & 6 & 1 & 50 & NA & NA\\
+1 & NA & 9 & 116 & 1 & NA & 48 & 4\\
 \hline
-2 & 147 & 5 & 1 & 2 & NA & 48 & 4\\
+2 & 4 & 54 & 6 & 2 & 50 & NA & NA\\
 \hline
-3 & NA & 9 & 116 & 3 & NA & 2 & 46\\
+3 & 147 & 5 & 1 & 3 & NA & 2 & 46\\
 \hline

### 2021-01-27 penguins.Rmd
ah updates

@@
 
- Edgar Anderson's famous 1935 *Iris* dataset, which exists in the base R **datasets** package as **iris**, is ubiquitous in statistics and data science education.
+ Edgar Anderson's famous **iris** dataset, which contains size measurements for three species of *Iris* collected in 1935, exists in the base R **datasets** package as **iris** and is ubiquitous in statistics and data science materials.
 Due to a lack of metadata, non-intuitive variables (e.g.
@@
 
+### Overview
+
+```{r, echo = FALSE}
+overview_compare <- tribble(
+  ~Feature, ~iris, ~penguins,
+  "Year(s) collected", "1935", "2007 - 2009",
+  "Dimensions (col x row)", "5 x 150", "8 x 344",
+  "Documentation", "minimal", "complete metadata",
+  "Variable classes", "double (4), factor (1)", "double (2), int (3), factor (3)",
+  "Missing values?", "no (n = 0; 0.0%)", "yes (n = 19; 0.7%)"
+)
+
+overview_compare %>% 
+  kable(caption = "Overview comparison of penguins and iris dataset features and characteristics.")
+```
+
 ### Data structure and sample size
 
-Both **iris** and **penguins** are in tidy format [ref] with each column denoting a single variable and each row containing measurements for a single *Iris* flower or penguin.
+Both **iris** and **penguins** are in tidy format [@wickham_tidy_2014] with each column denoting a single variable and each row containing measurements for a single *Iris* flower or penguin.
 The two datasets are comparable in size: dimensions (columns × rows) are 5 × 150 and 8 × 344 for **iris** and **penguins**, respectively, and sample sizes within species are similar (Table 1).

### 2021-01-27 penguins.tex
ah updates

@@
 % !TeX root = RJwrapper.tex
-\title{Palmer Archipelago Penguins Data in the palmerpenguins R Package - An
-Alternative to Anderson's \emph{Irises}}
+\title{Palmer Archipelago Penguins Data in the palmerpenguins R Package
+- An Alternative to Anderson's \emph{Irises}}
 \author{by Allison M. Horst, Alison P. Hill, Kristen B. Gorman}
@@
 \abstract{%
-Edgar Anderson's famous 1935 \emph{Iris} dataset, which exists in the
-base R \textbf{datasets} package as \textbf{iris}, is ubiquitous in
-statistics and data science education. Due to a lack of metadata,
-non-intuitive variables (e.g.~Sepal.Width), its publication in (and
-inextricable connection with) Fisher's 1936 paper in \emph{Annals of
-Eugenics}, and perfect structure (e.g.~no missing values)
+Edgar Anderson's famous \textbf{iris} dataset, which contains size
+measurements for three species of \emph{Iris} collected in 1935, exists
+in the base R \textbf{datasets} package as \textbf{iris} and is
+ubiquitous in statistics and data science materials. Due to a lack of
+metadata, non-intuitive variables (e.g.~Sepal.Width), its publication in
+(and inextricable connection with) Fisher's 1936 paper in \emph{Annals
+of Eugenics}, and perfect structure (e.g.~no missing values)
 unrepresentative of most real-world environmental data, a modern
@@
 
+\hypertarget{overview}{%
+\subsubsection{Overview}\label{overview}}
+
+\begin{Schunk}
+\begin{table}
+
+\caption{\label{tab:unnamed-chunk-1}Overview comparison of penguins and iris dataset features and characteristics.}
+\centering
+\begin{tabular}[t]{l|l|l}
+\hline
+Feature & iris & penguins\\
+\hline
+Year(s) collected & 1935 & 2007 - 2009\\
+\hline
+Dimensions (col x row) & 5 x 150 & 8 x 344\\
+\hline
+Documentation & minimal & complete metadata\\
+\hline
+Variable classes & double (4), factor (1) & double (2), int (3), factor (3)\\
+\hline
+Missing values? & no (n = 0; 0.0\%) & yes (n = 19; 0.7\%)\\
+\hline
+\end{tabular}
+\end{table}
+
+\end{Schunk}
+
 \hypertarget{data-structure-and-sample-size}{%
@@
 
-Both \textbf{iris} and \textbf{penguins} are in tidy format {[}ref{]}
-with each column denoting a single variable and each row containing
-measurements for a single \emph{Iris} flower or penguin. The two
-datasets are comparable in size: dimensions (columns × rows) are 5 × 150
-and 8 × 344 for \textbf{iris} and \textbf{penguins}, respectively, and
-sample sizes within species are similar (Table 1). Notably, sample sizes
-differ for the three penguin species, while sample sizes in
-\textbf{iris} are equal (n = 50 for each Iris species). Multiple factor
-variables in \textbf{penguins} (\emph{species}, \emph{island} and
-\emph{sex}) along with \emph{year} create additional opportunities for
-grouping, compared to the single factor (\emph{species}) in
-\textbf{iris}.
+Both \textbf{iris} and \textbf{penguins} are in tidy format
+\citep{wickham_tidy_2014} with each column denoting a single variable
+and each row containing measurements for a single \emph{Iris} flower or
+penguin. The two datasets are comparable in size: dimensions (columns ×
+rows) are 5 × 150 and 8 × 344 for \textbf{iris} and \textbf{penguins},
+respectively, and sample sizes within species are similar (Table 1).
+Notably, sample sizes differ for the three penguin species, while sample
+sizes in \textbf{iris} are equal (n = 50 for each Iris species).
+Multiple factor variables in \textbf{penguins} (\emph{species},
+\emph{island} and \emph{sex}) along with \emph{year} create additional
+opportunities for grouping, compared to the single factor
+(\emph{species}) in \textbf{iris}.
 
@@
 
-\caption{\label{tab:unnamed-chunk-1}Grouped sample size for iris (by species; n = 150 total) and penguins (by species and sex; n = 344 total).
+\caption{\label{tab:unnamed-chunk-2}Grouped sample size for iris (by species; n = 150 total) and penguins (by species and sex; n = 344 total).
 Penguins can be further grouped by variables for island and study year.}
@@
 
-\caption{\label{tab:unnamed-chunk-3}K-means cluster assignments by species based on penguin bill length (mm) and depth (mm), and iris petal length (cm) and width (cm).}
+\caption{\label{tab:unnamed-chunk-4}K-means cluster assignments by species based on penguin bill length (mm) and depth (mm), and iris petal length (cm) and width (cm).}
 \centering
@@
 \hline
-1 & NA & 9 & 116 & 1 & NA & 48 & 4\\
+1 & 147 & 5 & 1 & 1 & NA & 2 & 46\\
 \hline
-2 & 4 & 54 & 6 & 2 & 50 & NA & NA\\
+2 & NA & 9 & 116 & 2 & NA & 48 & 4\\
 \hline
-3 & 147 & 5 & 1 & 3 & NA & 2 & 46\\
+3 & 4 & 54 & 6 & 3 & 50 & NA & NA\\
 \hline

### 2021-03-22 penguins.Rmd

@@
     email:  ahorst@ucsb.edu
-  - name: Alison P. Hill
+  - name: Alison Presmanes Hill
     affiliation: RStudio, PBC
@@
 
- Edgar Anderson's famous **iris** dataset, which contains size measurements for three species of *Iris* collected in 1935, exists in the base R **datasets** package as **iris** and is ubiquitous in statistics and data science materials.
-Due to a lack of metadata, non-intuitive variables (e.g.
-Sepal.Width), its publication in (and inextricable connection with) Fisher's 1936 paper in *Annals of Eugenics*, and perfect structure (e.g.
-no missing values) unrepresentative of most real-world environmental data, a modern alternative to **iris** is needed.
-Penguin morphological measurements for three species of *Pygoscelis* penguins that breed on islands throughout the Palmer Archipelago, Antarctica, provide an approachable, charismatic and near drop-in replacement for the **iris** data.
-Here, we introduce the **palmerpenguins** R package and describe the included **penguins** data.
-We directly compare the two datasets for selected analyses to demonstrate that R users, in particular teachers and learners currently using the **iris** data, can switch to the Palmer Archipelago penguins for many use cases including data wrangling and visualization, correlation, regression, hypothesis testing, multivariate analysis (e.g.
-PCA), cluster analysis and classification (e.g.
-by k-means).
+ In 1935, Edgar Anderson collected size measurements for 150 flowers from three different species of Iris on the Gaspé Peninsula.
+Since then, Anderson's *Iris* dataset has become a classic in statistics, machine learning, and data science teaching materials.
+It is included in the base R **datasets** package as **iris**, making it easy for users to access without knowing much about it.
+Most notably, Anderson's *Iris* dataset appeared in a 1936 publication by R.
+A.
+Fisher in the *Annals of Eugenics*, linking it to a field of research rooted in racism and prejudice.
+Combined with the lack of dataset documentation, the presence of non-intuitive variables like "sepal width", and the perfectly balanced groups with zero missing values, *Iris* is an unsatisfactory dataset for teaching and illustrating concepts.
+Users would benefit from working with a more representative, real-world environmental dataset---one with a clean history and a clear link to modern scientific research.
+A modern alternative to the *Iris* dataset is needed.
+Here, we introduce the **palmerpenguins** R package, which includes morphological measurements for three species of *Pygoscelis* penguins that breed on islands throughout the Palmer Archipelago, Antarctica.
+The **penguins** dataset provides an approachable, charismatic, and near drop-in replacement for the **iris** data.
+We directly compare the two datasets for selected analyses to demonstrate that R users, in particular teachers and learners currently using the **iris** data, can switch to the Palmer Archipelago penguins for many use cases including data wrangling and visualization, correlation, regression, hypothesis testing, multivariate analysis (e.g., PCA), cluster analysis and classification (e.g., by k-means).
 
@@
 In 1935, American botanist Edgar Anderson measured petal and sepal structural dimensions (length and width) for 50 flowers from three Iris species: *Iris setosa*, *Iris versicolor*, and *Iris virginica* [@anderson_irises_1935].
-The manageable but non-trivial size (5 variables and 150 total observations) and characteristics of the data, including linear relationships and multivariate normality, make it amenable for introducing a wide range of statistical methods including data wrangling, visualization, regression, multivariate exploration, and machine learning.
+The manageable but non-trivial size (`r ncol(iris)` variables and `r nrow(iris)` total observations) and characteristics of the data, including linear relationships and multivariate normality, make it amenable for introducing a wide range of statistical methods including data wrangling, visualization, regression, multivariate exploration, and machine learning.
 Anderson's *Iris* dataset is built into a number of software packages including the auto-installed datasets package in R [@r_core_team_r_2019], Python's scikit-learn machine learning library [@pedregosa_scikit-learn_2011], and the SAS Sashelp library (SAS Institute, Cary NC), which have facilitated widespread use of the data.
-As a result, eighty-five years after the data were initially published, Anderson’s *Iris* measurements are ubiquitous in statistics, computational methods, and data science courses and materials.
+As a result, eighty-five years after the data were initially published, Anderson’s *Iris* measurements are ubiquitous in statistics, computational methods, software documentation, and data science courses and materials.
 
-There are a number of reasons that practitioners and educators may want to retire Anderson's *Iris* dataset in their modern data science materials.
-Anderson's *Iris* dataset lacks metadata almost entirely [@anderson_irises_1935], limiting meaningful interpretation and discussion of research methods, analyses, and outcomes.
-Of five variables in the dataset, two (`Sepal.Width` and `Sepal.Length`) are unintuitive for most non-botanists, and even with explanation the difference between *Iris* *petal* and *sepal* dimensions is not obvious.
-The *Iris* dataset contains only complete cases (there are no missing values) and equal sample size for all 3 species (*n* = 50 for all), which is cleaner than most real-world datasets learners are likely to encounter "in the wild." The single factor (`Species`) in the *Iris* data limits options for analyses.
-Finally, due to their publication in (and inextricable connection to) eugenicist and statistician R.A.
-Fisher's 1936 paper in *Annals of Eugenics* [@fisher_use_1936], the *Iris* dataset is burdened by a history in racist research.
+There are a number of reasons that modern data science practitioners and educators may want to look beyond Anderson's *Iris* dataset.
+First, Anderson's *Iris* dataset lacks metadata almost entirely [@anderson_irises_1935], limiting meaningful interpretation and discussion of research methods, analyses, and outcomes.
+Of five variables in the dataset, two (`Sepal.Width` and `Sepal.Length`) are not intuitive for most non-botanists.
+Even with explanation, the difference between *Iris* *petal* and *sepal* dimensions is not obvious.
+Second, the *Iris* dataset contains no missing values and equal sample sizes for each of the three species (*n* = 50), which is cleaner than most real-world datasets learners are likely to encounter "in the wild." Third, the single factor (`Species`) in the *Iris* data limits options for analyses.
+Finally, due to its direct link to a publication in the *Annals of Eugenics* by eugenicist and statistician R.A.
+Fisher's [@fisher_use_1936], the *Iris* dataset is burdened by a history in racist research.
 
@@
 
-Given these motivations to retire *Iris*, we sought to identify, then make readily available, a suitable alternative dataset.
-Considering the positive and negative characteristics of the *Iris* dataset in data science and statistics materials, we established the following criteria for a suitable alternative:
+Given the need for fresh data science-ready datasets, we wanted to find a suitable alternative dataset that we could make easy to access for everyone.
+After evaluating the positive and negative features of the *Iris* dataset in data science and statistics materials, we established the following criteria for a suitable alternative:
 
@@
 
-Here, we describe an alternative to Anderson's *Iris* data that largely satisfies these criteria: a refreshing, approachable and charismatic dataset containing real-world morphological data for three *Pygoscelis* penguin species that breed throughout the Western Antarctic Peninsula region, made available through the Long-Term Ecological Research Network (US LTER).
+Here, we describe an alternative to Anderson's *Iris* data that largely satisfies these criteria: a refreshing, approachable, and charismatic dataset containing real-world morphological data for three *Pygoscelis* penguin species that breed throughout the Western Antarctic Peninsula region, made available through the Long-Term Ecological Research Network (US LTER).
 By comparing data structure, size, and a range of analyses side-by-side for the two datasets, we demonstrate that the Palmer Archipelago penguin measurements are an ideal substitute for Anderson's *Iris* data for many use cases in statistics and data science education.
@@
 Kristen Gorman in collaboration with the [Palmer Station LTER](https://pal.lternet.edu/), part of the [US LTER Network](https://lternet.edu/).
-For complete data collection methods and published analyses see Gorman et al.
+For complete data collection methods and published analyses, see Gorman et al.
 (2014).
@@
 
-R users can install the **palmerpenguins** package from CRAN by:
+R users can install the **palmerpenguins** package from CRAN:
 
-      `install.packages("palmerpenguins")`
-
-Alternatively, the development version of the package can be installed from GitHub:
-
-      `remotes::install_github("allisonhorst/palmerpenguins")`
+    install.packages("palmerpenguins")
 
@@
 The **palmerpenguins** R package contains two data objects: **penguins_raw** and **penguins**.
-The **penguins_raw** data consists of all raw data for 17 variables (Appendix Table 1), recorded completely or in part for 344 individual penguins, accessed directly from EDI.
-As a direct alternative to Anderson's *Iris* data we recommend using the curated data in **penguins**, which is a subset of **penguins_raw** retaining all 344 observations, minimally updated (Appendix B) and reduced to the following eight variables:
+The **penguins_raw** data consists of all raw data for `r ncol(palmerpenguins::penguins_raw)` variables (Appendix Table 1), recorded completely or in part for `r nrow(palmerpenguins::penguins_raw)` individual penguins, accessed directly from EDI.
+We generally recommend using the curated data in **penguins**, which is a subset of **penguins_raw** retaining all `r nrow(palmerpenguins::penguins)` observations, minimally updated (Appendix B) and reduced to the following eight variables:
 
@@
 ```
-    library(tidyverse)
+    # install.packages("readr")
     library(palmerpenguins)
-    df <- read_csv(path_to_file(“penguins.csv”))
+    df <- readr::read_csv(path_to_file("penguins.csv"))
 ```
 
-will read in "penguins.csv" as if from an external file, thus automatically parsing *species*, *island* and *sex* variables as characters.
-This option allows users opportunities to practice or demonstrate reading in data from a CSV, then updating variable class (e.g.
-characters to factors).
+will read in "penguins.csv" as if from an external file, thus automatically parsing *species*, *island*, and *sex* variables as characters instead of factors.
+This option allows users opportunities to practice or demonstrate reading in data from a CSV, then updating variable class (e.g., characters to factors).
 
@@
 
-Both **iris** and **penguins** are in tidy format [@wickham_tidy_2014] with each column denoting a single variable and each row containing measurements for a single *Iris* flower or penguin.
-The two datasets are comparable in size: dimensions (columns × rows) are 5 × 150 and 8 × 344 for **iris** and **penguins**, respectively, and sample sizes within species are similar (Table 1).
-Notably, sample sizes differ for the three penguin species, while sample sizes in **iris** are equal (n = 50 for each Iris species).
-Multiple factor variables in **penguins** (*species*, *island* and *sex*) along with *year* create additional opportunities for grouping, compared to the single factor (*species*) in **iris**.
+Both **iris** and **penguins** are in tidy format [@wickham_tidy_2014] with each column denoting a single variable and each row containing measurements for a single *Iris* flower or penguin, respectively.
+The two datasets are comparable in size: dimensions (columns × rows) are `r ncol(iris)` × `r nrow(iris)` and `r ncol(penguins)` × `r nrow(penguins)` for **iris** and **penguins**, respectively, and sample sizes within species are similar (Table 1).
 
-Unlike **iris**, which contains only complete cases, the **penguins** dataset contains a small number of missing values (n~missing~ = 19, out of 2,752 total values).
+Notably, while sample sizes in **iris** across species are all the same, sample sizes in **penguins** differ across the three species.
+The inclusion of three factor variables in **penguins** (*species*, *island*, and *sex*), along with *year*, create additional opportunities for grouping and faceting, compared to the single factor (*Species*) in **iris**.
+
+Unlike **iris**, which contains only complete cases, the **penguins** dataset contains a small number of missing values (n~missing~ = 19, out of `r `ncol(palmerpenguins::penguins) * nrow(palmerpenguins::penguins)` total values).
 Missing values and unequal sample sizes are common in real-world data, and we believe add learning value to **penguins**.
@@
 
-Distributions, relationships between variables and clustering can be visually explored between species for the four structural size measurements in **penguins** (flipper length, body mass, bill length and depth; Figure \ref{fig:penguins-pairs}) and **iris** (sepal width and length, petal width and length; Figure \ref{fig:iris-pairs}).
+Distributions, relationships between variables, and clustering can be visually explored between species for the four structural size measurements in **penguins** (flipper length, body mass, bill length and depth; Figure \ref{fig:penguins-pairs}) and **iris** (sepal width and length, petal width and length; Figure \ref{fig:iris-pairs}).
 
@@
 
-There exist numerous opportunities to explore linear relationships and correlations in **penguins** and **iris** both within and across species (Figures \ref{fig:penguins-pairs} & \ref{fig:iris-pairs}).
-Here, we highlight one comparison that is uniquely similar: petal dimensions (petal length versus petal width) from **iris**, and penguin size (flipper length versus body mass) in **penguins** (\ref{fig:linear-example}).
+Both **penguins** and **iris** offer numerous opportunities to explore linear relationships and correlations, within and across species (Figures \ref{fig:penguins-pairs} & \ref{fig:iris-pairs}).
+A classic bivariate scatterplot made with the **iris** dataset reveals a clean linear relationship between petal length and petal width.
+Using **penguins** (\ref{fig:linear-example}), we can create a uniquely similar scatterplot with flipper length and body mass.
 The overall trend across all three species is approximately linear for both **iris** and **penguins**.
@@
 Principal component analysis (PCA) is a dimensional reduction method commonly used to explore patterns in multivariate data.
-PCA tutorials frequently employ the **iris** dataset, which is useful for teaching the method due to multivariate normality, and clear, approachable outcomes for variable loadings and clustering.
+The **iris** dataset frequently appears in PCA tutorials due to multivariate normality and clear interpretation of variable loadings and clustering.
 
@@
 
-Unsupervised clustering by k-means is a common and popular entryway to machine learning and classification, and frequently employs the **iris** data for introductory examples.
+Unsupervised clustering by k-means is a common and popular entryway to machine learning and classification, and again, the **iris** dataset is frequently used in introductory examples.
 The **penguins** data provides similar opportunities for introducing k-means clustering.
@@
 All variables are scaled prior to k-means.
-Three clusters (*k* = 3) are specified for each since there are three species of both *Iris* (*setosa*, *versicolor*, and *virginica*) and penguins (Adélie, chinstrap and gentoo).
+Three clusters (*k* = 3) are specified for each, since there are three species of both *Iris* (*setosa*, *versicolor*, and *virginica*) and penguins (Adélie, chinstrap and gentoo).
 
-K-means clustering with penguin bill dimensions and iris petal dimensions yields largely distinct clusters each dominated by one species (\ref{fig:kmeans}).
+K-means clustering with penguin bill dimensions and iris petal dimensions yields largely distinct clusters, each dominated by one species (\ref{fig:kmeans}).
 For iris petal dimensions, k-means yields a perfectly separated cluster (Cluster 1) containing all 50 setosa iris observations and zero misclassified virginica or versicolor irises (Table 2).

### 2021-03-24 penguins.Rmd
Update penguins.Rmd

@@
   - name: Kristen B. Gorman
-    affiliation: University of Alaska Fairbanks College of Fisheries and Ocean Sciences
+    affiliation: College of Fisheries and Ocean Sciences
     address:
+    - University of Alaska Fairbanks 
     - 2150 Koyukuk Drive
@@
 A.
-Fisher in the *Annals of Eugenics*, linking it to a field of research rooted in racism and prejudice.
+Fisher in the *Annals of Eugenics*, linking the dataset to a field of research rooted in racism and prejudice.
 Combined with the lack of dataset documentation, the presence of non-intuitive variables like "sepal width", and the perfectly balanced groups with zero missing values, *Iris* is an unsatisfactory dataset for teaching and illustrating concepts.
 Users would benefit from working with a more representative, real-world environmental dataset---one with a clean history and a clear link to modern scientific research.
-A modern alternative to the *Iris* dataset is needed.
+Thus, a modern alternative to the *Iris* dataset is needed.
 Here, we introduce the **palmerpenguins** R package, which includes morphological measurements for three species of *Pygoscelis* penguins that breed on islands throughout the Palmer Archipelago, Antarctica.
 The **penguins** dataset provides an approachable, charismatic, and near drop-in replacement for the **iris** data.
-We directly compare the two datasets for selected analyses to demonstrate that R users, in particular teachers and learners currently using the **iris** data, can switch to the Palmer Archipelago penguins for many use cases including data wrangling and visualization, correlation, regression, hypothesis testing, multivariate analysis (e.g., PCA), cluster analysis and classification (e.g., by k-means).
+We directly compare the two datasets for selected analyses to demonstrate that R users, in particular teachers and learners currently using the **iris** dataset, can switch to the Palmer Archipelago penguins for many use cases including data wrangling and visualization, correlation, regression, hypothesis testing, multivariate analysis (e.g., PCA), cluster analysis and classification (e.g., by k-means).
 
@@
 In 1935, American botanist Edgar Anderson measured petal and sepal structural dimensions (length and width) for 50 flowers from three Iris species: *Iris setosa*, *Iris versicolor*, and *Iris virginica* [@anderson_irises_1935].
-The manageable but non-trivial size (`r ncol(iris)` variables and `r nrow(iris)` total observations) and characteristics of the data, including linear relationships and multivariate normality, make it amenable for introducing a wide range of statistical methods including data wrangling, visualization, regression, multivariate exploration, and machine learning.
-Anderson's *Iris* dataset is built into a number of software packages including the auto-installed datasets package in R [@r_core_team_r_2019], Python's scikit-learn machine learning library [@pedregosa_scikit-learn_2011], and the SAS Sashelp library (SAS Institute, Cary NC), which have facilitated widespread use of the data.
+The manageable but non-trivial size (`r ncol(iris)` variables and `r nrow(iris)` total observations) and characteristics of the dataset, including linear relationships and multivariate normality, make it amenable for introducing a wide range of statistical methods including data wrangling, visualization, regression, multivariate exploration, and machine learning.
+Anderson's *Iris* dataset is built into a number of software packages including the auto-installed datasets package in R [@r_core_team_r_2019], Python's scikit-learn machine learning library [@pedregosa_scikit-learn_2011], and the SAS Sashelp library (SAS Institute, Cary NC), which have facilitated widespread use of the *Iris* data.
 As a result, eighty-five years after the data were initially published, Anderson’s *Iris* measurements are ubiquitous in statistics, computational methods, software documentation, and data science courses and materials.
@@
 
-Given the need for fresh data science-ready datasets, we wanted to find a suitable alternative dataset that we could make easy to access for everyone.
+Given the need for fresh data science-ready datasets, we sought to identify a suitable alternative dataset that could be made easily accessible for a broad audience.
 After evaluating the positive and negative features of the *Iris* dataset in data science and statistics materials, we established the following criteria for a suitable alternative:
@@
 - Available by appropriate license (ideally, CC0 "no rights reserved")
-- Features intuitive subjects and variables that are interesting and understandable to learners across fields
+- Features intuitive subjects and variables that are interesting and understandable to learners across disciplines
 - Complete metadata and documentation
@@
 - Real-world (not manufactured), modern data
-- Similar opportunities for teaching and learning R, data science, and statistical skills
+- Provides similar opportunities for teaching and learning R, data science, and statistical skills
 - Can easily replace *Iris* for most use cases
 
-Here, we describe an alternative to Anderson's *Iris* data that largely satisfies these criteria: a refreshing, approachable, and charismatic dataset containing real-world morphological data for three *Pygoscelis* penguin species that breed throughout the Western Antarctic Peninsula region, made available through the Long-Term Ecological Research Network (US LTER).
-By comparing data structure, size, and a range of analyses side-by-side for the two datasets, we demonstrate that the Palmer Archipelago penguin measurements are an ideal substitute for Anderson's *Iris* data for many use cases in statistics and data science education.
+Here, we describe an alternative to Anderson's *Iris* data that largely satisfies these criteria: a refreshing, approachable, and charismatic dataset containing real-world morphological data for three *Pygoscelis* penguin species that breed throughout the Western Antarctic Peninsula region, made available through the United States Long-Term Ecological Research (US LTER) Network.
+By comparing data structure, size, and a range of analyses side-by-side for the two datasets, we demonstrate that the Palmer Archipelago penguin data are an ideal substitute for Anderson's *Iris* dataset for many use cases in statistics and data science education.
 
@@
 
-Body size measurements, clutch (i.e., egg laying) observations (e.g., date of first egg laid, and clutch completion), and carbon (^13^C/^12^C, $\delta$^13^C) and nitrogen (^15^N/^14^N, $\delta$^15^N) stable isotope values of red blood cells for male and female adult Adélie (*P.
+Body size measurements, clutch (i.e., egg laying) observations (e.g., date of first egg laid, and clutch completion), and carbon (^13^C/^12^C, $\delta$^13^C) and nitrogen (^15^N/^14^N, $\delta$^15^N) stable isotope values of red blood cells for adult male and female Adélie (*P.
 adeliae*), chinstrap (*P.
 antarcticus*), and gentoo (*P.
-papua*) penguins on three islands (Biscoe, Dream and Torgersen) in the Palmer Archipelago were collected from 2007 - 2009 by Dr.
+papua*) penguins on three islands (Biscoe, Dream and Torgersen) within the Palmer Archipelago were collected from 2007 - 2009 by Dr.
 Kristen Gorman in collaboration with the [Palmer Station LTER](https://pal.lternet.edu/), part of the [US LTER Network](https://lternet.edu/).
@@
 - *species:* a factor denoting the penguin species (Adélie, chinstrap, or gentoo)
-- *island:* a factor denoting the island (in Palmer Archipelago, Antarctica) where observed (Biscoe, Dream or Torgersen)
+- *island:* a factor denoting the island (Palmer Archipelago, Antarctica) where penguins were observed (Biscoe Point, Dream Island or Torgersen Island)
 - *bill_length_mm:* a number denoting length of the dorsal ridge of a penguin bill (millimeters)
@@
 
-The **penguins** data in **palmerpenguins** is generally useful and approachable for data science and statistics education, and is uniquely well-suited to replace the **iris** dataset.
+The **penguins** data in **palmerpenguins** is useful and approachable for data science and statistics education, and is uniquely well-suited to replace the **iris** dataset.
 Comparisons presented are selected examples for common **iris** uses, and are not exhaustive.
@@
 In addition, teaching and learning opportunities in **penguins** are increased due to a greater number of variables, missing values, unequal sample sizes, and Simpson's paradox examples.
+Importantly, the **penguins** dataset encompasses real-world information derived from several charasmatic marine predator species with regional breeding populations notably responding to environmental change occuring throughout the Western Antarctic Peninsula region of the Southern Ocean (see Bestelmeyer et al.
+2011, Gorman et al.
+2014, 2015, 2017).
+Thus, the **penguins** dataset can facilitate discussions more broadly on biodiversity responses to global change - a contemporary and critical topic in ecology, evolution, and the environmental sciences.
+
+Bestelmeyer, B.T., A.M.
+Ellison, W.R.
+Fraser, K.B.
+Gorman, S.J.
+Holbrook, C.M.
+Laney, M.D.
+Ohman, D.P.C.
+Peters, F.C.
+Pillsbury, A.
+Rassweiler, R.
+Schmitt, and S.
+Sharma.
+2011.
+Analysis of abrupt transitions in ecological systems.
+Ecosphere 2(12):art129.
+
+Gorman, K.B.
+2015.
+Integrative studies of Southern Ocean food webs and Pygoscelis penguin demography: mechanisms of population response to environmental change.
+Ph.D.
+Thesis.
+Simon Fraser University, Burnaby, British Columbia, Canada.
+337 pp.
+
+Gorman, K.B., S.L.
+Talbot, S.A.
+Sonsthagen, G.K.
+Sage, M.C.
+Gravely, W.R.
+Fraser and T.D.
+Williams.
+2017.
+Population genetic structure and gene flow of Adélie penguins (Pygoscelis adeliae) breeding throughout the western Antarctic Peninsula.
+Antarctic Science 29(6):499-510.
+
+Gorman, K.B., T.D.
+Williams, and W.R.
+Fraser.
+2014.
+Ecological sexual dimorphism and environmental variability within a community of Antarctic penguins (genus Pygoscelis).
+PLoS ONE 9(3): e90081.
+***ALREADY REFERENCED
 

### 2021-04-26 R/penguins_code.Rmd
Allison fine tuning figures, text (#20)
* update refs, change NA in kmeans tables to 0, add penguins_raw summary table to app b

* update abstract to fresh data focus

* finalizing figures, text, formatting for R Journal

@@
 ---
-title: "Palmer penguins paper code"
-author: "Allison Horst"
-date: "7/2/2020"
-output: html_document
+title: 'Palmer Penguins manuscript code'
+output:
+  html_document:
+    number_sections: yes
+  pdf_document: default
 ---
 
-This document contains accompanying code for all figures, tables and analyses presented in "Palmer Archipelago (Antarctica) penguins data: an alternative to Edgar Anderson's *Iris* data."
-
-## Set-up and required packages:
-
-```{r setup, include = TRUE, message = FALSE, warning = FALSE}
-
-# ---- Code options ---- #
-knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
-
-# ---- Packages required ---- # 
+```{r setup, include=FALSE}
 library(tidyverse)
@@
 library(shadowtext)
+library(broom)
+knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
 
-# ---- Set default ggplot theme to theme_minimal ---- #
 theme_set(theme_minimal())
-
 ```
 
-## Table 1: *iris* sample
+# MANUSCRIPT CODE
 
-```{r iris_sample}
-set.seed(1900)
-sample_n(iris, 6) %>% 
-  kable() %>% 
-  kable_styling(full_width = FALSE)
+## Table 1: Overview summary of penguins and iris datasets
+
+```{r, echo = TRUE}
+# Features overview for penguins and iris: 
+overview_compare <- tribble(
+  ~Feature, ~iris, ~penguins,
+  "Year(s) collected", "1935", "2007 - 2009",
+  "Dimensions (col x row)", "5 x 150", "8 x 344",
+  "Documentation", "minimal", "complete metadata",
+  "Variable classes", "double (4), factor (1)", "double (2), int (3), factor (3)",
+  "Missing values?", "no (n = 0; 0.0%)", "yes (n = 19; 0.7%)"
+)
+
+overview_compare %>% 
+  kable(caption = "Overview comparison of penguins and iris dataset features and characteristics.")
 ```
 
-## Table 2: *penguins* sample
+## Table 2: Counts comparison between iris and penguins
 
-```{r penguin_sample}
-set.seed(1900)
-sample_n(penguins, 6) %>% 
-  kable() %>% 
-  kable_styling(full_width = FALSE)
-```
+```{r compare_counts, echo = TRUE}
 
-## Table 3: Sample size, *iris* (by species) and *penguins* (by species and sex)
-
-### NOTE: I am not happy about this method to convert NA values (for missing factor levels) to 0.
-
-```{r compare_counts}
-
-# ---- Iris counts by species ---- #
+# Count iris observations by species:
 iris_counts <- iris %>% 
@@
 
-# ---- Penguin counts by species and sex ---- #
+# Count penguins observations by species and sex:
 penguin_counts <- penguins %>% 
+  mutate(species = as.character(species)) %>% 
+  mutate(species = case_when(
+    species == "Adelie" ~ "Adélie", 
+    TRUE ~ species
+  )) %>% 
   mutate(sex = str_to_title(sex)) %>% 
-  count(species, sex) %>% 
+  group_by(species, sex) %>% 
+  tally() %>% 
   pivot_wider(names_from = sex, values_from = n) %>% 
@@
 
-# ---- Combine iris and penguin counts ---- #
-
+# Combine into a single table
 iris_penguin_n <- cbind(iris_counts, penguin_counts) %>%
@@
 
-# ---- Create a finalized table using `gt` package ---- #
+# Finalize combined counts table:
 iris_penguin_n %>% 
-  gt() %>%
-  tab_spanner(
-    label = "Iris sample size (by species)",
-    columns = vars(`Iris species`, `Sample size`)
-  ) %>%
-  tab_spanner(
-    label = "Penguin sample size (by species and sex)",
-    columns = vars(`Penguin species`, `Female`, `Male`, `NA`)
-  ) %>%
-  tab_footnote(
-    footnote = "Note: Here, penguin sample sizes are split by species and sex, and could be further divided by island. NA column for penguins indicate observations where sex was not recorded.",
-    location = cells_column_labels(
-      columns = vars(`Penguin species`)
-    )
-  )
-
+   kable(caption = "Grouped sample size for iris (by species; n = 150 total) and penguins (by species and sex; n = 344 total).
+Penguins can be further grouped by variables for island and study year.", align = "lclccc") %>%
+  add_header_above(c("Iris sample size (by species)" = 2, "Penguin sample size (by species and sex)" = 4 ))
 ```
 
-## Figure 1: *penguins* pairs plot
+## Figures 1 & 2: Overview comparison: numeric variables (pairs plots with ggpairs)
 
-```{r penguin_pairs}
-penguins %>%
+```{r, echo = TRUE, fig.align = "center"}
+### Pairs plots for penguins and iris (with ggpairs)
+
+# Penguin pairs plot:
+
+png(here("fig", "penguin_pairs.png"), units="in", width = 6, height = 4.5, res = 500)
+penguin_pairs <- penguins %>%
+  mutate(species = as.character(species)) %>% 
+  mutate(species = case_when(
+    species == "Adelie" ~ "Adélie", 
+    TRUE ~ species
+  )) %>% 
   select(species, where(is.numeric)) %>% 
@@
         )
-```
+  print(penguin_pairs)
+dev.off()
 
-## Figure 2: *iris* pairs plot
+# Iris pairs plot:
 
-```{r iris_pairs}
-iris %>%
-  ggpairs(aes(color = Species),
+png(here("fig", "iris_pairs.png"), units="in", width = 6, height = 4.5, res = 500)
+iris_pairs <- iris %>%
+  ggpairs(aes(color = Species, shape = Species),
           columns = c("Petal.Length", "Petal.Width", 
@@
           columnLabels = c("Petal length (cm)","Petal width (cm)", "Sepal length (cm)", "Sepal width (cm)"),
-          upper = list(continuous = wrap("cor", size = 3))) +
-  scale_colour_manual(values = c("darkorange","purple","cyan4")) +
-  scale_fill_manual(values = c("darkorange","purple","cyan4")) +
+          upper = list(continuous = wrap("cor", size = 2.7, color = "black"))) +
+  scale_colour_manual(values = c("gray70","gray40","black")) +
+  scale_fill_manual(values = c("gray70","gray40","black")) +
   theme_minimal() +
   theme(
-        text = element_text(size = 10, family = "Arial"),
+        text = element_text(size = 9),
         panel.grid.major = element_line(colour = NA),
         panel.grid.minor = element_blank(),
-        panel.background = element_rect(fill = "gray90", color = "gray90")
+        panel.background = element_rect(fill = "white", color = "gray80")
         )
+  print(iris_pairs)
+dev.off()
+
 ```
 
-## Figure 3: Visualize selected linear relationships
+## Figure 3: Example linear relationships
 
-```{r linear_viz, warning = FALSE, message = FALSE, fig.height = 6, fig.width = 6}
+```{r, echo = TRUE, fig.align = "center"}
 
-# ---- Penguin scatterplot, species included ---- #
+# Penguin linear relationships (flipper length versus body mass):
 
-penguin_flip_mass_scatter <- ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
-  geom_point(aes(color = species, shape = species), size = 1.5, alpha = 0.5) +
+penguin_flip_mass_scatter <- 
+  penguins %>% 
+  mutate(species = as.character(species)) %>% 
+  mutate(species = case_when(
+    species == "Adelie" ~ "Adélie", 
+    TRUE ~ species
+  )) %>%
+  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
+  geom_point(aes(color = species, shape = species), size = 1.5) +
   geom_smooth(method = "lm", aes(group = species, color = species), se = FALSE, show.legend = FALSE) +
   theme_minimal() +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme(legend.position = "bottom") +
+  scale_color_manual(values = c("darkorange","#0072b2","#009e73")) +
+  theme(legend.position = c(0.2,0.85)) +
   labs(x = "Flipper length (mm)", 
@@
   theme(plot.title.position = "plot",
-        panel.border = element_rect(fill = NA, color = "gray70"))
+        panel.grid = element_blank(),
+        panel.border = element_rect(fill = NA, color = "gray70"),
+        legend.title = element_blank())
 
-# ----- Penguin scatterplot, species omitted ---- #
-
+# To compare without penguin species as a variable (not in manuscript):
 penguin_flip_mass_scatter_all <- ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
-  geom_point(color = "gray50", size = 1.5, alpha = 0.5) +
+  geom_point(color = "gray50", size = 1.5) +
   geom_smooth(method = "lm", se = FALSE, color = "black") +
@@
   theme(plot.title.position = "plot",
+        panel.grid = element_blank(),
         panel.border = element_rect(fill = NA, color = "gray70"))
 
-# ---- Iris scatterplot, species included ---- #
-
+# Iris linear relationships (petal dimensions):
 iris_petal_scatter <- ggplot(iris, aes(x = Petal.Length, y = Petal.Width)) +
-  geom_point(aes(color = Species, size = Species, shape = Species), size = 1.5, alpha = 0.5) +
+  geom_point(aes(color = Species, size = Species, shape = Species), size = 1.5) +
   geom_smooth(aes(group = Species, color = Species), method = "lm", se = FALSE, show.legend = FALSE) +
   theme_minimal() +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
-  theme(legend.position = "bottom") +
+  scale_color_manual(values = c("gray70","gray40","black")) +
+  theme(legend.position = c(0.2, 0.85)) +
   labs(x = "Petal length (cm)",
@@
   theme(plot.title.position = "plot",
-        panel.background = element_rect(fill = "gray90", color = "gray90"))
+        panel.grid = element_blank(),
+        panel.background = element_rect(fill = NA, color = "gray70"),
+        legend.title = element_blank())
 
-# ---- Iris scatterplot, species omitted ---- #
-
+# To compare without iris species as a variable (not in manuscript):
 iris_petal_scatter_all <- ggplot(iris, aes(x = Petal.Length, y = Petal.Width)) +
-  geom_point(color = "gray50", size = 1.5, alpha = 0.5) +
+  geom_point(color = "gray50", size = 1.5) +
   geom_smooth(method = "lm", se = FALSE, color = "black") +
@@
   theme(plot.title.position = "plot",
-        panel.background = element_rect(fill = "gray90", color = "gray90"))
+        panel.grid = element_blank(),
+        panel.background = element_rect(fill = NA, color = "gray70"))
 
-(penguin_flip_mass_scatter_all + iris_petal_scatter_all) / (penguin_flip_mass_scatter + iris_petal_scatter) + plot_annotation(tag_levels = 'A')
+# Combine and save image: 
+png(here("fig", "linear_example.png"), units="in", width = 6, height = 3, res = 500)
+linear_example <- (penguin_flip_mass_scatter + iris_petal_scatter) + plot_annotation(tag_levels = 'A')
+print(linear_example)
+dev.off()
+
+# ggsave(here("fig","linear_example.png"), width = 6, height = 3)
 ```
 
-## Figure 4: Simpson's Paradox example
+## Figure 4: Simpson's Paradox
 
-```{r simpsons_paradox, fig.width = 8, fig.height = 4}
-# Simpson's Paradox example (bill dimensions)
+```{r, echo = TRUE}
+# Simpson's Paradox example (bill dimensions, omitting species):
 simpson_nospecies <- ggplot(data = penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
@@
 
-simpson_wspecies <- ggplot(data = penguins, aes(x = bill_length_mm, y = bill_depth_mm, group = species)) +
-  geom_point(aes(color = species, shape = species), size = 2, alpha = 0.5) +
+# Bill dimensions, including species:
+simpson_wspecies <- 
+  penguins %>% 
+  mutate(species = as.character(species)) %>% 
+  mutate(species = case_when(
+    species == "Adelie" ~ "Adélie",
+    TRUE ~ species)
+    ) %>% 
+    ggplot(aes(x = bill_length_mm, y = bill_depth_mm, group = species)) +
+  geom_point(aes(color = species, shape = species), size = 2, alpha = 0.8) +
   geom_smooth(method = "lm", se = FALSE, aes(color = species)) +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
+  scale_color_manual(values = c("darkorange","#0072b2","#009e73")) +
   theme_minimal() +
-  theme(panel.border = element_rect(fill = NA, color = "gray70"),
-        legend.position = c(0.16,0.14)) +
-  labs(x = "Bill length (mm)", y = "Bill depth (mm)")
+  theme(panel.border = element_rect(fill = NA, color = "gray70")) +
+  labs(x = "Bill length (mm)", y = "Bill depth (mm)") +
+  guides(color = guide_legend("Species"), 
+           shape = guide_legend("Species"))
 
-(simpson_nospecies | simpson_wspecies) + plot_annotation(tag_levels = 'A')
+# Combining into a compound figure:
+simpson_gg <- (simpson_nospecies | simpson_wspecies) + plot_annotation(tag_levels = 'A')
 
+ggsave(here("fig","simpson_gg.png"), width = 6, height = 2.5, dpi = 500)
 ```
@@
 
-```{r pca, fig.height = 8, fig.width = 8}
+```{r, echo = TRUE, fig.align = "center"}
+# From Alison Hill's post for palmerpenguins pkgdown site
 
-# ---- Penguins data prep ---- #
+### PENGUINS PCA:
+
+# Omit year
+penguins_noyr <- penguins %>% 
+  select(-year) %>% 
+  mutate(species = as.character(species)) %>% 
+  mutate(species = case_when(
+    species == "Adelie" ~ "Adélie",
+    TRUE ~ species
+  )) %>% 
+  mutate(species = as.factor(species))
+
 penguin_recipe <-
-  recipe(~., data = penguins) %>% 
+  recipe(~., data = penguins_noyr) %>% 
   update_role(species, island, sex, new_role = "id") %>% 
@@
 
-# ---- Penguin PCA ---- #
 penguin_pca <- 
@@
   tidy(id = "pca")
-
-# ---- Get penguins % variance ---- #
 
@@
 
-# ---- Clean penguins PCA outcome ---- #
+# Make the penguins PCA biplot:
 
+# Get pca loadings into wider format
 pca_wider <- penguin_pca %>% 
@@
 
-# ---- Define arrow style for biplot ---- #
+# define arrow style:
 arrow_style <- arrow(length = unit(.05, "inches"),
@@
 
-# ---- Make penguins biplot ---- #
+# Make the penguins PCA biplot:
 pca_plot <-
@@
   geom_point(aes(color = species, shape = species), 
-             alpha = 0.5, 
+             alpha = 0.7, 
              size = 2) +
-  scale_colour_manual(values = c("darkorange","purple","cyan4"))
+  scale_color_manual(values = c("darkorange","#0072b2","#009e73")) +
+  guides(color = guide_legend("Species"), 
+           shape = guide_legend("Species"))
 
@@
 
-# ---- Penguins scree plot ---- #
 penguin_screeplot <- penguin_percvar %>% 
@@
 
-#############################
-
-# Iris PCA
-
-# ---- Prep iris data for PCA ---- #
+### IRIS PCA:
 
@@
   prep()
-
-# ---- Iris PCA ---- #
-
 iris_pca <- 
@@
 
-# ---- Iris PCA loadings into wider format ---- #
+### Iris PCA biplot:
+
+# Get pca loadings into wider format
 iris_wider <- iris_pca %>% 
@@
 
-# ---- Make iris PCA biplot ---- #
+# define arrow style:
+arrow_style <- arrow(length = unit(.05, "inches"),
+                     type = "closed")
 
+# Make the iris PCA biplot:
 iris_pca_plot <-
@@
             bg.color = "white") +
-  theme(panel.background = element_rect(fill = "gray90", color = "gray90"),
+  theme(panel.background = element_rect(fill = NA, color = "gray70"),
         legend.position = "bottom")
 
-# ---- Get iris % variance ---- #
 iris_percvar <- iris_recipe %>% 
@@
 
-# ---- Make iris scree plot ---- #
+# Iris screeplot:
+
 iris_screeplot <- iris_percvar %>% 
@@
   geom_text(aes(label = round(value,2)), vjust=-0.25) +
-  theme(panel.background = element_rect(fill = "gray90", color = "gray90"))
+  theme(panel.border = element_rect(color = "gray70", fill = NA))
 
-# ---- Create compound figure (uses patchwork) ---- #
+# Combine biplots and screeplots for both iris & penguins:
+
 (penguins_biplot | iris_biplot) / (penguin_screeplot | iris_screeplot) + plot_annotation(tag_levels = 'A')
+
+ggsave(here("fig","pca_plots.png"), width = 8, height = 8, dpi = 500)
 ```
 
-## Figure 6: k-means
+## Figure 6 and Table 3: K-means clustering
 
-Simplified to two variables for each dataset (could be performed for all 4 numeric variables instead).
+```{r, echo = TRUE, fig.align = "center"}
 
-### NOTE: I am not happy about this method to convert NA values (for missing factor levels) to 0.
+# TWO VARIABLE k-means comparison
+# Penguins: Bill length vs. bill depth
+pb_species <- penguins %>% 
+  select(species, starts_with("bill")) %>% 
+  drop_na() %>% 
+  mutate(species = as.character(species)) %>% 
+  mutate(species = case_when(
+    species == "Adelie" ~ "Adélie",
+    TRUE ~ species
+  )) %>% 
+  mutate(species = as.factor(species))
 
-```{r penguin_kmeans}
+# Prep penguins for k-means:
+pb_nospecies <- pb_species %>% 
+  select(-species) %>% 
+  recipe() %>% 
+  step_normalize(all_numeric()) %>% 
+  prep() %>% 
+  juice()
 
-# ---- PENGUIN BILL DIMENSIONS K-MEANS ---- #
+# Perform k-means on penguin bill dimensions (k = 3, w/20 centroid starts)
 
-# ---- Simplify & drop missings ---- #
+# Save augmented data
+set.seed(100)
+pb_clust <- 
+  pb_nospecies %>% 
+  kmeans(centers = 3, nstart = 20) %>% 
+  broom::augment(pb_species)
 
-p_bills <- penguins %>% 
-  select(species, bill_length_mm, bill_depth_mm) %>% 
-  drop_na()
+# Plot penguin k-means clusters:
+pb_kmeans_gg <- 
+  pb_clust %>% 
+  ggplot(aes(x = bill_length_mm, y = bill_depth_mm)) +
+  geom_text(aes(label = .cluster, color = species),
+            key_glyph = draw_key_rect) +
+  scale_color_manual(values = c("darkorange","#0072b2","#009e73"))+
+  theme(legend.position = "bottom",
+         panel.border = element_rect(fill = NA, color = "gray70")) +
+  labs(x = "Bill length (mm)",
+       y = "Bill depth (mm)",
+       color = "Species")
 
-# ---- Remove species factor & scale ---- # 
-p_bills_nospecies <- p_bills %>% 
-  select(-species) %>% 
-  scale() %>% 
-  as.data.frame()
+# Get counts in each cluster by species
+pb_clust_n <- pb_clust %>% 
+  count(species, .cluster) %>% 
+  pivot_wider(names_from = species, values_from = n, names_sort = TRUE) %>% 
+  arrange(.cluster) %>% 
+  replace_na(list(`Adelie` = 0))
 
-# ---- K-means on penguin bill dimensions (k = 3, w/20 centroid starts) ---- #
+### Iris k-means
+# Iris: petal length vs. petal width
 
-p_bills_kmeans <- kmeans(p_bills_nospecies, centers = 3, nstart = 20)
+# Select only petal dimensions:
+ip_species <- iris %>% 
+  select(Species, starts_with("Petal"))
 
-# ---- Bind cluster assignments to the data ---- #
+# Remove species factor & scale petal dimensions: 
+ip_nospecies <- ip_species %>% 
+  select(-Species) %>% 
+  recipe() %>% 
+  step_normalize(all_numeric()) %>% 
+  prep() %>% 
+  juice()
 
-p_kmeans_clusters <- cbind(p_bills_nospecies, cluster = as.factor(p_bills_kmeans$cluster), species = p_bills$species)
+# Perform k-means on iris petal dimensions (k = 3, w/20 centroid starts):
+set.seed(100)
+ip_clust <- 
+  ip_nospecies %>% 
+  kmeans(centers = 3, nstart = 20) %>% 
+  broom::augment(ip_species)
 
-# ---- Plot clusters ---- #
+# Plot iris k-means clusters:
+ip_kmeans_gg <- 
+  ip_clust %>% 
+  ggplot(aes(x = Petal.Length, y = Petal.Width)) +
+  geom_text(aes(label = .cluster, color = Species), 
+            key_glyph = draw_key_rect) +
+  theme(legend.position = "bottom",
+         panel.border = element_rect(fill = NA, color = "gray70")) +
+  scale_color_manual(values = c("gray70","gray50","black"))  +
+  labs(x = "Petal length (cm)",
+       y = "Petal width (cm)")
 
-p_kmeans_gg <- ggplot(p_kmeans_clusters, aes(x = bill_length_mm, y = bill_depth_mm)) +
-  geom_text(aes(label = cluster, color = species)) +
-  theme(legend.position = "bottom") +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4"))
+# Get iris counts in each cluster by species:
+ip_clust_n <- ip_clust %>% 
+  count(Species, .cluster) %>% 
+  pivot_wider(names_from = Species, values_from = n, names_sort = TRUE) %>% 
+  arrange(.cluster) %>% 
+  replace_na(list(`setosa` = 0, `versicolor` = 0, `virginica` = 0))
 
-# ---- Get counts in each cluster by species ---- #
-p_cluster_counts <- p_kmeans_clusters %>% 
-  count(species, cluster) %>% 
-  pivot_wider(names_from = species, values_from = n) %>% 
-  arrange(cluster)
+# Combine k-means plots for penguins & iris:
+(pb_kmeans_gg | ip_kmeans_gg) + plot_annotation(tag_levels = "A")
 
+# Save image:
+ggsave(here("fig","kmeans.png"), width = 8, height = 4.5, dpi = 500)
+
+# Make cluster assignments table:
+kmeans_2var_table <- cbind(pb_clust_n, ip_clust_n) %>% 
+  kable(col.names = c("Cluster", "Adélie", "Chinstrap", "Gentoo", "Cluster", "setosa", "versicolor", "virginica"),
+        caption = "K-means cluster assignments by species based on penguin bill length (mm) and depth (mm), and iris petal length (cm) and width (cm).", align = "cccccc") %>% 
+  kable_styling(full_width = FALSE) %>% 
+  add_header_above(c("Penguins cluster assignments" = 4, "Iris cluster assignments" = 4))
 ```
-
-```{r iris_kmeans}
-
-# ---- IRIS PETALS K-MEANS ---- #
-
-# ---- Subset iris species & petal dimensions ---- #
-iris_petals <- iris %>% 
-  select(Species, Petal.Length, Petal.Width)
-
-# ---- remove species factor & scale petal dimensions ---- #
-
-ip_nospecies <- iris_petals %>% 
-  select(-Species) %>% 
-  scale() %>% 
-  as.data.frame()
-
-# ---- k-means on iris petal dimensions (k = 3, w/20 centroid starts) ---- #
-
-set.seed(1967)
-ip_kmeans <- kmeans(ip_nospecies, centers = 3, nstart = 20)
-
-# ---- Bind clusters to the data ---- #
-
-iris_kmeans_clusters <- cbind(ip_nospecies, cluster = as.factor(ip_kmeans$cluster), species = iris$Species)
-
-# ---- Plot clusters ---- #  
-iris_kmeans_gg <- ggplot(iris_kmeans_clusters, aes(x = Petal.Length, y = Petal.Width)) +
-  geom_text(aes(label = cluster, color = species)) +
-   theme(legend.position = "bottom",
-         panel.background = element_rect(fill = "gray92", color = "gray90")) +
-  scale_color_manual(values = c("darkorange","darkorchid","cyan4"))
-
-# ---- Get counts in each cluster by species ---- #
-iris_cluster_counts <- iris_kmeans_clusters %>% 
-  count(species, cluster) %>% 
-  pivot_wider(names_from = species, values_from = n) %>% 
-  arrange(cluster)
-
-```
-
-```{r kmeans_figs}
-
-# ---- Combine k-means plots with patchwork ---- #
-
-(p_kmeans_gg | iris_kmeans_gg) + plot_annotation(tag_levels = "A")
-
-# ---- Make table of cluster assignments ---- #
-
-kmeans_2var_table <- cbind(p_cluster_counts, iris_cluster_counts) %>% 
-  replace_na(list(`Adelie` = 0, `setosa` = 0, `versicolor` = 0, `virginica` = 0)) %>% 
-  kable() %>% 
-  kable_styling(full_width = FALSE) %>% 
-  column_spec(c(5:8), background = "gainsboro") %>% 
-  add_header_above(c("Penguins cluster assignments" = 4, "Iris cluster assignments" = 4))
-
-# ---- Return table ---- #
-
-kmeans_2var_table
-```

### 2021-04-26 penguins.Rmd
Allison fine tuning figures, text (#20)
* update refs, change NA in kmeans tables to 0, add penguins_raw summary table to app b

* update abstract to fresh data focus

* finalizing figures, text, formatting for R Journal

@@
 ---
-title: Palmer Archipelago Penguins Data in the palmerpenguins R Package - An Alternative to Anderson's *Irises*
+title: Palmer Archipelago Penguins Data in the palmerpenguins R Package - An Alternative to Anderson's Irises
 author:
@@
 
- In 1935, Edgar Anderson collected size measurements for 150 flowers from three different species of Iris on the Gaspé Peninsula.
-Since then, Anderson's *Iris* dataset has become a classic in statistics, machine learning, and data science teaching materials.
+ In 1935, Edgar Anderson collected size measurements for 150 flowers from three species of *Iris* on the Gaspé Peninsula in Quebec, Canada.
+Since then, Anderson's *Iris* observations have become a classic dataset in statistics, machine learning, and data science teaching materials.
 It is included in the base R **datasets** package as **iris**, making it easy for users to access without knowing much about it.
-Most notably, Anderson's *Iris* dataset appeared in a 1936 publication by R.
+However, the lack of data documentation, presence of non-intuitive variables (e.g.
+"sepal width"), and perfectly balanced groups with zero missing values make **iris** an inadequate and stale dataset for teaching and learning modern data science skills.
+Users would benefit from working with a more representative, real-world environmental dataset with a clear link to current scientific research.
+Importantly, Anderson’s *Iris* dataset appeared in a 1936 publication by R.
 A.
-Fisher in the *Annals of Eugenics*, linking the dataset to a field of research rooted in racism and prejudice.
-Combined with the lack of dataset documentation, the presence of non-intuitive variables like "sepal width", and the perfectly balanced groups with zero missing values, *Iris* is an unsatisfactory dataset for teaching and illustrating concepts.
-Users would benefit from working with a more representative, real-world environmental dataset---one with a clean history and a clear link to modern scientific research.
-Thus, a modern alternative to the *Iris* dataset is needed.
-Here, we introduce the **palmerpenguins** R package, which includes morphological measurements for three species of *Pygoscelis* penguins that breed on islands throughout the Palmer Archipelago, Antarctica.
-The **penguins** dataset provides an approachable, charismatic, and near drop-in replacement for the **iris** data.
-We directly compare the two datasets for selected analyses to demonstrate that R users, in particular teachers and learners currently using the **iris** dataset, can switch to the Palmer Archipelago penguins for many use cases including data wrangling and visualization, correlation, regression, hypothesis testing, multivariate analysis (e.g., PCA), cluster analysis and classification (e.g., by k-means).
+Fisher in the *Annals of Eugenics* (which is often the first-listed citation for the dataset), inextricably linking **iris** to racist research and theory.
+Thus, a modern alternative to the **iris** dataset is needed.
+ Here, we introduce the **palmerpenguins** R package, which includes morphological measurements collected from 2007 - 2009 for three species of *Pygoscelis* penguins that breed on islands throughout the Palmer Archipelago, Antarctica.
+The **penguins** data provides an approachable, charismatic, and near drop-in replacement for **iris** with topical relevance global climate change and environmental impacts on marine predators.
+Since being added to CRAN in July 2020, the **palmerpenguins** package has been downloaded over 102,000 times, highlighting the demand and widespread adoption of this viable **iris** alternative.
+ We directly compare the **iris** and **penguins** datasets for selected analyses to demonstrate that R users, in particular teachers and learners currently using **iris**, can switch to the Palmer Archipelago penguins for many use cases including data wrangling and visualization, correlation, regression, multivariate analysis (e.g., PCA), cluster analysis and classification (e.g., by k-means).
 
@@
 
-In 1935, American botanist Edgar Anderson measured petal and sepal structural dimensions (length and width) for 50 flowers from three Iris species: *Iris setosa*, *Iris versicolor*, and *Iris virginica* [@anderson_irises_1935].
-The manageable but non-trivial size (`r ncol(iris)` variables and `r nrow(iris)` total observations) and characteristics of the dataset, including linear relationships and multivariate normality, make it amenable for introducing a wide range of statistical methods including data wrangling, visualization, regression, multivariate exploration, and machine learning.
-Anderson's *Iris* dataset is built into a number of software packages including the auto-installed datasets package in R [@r_core_team_r_2019], Python's scikit-learn machine learning library [@pedregosa_scikit-learn_2011], and the SAS Sashelp library (SAS Institute, Cary NC), which have facilitated widespread use of the *Iris* data.
-As a result, eighty-five years after the data were initially published, Anderson’s *Iris* measurements are ubiquitous in statistics, computational methods, software documentation, and data science courses and materials.
+In 1935, American botanist Edgar Anderson measured petal and sepal structural dimensions (length and width) for 50 flowers from three *Iris* species: *Iris setosa*, *Iris versicolor*, and *Iris virginica* [@anderson_irises_1935].
+The manageable but non-trivial size (`r ncol(iris)` variables and `r nrow(iris)` total observations) and characteristics of the dataset, including linear relationships and multivariate normality, have made the dataset amenable for introducing a wide range of statistical methods including data wrangling, visualization, regression, multivariate exploration, and machine learning.
+Anderson's *Iris* data is built into a number of software packages including the auto-installed datasets package in R [@r_core_team_r_2019], Python's scikit-learn machine learning library [@pedregosa_scikit-learn_2011], and the SAS Sashelp library (SAS Institute, Cary NC), which have facilitated widespread use of the *Iris* data.
+As a result, eighty-six years after the data were initially published, Anderson’s *Iris* measurements are ubiquitous in statistics, computational methods, software documentation, and data science courses and materials.
 
-There are a number of reasons that modern data science practitioners and educators may want to look beyond Anderson's *Iris* dataset.
-First, Anderson's *Iris* dataset lacks metadata almost entirely [@anderson_irises_1935], limiting meaningful interpretation and discussion of research methods, analyses, and outcomes.
-Of five variables in the dataset, two (`Sepal.Width` and `Sepal.Length`) are not intuitive for most non-botanists.
+There are a number of reasons that modern data science practitioners and educators may want to move on from Anderson's *Iris* dataset.
+First, Anderson's *Iris* dataset lacks metadata almost entirely [@anderson_irises_1935], which does not reinforce best practices and limits meaningful interpretation and discussion of research methods, analyses, and outcomes.
+Of the five variables in the dataset, two (`Sepal.Width` and `Sepal.Length`) are not intuitive for most non-botanists.
 Even with explanation, the difference between *Iris* *petal* and *sepal* dimensions is not obvious.
-Second, the *Iris* dataset contains no missing values and equal sample sizes for each of the three species (*n* = 50), which is cleaner than most real-world datasets learners are likely to encounter "in the wild." Third, the single factor (`Species`) in the *Iris* data limits options for analyses.
-Finally, due to its direct link to a publication in the *Annals of Eugenics* by eugenicist and statistician R.A.
-Fisher's [@fisher_use_1936], the *Iris* dataset is burdened by a history in racist research.
+Second, the *Iris* dataset contains no missing values and sample sizes for each of the three species (*n* = 50) are equal, which is cleaner than most real-world datasets that learners are likely to encounter "in the wild." Third, the single factor (`Species`) in the *Iris* data limits options for analyses.
+Finally, due to its publication in the *Annals of Eugenics* by eugenicist and statistician R.A.
+Fisher [@fisher_use_1936], the *Iris* dataset is burdened by a history in scientific racism, which we are committed to addressing through the development of new data science education products as described below.
 
-<!---**Sounds great. Why consider replacing iris?**
-
-In 1936, one year after being described in the Bulletin of the American Iris Society, Anderson’s *Iris* measurements were published in full by eugenicist and statistician Ronald A.
-Fisher in *Annals of Eugenics* [@fisher_use_1936].
-Regardless of Anderson's initial motivation, the data today remain inextricably linked to Fisher’s eugenics research and are even commonly, if unfairly, referred to as "Fisher's iris data.” For example, an August 2020 search for “Iris flower data set” in Wikipedia returns an article beginning with “The iris flower data set or Fisher’s iris data…”, then later in the article: “It is sometimes called Anderson's Iris data set” [@wikipedia_iris_2020].
-The *Iris* dataset is similarly credited to Fisher in statistical computing literature [@trendafilov_simple_2009, @wang_matlab_2015, @woods_how_2015, @chen_unsupervised_2018].
-
-There is recent momentum to address Fisher’s racist legacy in statistics.
-In June 2020, the Committee of Presidents of Statistical Societies’ R.A.
-Fisher Award and Lectureship was replaced by the Distinguished Achievement Award and Lectureship, to “advance a more just, equitable, diverse, and inclusive statistical community” [@noauthor_institute_2020].
-Cambridge University’s Gonville and Caius College recently announced plans to remove a stained-glass window celebrating Fisher from a campus dining hall [@noauthor_sir_2020].
-In the R community, there are calls to (1) address the *Iris* dataset's use in Fisher’s eugenics work, and/or (2) consider using a different dataset [@poisot_timothee_its_2020, @garrick_aden-buie_lets_2020].
---->
-
-Given the need for fresh data science-ready datasets, we sought to identify a suitable alternative dataset that could be made easily accessible for a broad audience.
+Given the growing need for fresh data science-ready datasets, we sought to identify an alternative dataset that could be made easily accessible for a broad audience.
 After evaluating the positive and negative features of the *Iris* dataset in data science and statistics materials, we established the following criteria for a suitable alternative:
@@
 - Available by appropriate license (ideally, CC0 "no rights reserved")
-- Features intuitive subjects and variables that are interesting and understandable to learners across disciplines
+- Feature intuitive subjects and variables that are interesting and understandable to learners across disciplines
 - Complete metadata and documentation
@@
 Kristen Gorman in collaboration with the [Palmer Station LTER](https://pal.lternet.edu/), part of the [US LTER Network](https://lternet.edu/).
-For complete data collection methods and published analyses, see Gorman et al.
-(2014).
+For complete data collection methods and published analyses, see @gorman_ecological_2014.
+Throughout this paper, penguins species referred to as "Adélie", "Chinstrap", and "Gentoo."
 
-The data in the **palmerpenguins** R package are available for use by CC0 license ("No Rights Reserved") in accordance with the [Palmer Station LTER Data Policy](https://pal.lternet.edu/data/policies) and the [LTER Data Access Policy](https://lternet.edu/data-access-policy/), and were imported from the [Environmental Data Initiative (EDI) Data Portal](https://environmentaldatainitiative.org/) at the links below:
+The data in the \CRANpkg{palmerpenguins} R package are available for use by CC0 license ("No Rights Reserved") in accordance with the [Palmer Station LTER Data Policy](https://pal.lternet.edu/data/policies) and the [LTER Data Access Policy](https://lternet.edu/data-access-policy/), and were imported from the [Environmental Data Initiative (EDI) Data Portal](https://environmentaldatainitiative.org/) at the links below:
 
@@
 
-## R package: **palmerpenguins**
+## R package: palmerpenguins
 
-R users can install the **palmerpenguins** package from CRAN:
+R users can install the \CRANpkg{palmerpenguins} package from CRAN:
 
@@
 
-Information, examples, and links to community-contributed materials are available on the **palmerpenguins** package website: https://allisonhorst.github.io/palmerpenguins/.
+Information, examples, and links to community-contributed materials are available on the \CRANpkg{palmerpenguins} package website: \url{https://allisonhorst.github.io/palmerpenguins/}.
 
-The **palmerpenguins** R package contains two data objects: **penguins_raw** and **penguins**.
-The **penguins_raw** data consists of all raw data for `r ncol(palmerpenguins::penguins_raw)` variables (Appendix Table 1), recorded completely or in part for `r nrow(palmerpenguins::penguins_raw)` individual penguins, accessed directly from EDI.
-We generally recommend using the curated data in **penguins**, which is a subset of **penguins_raw** retaining all `r nrow(palmerpenguins::penguins)` observations, minimally updated (Appendix B) and reduced to the following eight variables:
+The \CRANpkg{palmerpenguins} R package contains two data objects: **penguins_raw** and **penguins**.
+The **penguins_raw** data consists of all raw data for `r ncol(palmerpenguins::penguins_raw)` variables, recorded completely or in part for `r nrow(palmerpenguins::penguins_raw)` individual penguins, accessed directly from EDI (**penguins_raw** properties are summarized in Appendix B).
+We generally recommend using the curated data in **penguins**, which is a subset of **penguins_raw** retaining all `r nrow(palmerpenguins::penguins)` observations, minimally updated (Appendix A) and reduced to the following eight variables:
 
-- *species:* a factor denoting the penguin species (Adélie, chinstrap, or gentoo)
+- *species:* a factor denoting the penguin species (Adélie, Chinstrap, or Gentoo)
 - *island:* a factor denoting the island (Palmer Archipelago, Antarctica) where penguins were observed (Biscoe Point, Dream Island or Torgersen Island)
@@
 
-The same data exist as comma-separated value (CSV) files in the package ("penguins_raw.csv" and "penguins.csv"), and can be read in using the built-in `path_to_file()` function in **palmerpenguins**.
+The same data exist as comma-separated value (CSV) files in the package ("penguins_raw.csv" and "penguins.csv"), and can be read in using the built-in `path_to_file()` function in \CRANpkg{palmerpenguins}.
 For example,
@@
 
-Python: Python users can access the penguins data in the **seaborn** data visualization library (Waskom et al. 2017). Example code to load the data in Python:
+Python: Python users can access the penguins data in the **seaborn** data visualization library [@michael_waskom_2017_883859]. Example code to load the data in Python:
 
@@
 
-Julia: Julia users can access the penguins data in the **PalmerPenguins.jl** package. Example code to import the penguins data through **PalmerPenguins.jl**:
+Julia: Julia users can access the penguins data in the **PalmerPenguins.jl** package.
+Example code to import the penguins data through **PalmerPenguins.jl** (more information on **PalmerPenguins.jl** from David Widmann can be found [here](https://github.com/devmotion/PalmerPenguins.jl)):
 
 ```
-    julia> using DataFrames
-    julia> df = DataFrame(table)
+    julia> using PalmerPenguins
+    julia> table = PalmerPenguins.load()
 ```
@@
 
-All analyses were performed in the R language environment using version 3.6.2 (R Core Team 2019).
+All analyses were performed in the R language environment using version 3.6.2 [@r_core_team_r_2019].
 Complete code for this paper is shared in the Supplemental Material.
@@
 
-- **tidyverse** (Wickham et al. 2019), for data import and cleaning
-- **ggplot2** (Wickham 2016:2), for data visualizations
-- **here** (Müller 2017), for file path control
-- **kableExtra** (Zhu 2019) for finalized tables
-- **gt** (Iannone, Cheng, and Schloerke 2020) for finalized tables
-- **GGally** (Schloerke et al. 2020), for pairs plots
-- **patchwork** (Pedersen 2019), for compound figures
-- **shadowtext** (Yu 2019), to add a background color to text labels
-- **recipes** (Kuhn and Wickham 2020), for data pre-processing
-- **base** and **stats** (R Core Team 2019) for various analyses throughout
-- **pkgdown** (ref) was used to build the package website
+- \CRANpkg{tidyverse} [@wickham_welcome_2019]: for data import and cleaning
+- \CRANpkg{ggplot2} [@wickham_ggplot2_2016]: for data visualizations
+- \CRANpkg{here} [@muller_here_2017]: for file path control
+- \CRANpkg{kableExtra} [@zhu_kableextra_2019]: for finalized tables
+- \CRANpkg{gt} [@iannone_gt_2020]: for finalized tables
+- \CRANpkg{GGally} [@schloerke_ggally_2020]: for pairs plots
+- \CRANpkg{patchwork} [@pedersen_patchwork_2019]: for compound figures
+- \CRANpkg{shadowtext} [@yu_shadowtext_2019]: to add a background color to text labels
+- \CRANpkg{recipes} [@kuhn_recipes_2020]: for data pre-processing
+- \CRANpkg{pkgdown} [@wickham_pkgdown_2020]: to build the package website
+- \pkg{base} and \pkg{stats} [@r_core_team_r_2019]: for various analyses throughout
 
@@
 
-The **penguins** data in **palmerpenguins** is useful and approachable for data science and statistics education, and is uniquely well-suited to replace the **iris** dataset.
+The **penguins** data in \CRANpkg{palmerpenguins} is useful and approachable for data science and statistics education, and is uniquely well-suited to replace the **iris** dataset.
 Comparisons presented are selected examples for common **iris** uses, and are not exhaustive.
@@
 Notably, while sample sizes in **iris** across species are all the same, sample sizes in **penguins** differ across the three species.
-The inclusion of three factor variables in **penguins** (*species*, *island*, and *sex*), along with *year*, create additional opportunities for grouping and faceting, compared to the single factor (*Species*) in **iris**.
+The inclusion of three factor variables in **penguins** (*species*, *island*, and *sex*), along with *year*, create additional opportunities for grouping, faceting, and analysis compared to the single factor (*Species*) in **iris**.
 
-Unlike **iris**, which contains only complete cases, the **penguins** dataset contains a small number of missing values (n~missing~ = 19, out of `r `ncol(palmerpenguins::penguins) * nrow(palmerpenguins::penguins)` total values).
-Missing values and unequal sample sizes are common in real-world data, and we believe add learning value to **penguins**.
+Unlike **iris**, which contains only complete cases, the **penguins** dataset contains a small number of missing values (n~missing~ = 19, out of `r ncol(palmerpenguins::penguins) * nrow(palmerpenguins::penguins)` total values).
+Missing values and unequal sample sizes are common in real-world data, and adds learning opportunities to the **penguins** dataset.
 
@@
     species == "Adelie" ~ "Adélie", 
-    species == "Chinstrap" ~ "chinstrap",
-    species == "Gentoo" ~ "gentoo",
     TRUE ~ species
@@
 
-```{r, penguins-pairs, fig.cap = "Distribution and correlations for numeric variables in the penguins data (flipper length (mm), body mass, (g) bill length (mm) and bill depth (mm)) for the three observed species: gentoo (green, triangles); chinstrap (blue, circles); and Adélie (orange, squares).
-Correlations are Pearson's r (*p < 0.05; **p < 0.01; ***p < 0.001).", warning = FALSE, message = FALSE, echo = FALSE}
-penguin_pairs <- penguins %>%
-  mutate(species = as.character(species)) %>% 
-  mutate(species = case_when(
-    species == "Adelie" ~ "Adélie", 
-    TRUE ~ species
-  )) %>% 
-  select(species, where(is.numeric)) %>% 
-  ggpairs(aes(color = species, shape = species),
-          columns = c("flipper_length_mm", "body_mass_g", 
-                      "bill_length_mm", "bill_depth_mm"),
-          columnLabels = c("Flipper length (mm)","Body mass (g)", "Bill length (mm)", "Bill depth (mm)"),
-          upper = list(continuous = wrap("cor", size = 2.7)),
-          lower = list(continuous = wrap(ggally_points, size = 1.3))) +
-  scale_fill_manual(values = c("darkorange","#0072b2","#009e73")) +
-  scale_color_manual(values = c("darkorange","#0072b2","#009e73")) +
-  scale_shape_manual(values = c(15,16,17)) +
-  theme_minimal() +
-  theme(
-        text = element_text(size = 9),
-        panel.grid.major = element_line(colour = NA),
-        panel.grid.minor = element_blank(),
-        panel.border = element_rect(color = "gray80", fill = NA)
-        )
+```{r, penguins-pairs, out.width = "100%", fig.cap = "Distributions and correlations for numeric variables in the penguins data (flipper length (mm), body mass (g), bill length (mm) and bill depth (mm)) for the three observed species: Gentoo (green, triangles); Chinstrap (blue, circles); and Adélie (orange, squares).
+Significance levels indicated for bivariate correlations are: *p < 0.05; **p < 0.01; ***p < 0.001.", warning = FALSE, message = FALSE, echo = FALSE}
 
-penguin_pairs
+knitr::include_graphics(here("fig", "penguin_pairs.png"))
+
 ```
 
-```{r, iris-pairs, fig.cap = "Distribution and correlations for numeric variables in the iris data (petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm)) for the three included Iris species: setosa (light gray, circles); versicolor (dark gray, triangles); and virginica (black, squares).
-Correlations are Pearson's r (*p < 0.05; **p < 0.01; ***p < 0.001).", warning = FALSE, message = FALSE, echo = FALSE}
-iris_pairs <- iris %>%
-  ggpairs(aes(color = Species, shape = Species),
-          columns = c("Petal.Length", "Petal.Width", 
-                      "Sepal.Length", "Sepal.Width"),
-          columnLabels = c("Petal length (cm)","Petal width (cm)", "Sepal length (cm)", "Sepal width (cm)"),
-          upper = list(continuous = wrap("cor", size = 2.7, color = "black"))) +
-  scale_colour_manual(values = c("gray70","gray40","black")) +
-  scale_fill_manual(values = c("gray70","gray40","black")) +
-  theme_minimal() +
-  theme(
-        text = element_text(size = 9),
-        panel.grid.major = element_line(colour = NA),
-        panel.grid.minor = element_blank(),
-        panel.background = element_rect(fill = "white", color = "gray80")
-        )
+```{r, iris-pairs, out.width = "100%", fig.cap = "Distributions and correlations for numeric variables in the iris data (petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm)) for the three included Iris species: setosa (light gray, circles); versicolor (dark gray, triangles); and virginica (black, squares).
+Significance levels indicated for bivariate correlations are: *p < 0.05; **p < 0.01; ***p < 0.001.", warning = FALSE, message = FALSE, echo = FALSE}
 
-iris_pairs
+knitr::include_graphics(here("fig", "iris_pairs.png"))
 
@@
 Both **penguins** and **iris** offer numerous opportunities to explore linear relationships and correlations, within and across species (Figures \ref{fig:penguins-pairs} & \ref{fig:iris-pairs}).
-A classic bivariate scatterplot made with the **iris** dataset reveals a clean linear relationship between petal length and petal width.
+A bivariate scatterplot made with the **iris** dataset reveals a clear linear relationship between petal length and petal width.
 Using **penguins** (\ref{fig:linear-example}), we can create a uniquely similar scatterplot with flipper length and body mass.
@@
 
-```{r, linear-example, fig.cap = "Representative linear relationships for penguin flipper length (mm) and body mass (g) for Adélie (orange circles), chinstrap (blue triangles), and gentoo (green squares) penguins (A), and iris petal length (cm) and width (cm) for setosa (light gray circles), versicolor (dark gray triangles) and virginica (black squares) irises (B).
-Within-species linear model is visualized for each penguin or iris species.", echo = FALSE, out.width = '6in', fig.align='center', fig.pos='htbp'}
+```{r, linear-example, out.width = "100%", fig.cap = "Representative linear relationships for (A): penguin flipper length (mm) and body mass (g) for Adélie (orange circles), Chinstrap (blue triangles), and Gentoo (green squares) penguins; (B): iris petal length (cm) and width (cm) for setosa (light gray circles), versicolor (dark gray triangles) and virginica (black squares) irises.
+Within-species linear model is visualized for each penguin or iris species.", echo = FALSE, fig.align='center', fig.pos='htbp'}
 
@@
 
-Notably, distinctions between species are clearer for iris petals - particularly, the much smaller petals for *Iris setosa* - compared to penguins, in which Adélie and chinstrap penguins are largely overlapping in body size (body mass and flipper length), and are both generally smaller than gentoos.
+Notably, distinctions between species are clearer for iris petals - particularly, the much smaller petals for *Iris setosa* - compared to penguins, in which Adélie and Chinstrap penguins are largely overlapping in body size (body mass and flipper length), and are both generally smaller than Gentoo penguins.
 
 Simpson’s Paradox is a data phenomenon in which a trend observed between variables is reversed when data are pooled, omitting a meaningful variable.
-While often taught and discussed in statistics courses, finding a real-world and approachable example of Simpson’s paradox can be a challenge.
-Here, we show one (of several possible; see Figure \ref{fig:penguins-pairs}) Simpson’s Paradox examples in **penguins**: exploring bill dimensions with and without *species* included (Figure \ref{fig:simpsons}).
-When penguin *species* is omitted (Figure \ref{fig:simpsons} A), bill length and depth appear negatively correlated overall.
-The trend is reversed when *species* is included, revealing an obviously positive correlation between bill length and bill depth within species (Figure \ref{fig:simpsons} B).
+While often taught and discussed in statistics courses, finding a real-world and approachable example of Simpson’s Paradox can be a challenge.
+Here, we show one (of several possible - see Figure \ref{fig:penguins-pairs}) Simpson’s Paradox examples in **penguins**: exploring bill dimensions with and without species included (Figure \ref{fig:simpsons}).
+When penguin species is omitted (Figure \ref{fig:simpsons} A), bill length and depth appear negatively correlated overall.
+The trend is reversed when species is included, revealing an obviously positive correlation between bill length and bill depth within species (Figure \ref{fig:simpsons} B).
 
-```{r, simpsons, fig.cap = "Trends for penguin bill dimensions (bill length and bill depth, millimeters) if the ‘species’ variable is excluded (A) or included (B), illustrating Simpson’s Paradox.
-Note: linear regression for bill dimensions without including species in (A) is ill-advised; the linear trendline is only included to visualize trend reversal for Simpson’s Paradox when compared to (B).", echo = FALSE, out.width = '6in', fig.align='center', fig.pos='htbp'}
+```{r, simpsons, out.width = "100%", fig.cap = "Trends for penguin bill dimensions (bill length and bill depth, millimeters) if the species variable is excluded (A) or included (B), illustrating Simpson’s Paradox.
+Note: linear regression for bill dimensions without including species in (A) is ill-advised; the linear trendline is only included to visualize trend reversal for Simpson’s Paradox when compared to (B).", echo = FALSE, fig.align='center', fig.pos='htbp'}
 
@@
 A comparison of PCA with the four variables of structural size measurements in **penguins** and **iris** (both normalized prior to PCA) reveals highly similar results (Figure \ref{fig:pca}).
-For both datasets, one species is distinct (gentoo penguins, and setosa irises) while the other two species (chinstrap/Adélie and versicolor/virginica) appear somewhat overlapping in the first two principal components (Figure \ref{fig:pca} A,B).
-Variance explained by each principal component (PC) is similar, particularly for PC1 and PC2: for **penguins**, 88.15% of total variance is captured by the first two PCs, compared to 95.81% for **iris**, with a similarly large percentage of variance captured by PC1 and PC2 in each (Figure \ref{fig:pca} C,D).
+For both datasets, one species is distinct (Gentoo penguins, and setosa irises) while the other two species (Chinstrap/Adélie and versicolor/virginica) appear somewhat overlapping in the first two principal components (Figure \ref{fig:pca} A,B).
+Screeplots reveal that the variance explained by each principal component (PC) is very similar across the two datasets, particularly for PC1 and PC2: for **penguins**, 88.15% of total variance is captured by the first two PCs, compared to 95.81% for **iris**, with a similarly large percentage of variance captured by PC1 and PC2 in each (Figure \ref{fig:pca} C,D).
 
-```{r, pca, fig.cap = "Principal component analysis biplots and scree plots for structural size measurements in penguins (A,C) and iris (B,D), revealing similarities in multivariate patterns, variable loadings, and variance explained by each component.
-For penguins, variables are flipper length (mm), body mass (g), bill length (mm) and bill depth (mm); groups are visualized by species (Adélie = orange circles, chinstrap = blue triangles, gentoo = green squares).
+```{r, pca, out.width = "100%", fig.cap = "Principal component analysis biplots and screeplots for structural size measurements in penguins (A,C) and iris (B,D), revealing similarities in multivariate patterns, variable loadings, and variance explained by each component.
+For penguins, variables are flipper length (mm), body mass (g), bill length (mm) and bill depth (mm); groups are visualized by species (Adélie = orange circles, Chinstrap = blue triangles, Gentoo = green squares).
 For iris, variables are petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm); groups are visualized by species (setosa = light gray circles, versicolor = dark gray triangles, virginica = black squares).
-Values above scree plot columns (C,D) indicate percent of total variance explained by each of the four principal components.", echo = FALSE, out.width = '6in', fig.align='center', fig.pos='htbp'}
+Values above screeplot columns (C,D) indicate percent of total variance explained by each of the four principal components.", echo = FALSE, fig.align='center', fig.pos='htbp'}
 
@@
 All variables are scaled prior to k-means.
-Three clusters (*k* = 3) are specified for each, since there are three species of both *Iris* (*setosa*, *versicolor*, and *virginica*) and penguins (Adélie, chinstrap and gentoo).
+Three clusters (*k* = 3) are specified for each, since there are three species of both *Iris* (*setosa*, *versicolor*, and *virginica*) and penguins (Adélie, Chinstrap and Gentoo).
 
 K-means clustering with penguin bill dimensions and iris petal dimensions yields largely distinct clusters, each dominated by one species (\ref{fig:kmeans}).
-For iris petal dimensions, k-means yields a perfectly separated cluster (Cluster 1) containing all 50 setosa iris observations and zero misclassified virginica or versicolor irises (Table 2).
+For iris petal dimensions, k-means yields a perfectly separated cluster (Cluster 3) containing all 50 setosa iris observations and zero misclassified virginica or versicolor irises (Table 2).
 While clustering is not perfectly distinct for any penguin species, each species is largely contained within a single cluster, with little overlap from the other two species.
-For example, considering Adélie penguins (orange observations in Figure \ref{fig:kmeans} A): 147 (out of 151) Adélie penguins are assigned to Cluster 1, zero are assigned to Cluster 2, and 4 are assigned to the chinstrap-dominated Cluster 3 (Table 2).
-Only 5 (of 68) chinstrap penguins and 1 (of 123) gentoo penguins are assigned to the Adélie-dominated Cluster 1 (Table 2).
+For example, considering Adélie penguins (orange observations in Figure \ref{fig:kmeans} A): 147 (out of 151) Adélie penguins are assigned to Cluster 3, zero are assigned to Cluster 1, and 4 are assigned to the Chinstrap-dominated Cluster 2 (Table 2).
+Only 5 (of 68) Chinstrap penguins and 1 (of 123) Gentoo penguins are assigned to the Adélie-dominated Cluster 3 (Table 2).
 
 ```{r, include = FALSE, echo = FALSE}
-# Including so kable table can be created here
-
-# TWO VARIABLE k-means comparison
+# Included to create kable table of cluster assignment
+# k-means comparison
 # Penguins: Bill length vs. bill depth
@@
   select(species, starts_with("bill")) %>% 
-  drop_na()
+  drop_na() %>% 
+  mutate(species = as.character(species)) %>% 
+  mutate(species = case_when(
+    species == "Adelie" ~ "Adélie",
+    TRUE ~ species
+  )) %>% 
+  mutate(species = as.factor(species))
 
@@
 # save augmented data
+set.seed(100)
 pb_clust <- 
@@
 
-# Plot clusters
-pb_kmeans_gg <- 
-  pb_clust %>% 
-  ggplot(aes(x = bill_length_mm, y = bill_depth_mm)) +
-  geom_text(aes(label = .cluster, color = species),
-            key_glyph = draw_key_rect) +
-  scale_color_manual(values = c("darkorange","#0072b2","#009e73"))+
-  theme(legend.position = "bottom",
-         panel.border = element_rect(fill = NA, color = "gray70"))
-
 # Get counts in each cluster by species
@@
   pivot_wider(names_from = species, values_from = n, names_sort = TRUE) %>% 
-  arrange(.cluster)
+  arrange(.cluster) %>% 
+  replace_na(list(`Adélie` = 0))
 
@@
 # Perform k-means on iris petal dimensions (k = 3, w/20 centroid starts)
+set.seed(100)
 ip_clust <- 
@@
 
-# Plot clusters
-ip_kmeans_gg <- 
-  ip_clust %>% 
-  ggplot(aes(x = Petal.Length, y = Petal.Width)) +
-  geom_text(aes(label = .cluster, color = Species), 
-            key_glyph = draw_key_rect) +
-  theme(legend.position = "bottom",
-         panel.border = element_rect(fill = NA, color = "gray70")) +
-  scale_color_manual(values = c("gray70","gray50","black"))
-
 # Get counts in each cluster by species
@@
   pivot_wider(names_from = Species, values_from = n, names_sort = TRUE) %>% 
-  arrange(.cluster)
+  arrange(.cluster) %>% 
+  replace_na(list(`setosa` = 0, `versicolor` = 0, `virginica` = 0))
 
-# Combine kmeans plots for penguins & iris
-(pb_kmeans_gg | ip_kmeans_gg) + plot_annotation(tag_levels = "A")
-
-# ggsave(here("fig","kmeans.png"), width = 8, height = 4.5)
-
+# Making cluster assignments table
 kmeans_2var_table <- cbind(pb_clust_n, ip_clust_n) %>% 
-  kable(caption = "Grouped sample size for iris (by species; n = 150 total) and penguins (by species and sex; n = 344 total).
-Penguins can be further grouped by variables for island and study year.", align = "lclccc") %>%
+  kable(col.names = c("Cluster", "Adélie", "Chinstrap", "Gentoo", "Cluster", "setosa", "versicolor", "virginica"),
+        caption = "K-means cluster assignments by species based on penguin bill length (mm) and depth (mm), and iris petal length (cm) and width (cm).", align = "cccccc") %>% 
   kable_styling(full_width = FALSE) %>% 
-  column_spec(c(5:8), background = "gainsboro") %>% 
   add_header_above(c("Penguins cluster assignments" = 4, "Iris cluster assignments" = 4))
@@
 
-```{r, kmeans, fig.cap = "K-means clustering outcomes for penguin bill dimensions (A) and iris petal dimensions (B).
-Numbers indicate the cluster to which an observation was assigned, revealing a high degree of separation between species for both penguins and iris.", echo = FALSE, out.width = '6in', fig.align='center', fig.pos='htbp'}
+```{r, kmeans, out.width = "100%", fig.cap = "K-means clustering outcomes for penguin bill dimensions (A) and iris petal dimensions (B).
+Numbers indicate the cluster to which an observation was assigned, revealing a high degree of separation between species for both penguins and iris.", echo = FALSE, fig.align='center', fig.pos='htbp'}
 
@@
 ```{r, echo = FALSE}
-kmeans_2var_table <- cbind(pb_clust_n, ip_clust_n) %>% 
-  kable(col.names = c("Cluster", "Adélie", "chinstrap", "gentoo", "Cluster", "setosa", "versicolor", "virginica"),
-        caption = "K-means cluster assignments by species based on penguin bill length (mm) and depth (mm), and iris petal length (cm) and width (cm).", align = "cccccc") %>% 
-  kable_styling(full_width = FALSE) %>% 
-  add_header_above(c("Penguins cluster assignments" = 4, "Iris cluster assignments" = 4))
-
 kmeans_2var_table
@@
 
-Here, we have shown that structural size measurements for Palmer Archipelago *Pygoscelis* penguins, available as **penguins** in the **palmerpenguins** R package, offer a near drop-in replacement for **iris** in a number of common use cases for data science and statistics education including exploratory data visualization, linear correlation and regression, PCA, and clustering by k-means.
+Here, we have shown that structural size measurements for Palmer Archipelago *Pygoscelis* penguins, available as **penguins** in the \CRANpkg{palmerpenguins} R package, offer a near drop-in replacement for **iris** in a number of common use cases for data science and statistics education including exploratory data visualization, linear correlation and regression, PCA, and clustering by k-means.
 In addition, teaching and learning opportunities in **penguins** are increased due to a greater number of variables, missing values, unequal sample sizes, and Simpson's paradox examples.
-Importantly, the **penguins** dataset encompasses real-world information derived from several charasmatic marine predator species with regional breeding populations notably responding to environmental change occuring throughout the Western Antarctic Peninsula region of the Southern Ocean (see Bestelmeyer et al.
-2011, Gorman et al.
-2014, 2015, 2017).
+Importantly, the **penguins** dataset encompasses real-world information derived from several charismatic marine predator species with regional breeding populations notably responding to environmental change occurring throughout the Western Antarctic Peninsula region of the Southern Ocean (see @bestelmeyer_analysis_2011, @gorman_ecological_2014, @gorman_integrative_2015, @gorman_population_2017).
 Thus, the **penguins** dataset can facilitate discussions more broadly on biodiversity responses to global change - a contemporary and critical topic in ecology, evolution, and the environmental sciences.
-
-Bestelmeyer, B.T., A.M.
-Ellison, W.R.
-Fraser, K.B.
-Gorman, S.J.
-Holbrook, C.M.
-Laney, M.D.
-Ohman, D.P.C.
-Peters, F.C.
-Pillsbury, A.
-Rassweiler, R.
-Schmitt, and S.
-Sharma.
-2011.
-Analysis of abrupt transitions in ecological systems.
-Ecosphere 2(12):art129.
-
-Gorman, K.B.
-2015.
-Integrative studies of Southern Ocean food webs and Pygoscelis penguin demography: mechanisms of population response to environmental change.
-Ph.D.
-Thesis.
-Simon Fraser University, Burnaby, British Columbia, Canada.
-337 pp.
-
-Gorman, K.B., S.L.
-Talbot, S.A.
-Sonsthagen, G.K.
-Sage, M.C.
-Gravely, W.R.
-Fraser and T.D.
-Williams.
-2017.
-Population genetic structure and gene flow of Adélie penguins (Pygoscelis adeliae) breeding throughout the western Antarctic Peninsula.
-Antarctic Science 29(6):499-510.
-
-Gorman, K.B., T.D.
-Williams, and W.R.
-Fraser.
-2014.
-Ecological sexual dimorphism and environmental variability within a community of Antarctic penguins (genus Pygoscelis).
-PLoS ONE 9(3): e90081.
-***ALREADY REFERENCED
 
@@
 ### Appendix A
-### Appendix B
 
@@
 
-- All variable names are converted to lower snake case
-- Entries in *species* are truncated to only include the common name (e.g. "gentoo", instead of "gentoo penguin (Pygoscelis papua)")
+- All variable names are converted to lower snake case (e.g. from `Flipper Length (mm)` to `flipper_length_mm`)
+- Entries in *species* are truncated to only include the common name (e.g. "Gentoo", instead of "gentoo penguin (Pygoscelis papua)")
 - Recorded sex for penguin N36A1, originally recorded as ".", is updated to `NA`
@@
 - Variable *year* was pulled from clutch observations
+
+### Appendix B
+
+Summary table, **penguins_raw** dataset in the **palmerpenguins** package.
+
+```{r, echo = FALSE}
+overview_penguins_raw <- tribble(
+  ~Feature, ~penguins_raw,
+  "Year(s) collected", "2007 - 2009",
+  "Dimensions (col x row)", "17 x 344",
+  "Documentation", "complete metadata",
+  "Variable classes", "character (9), Date (1), numeric (7)",
+  "Missing values?", "yes (n = 336; 5.7%)"
+)
+
+overview_penguins_raw %>% 
+  kable()
+```
 

### 2021-04-27 penguins.Rmd
Aph deux (#21)
* repurposed R/penguins_code.Rmd with penguins_plots.R to reduce duplication

* rm fig dir

* start keeping these files? not sure if we want to separate

* commit those files

* all png files now go here when you knit

* this is a purled script

* add tex file to git repo

* rework manuscript for better reproducibility

* don't show wizard behind curtain

* booktabs

* small edits

* updates

* no more gt?

* I don't even know

@@
 Thus, a modern alternative to the **iris** dataset is needed.
- Here, we introduce the **palmerpenguins** R package, which includes morphological measurements collected from 2007 - 2009 for three species of *Pygoscelis* penguins that breed on islands throughout the Palmer Archipelago, Antarctica.
+In this paper, we introduce the **palmerpenguins** R package, which includes morphological measurements collected from 2007 - 2009 for three species of *Pygoscelis* penguins that breed on islands throughout the Palmer Archipelago, Antarctica.
 The **penguins** data provides an approachable, charismatic, and near drop-in replacement for **iris** with topical relevance global climate change and environmental impacts on marine predators.
-Since being added to CRAN in July 2020, the **palmerpenguins** package has been downloaded over 102,000 times, highlighting the demand and widespread adoption of this viable **iris** alternative.
- We directly compare the **iris** and **penguins** datasets for selected analyses to demonstrate that R users, in particular teachers and learners currently using **iris**, can switch to the Palmer Archipelago penguins for many use cases including data wrangling and visualization, correlation, regression, multivariate analysis (e.g., PCA), cluster analysis and classification (e.g., by k-means).
+Since the release on CRAN in July 2020, the **palmerpenguins** package has been downloaded over 102,000 times, highlighting the demand and widespread adoption of this viable **iris** alternative.
+ We directly compare the **iris** and **penguins** datasets for selected analyses to demonstrate that R users, in particular teachers and learners currently using **iris**, can switch to the Palmer Archipelago penguins for many use cases including data wrangling, visualization, linear modeling, multivariate analysis (e.g., PCA), cluster analysis and classification (e.g., by k-means).
 
@@
 output: rticles::rjournal_article
+bibiography: packages.bib
 ---
 
-```{r setup, include=FALSE}
+```{r setup, include=FALSE, purl=FALSE}
+knitr::opts_chunk$set(echo = FALSE, 
+                      warning = FALSE, 
+                      message = FALSE, 
+                      purl = TRUE,
+                      out.width = "100%",
+                      dpi = 300, 
+                      fig.align = "center",
+                      fig.path = "figs/",
+                      dev = "png")
+```
+
+```{r pkgs, include=FALSE, purl=TRUE}
 library(tidyverse)
-library(here)
 library(palmerpenguins)
+
+# for tables
 library(kableExtra)
+
+# for plots
+library(paletteer)
 library(patchwork)
-library(gt)
 library(GGally)
+library(shadowtext)
+ggplot2::theme_set(ggplot2::theme_minimal())
+
+# for modeling
 library(recipes)
-library(shadowtext)
 library(broom)
-knitr::opts_chunk$set(echo = TRUE)
 
-theme_set(theme_minimal())
+knitr::read_chunk("penguins_plots.R")
+```
+
+```{r write-bib, purl=FALSE, include=FALSE}
+knitr::write_bib(.packages(), "packages.bib")
 ```
@@
 In 1935, American botanist Edgar Anderson measured petal and sepal structural dimensions (length and width) for 50 flowers from three *Iris* species: *Iris setosa*, *Iris versicolor*, and *Iris virginica* [@anderson_irises_1935].
-The manageable but non-trivial size (`r ncol(iris)` variables and `r nrow(iris)` total observations) and characteristics of the dataset, including linear relationships and multivariate normality, have made the dataset amenable for introducing a wide range of statistical methods including data wrangling, visualization, regression, multivariate exploration, and machine learning.
+The manageable but non-trivial size (`r ncol(iris)` variables and `r nrow(iris)` total observations) and characteristics of the dataset, including linear relationships and multivariate normality, have made the dataset amenable for introducing a wide range of statistical methods including data wrangling, visualization, linear modeling, multivariate analyses, and machine learning.
 Anderson's *Iris* data is built into a number of software packages including the auto-installed datasets package in R [@r_core_team_r_2019], Python's scikit-learn machine learning library [@pedregosa_scikit-learn_2011], and the SAS Sashelp library (SAS Institute, Cary NC), which have facilitated widespread use of the *Iris* data.
@@
 Even with explanation, the difference between *Iris* *petal* and *sepal* dimensions is not obvious.
-Second, the *Iris* dataset contains no missing values and sample sizes for each of the three species (*n* = 50) are equal, which is cleaner than most real-world datasets that learners are likely to encounter "in the wild." Third, the single factor (`Species`) in the *Iris* data limits options for analyses.
+Second, the *Iris* dataset contains equal sample sizes for each of the three species (*n* = 50) with no missing values, which is cleaner than most real-world datasets that learners are likely to encounter "in the wild." Third, the single factor (`Species`) in the *Iris* data limits options for analyses.
 Finally, due to its publication in the *Annals of Eugenics* by eugenicist and statistician R.A.
@@
 - Minimal data cleaning and pre-processing required for most analyses
-- Real-world (not manufactured), modern data
+- Real-world (not manufactured) modern data
 - Provides similar opportunities for teaching and learning R, data science, and statistical skills
@@
 Information, examples, and links to community-contributed materials are available on the \CRANpkg{palmerpenguins} package website: \url{https://allisonhorst.github.io/palmerpenguins/}.
+See the Appendix for how Python and Julia users can access the same data.
 
@@
 - *species:* a factor denoting the penguin species (Adélie, Chinstrap, or Gentoo)
-- *island:* a factor denoting the island (Palmer Archipelago, Antarctica) where penguins were observed (Biscoe Point, Dream Island or Torgersen Island)
+- *island:* a factor denoting the Palmer Archipelago island in Antarctica where each penguin was observed (Biscoe Point, Dream Island, or Torgersen Island)
 - *bill_length_mm:* a number denoting length of the dorsal ridge of a penguin bill (millimeters)
@@
 - *sex:* a factor denoting the sex of a penguin sex (male, female) based on molecular data
-- *year:* an integer denoting the year of study (2007, 2008 or 2009)
+- *year:* an integer denoting the year of study (2007, 2008, or 2009)
+
+\newpage
 
@@
 ```
-    # install.packages("readr")
     library(palmerpenguins)
-    df <- readr::read_csv(path_to_file("penguins.csv"))
+    df <- read.csv(path_to_file("penguins.csv"))
 ```
@@
 
-## Other data access options
-
-Python: Python users can access the penguins data in the **seaborn** data visualization library [@michael_waskom_2017_883859]. Example code to load the data in Python:
-
-```
-    import seaborn as sns
-    df = sns.load_dataset('penguins') 
-```
-
-Julia: Julia users can access the penguins data in the **PalmerPenguins.jl** package.
-Example code to import the penguins data through **PalmerPenguins.jl** (more information on **PalmerPenguins.jl** from David Widmann can be found [here](https://github.com/devmotion/PalmerPenguins.jl)):
-
-```
-    julia> using PalmerPenguins
-    julia> table = PalmerPenguins.load()
-```
-
-## Software and code
-
-All analyses were performed in the R language environment using version 3.6.2 [@r_core_team_r_2019].
-Complete code for this paper is shared in the Supplemental Material.
-We acknowledge the following R packages used in analyses, with gratitude to developers and contributors:
-
-- \CRANpkg{tidyverse} [@wickham_welcome_2019]: for data import and cleaning
-- \CRANpkg{ggplot2} [@wickham_ggplot2_2016]: for data visualizations
-- \CRANpkg{here} [@muller_here_2017]: for file path control
-- \CRANpkg{kableExtra} [@zhu_kableextra_2019]: for finalized tables
-- \CRANpkg{gt} [@iannone_gt_2020]: for finalized tables
-- \CRANpkg{GGally} [@schloerke_ggally_2020]: for pairs plots
-- \CRANpkg{patchwork} [@pedersen_patchwork_2019]: for compound figures
-- \CRANpkg{shadowtext} [@yu_shadowtext_2019]: to add a background color to text labels
-- \CRANpkg{recipes} [@kuhn_recipes_2020]: for data pre-processing
-- \CRANpkg{pkgdown} [@wickham_pkgdown_2020]: to build the package website
-- \pkg{base} and \pkg{stats} [@r_core_team_r_2019]: for various analyses throughout
-
-## Selected comparisons between **iris** and **penguins**
+## Comparing **iris** and **penguins**
 
@@
 
-### Overview
-
-```{r, echo = FALSE}
+```{r overview-tbl}
 overview_compare <- tribble(
@@
 overview_compare %>% 
-  kable(caption = "Overview comparison of penguins and iris dataset features and characteristics.")
+  kable(caption = "Overview comparison of penguins and iris dataset features and characteristics.",
+        align = "lclccc",
+        booktabs = TRUE)
 ```
@@
 
-```{r, echo = FALSE}
+```{r counts-tbl}
 iris_counts <- iris %>% 
@@
    kable(caption = "Grouped sample size for iris (by species; n = 150 total) and penguins (by species and sex; n = 344 total).
-Penguins can be further grouped by variables for island and study year.", align = "lclccc") %>%
+Penguins can be further grouped by variables for island and study year.",
+         align = "lclccc",
+         booktabs = TRUE) %>% 
   add_header_above(c("Iris sample size (by species)" = 2, "Penguin sample size (by species and sex)" = 4 ))
@@
 
-Distributions, relationships between variables, and clustering can be visually explored between species for the four structural size measurements in **penguins** (flipper length, body mass, bill length and depth; Figure \ref{fig:penguins-pairs}) and **iris** (sepal width and length, petal width and length; Figure \ref{fig:iris-pairs}).
+Distributions, relationships between variables, and clustering can be visually explored between species for the four structural size measurements in **penguins** (flipper length, body mass, bill length and depth; Figure \ref{fig:penguin-pairs}) and **iris** (sepal width and length, petal width and length; Figure \ref{fig:iris-pairs}).
 
-```{r, penguins-pairs, out.width = "100%", fig.cap = "Distributions and correlations for numeric variables in the penguins data (flipper length (mm), body mass (g), bill length (mm) and bill depth (mm)) for the three observed species: Gentoo (green, triangles); Chinstrap (blue, circles); and Adélie (orange, squares).
-Significance levels indicated for bivariate correlations are: *p < 0.05; **p < 0.01; ***p < 0.001.", warning = FALSE, message = FALSE, echo = FALSE}
-
-knitr::include_graphics(here("fig", "penguin_pairs.png"))
-
+```{r, penguin-pairs, width = 6, height = 4.5, fig.cap = "Distributions and correlations for numeric variables in the penguins data (flipper length (mm), body mass (g), bill length (mm) and bill depth (mm)) for the three observed species: Gentoo (green, triangles); Chinstrap (blue, circles); and Adélie (orange, squares).
+Significance levels indicated for bivariate correlations are: *p < 0.05; **p < 0.01; ***p < 0.001."}
 ```
 
-```{r, iris-pairs, out.width = "100%", fig.cap = "Distributions and correlations for numeric variables in the iris data (petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm)) for the three included Iris species: setosa (light gray, circles); versicolor (dark gray, triangles); and virginica (black, squares).
-Significance levels indicated for bivariate correlations are: *p < 0.05; **p < 0.01; ***p < 0.001.", warning = FALSE, message = FALSE, echo = FALSE}
-
-knitr::include_graphics(here("fig", "iris_pairs.png"))
-
+```{r, iris-pairs, width = 6, height = 4.5, fig.cap = "Distributions and correlations for numeric variables in the iris data (petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm)) for the three included Iris species: setosa (light gray, circles); versicolor (dark gray, triangles); and virginica (black, squares).
+Significance levels indicated for bivariate correlations are: *p < 0.05; **p < 0.01; ***p < 0.001."}
 ```
 
-Both **penguins** and **iris** offer numerous opportunities to explore linear relationships and correlations, within and across species (Figures \ref{fig:penguins-pairs} & \ref{fig:iris-pairs}).
+Both **penguins** and **iris** offer numerous opportunities to explore linear relationships and correlations, within and across species (Figures \ref{fig:penguin-pairs} & \ref{fig:iris-pairs}).
 A bivariate scatterplot made with the **iris** dataset reveals a clear linear relationship between petal length and petal width.
@@
 
-```{r, linear-example, out.width = "100%", fig.cap = "Representative linear relationships for (A): penguin flipper length (mm) and body mass (g) for Adélie (orange circles), Chinstrap (blue triangles), and Gentoo (green squares) penguins; (B): iris petal length (cm) and width (cm) for setosa (light gray circles), versicolor (dark gray triangles) and virginica (black squares) irises.
-Within-species linear model is visualized for each penguin or iris species.", echo = FALSE, fig.align='center', fig.pos='htbp'}
-
-knitr::include_graphics('fig/linear_example.png')
-
+```{r, linear-example, width = 6, height = 3, fig.cap = "Representative linear relationships for (A): penguin flipper length (mm) and body mass (g) for Adélie (orange circles), Chinstrap (blue triangles), and Gentoo (green squares) penguins; (B): iris petal length (cm) and width (cm) for setosa (light gray circles), versicolor (dark gray triangles) and virginica (black squares) irises.
+Within-species linear model is visualized for each penguin or iris species.", fig.pos='htbp'}
 ```
@@
 While often taught and discussed in statistics courses, finding a real-world and approachable example of Simpson’s Paradox can be a challenge.
-Here, we show one (of several possible - see Figure \ref{fig:penguins-pairs}) Simpson’s Paradox examples in **penguins**: exploring bill dimensions with and without species included (Figure \ref{fig:simpsons}).
+Here, we show one (of several possible - see Figure \ref{fig:penguin-pairs}) Simpson’s Paradox examples in **penguins**: exploring bill dimensions with and without species included (Figure \ref{fig:simpsons}).
 When penguin species is omitted (Figure \ref{fig:simpsons} A), bill length and depth appear negatively correlated overall.
@@
 
-```{r, simpsons, out.width = "100%", fig.cap = "Trends for penguin bill dimensions (bill length and bill depth, millimeters) if the species variable is excluded (A) or included (B), illustrating Simpson’s Paradox.
-Note: linear regression for bill dimensions without including species in (A) is ill-advised; the linear trendline is only included to visualize trend reversal for Simpson’s Paradox when compared to (B).", echo = FALSE, fig.align='center', fig.pos='htbp'}
-
-knitr::include_graphics('fig/simpson_gg.png')
-
+```{r, simpsons, width = 6, height = 2.5, fig.cap = "Trends for penguin bill dimensions (bill length and bill depth, millimeters) if the species variable is excluded (A) or included (B), illustrating Simpson’s Paradox.
+Note: linear regression for bill dimensions without including species in (A) is ill-advised; the linear trendline is only included to visualize trend reversal for Simpson’s Paradox when compared to (B).", fig.pos='htbp'}
 ```
@@
 
-```{r, pca, out.width = "100%", fig.cap = "Principal component analysis biplots and screeplots for structural size measurements in penguins (A,C) and iris (B,D), revealing similarities in multivariate patterns, variable loadings, and variance explained by each component.
+```{r, pca, width = 8, height = 8, fig.cap = "Principal component analysis biplots and screeplots for structural size measurements in penguins (A,C) and iris (B,D), revealing similarities in multivariate patterns, variable loadings, and variance explained by each component.
 For penguins, variables are flipper length (mm), body mass (g), bill length (mm) and bill depth (mm); groups are visualized by species (Adélie = orange circles, Chinstrap = blue triangles, Gentoo = green squares).
 For iris, variables are petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm); groups are visualized by species (setosa = light gray circles, versicolor = dark gray triangles, virginica = black squares).
-Values above screeplot columns (C,D) indicate percent of total variance explained by each of the four principal components.", echo = FALSE, fig.align='center', fig.pos='htbp'}
-
-knitr::include_graphics('fig/pca_plots.png')
-
+Values above screeplot columns (C,D) indicate percent of total variance explained by each of the four principal components.", fig.pos='htbp'}
 ```
@@
 
-```{r, include = FALSE, echo = FALSE}
+```{r}
 # Included to create kable table of cluster assignment
@@
   kable(col.names = c("Cluster", "Adélie", "Chinstrap", "Gentoo", "Cluster", "setosa", "versicolor", "virginica"),
-        caption = "K-means cluster assignments by species based on penguin bill length (mm) and depth (mm), and iris petal length (cm) and width (cm).", align = "cccccc") %>% 
+        caption = "K-means cluster assignments by species based on penguin bill length (mm) and depth (mm), and iris petal length (cm) and width (cm).", 
+        align = "cccccc",
+        booktabs = TRUE) %>% 
   kable_styling(full_width = FALSE) %>% 
@@
 
-```{r, kmeans, out.width = "100%", fig.cap = "K-means clustering outcomes for penguin bill dimensions (A) and iris petal dimensions (B).
-Numbers indicate the cluster to which an observation was assigned, revealing a high degree of separation between species for both penguins and iris.", echo = FALSE, fig.align='center', fig.pos='htbp'}
-
-knitr::include_graphics('fig/kmeans.png')
+```{r, kmeans, width = 8, height = 4.5, fig.cap = "K-means clustering outcomes for penguin bill dimensions (A) and iris petal dimensions (B).
+Numbers indicate the cluster to which an observation was assigned, revealing a high degree of separation between species for both penguins and iris.", fig.pos='htbp'}
 ```
 
-```{r, echo = FALSE}
+```{r}
 kmeans_2var_table
@@
 
-## Appendix
+# Appendix {-}
 
-### Appendix A
+## Penguins data processing
 
@@
 
-### Appendix B
-
-Summary table, **penguins_raw** dataset in the **palmerpenguins** package.
+## Summary of the **penguins_raw** dataset
 
@@
 overview_penguins_raw %>% 
-  kable()
+  kable(booktabs = TRUE)
 ```
 
-<!---This section may contain a figure such as Figure \ref{fig:Rlogo}.
+## Other data sources
 
-```{r, Rlogo, echo=FALSE, fig.cap='The logo of R.', out.width='2in', fig.align='center', fig.pos='htbp'}
-knitr::include_graphics('palmerpenguins-logo.png')
+Python: Python users can access the penguins data in the **seaborn** data visualization library [@michael_waskom_2017_883859]. Example code to load the data in Python:
+
 ```
---->
+    import seaborn as sns
+    df = sns.load_dataset('penguins') 
+```
 
-<!---
-## Another section
+Julia: Julia users can access the penguins data in the **PalmerPenguins.jl** package.
+Example code to import the penguins data through **PalmerPenguins.jl** (more information on **PalmerPenguins.jl** from David Widmann can be found [here](https://github.com/devmotion/PalmerPenguins.jl)):
 
-There will likely be several sections, perhaps including code snippets, such as:
+```
+    julia> using PalmerPenguins
+    julia> table = PalmerPenguins.load()
+```
 
-```{r}
-x <- 1:10
-x
-```
---->
+## Acknowledgements
 
-<!----
-## Summary
+All analyses were performed in the R language environment using version 3.6.2 [@r_core_team_r_2019].
+Complete code for this paper is shared in the Supplemental Material.
+We acknowledge the following R packages used in analyses, with gratitude to developers and contributors:
 
-This file is only a basic article template.
-For full details of _The R Journal_ style and information on how to prepare your article for submission, see the [Instructions for Authors](https://journal.r-project.org/share/author-guide.pdf).
-
---->
+- \CRANpkg{tidyverse} [@wickham_welcome_2019]: for data import and cleaning
+- \CRANpkg{ggplot2} [@wickham_ggplot2_2016]: for data visualizations
+- \CRANpkg{kableExtra} [@zhu_kableextra_2019]: for finalized tables
+- \CRANpkg{GGally} [@schloerke_ggally_2020]: for pairs plots
+- \CRANpkg{patchwork} [@pedersen_patchwork_2019]: for compound figures
+- \CRANpkg{shadowtext} [@yu_shadowtext_2019]: to add a background color to text labels
+- \CRANpkg{recipes} [@kuhn_recipes_2020]: for data pre-processing
+- \CRANpkg{pkgdown} [@wickham_pkgdown_2020]: to build the package website
+- \pkg{base} and \pkg{stats} [@r_core_team_r_2019]: for various analyses throughout
 

### 2021-04-27 penguins.tex
Aph deux (#21)
* repurposed R/penguins_code.Rmd with penguins_plots.R to reduce duplication

* rm fig dir

* start keeping these files? not sure if we want to separate

* commit those files

* all png files now go here when you knit

* this is a purled script

* add tex file to git repo

* rework manuscript for better reproducibility

* don't show wizard behind curtain

* booktabs

* small edits

* updates

* no more gt?

* I don't even know

--- 2020-09-12 penguins.tex
@@
 % !TeX root = RJwrapper.tex
-\title{Palmer Archipelago Penguins Data in the palmerpenguins R Package - An
-Alternative to Anderson's \emph{Irises}}
-\author{by Allison M. Horst, Alison P. Hill, Kristen B. Gorman}
+\title{Palmer Archipelago Penguins Data in the palmerpenguins R Package
+- An Alternative to Anderson's Irises}
+\author{by Allison M. Horst, Alison Presmanes Hill, and Kristen B. Gorman}
 
@@
 \abstract{%
-Edgar Anderson's famous 1935 \emph{Iris} dataset, which exists in the
-base R \textbf{datasets} package as \textbf{iris}, is ubiquitous in
-statistics and data science education. Due to a lack of metadata,
-non-intuitive variables (e.g.~Sepal.Width), its publication in (and
-inextricable connection with) Fisher's 1936 paper in \emph{Annals of
-Eugenics}, and perfect structure (e.g.~no missing values)
-unrepresentative of most real-world environmental data, a modern
-alternative to \textbf{iris} is needed. Penguin morphological
-measurements for three species of \emph{Pygoscelis} penguins that breed
-on islands throughout the Palmer Archipelago, Antarctica, provide an
-approachable, charismatic and near drop-in replacement for the
-\textbf{iris} data. Here, we introduce the \textbf{palmerpenguins} R
-package and describe the included \textbf{penguins} data. We directly
-compare the two datasets for selected analyses to demonstrate that R
-users, in particular teachers and learners currently using the
-\textbf{iris} data, can switch to the Palmer Archipelago penguins for
-many use cases including data wrangling and visualization, correlation,
-regression, hypothesis testing, multivariate analysis (e.g.~PCA),
-cluster analysis and classification (e.g.~by k-means).
+In 1935, Edgar Anderson collected size measurements for 150 flowers from
+three species of \emph{Iris} on the Gaspé Peninsula in Quebec, Canada.
+Since then, Anderson's \emph{Iris} observations have become a classic
+dataset in statistics, machine learning, and data science teaching
+materials. It is included in the base R \textbf{datasets} package as
+\textbf{iris}, making it easy for users to access without knowing much
+about it. However, the lack of data documentation, presence of
+non-intuitive variables (e.g.~``sepal width''), and perfectly balanced
+groups with zero missing values make \textbf{iris} an inadequate and
+stale dataset for teaching and learning modern data science skills.
+Users would benefit from working with a more representative, real-world
+environmental dataset with a clear link to current scientific research.
+Importantly, Anderson's \emph{Iris} dataset appeared in a 1936
+publication by R. A. Fisher in the \emph{Annals of Eugenics} (which is
+often the first-listed citation for the dataset), inextricably linking
+\textbf{iris} to racist research and theory. Thus, a modern alternative
+to the \textbf{iris} dataset is needed. In this paper, we introduce the
+\textbf{palmerpenguins} R package, which includes morphological
+measurements collected from 2007 - 2009 for three species of
+\emph{Pygoscelis} penguins that breed on islands throughout the Palmer
+Archipelago, Antarctica. The \textbf{penguins} data provides an
+approachable, charismatic, and near drop-in replacement for
+\textbf{iris} with topical relevance global climate change and
+environmental impacts on marine predators. Since the release on CRAN in
+July 2020, the \textbf{palmerpenguins} package has been downloaded over
+102,000 times, highlighting the demand and widespread adoption of this
+viable \textbf{iris} alternative. We directly compare the \textbf{iris}
+and \textbf{penguins} datasets for selected analyses to demonstrate that
+R users, in particular teachers and learners currently using
+\textbf{iris}, can switch to the Palmer Archipelago penguins for many
+use cases including data wrangling, visualization, linear modeling,
+multivariate analysis (e.g., PCA), cluster analysis and classification
+(e.g., by k-means).
 }
@@
 In 1935, American botanist Edgar Anderson measured petal and sepal
-structural dimensions (length and width) for 50 flowers from three Iris
-species: \emph{Iris setosa}, \emph{Iris versicolor}, and \emph{Iris
-virginica} \citep{anderson_irises_1935}. The manageable but non-trivial
-size (5 variables and 150 total observations) and characteristics of the
-data make it amenable for introducing a wide range of statistical
-methods including data wrangling, visualization, regression,
-multivariate exploration, and machine learning. Anderson's \emph{Iris}
-dataset is built into a number of software packages including the
-auto-installed datasets package in R \citep{r_core_team_r_2019},
-Python's scikit-learn machine learning library
-\citep{pedregosa_scikit-learn_2011}, and the SAS Sashelp library (SAS
-Institute, Cary NC), which have facilitated widespread use of the data.
-As a result, eighty-five years after the data were initially published,
-Anderson's \emph{Iris} measurements are ubiquitous in statistics,
-computational methods, and data science courses and materials.
+structural dimensions (length and width) for 50 flowers from three
+\emph{Iris} species: \emph{Iris setosa}, \emph{Iris versicolor}, and
+\emph{Iris virginica} \citep{anderson_irises_1935}. The manageable but
+non-trivial size (5 variables and 150 total observations) and
+characteristics of the dataset, including linear relationships and
+multivariate normality, have made the dataset amenable for introducing a
+wide range of statistical methods including data wrangling,
+visualization, linear modeling, multivariate analyses, and machine
+learning. Anderson's \emph{Iris} data is built into a number of software
+packages including the auto-installed datasets package in R
+\citep{r_core_team_r_2019}, Python's scikit-learn machine learning
+library \citep{pedregosa_scikit-learn_2011}, and the SAS Sashelp library
+(SAS Institute, Cary NC), which have facilitated widespread use of the
+\emph{Iris} data. As a result, eighty-six years after the data were
+initially published, Anderson's \emph{Iris} measurements are ubiquitous
+in statistics, computational methods, software documentation, and data
+science courses and materials.
 
-There are several reasons that practitioners and educators may want to
-retire Anderson's \emph{Iris} dataset in modern data science materials,
-including:
+There are a number of reasons that modern data science practitioners and
+educators may want to move on from Anderson's \emph{Iris} dataset.
+First, Anderson's \emph{Iris} dataset lacks metadata almost entirely
+\citep{anderson_irises_1935}, which does not reinforce best practices
+and limits meaningful interpretation and discussion of research methods,
+analyses, and outcomes. Of the five variables in the dataset, two
+(\texttt{Sepal.Width} and \texttt{Sepal.Length}) are not intuitive for
+most non-botanists. Even with explanation, the difference between
+\emph{Iris} \emph{petal} and \emph{sepal} dimensions is not obvious.
+Second, the \emph{Iris} dataset contains equal sample sizes for each of
+the three species (\emph{n} = 50) with no missing values, which is
+cleaner than most real-world datasets that learners are likely to
+encounter ``in the wild.'' Third, the single factor (\texttt{Species})
+in the \emph{Iris} data limits options for analyses. Finally, due to its
+publication in the \emph{Annals of Eugenics} by eugenicist and
+statistician R.A. Fisher \citep{fisher_use_1936}, the \emph{Iris}
+dataset is burdened by a history in scientific racism, which we are
+committed to addressing through the development of new data science
+education products as described below.
 
-\begin{itemize}
-\tightlist
-\item
-  Anderson's \emph{Iris} dataset entirely lacks metadata, limiting
-  meaningful interpretation and discussion of research methods,
-  analyses, and outcomes
-\item
-  Two of its five variables, \texttt{Sepal.Width} and
-  \texttt{Sepal.Length}, are not intuitive to most users
-\item
-  In 1936, the \emph{Iris} data were published by eugenicist R.A. Fisher
-  in \emph{Annals of Eugenics}, and are now inextricably linked to
-  Fisher's eugenics work; in fact, the dataset is commonly referred to
-  as \emph{Fisher's Iris data}
-\item
-  The \emph{Iris} dataset contains only complete cases (i.e.~no missing
-  values) and equal sample sizes (\emph{n} = 50 for each \emph{Iris}
-  species) - more perfect than real-world datasets learners are likely
-  to encounter ``in the wild''
-\item
-  The single categorical variable (\texttt{Species}) limits options for
-  analyses
-\end{itemize}
-
-Given these motivations to retire \emph{Iris}, we sought to identify,
-then make readily available, a suitable alternative dataset. Considering
-\emph{Iris} usage in data science and statistics materials, we
-established the following criteria for a suitable alternative:
+Given the growing need for fresh data science-ready datasets, we sought
+to identify an alternative dataset that could be made easily accessible
+for a broad audience. After evaluating the positive and negative
+features of the \emph{Iris} dataset in data science and statistics
+materials, we established the following criteria for a suitable
+alternative:
 
@@
 \item
-  Features intuitive subjects and variables that are interesting and
-  understandable to learners across fields
+  Feature intuitive subjects and variables that are interesting and
+  understandable to learners across disciplines
 \item
@@
 \item
-  Real-world (not manufactured), modern data
+  Real-world (not manufactured) modern data
 \item
-  Similar opportunities for teaching and learning R, data science, and
-  statistical skills
+  Provides similar opportunities for teaching and learning R, data
+  science, and statistical skills
 \item
@@
 Here, we describe an alternative to Anderson's \emph{Iris} data that
-largely satisfies these criteria: a refreshing, approachable and
+largely satisfies these criteria: a refreshing, approachable, and
 charismatic dataset containing real-world morphological data for three
 \emph{Pygoscelis} penguin species that breed throughout the Western
-Antarctic Peninsula region, made available through the Long-Term
-Ecological Research Network (US LTER). By comparing data structure,
-size, and a range of analyses side-by-side for the two datasets, we
-demonstrate that the Palmer Archipelago penguin measurements are an
-ideal substitute for Anderson's \emph{Iris} data for many use cases in
-statistics and data science education.
+Antarctic Peninsula region, made available through the United States
+Long-Term Ecological Research (US LTER) Network. By comparing data
+structure, size, and a range of analyses side-by-side for the two
+datasets, we demonstrate that the Palmer Archipelago penguin data are an
+ideal substitute for Anderson's \emph{Iris} dataset for many use cases
+in statistics and data science education.
 
@@
 \(\delta\)\textsuperscript{15}N) stable isotope values of red blood
-cells for male and female adult Adélie (\emph{P. adeliae}), chinstrap
+cells for adult male and female Adélie (\emph{P. adeliae}), chinstrap
 (\emph{P. antarcticus}), and gentoo (\emph{P. papua}) penguins on three
-islands (Biscoe, Dream and Torgersen) in the Palmer Archipelago were
+islands (Biscoe, Dream and Torgersen) within the Palmer Archipelago were
 collected from 2007 - 2009 by Dr.~Kristen Gorman in collaboration with
@@
 \href{https://lternet.edu/}{US LTER Network}. For complete data
-collection methods and published analyses see Gorman et al.~(2014).
+collection methods and published analyses, see
+\citet{gorman_ecological_2014}. Throughout this paper, penguins species
+referred to as ``Adélie'', ``Chinstrap'', and ``Gentoo.''
 
-The data in the \textbf{palmerpenguins} R package are available for use
+The data in the \CRANpkg{palmerpenguins} R package are available for use
 by CC0 license (``No Rights Reserved'') in accordance with the
@@
 \hypertarget{r-package-palmerpenguins}{%
-\subsection{\texorpdfstring{R package:
-\textbf{palmerpenguins}}{R package: palmerpenguins}}\label{r-package-palmerpenguins}}
+\subsection{R package: palmerpenguins}\label{r-package-palmerpenguins}}
 
-R users can install the \textbf{palmerpenguins} package from CRAN by:
+R users can install the \CRANpkg{palmerpenguins} package from CRAN:
 
 \begin{verbatim}
-  `install.packages("palmerpenguins")`
-\end{verbatim}
-
-Alternatively, the development version of the package can be installed
-from GitHub:
-
-\begin{verbatim}
-  `remotes::install_github("allisonhorst/palmerpenguins")`
+install.packages("palmerpenguins")
 \end{verbatim}
@@
 Information, examples, and links to community-contributed materials are
-available on the \textbf{palmerpenguins} package website:
-\url{https://allisonhorst.github.io/palmerpenguins/}.
+available on the \CRANpkg{palmerpenguins} package website:
+\url{https://allisonhorst.github.io/palmerpenguins/}. See the Appendix
+for how Python and Julia users can access the same data.
 
-The \textbf{palmerpenguins} R package contains two data objects:
+The \CRANpkg{palmerpenguins} R package contains two data objects:
 \textbf{penguins\_raw} and \textbf{penguins}. The \textbf{penguins\_raw}
-data consists of all raw data for 17 variables (Appendix Table 1),
-recorded completely or in part for 344 individual penguins, accessed
-directly from EDI. As a direct alternative to Anderson's \emph{Iris}
-data we recommend using the curated data in \textbf{penguins}, which is
-a subset of \textbf{penguins\_raw} retaining all 344 observations,
-minimally updated (Appendix B) and reduced to the following eight
+data consists of all raw data for 17 variables, recorded completely or
+in part for 344 individual penguins, accessed directly from EDI
+(\textbf{penguins\_raw} properties are summarized in Appendix B). We
+generally recommend using the curated data in \textbf{penguins}, which
+is a subset of \textbf{penguins\_raw} retaining all 344 observations,
+minimally updated (Appendix A) and reduced to the following eight
 variables:
@@
   \emph{species:} a factor denoting the penguin species (Adélie,
-  chinstrap, or gentoo)
+  Chinstrap, or Gentoo)
 \item
-  \emph{island:} a factor denoting the island (in Palmer Archipelago,
-  Antarctica) where observed (Biscoe, Dream or Torgersen)
+  \emph{island:} a factor denoting the Palmer Archipelago island in
+  Antarctica where each penguin was observed (Biscoe Point, Dream
+  Island, or Torgersen Island)
 \item
@@
 \item
-  \emph{year:} an integer denoting the year of study (2007, 2008 or
+  \emph{year:} an integer denoting the year of study (2007, 2008, or
   2009)
 \end{itemize}
+
+\newpage
 
@@
 the built-in \texttt{path\_to\_file()} function in
-\textbf{palmerpenguins}. For example,
+\CRANpkg{palmerpenguins}. For example,
 
 \begin{verbatim}
-    library(tidyverse)
     library(palmerpenguins)
-    df <- read_csv(path_to_file(“penguins.csv”))
+    df <- read.csv(path_to_file("penguins.csv"))
 \end{verbatim}
@@
 will read in ``penguins.csv'' as if from an external file, thus
-automatically parsing \emph{species}, \emph{island} and \emph{sex}
-variables as characters. This option allows users opportunities to
-practice or demonstrate reading in data from a CSV, then updating
-variable class (e.g.~characters to factors).
+automatically parsing \emph{species}, \emph{island}, and \emph{sex}
+variables as characters instead of factors. This option allows users
+opportunities to practice or demonstrate reading in data from a CSV,
+then updating variable class (e.g., characters to factors).
 
-\hypertarget{other-data-access-options}{%
-\subsection{Other data access options}\label{other-data-access-options}}
+\hypertarget{comparing-iris-and-penguins}{%
+\subsection{\texorpdfstring{Comparing \textbf{iris} and
+\textbf{penguins}}{Comparing iris and penguins}}\label{comparing-iris-and-penguins}}
 
-Python: Python users can access the penguins data in the
-\textbf{seaborn} data visualization library (Waskom et al.~2017).
-Example code to load the data in Python:
+The \textbf{penguins} data in \CRANpkg{palmerpenguins} is useful and
+approachable for data science and statistics education, and is uniquely
+well-suited to replace the \textbf{iris} dataset. Comparisons presented
+are selected examples for common \textbf{iris} uses, and are not
+exhaustive.
 
-\begin{verbatim}
-    import seaborn as sns
-    df = sns.load_dataset('penguins') 
-\end{verbatim}
+\begin{Schunk}
+\begin{table}
 
-Julia: Julia users can access the penguins data in the
-\textbf{PalmerPenguins.jl} package. Example code to import the penguins
-data through \textbf{PalmerPenguins.jl}:
+\caption{\label{tab:overview-tbl}Overview comparison of penguins and iris dataset features and characteristics.}
+\centering
+\begin{tabular}[t]{lcl}
+\toprule
+Feature & iris & penguins\\
+\midrule
+Year(s) collected & 1935 & 2007 - 2009\\
+Dimensions (col x row) & 5 x 150 & 8 x 344\\
+Documentation & minimal & complete metadata\\
+Variable classes & double (4), factor (1) & double (2), int (3), factor (3)\\
+Missing values? & no (n = 0; 0.0\%) & yes (n = 19; 0.7\%)\\
+\bottomrule
+\end{tabular}
+\end{table}
 
-\begin{verbatim}
-    julia> using DataFrames
-    julia> df = DataFrame(table)
-\end{verbatim}
-
-\hypertarget{software-and-code}{%
-\subsection{Software and code}\label{software-and-code}}
-
-All analyses were performed in the R language environment using version
-3.6.2 (R Core Team 2019). Complete code for this paper is shared in the
-Supplemental Material. We acknowledge the following R packages used in
-analyses, with gratitude to developers and contributors:
-
-\begin{itemize}
-\tightlist
-\item
-  \textbf{tidyverse} (Wickham et al.~2019), for data import and cleaning
-\item
-  \textbf{ggplot2} (Wickham 2016:2), for data visualizations
-\item
-  \textbf{here} (Müller 2017), for file path control
-\item
-  \textbf{kableExtra} (Zhu 2019) for finalized tables
-\item
-  \textbf{gt} (Iannone, Cheng, and Schloerke 2020) for finalized tables
-\item
-  \textbf{GGally} (Schloerke et al.~2020), for pairs plots
-\item
-  \textbf{patchwork} (Pedersen 2019), for compound figures
-\item
-  \textbf{shadowtext} (Yu 2019), to add a background color to text
-  labels
-\item
-  \textbf{recipes} (Kuhn and Wickham 2020), for data pre-processing
-\item
-  \textbf{base} and \textbf{stats} (R Core Team 2019) for various
-  analyses throughout
-\item
-  \textbf{pkgdown} (ref) was used to build the package website
-\end{itemize}
-
-\hypertarget{selected-comparisons-between-iris-and-penguins}{%
-\subsection{\texorpdfstring{Selected comparisons between \textbf{iris}
-and
-\textbf{penguins}}{Selected comparisons between iris and penguins}}\label{selected-comparisons-between-iris-and-penguins}}
-
-The \textbf{penguins} data in \textbf{palmerpenguins} is generally
-useful and approachable for data science and statistics education, and
-is uniquely well-suited to replace the \textbf{iris} dataset.
-Comparisons presented are selected examples for common \textbf{iris}
-uses, and are not exhaustive.
+\end{Schunk}
 
@@
 
-Both \textbf{iris} and \textbf{penguins} are in tidy format {[}ref{]}
-with each column denoting a single variable and each row containing
-measurements for a single \emph{Iris} flower or penguin. The two
-datasets are comparable in size: dimensions (columns × rows) are 5 × 150
-and 8 × 344 for \textbf{iris} and \textbf{penguins}, respectively, and
-sample sizes within species are similar (Table 1). Notably, sample sizes
-differ for the three penguin species, while sample sizes in
-\textbf{iris} are equal (n = 50 for each Iris species). Multiple factor
-variables in \textbf{penguins} (\emph{species}, \emph{island} and
-\emph{sex}) along with \emph{year} create additional opportunities for
-grouping, compared to the single factor (\emph{species}) in
-\textbf{iris}.
+Both \textbf{iris} and \textbf{penguins} are in tidy format
+\citep{wickham_tidy_2014} with each column denoting a single variable
+and each row containing measurements for a single \emph{Iris} flower or
+penguin, respectively. The two datasets are comparable in size:
+dimensions (columns × rows) are 5 × 150 and 8 × 344 for \textbf{iris}
+and \textbf{penguins}, respectively, and sample sizes within species are
+similar (Table 1).
+
+Notably, while sample sizes in \textbf{iris} across species are all the
+same, sample sizes in \textbf{penguins} differ across the three species.
+The inclusion of three factor variables in \textbf{penguins}
+(\emph{species}, \emph{island}, and \emph{sex}), along with \emph{year},
+create additional opportunities for grouping, faceting, and analysis
+compared to the single factor (\emph{Species}) in \textbf{iris}.
 
@@
 \textbf{penguins} dataset contains a small number of missing values
-(n\textsubscript{missing} = 19, out of 2,752 total values). Missing
-values and unequal sample sizes are common in real-world data, and we
-believe add learning value to \textbf{penguins}.
+(n\textsubscript{missing} = 19, out of 2752 total values). Missing
+values and unequal sample sizes are common in real-world data, and adds
+learning opportunities to the \textbf{penguins} dataset.
 
@@
 
-\caption{\label{tab:unnamed-chunk-1}Grouped sample size for iris (by species; n = 150 total) and penguins (by species and sex; n = 344 total).
+\caption{\label{tab:counts-tbl}Grouped sample size for iris (by species; n = 150 total) and penguins (by species and sex; n = 344 total).
 Penguins can be further grouped by variables for island and study year.}
 \centering
-\begin{tabular}[t]{l|c|l|c|c|c}
-\hline
-\multicolumn{2}{c|}{Iris sample size (by species)} & \multicolumn{4}{c}{Penguin sample size (by species and sex)} \\
-\cline{1-2} \cline{3-6}
+\begin{tabular}[t]{lclccc}
+\toprule
+\multicolumn{2}{c}{Iris sample size (by species)} & \multicolumn{4}{c}{Penguin sample size (by species and sex)} \\
+\cmidrule(l{3pt}r{3pt}){1-2} \cmidrule(l{3pt}r{3pt}){3-6}
 Iris species & Sample size & Penguin species & Female & Male & NA\\
-\hline
+\midrule
 setosa & 50 & Adélie & 73 & 73 & 6\\
-\hline
-versicolor & 50 & chinstrap & 34 & 34 & 0\\
-\hline
-virginica & 50 & gentoo & 58 & 61 & 5\\
-\hline
+versicolor & 50 & Chinstrap & 34 & 34 & 0\\
+virginica & 50 & Gentoo & 58 & 61 & 5\\
+\bottomrule
 \end{tabular}
@@
 
-Distributions, relationships between variables and clustering can be
+Distributions, relationships between variables, and clustering can be
 visually explored between species for the four structural size
 measurements in \textbf{penguins} (flipper length, body mass, bill
-length and depth; Figure \ref{fig:penguins-pairs}) and \textbf{iris}
+length and depth; Figure \ref{fig:penguin-pairs}) and \textbf{iris}
 (sepal width and length, petal width and length; Figure
@@
 \begin{figure}
-\includegraphics{penguins_files/figure-latex/penguins-pairs-1} \caption[Distribution and correlations for numeric variables in the penguins data (flipper length (mm), body mass, (g) bill length (mm) and bill depth (mm)) for the three observed species]{Distribution and correlations for numeric variables in the penguins data (flipper length (mm), body mass, (g) bill length (mm) and bill depth (mm)) for the three observed species: gentoo (green, triangles); chinstrap (blue, circles); and Adélie (orange, squares).
-Correlations are Pearson's r (*p < 0.05; **p < 0.01; ***p < 0.001).}\label{fig:penguins-pairs}
+
+{\centering \includegraphics[width=1\linewidth]{figs/penguin-pairs-1}
+
+}
+
+\caption[Distributions and correlations for numeric variables in the penguins data (flipper length (mm), body mass (g), bill length (mm) and bill depth (mm)) for the three observed species]{Distributions and correlations for numeric variables in the penguins data (flipper length (mm), body mass (g), bill length (mm) and bill depth (mm)) for the three observed species: Gentoo (green, triangles); Chinstrap (blue, circles); and Adélie (orange, squares).
+Significance levels indicated for bivariate correlations are: *p < 0.05; **p < 0.01; ***p < 0.001.}\label{fig:penguin-pairs}
 \end{figure}
@@
 \begin{figure}
-\includegraphics{penguins_files/figure-latex/iris-pairs-1} \caption[Distribution and correlations for numeric variables in the iris data (petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm)) for the three included Iris species]{Distribution and correlations for numeric variables in the iris data (petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm)) for the three included Iris species: setosa (light gray, circles); versicolor (dark gray, triangles); and virginica (black, squares).
-Correlations are Pearson's r (*p < 0.05; **p < 0.01; ***p < 0.001).}\label{fig:iris-pairs}
+
+{\centering \includegraphics[width=1\linewidth]{figs/iris-pairs-1}
+
+}
+
+\caption[Distributions and correlations for numeric variables in the iris data (petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm)) for the three included Iris species]{Distributions and correlations for numeric variables in the iris data (petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm)) for the three included Iris species: setosa (light gray, circles); versicolor (dark gray, triangles); and virginica (black, squares).
+Significance levels indicated for bivariate correlations are: *p < 0.05; **p < 0.01; ***p < 0.001.}\label{fig:iris-pairs}
 \end{figure}
@@
 
-There exist numerous opportunities to explore linear relationships and
-correlations in \textbf{penguins} and \textbf{iris} both within and
-across species (Figures \ref{fig:penguins-pairs} \&
-\ref{fig:iris-pairs}). Here, we highlight one comparison that is
-uniquely similar: petal dimensions (petal length versus petal width)
-from \textbf{iris}, and penguin size (flipper length versus body mass)
-in \textbf{penguins} (\ref{fig:linear-example}). The overall trend
+Both \textbf{penguins} and \textbf{iris} offer numerous opportunities to
+explore linear relationships and correlations, within and across species
+(Figures \ref{fig:penguin-pairs} \& \ref{fig:iris-pairs}). A bivariate
+scatterplot made with the \textbf{iris} dataset reveals a clear linear
+relationship between petal length and petal width. Using
+\textbf{penguins} (\ref{fig:linear-example}), we can create a uniquely
+similar scatterplot with flipper length and body mass. The overall trend
 across all three species is approximately linear for both \textbf{iris}
@@
 
-{\centering \includegraphics[width=6in]{fig/linear_example}
+{\centering \includegraphics[width=1\linewidth]{figs/linear-example-1}
 
@@
 
-\caption[Representative linear relationships for penguin flipper length (mm) and body mass (g) for Adélie (orange circles), chinstrap (blue triangles), and gentoo (green squares) penguins (A), and iris petal length (cm) and width (cm) for setosa (light gray circles), versicolor (dark gray triangles) and virginica (black squares) irises (B)]{Representative linear relationships for penguin flipper length (mm) and body mass (g) for Adélie (orange circles), chinstrap (blue triangles), and gentoo (green squares) penguins (A), and iris petal length (cm) and width (cm) for setosa (light gray circles), versicolor (dark gray triangles) and virginica (black squares) irises (B).
+\caption[Representative linear relationships for (A)]{Representative linear relationships for (A): penguin flipper length (mm) and body mass (g) for Adélie (orange circles), Chinstrap (blue triangles), and Gentoo (green squares) penguins; (B): iris petal length (cm) and width (cm) for setosa (light gray circles), versicolor (dark gray triangles) and virginica (black squares) irises.
 Within-species linear model is visualized for each penguin or iris species.}\label{fig:linear-example}
@@
 particularly, the much smaller petals for \emph{Iris setosa} - compared
-to penguins, in which Adélie and chinstrap penguins are largely
+to penguins, in which Adélie and Chinstrap penguins are largely
 overlapping in body size (body mass and flipper length), and are both
-generally smaller than gentoos.
+generally smaller than Gentoo penguins.
 
@@
 variable. While often taught and discussed in statistics courses,
-finding a real-world and approachable example of Simpson's paradox can
-be a challenge. Here, we show one (of several possible; see Figure
-\ref{fig:penguins-pairs}) Simpson's Paradox examples in
-\textbf{penguins}: exploring bill dimensions with and without
-\emph{species} included (Figure \ref{fig:simpsons}). When penguin
-\emph{species} is omitted (Figure \ref{fig:simpsons} A), bill length and
-depth appear negatively correlated overall. The trend is reversed when
-\emph{species} is included, revealing an obviously positive correlation
-between bill length and bill depth within species (Figure
-\ref{fig:simpsons} B).
+finding a real-world and approachable example of Simpson's Paradox can
+be a challenge. Here, we show one (of several possible - see Figure
+\ref{fig:penguin-pairs}) Simpson's Paradox examples in
+\textbf{penguins}: exploring bill dimensions with and without species
+included (Figure \ref{fig:simpsons}). When penguin species is omitted
+(Figure \ref{fig:simpsons} A), bill length and depth appear negatively
+correlated overall. The trend is reversed when species is included,
+revealing an obviously positive correlation between bill length and bill
+depth within species (Figure \ref{fig:simpsons} B).
 
@@
 
-{\centering \includegraphics[width=6in]{fig/simpson_gg}
+{\centering \includegraphics[width=1\linewidth]{figs/simpsons-1}
 
@@
 
-\caption[Trends for penguin bill dimensions (bill length and bill depth, millimeters) if the ‘species’ variable is excluded (A) or included (B), illustrating Simpson’s Paradox]{Trends for penguin bill dimensions (bill length and bill depth, millimeters) if the ‘species’ variable is excluded (A) or included (B), illustrating Simpson’s Paradox.
+\caption[Trends for penguin bill dimensions (bill length and bill depth, millimeters) if the species variable is excluded (A) or included (B), illustrating Simpson’s Paradox]{Trends for penguin bill dimensions (bill length and bill depth, millimeters) if the species variable is excluded (A) or included (B), illustrating Simpson’s Paradox.
 Note: linear regression for bill dimensions without including species in (A) is ill-advised; the linear trendline is only included to visualize trend reversal for Simpson’s Paradox when compared to (B).}\label{fig:simpsons}
@@
 Principal component analysis (PCA) is a dimensional reduction method
-commonly used to explore patterns in multivariate data. PCA tutorials
-frequently employ the \textbf{iris} dataset, which is useful for
-teaching the method due to multivariate normality, and clear,
-approachable outcomes for variable loadings and clustering.
+commonly used to explore patterns in multivariate data. The
+\textbf{iris} dataset frequently appears in PCA tutorials due to
+multivariate normality and clear interpretation of variable loadings and
+clustering.
 
@@
 prior to PCA) reveals highly similar results (Figure \ref{fig:pca}). For
-both datasets, one species is distinct (gentoo penguins, and setosa
-irises) while the other two species (chinstrap/Adélie and
+both datasets, one species is distinct (Gentoo penguins, and setosa
+irises) while the other two species (Chinstrap/Adélie and
 versicolor/virginica) appear somewhat overlapping in the first two
-principal components (Figure \ref{fig:pca} A,B). Variance explained by
-each principal component (PC) is similar, particularly for PC1 and PC2:
-for \textbf{penguins}, 88.15\% of total variance is captured by the
-first two PCs, compared to 95.81\% for \textbf{iris}, with a similarly
-large percentage of variance captured by PC1 and PC2 in each (Figure
+principal components (Figure \ref{fig:pca} A,B). Screeplots reveal that
+the variance explained by each principal component (PC) is very similar
+across the two datasets, particularly for PC1 and PC2: for
+\textbf{penguins}, 88.15\% of total variance is captured by the first
+two PCs, compared to 95.81\% for \textbf{iris}, with a similarly large
+percentage of variance captured by PC1 and PC2 in each (Figure
 \ref{fig:pca} C,D).
@@
 
-{\centering \includegraphics[width=6in]{fig/pca_plots}
+{\centering \includegraphics[width=1\linewidth]{figs/pca-1}
 
@@
 
-\caption[Principal component analysis biplots and scree plots for structural size measurements in penguins (A,C) and iris (B,D), revealing similarities in multivariate patterns, variable loadings, and variance explained by each component]{Principal component analysis biplots and scree plots for structural size measurements in penguins (A,C) and iris (B,D), revealing similarities in multivariate patterns, variable loadings, and variance explained by each component.
-For penguins, variables are flipper length (mm), body mass (g), bill length (mm) and bill depth (mm); groups are visualized by species (Adélie = orange circles, chinstrap = blue triangles, gentoo = green squares).
+\caption[Principal component analysis biplots and screeplots for structural size measurements in penguins (A,C) and iris (B,D), revealing similarities in multivariate patterns, variable loadings, and variance explained by each component]{Principal component analysis biplots and screeplots for structural size measurements in penguins (A,C) and iris (B,D), revealing similarities in multivariate patterns, variable loadings, and variance explained by each component.
+For penguins, variables are flipper length (mm), body mass (g), bill length (mm) and bill depth (mm); groups are visualized by species (Adélie = orange circles, Chinstrap = blue triangles, Gentoo = green squares).
 For iris, variables are petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm); groups are visualized by species (setosa = light gray circles, versicolor = dark gray triangles, virginica = black squares).
-Values above scree plot columns (C,D) indicate percent of total variance explained by each of the four principal components.}\label{fig:pca}
+Values above screeplot columns (C,D) indicate percent of total variance explained by each of the four principal components.}\label{fig:pca}
 \end{figure}
@@
 Unsupervised clustering by k-means is a common and popular entryway to
-machine learning and classification, and frequently employs the
-\textbf{iris} data for introductory examples. The \textbf{penguins} data
-provides similar opportunities for introducing k-means clustering. For
-simplicity, we compare k-means clustering using only two variables for
-each dataset: for \textbf{iris}, petal width and petal length, and for
-\textbf{penguins}, bill length and bill depth. All variables are scaled
-prior to k-means. Three clusters (\emph{k} = 3) are specified for each
-since there are three species of both \emph{Iris} (\emph{setosa},
-\emph{versicolor}, and \emph{virginica}) and penguins (Adélie, chinstrap
-and gentoo).
+machine learning and classification, and again, the \textbf{iris}
+dataset is frequently used in introductory examples. The
+\textbf{penguins} data provides similar opportunities for introducing
+k-means clustering. For simplicity, we compare k-means clustering using
+only two variables for each dataset: for \textbf{iris}, petal width and
+petal length, and for \textbf{penguins}, bill length and bill depth. All
+variables are scaled prior to k-means. Three clusters (\emph{k} = 3) are
+specified for each, since there are three species of both \emph{Iris}
+(\emph{setosa}, \emph{versicolor}, and \emph{virginica}) and penguins
+(Adélie, Chinstrap and Gentoo).
 
 K-means clustering with penguin bill dimensions and iris petal
-dimensions yields largely distinct clusters each dominated by one
+dimensions yields largely distinct clusters, each dominated by one
 species (\ref{fig:kmeans}). For iris petal dimensions, k-means yields a
-perfectly separated cluster (Cluster 1) containing all 50 setosa iris
+perfectly separated cluster (Cluster 3) containing all 50 setosa iris
 observations and zero misclassified virginica or versicolor irises
@@
 Adélie penguins (orange observations in Figure \ref{fig:kmeans} A): 147
-(out of 151) Adélie penguins are assigned to Cluster 1, zero are
-assigned to Cluster 2, and 4 are assigned to the chinstrap-dominated
-Cluster 3 (Table 2). Only 5 (of 68) chinstrap penguins and 1 (of 123)
-gentoo penguins are assigned to the Adélie-dominated Cluster 1 (Table
+(out of 151) Adélie penguins are assigned to Cluster 3, zero are
+assigned to Cluster 1, and 4 are assigned to the Chinstrap-dominated
+Cluster 2 (Table 2). Only 5 (of 68) Chinstrap penguins and 1 (of 123)
+Gentoo penguins are assigned to the Adélie-dominated Cluster 3 (Table
 2).
@@
 
-{\centering \includegraphics[width=6in]{fig/kmeans}
+{\centering \includegraphics[width=1\linewidth]{figs/kmeans-1}
 
@@
 
-\caption{\label{tab:unnamed-chunk-3}K-means cluster assignments by species based on penguin bill length (mm) and depth (mm), and iris petal length (cm) and width (cm).}
+\caption{\label{tab:unnamed-chunk-1}K-means cluster assignments by species based on penguin bill length (mm) and depth (mm), and iris petal length (cm) and width (cm).}
 \centering
-\begin{tabular}[t]{c|c|c|c|c|c|c|c}
-\hline
-\multicolumn{4}{c|}{Penguins cluster assignments} & \multicolumn{4}{c}{Iris cluster assignments} \\
-\cline{1-4} \cline{5-8}
-Cluster & Adélie & chinstrap & gentoo & Cluster & setosa & versicolor & virginica\\
-\hline
-1 & 4 & 54 & 6 & 1 & 50 & NA & NA\\
-\hline
-2 & 147 & 5 & 1 & 2 & NA & 48 & 4\\
-\hline
-3 & NA & 9 & 116 & 3 & NA & 2 & 46\\
-\hline
+\begin{tabular}[t]{cccccccc}
+\toprule
+\multicolumn{4}{c}{Penguins cluster assignments} & \multicolumn{4}{c}{Iris cluster assignments} \\
+\cmidrule(l{3pt}r{3pt}){1-4} \cmidrule(l{3pt}r{3pt}){5-8}
+Cluster & Adélie & Chinstrap & Gentoo & Cluster & setosa & versicolor & virginica\\
+\midrule
+1 & 0 & 9 & 116 & 1 & 0 & 2 & 46\\
+2 & 4 & 54 & 6 & 2 & 0 & 48 & 4\\
+3 & 147 & 5 & 1 & 3 & 50 & 0 & 0\\
+\bottomrule
 \end{tabular}
@@
 Archipelago \emph{Pygoscelis} penguins, available as \textbf{penguins}
-in the \textbf{palmerpenguins} R package, offer a near drop-in
+in the \CRANpkg{palmerpenguins} R package, offer a near drop-in
 replacement for \textbf{iris} in a number of common use cases for data
@@
 missing values, unequal sample sizes, and Simpson's paradox examples.
+Importantly, the \textbf{penguins} dataset encompasses real-world
+information derived from several charismatic marine predator species
+with regional breeding populations notably responding to environmental
+change occurring throughout the Western Antarctic Peninsula region of
+the Southern Ocean (see \citet{bestelmeyer_analysis_2011},
+\citet{gorman_ecological_2014}, \citet{gorman_integrative_2015},
+\citet{gorman_population_2017}). Thus, the \textbf{penguins} dataset can
+facilitate discussions more broadly on biodiversity responses to global
+change - a contemporary and critical topic in ecology, evolution, and
+the environmental sciences.
 
 \hypertarget{appendix}{%
-\subsection{Appendix}\label{appendix}}
+\section*{Appendix}\label{appendix}}
+\addcontentsline{toc}{section}{Appendix}
 
-\hypertarget{appendix-a}{%
-\subsubsection{Appendix A}\label{appendix-a}}
-
-\hypertarget{appendix-b}{%
-\subsubsection{Appendix B}\label{appendix-b}}
+\hypertarget{penguins-data-processing}{%
+\subsection{Penguins data processing}\label{penguins-data-processing}}
 
@@
 \item
-  All variable names are converted to lower snake case
+  All variable names are converted to lower snake case (e.g.~from
+  \texttt{Flipper\ Length\ (mm)} to \texttt{flipper\_length\_mm})
 \item
   Entries in \emph{species} are truncated to only include the common
-  name (e.g.~``gentoo'', instead of ``gentoo penguin (Pygoscelis
+  name (e.g.~``Gentoo'', instead of ``gentoo penguin (Pygoscelis
   papua)'')
@@
 
+\hypertarget{summary-of-the-penguins_raw-dataset}{%
+\subsection{\texorpdfstring{Summary of the \textbf{penguins\_raw}
+dataset}{Summary of the penguins\_raw dataset}}\label{summary-of-the-penguins_raw-dataset}}
+
+\begin{Schunk}
+
+\begin{tabular}{ll}
+\toprule
+Feature & penguins\_raw\\
+\midrule
+Year(s) collected & 2007 - 2009\\
+Dimensions (col x row) & 17 x 344\\
+Documentation & complete metadata\\
+Variable classes & character (9), Date (1), numeric (7)\\
+Missing values? & yes (n = 336; 5.7\%)\\
+\bottomrule
+\end{tabular}
+
+\end{Schunk}
+
+\hypertarget{other-data-sources}{%
+\subsection{Other data sources}\label{other-data-sources}}
+
+Python: Python users can access the penguins data in the
+\textbf{seaborn} data visualization library
+\citep{michael_waskom_2017_883859}. Example code to load the data in
+Python:
+
+\begin{verbatim}
+    import seaborn as sns
+    df = sns.load_dataset('penguins') 
+\end{verbatim}
+
+Julia: Julia users can access the penguins data in the
+\textbf{PalmerPenguins.jl} package. Example code to import the penguins
+data through \textbf{PalmerPenguins.jl} (more information on
+\textbf{PalmerPenguins.jl} from David Widmann can be found
+\href{https://github.com/devmotion/PalmerPenguins.jl}{here}):
+
+\begin{verbatim}
+    julia> using PalmerPenguins
+    julia> table = PalmerPenguins.load()
+\end{verbatim}
+
+\hypertarget{acknowledgements}{%
+\subsection{Acknowledgements}\label{acknowledgements}}
+
+All analyses were performed in the R language environment using version
+3.6.2 \citep{r_core_team_r_2019}. Complete code for this paper is shared
+in the Supplemental Material. We acknowledge the following R packages
+used in analyses, with gratitude to developers and contributors:
+
+\begin{itemize}
+\tightlist
+\item
+  \CRANpkg{tidyverse} \citep{wickham_welcome_2019}: for data import and
+  cleaning
+\item
+  \CRANpkg{ggplot2} \citep{wickham_ggplot2_2016}: for data
+  visualizations
+\item
+  \CRANpkg{kableExtra} \citep{zhu_kableextra_2019}: for finalized tables
+\item
+  \CRANpkg{GGally} \citep{schloerke_ggally_2020}: for pairs plots
+\item
+  \CRANpkg{patchwork} \citep{pedersen_patchwork_2019}: for compound
+  figures
+\item
+  \CRANpkg{shadowtext} \citep{yu_shadowtext_2019}: to add a background
+  color to text labels
+\item
+  \CRANpkg{recipes} \citep{kuhn_recipes_2020}: for data pre-processing
+\item
+  \CRANpkg{pkgdown} \citep{wickham_pkgdown_2020}: to build the package
+  website
+\item
+  \pkg{base} and \pkg{stats} \citep{r_core_team_r_2019}: for various
+  analyses throughout
+\end{itemize}
+
 \bibliography{RJreferences}
@@
 Allison M. Horst\\
-Bren School of Environmental Science and Management\\
+Bren School of Environmental Science and Management\\%
 University of California, Santa Barbara\\ Santa Barbara, CA 93106-5131\\
+%
+%
+%
+\\\href{mailto:ahorst@ucsb.edu}{\nolinkurl{ahorst@ucsb.edu}}
 }
-\href{mailto:ahorst@ucsb.edu}{\nolinkurl{ahorst@ucsb.edu}}
 
 \address{%
-Alison P. Hill\\
-RStudio, PBC\\
+Alison Presmanes Hill\\
+RStudio, PBC\\%
 250 Northern Ave\\ Boston, MA 02210\\
+%
+%
+%
+\\\href{mailto:alison@rstudio.com}{\nolinkurl{alison@rstudio.com}}
 }
-\href{mailto:alison@rstudio.com}{\nolinkurl{alison@rstudio.com}}
 
@@
 Kristen B. Gorman\\
-University of Alaska Fairbanks College of Fisheries and Ocean Sciences\\
-2150 Koyukuk Drive\\ 245 O'Neill Building\\ Fairbanks, AK 99775-7220\\
+College of Fisheries and Ocean Sciences\\%
+University of Alaska Fairbanks\\ 2150 Koyukuk Drive\\ 245 O'Neill
+Building\\ Fairbanks, AK 99775-7220\\
+%
+%
+%
+\\\href{mailto:kbgorman@alaska.edu}{\nolinkurl{kbgorman@alaska.edu}}
 }
-\href{mailto:kbgorman@alaska.edu}{\nolinkurl{kbgorman@alaska.edu}}

### 2021-04-27 penguins.R

@@ -1,20 +1,4 @@
-## ----setup, include=FALSE-----------------------------------------------------
-library(tidyverse)
-library(here)
-library(palmerpenguins)
-library(kableExtra)
-library(patchwork)
-library(gt)
-library(GGally)
-library(recipes)
-library(shadowtext)
-library(broom)
-knitr::opts_chunk$set(echo = TRUE)
-
-theme_set(theme_minimal())
-
-
-## ---- echo = FALSE------------------------------------------------------------
+## ----overview-tbl-------------------------------------------------------------
 overview_compare <- tribble(
   ~Feature, ~iris, ~penguins,
   "Year(s) collected", "1935", "2007 - 2009",
@@ -28,7 +12,7 @@ overview_compare %>%
   kable(caption = "Overview comparison of penguins and iris dataset features and characteristics.")
 
 
-## ---- echo = FALSE------------------------------------------------------------
+## ----counts-tbl---------------------------------------------------------------
 iris_counts <- iris %>% 
   count(Species) %>% 
   rename(`Iris species` = Species)
@@ -37,8 +21,6 @@ penguin_counts <- penguins %>%
   mutate(species = as.character(species)) %>% 
   mutate(species = case_when(
     species == "Adelie" ~ "Adélie", 
-    species == "Chinstrap" ~ "chinstrap",
-    species == "Gentoo" ~ "gentoo",
     TRUE ~ species
   )) %>% 
   mutate(sex = str_to_title(sex)) %>% 
@@ -59,83 +41,36 @@ iris_penguin_n %>%
 
 
 
-## ---- penguins-pairs, fig.cap = "Distribution and correlations for numeric variables in the penguins data (flipper length (mm), body mass, (g) bill length (mm) and bill depth (mm)) for the three observed species: gentoo (green, triangles); chinstrap (blue, circles); and Adélie (orange, squares). Correlations are Pearson's r (*p < 0.05; **p < 0.01; ***p < 0.001).", warning = FALSE, message = FALSE, echo = FALSE----
-penguin_pairs <- penguins %>%
-  mutate(species = as.character(species)) %>% 
-  mutate(species = case_when(
-    species == "Adelie" ~ "Adélie", 
-    TRUE ~ species
-  )) %>% 
-  select(species, where(is.numeric)) %>% 
-  ggpairs(aes(color = species, shape = species),
-          columns = c("flipper_length_mm", "body_mass_g", 
-                      "bill_length_mm", "bill_depth_mm"),
[…]

### 2021-04-27 penguins_plots.R

@@ -1,103 +1,23 @@
----
-title: 'Palmer Penguins manuscript code'
-output:
-  html_document:
-    number_sections: yes
-  pdf_document: default
----
-
-```{r setup, include=FALSE}
-library(tidyverse)
-library(here)
-library(palmerpenguins)
-library(kableExtra)
-library(patchwork)
-library(gt)
-library(GGally)
-library(recipes)
-library(shadowtext)
-library(broom)
-knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
-
-theme_set(theme_minimal())
-```
-
-# MANUSCRIPT CODE
-
-## Table 1: Overview summary of penguins and iris datasets
-
-```{r, echo = TRUE}
-# Features overview for penguins and iris: 
-overview_compare <- tribble(
-  ~Feature, ~iris, ~penguins,
-  "Year(s) collected", "1935", "2007 - 2009",
-  "Dimensions (col x row)", "5 x 150", "8 x 344",
-  "Documentation", "minimal", "complete metadata",
-  "Variable classes", "double (4), factor (1)", "double (2), int (3), factor (3)",
-  "Missing values?", "no (n = 0; 0.0%)", "yes (n = 19; 0.7%)"
-)
-
-overview_compare %>% 
-  kable(caption = "Overview comparison of penguins and iris dataset features and characteristics.")
-```
-
-## Table 2: Counts comparison between iris and penguins
-
-```{r compare_counts, echo = TRUE}
-
-# Count iris observations by species:
-iris_counts <- iris %>% 
-  count(Species) %>% 
-  rename(`Iris species` = Species)
-
-# Count penguins observations by species and sex:
-penguin_counts <- penguins %>% 
-  mutate(species = as.character(species)) %>% 
-  mutate(species = case_when(
-    species == "Adelie" ~ "Adélie", 
-    TRUE ~ species
-  )) %>% 
-  mutate(sex = str_to_title(sex)) %>% 
-  group_by(species, sex) %>% 
-  tally() %>% 
-  pivot_wider(names_from = sex, values_from = n) %>% 
-  replace_na(list(`NA` = 0)) %>% 
-  as.data.frame() %>% 
-  rename(`Penguin species` = species)
-
-# Combine into a single table
-iris_penguin_n <- cbind(iris_counts, penguin_counts) %>%
-   rename(`Sample size` = n)
-
-# Finalize combined counts table:
-iris_penguin_n %>% 
[…]

### 2021-04-27 penguins.Rmd
Reformat PCA figure dimensions, update Python import, add LTER & Gorman citations for data (#22)
* resize PCA figure

* update LTER and Gorman citations for datasets, and change to PyPI palmerpenguins for Python rec

@@
 
-- Adélie penguin data (LTER and Gorman 2020a): [KNB-LTER Data Package 219.5](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.219.5)
-- Chinstrap penguin data (LTER and Gorman 2020b): [KNB-LTER Data Package 221.5](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.221.6)
-- Gentoo penguin data (LTER and Gorman 2020c): [KNB-LTER Data Package 220.5](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.220.5)
+- Adélie penguin data [@lter_structural_2020]: [KNB-LTER Data Package 219.5](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.219.5)
+- Gentoo penguin data [@lter_structural_2020-1]: [KNB-LTER Data Package 220.5](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.220.5)
+- Chinstrap penguin data [@lter_structural_2020-2]: [KNB-LTER Data Package 221.5](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.221.6)
 
@@
 
-```{r, pca, width = 8, height = 8, fig.cap = "Principal component analysis biplots and screeplots for structural size measurements in penguins (A,C) and iris (B,D), revealing similarities in multivariate patterns, variable loadings, and variance explained by each component.
+```{r, pca, fig.width = 8, fig.height = 8, fig.cap = "Principal component analysis biplots and screeplots for structural size measurements in penguins (A,C) and iris (B,D), revealing similarities in multivariate patterns, variable loadings, and variance explained by each component.
 For penguins, variables are flipper length (mm), body mass (g), bill length (mm) and bill depth (mm); groups are visualized by species (Adélie = orange circles, Chinstrap = blue triangles, Gentoo = green squares).
@@
 
-## Other data sources
+## palmerpenguins for other programming languages
 
-Python: Python users can access the penguins data in the **seaborn** data visualization library [@michael_waskom_2017_883859]. Example code to load the data in Python:
+Python: Python users can load the palmerpenguins datasets into their Python environment using the following code to install and access data in the [palmerpenguins Python package](https://pypi.org/project/palmerpenguins/):
 
 ```
-    import seaborn as sns
-    df = sns.load_dataset('penguins') 
+    pip install palmerpenguins
+    from palmerpenguins import load_penguins
+    penguins = load_penguins()
 ```

### 2021-04-27 penguins.tex
Reformat PCA figure dimensions, update Python import, add LTER & Gorman citations for data (#22)
* resize PCA figure

* update LTER and Gorman citations for datasets, and change to PyPI palmerpenguins for Python rec

@@
 \item
-  Adélie penguin data (LTER and Gorman 2020a):
+  Adélie penguin data \citep{lter_structural_2020}:
   \href{https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.219.5}{KNB-LTER
@@
 \item
-  Chinstrap penguin data (LTER and Gorman 2020b):
+  Gentoo penguin data \citep{lter_structural_2020-1}:
+  \href{https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.220.5}{KNB-LTER
+  Data Package 220.5}
+\item
+  Chinstrap penguin data \citep{lter_structural_2020-2}:
   \href{https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.221.6}{KNB-LTER
   Data Package 221.5}
-\item
-  Gentoo penguin data (LTER and Gorman 2020c):
-  \href{https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.220.5}{KNB-LTER
-  Data Package 220.5}
 \end{itemize}
@@
 
-\hypertarget{other-data-sources}{%
-\subsection{Other data sources}\label{other-data-sources}}
+\hypertarget{palmerpenguins-for-other-programming-languages}{%
+\subsection{palmerpenguins for other programming
+languages}\label{palmerpenguins-for-other-programming-languages}}
 
-Python: Python users can access the penguins data in the
-\textbf{seaborn} data visualization library
-\citep{michael_waskom_2017_883859}. Example code to load the data in
-Python:
+Python: Python users can load the palmerpenguins datasets into their
+Python environment using the following code to install and access data
+in the \href{https://pypi.org/project/palmerpenguins/}{palmerpenguins
+Python package}:
 
 \begin{verbatim}
-    import seaborn as sns
-    df = sns.load_dataset('penguins') 
+    pip install palmerpenguins
+    from palmerpenguins import load_penguins
+    penguins = load_penguins()
 \end{verbatim}

### 2021-04-27 penguins.Rmd
fix Palmer Station LTER bibliography entries (#23)

@@
 - Gentoo penguin data [@lter_structural_2020-1]: [KNB-LTER Data Package 220.5](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.220.5)
-- Chinstrap penguin data [@lter_structural_2020-2]: [KNB-LTER Data Package 221.5](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.221.6)
+- Chinstrap penguin data [@lter_structural_2020-2]: [KNB-LTER Data Package 221.6](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.221.6)
 

### 2021-04-29 penguins.Rmd
Minor changes before submitting (#24)
* tiny proofreading edits

* final commit before submission

@@
 The **penguins** data provides an approachable, charismatic, and near drop-in replacement for **iris** with topical relevance global climate change and environmental impacts on marine predators.
-Since the release on CRAN in July 2020, the **palmerpenguins** package has been downloaded over 102,000 times, highlighting the demand and widespread adoption of this viable **iris** alternative.
+Since the release on CRAN in July 2020, the **palmerpenguins** package has been downloaded over 105,000 times, highlighting the demand and widespread adoption of this viable **iris** alternative.
  We directly compare the **iris** and **penguins** datasets for selected analyses to demonstrate that R users, in particular teachers and learners currently using **iris**, can switch to the Palmer Archipelago penguins for many use cases including data wrangling, visualization, linear modeling, multivariate analysis (e.g., PCA), cluster analysis and classification (e.g., by k-means).
@@
 Unlike **iris**, which contains only complete cases, the **penguins** dataset contains a small number of missing values (n~missing~ = 19, out of `r ncol(palmerpenguins::penguins) * nrow(palmerpenguins::penguins)` total values).
-Missing values and unequal sample sizes are common in real-world data, and adds learning opportunities to the **penguins** dataset.
+Missing values and unequal sample sizes are common in real-world data, and create added learning opportunity to the **penguins** dataset.
 
@@
 The overall trend across all three species is approximately linear for both **iris** and **penguins**.
-Teachers may encourage students to explore how simple linear regression results and predictions differ when the species variable is omitted, compared to multiple linear regression with species included (\ref{fig:linear-example}).
+Teachers may encourage students to explore how simple linear regression results and predictions differ when the species variable is omitted, compared to, for example, multiple linear regression with species included (\ref{fig:linear-example}).
 
@@
 Here, we have shown that structural size measurements for Palmer Archipelago *Pygoscelis* penguins, available as **penguins** in the \CRANpkg{palmerpenguins} R package, offer a near drop-in replacement for **iris** in a number of common use cases for data science and statistics education including exploratory data visualization, linear correlation and regression, PCA, and clustering by k-means.
-In addition, teaching and learning opportunities in **penguins** are increased due to a greater number of variables, missing values, unequal sample sizes, and Simpson's paradox examples.
+In addition, teaching and learning opportunities in **penguins** are increased due to a greater number of variables, missing values, unequal sample sizes, and Simpson's Paradox examples.
 Importantly, the **penguins** dataset encompasses real-world information derived from several charismatic marine predator species with regional breeding populations notably responding to environmental change occurring throughout the Western Antarctic Peninsula region of the Southern Ocean (see @bestelmeyer_analysis_2011, @gorman_ecological_2014, @gorman_integrative_2015, @gorman_population_2017).

### 2021-04-29 penguins.tex
fix Palmer Station LTER bibliography entries (#23)
Minor changes before submitting (#24)
* tiny proofreading edits

* final commit before submission

@@
 July 2020, the \textbf{palmerpenguins} package has been downloaded over
-102,000 times, highlighting the demand and widespread adoption of this
+105,000 times, highlighting the demand and widespread adoption of this
 viable \textbf{iris} alternative. We directly compare the \textbf{iris}
@@
   \href{https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.221.6}{KNB-LTER
-  Data Package 221.5}
+  Data Package 221.6}
 \end{itemize}
@@
 (n\textsubscript{missing} = 19, out of 2752 total values). Missing
-values and unequal sample sizes are common in real-world data, and adds
-learning opportunities to the \textbf{penguins} dataset.
+values and unequal sample sizes are common in real-world data, and
+create added learning opportunity to the \textbf{penguins} dataset.
 
@@
 simple linear regression results and predictions differ when the species
-variable is omitted, compared to multiple linear regression with species
-included (\ref{fig:linear-example}).
+variable is omitted, compared to, for example, multiple linear
+regression with species included (\ref{fig:linear-example}).
 
@@
 \textbf{penguins} are increased due to a greater number of variables,
-missing values, unequal sample sizes, and Simpson's paradox examples.
+missing values, unequal sample sizes, and Simpson's Paradox examples.
 Importantly, the \textbf{penguins} dataset encompasses real-world

### 2021-04-30 penguins.Rmd
change RJbibliography name to penguins.bib for consistency (#25)
reknit and repurl (#26)

@@
 abstract: > 
-
  In 1935, Edgar Anderson collected size measurements for 150 flowers from three species of *Iris* on the Gaspé Peninsula in Quebec, Canada.
@@
  We directly compare the **iris** and **penguins** datasets for selected analyses to demonstrate that R users, in particular teachers and learners currently using **iris**, can switch to the Palmer Archipelago penguins for many use cases including data wrangling, visualization, linear modeling, multivariate analysis (e.g., PCA), cluster analysis and classification (e.g., by k-means).
-
 preamble: |
@@
 output: rticles::rjournal_article
-bibiography: packages.bib
+bibiography: penguins.bib
 ---
@@
 
-\bibliography{RJreferences}
+\bibliography{penguins}

### 2021-04-30 penguins.tex
change RJbibliography name to penguins.bib for consistency (#25)
reknit and repurl (#26)

@@
 
-\bibliography{RJreferences}
+\bibliography{penguins}
 
@@
 %
-\\\href{mailto:ahorst@ucsb.edu}{\nolinkurl{ahorst@ucsb.edu}}
+\href{mailto:ahorst@ucsb.edu}{\nolinkurl{ahorst@ucsb.edu}}%
 }
@@
 %
-\\\href{mailto:alison@rstudio.com}{\nolinkurl{alison@rstudio.com}}
+\href{mailto:alison@rstudio.com}{\nolinkurl{alison@rstudio.com}}%
 }
@@
 %
-\\\href{mailto:kbgorman@alaska.edu}{\nolinkurl{kbgorman@alaska.edu}}
+\href{mailto:kbgorman@alaska.edu}{\nolinkurl{kbgorman@alaska.edu}}%
 }

### 2021-04-30 penguins.R

@@ -1,3 +1,26 @@
+## ----pkgs, include=FALSE, purl=TRUE-------------------------------------------
+library(tidyverse)
+library(palmerpenguins)
+
+# for tables
+library(kableExtra)
+
+# for plots
+library(paletteer)
+library(patchwork)
+library(GGally)
+library(shadowtext)
+ggplot2::theme_set(ggplot2::theme_minimal())
+
+# for modeling
+library(recipes)
+library(broom)
+
+knitr::read_chunk("penguins_plots.R")
+
+
+
+
 ## ----overview-tbl-------------------------------------------------------------
 overview_compare <- tribble(
   ~Feature, ~iris, ~penguins,
@@ -9,7 +32,9 @@ overview_compare <- tribble(
 )
 
 overview_compare %>% 
-  kable(caption = "Overview comparison of penguins and iris dataset features and characteristics.")
+  kable(caption = "Overview comparison of penguins and iris dataset features and characteristics.",
+        align = "lclccc",
+        booktabs = TRUE)
 
 
 ## ----counts-tbl---------------------------------------------------------------
@@ -36,24 +61,313 @@ iris_penguin_n <- cbind(iris_counts, penguin_counts) %>%
    rename(`Sample size` = n)
 
 iris_penguin_n %>% 
-   kable(caption = "Grouped sample size for iris (by species; n = 150 total) and penguins (by species and sex; n = 344 total). Penguins can be further grouped by variables for island and study year.", align = "lclccc") %>% 
+   kable(caption = "Grouped sample size for iris (by species; n = 150 total) and penguins (by species and sex; n = 344 total). Penguins can be further grouped by variables for island and study year.", 
+         align = "lclccc",
+         booktabs = TRUE) %>% 
   add_header_above(c("Iris sample size (by species)" = 2, "Penguin sample size (by species and sex)" = 4 ))
 
 
 
 ## ---- penguin-pairs, width = 6, height = 4.5, fig.cap = "Distributions and correlations for numeric variables in the penguins data (flipper length (mm), body mass (g), bill length (mm) and bill depth (mm)) for the three observed species: Gentoo (green, triangles); Chinstrap (blue, circles); and Adélie (orange, squares). Significance levels indicated for bivariate correlations are: *p < 0.05; **p < 0.01; ***p < 0.001."----
+### Pairs plots for penguins and iris (with ggpairs)
+
+# Penguin pairs plot:
+
[…]

### 2021-09-23 penguins.Rmd
add distill output format

@@
   % Any extra LaTeX you need in the preamble
-output: rticles::rjournal_article
+output: 
+  rjdistill::rjournal_web_article: default
+  rticles::rjournal_article: default
 bibiography: penguins.bib
@@
 
+```{r}
+library(crosstalk)
+shared_penguins <- SharedData$new(penguins, ~species)
+
+simpson_nospecies <- ggplot(data = shared_penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
+  geom_point(color = "gray40", alpha = 0.6, size = 2) +
+  geom_smooth(method = "lm", se = FALSE, color = "black") +
+  theme_minimal() +
+  theme(panel.border = element_rect(fill = NA, color = "gray70")) +
+  labs(x = "Bill length (mm)", y = "Bill depth (mm)")
+
+simpson_wspecies <-
+  ggplot(data = shared_penguins, aes(x = bill_length_mm, y = bill_depth_mm, group = species)) +
+  geom_point(aes(color = species, shape = species), size = 2, alpha = 0.8) +
+  geom_smooth(method = "lm", se = FALSE, aes(color = species), show.legend = FALSE) +
+  scale_color_paletteer_d("colorblindr::OkabeIto") +
+  theme_minimal() +
+  theme(panel.border = element_rect(fill = NA, color = "gray70")) +
+  labs(x = "Bill length (mm)", y = "Bill depth (mm)") +
+  guides(color = guide_legend("species"),
+         shape = "none")
+
+bscols(widths = c(2,NA),
+  list(
+    filter_checkbox("species", "Species", shared_penguins, ~species, inline = TRUE)
+  ),
+  plotly::ggplotly(simpson_wspecies)
+)
+```
+
 ### Principal component analysis
